---
title: FT-NIRS Otolith Age Prediction - IBM
output: 
  html_document:
    theme: simplex
    toc: yes
    toc_float:
      collapsed: true
---

# Importing FT-NIRS Spectra

### Metadata

```{r, cache = T}
ibm_meta<-read.csv("metadata/IBM_metadata.csv", header = TRUE)
ibm_meta<-ibm_meta[ibm_meta$file_name !="",] # 8 specimens removed (missing scans?)
ibm_meta$file_path<-paste0(getwd(), "/FT-NIRS_spectra/IBM_scans/", ibm_meta$file_name)
Opusfiles<-as.vector(ibm_meta$file_path)
exists<-as.vector(lapply(Opusfiles, file.exists))
ibm_meta$exists<-exists
meta1<-ibm_meta[ibm_meta$exists=="TRUE",]
Opusfiles<-as.vector(meta1$file_path)
rm(exists)
rm(meta1)
```

### Opus files

```{r, cache = T}
file <- Opusfiles[1]
data_list <- read_opus(dsn = file)
rm(data_list)
rm(file)
SPCfiles_nooffset<-lapply(Opusfiles,read_opus)
spectra<-lapply(SPCfiles_nooffset, function (x) x[[1]]$ab$data)
species<-lapply(SPCfiles_nooffset,function (x) x$lab_and_process_param_raw$parameters$FC2$parameter_value)
file_id<-lapply(SPCfiles_nooffset,function (x) x$lab_and_process_param_raw$parameters$FD1$parameter_value)
instrument<-lapply(SPCfiles_nooffset,function (x) x[[1]]$instrument_ref$parameters$INS$parameter_value)
wavenumber<-lapply(SPCfiles_nooffset,function (x) x[[1]]$ab$wavenumbers)
spectra<-lapply(spectra,as.data.frame)
for (i in 1:length(spectra)){
  colnames(spectra[[i]])<-wavenumber[[i]]
}
for (i in 1:length(spectra)){
  spectra[[i]]$species<-species[[i]]
}
for (i in 1:length(spectra)){
  spectra[[i]]$file_id<-file_id[[i]]
}
for (i in 1:length(spectra)){
  spectra[[i]]$instrument<-instrument[[i]]
}
for (i in 1:length(spectra)){
  spectra[[i]]$file_path<-Opusfiles[[i]]
}
file_name<-lapply(spectra, function (x) splitstackshape::cSplit(as.data.frame(x$file_path),sep="/",splitCols="x$file_path",type.convert=FALSE)%>%select(tail(names(.), 1)))
for (i in 1:length(spectra)){
  spectra[[i]]$file_name<-file_name[[i]][[1,1]]
}
```

### Combine dataframes

```{r, cache = T}
df <- as.data.frame(do.call(rbind,spectra))
dfmeta_IBM<-dplyr::left_join(ibm_meta,df,by=c("file_name","file_path"))
dfmeta_IBM<-dfmeta_IBM%>%dplyr::select(.,-c(exists, instrument))
rm(ibm_meta, df, wavenumber, spectra, instrument, species, SPCfiles_nooffset,file_id)
colnames(dfmeta_IBM)<-as.character(colnames(dfmeta_IBM))
dfmeta_IBM$run_number <- as.factor(dfmeta_IBM$run_number)
# saveRDS(dfmeta_IBM, "RDS_dataframes/IBM_dfmeta.RDS")
dfmeta_IBM_long<-tidyr::pivot_longer(dfmeta_IBM, cols= -c(1:24)) #make it long format to plot it
dfmeta_IBM_long<-dfmeta_IBM_long%>%rename(.,"wavenumber"="name")
dfmeta_IBM_long$wavenumber<-as.numeric(as.character(dfmeta_IBM_long$wavenumber))
rm(Opusfiles, file_name)
```

### Plot check

```{r, cache = T}
ggplot()+
  geom_path(data=dfmeta_IBM_long[dfmeta_IBM_long$specimen %in% c(423801,423812,423818,423902, 423904,423914,423918), ] ,aes(x=wavenumber,y=value,group=file_name),linewidth=.5)+ 
  scale_x_reverse()+
  labs(y="Absorbance units",x= expression(paste("Wavenumber ", cm^-1)))+
  theme(axis.text =element_text(size=10),
        axis.title=element_text(size=12),
        strip.text = element_text(size=14),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))+
  facet_wrap(~specimen)
```

# Exploratory Analysis

### Spectra - Scans 1 vs 2

```{r}
ggplot(dfmeta_IBM_long) + geom_path(aes(x = wavenumber, y = value, color = run_number, group = file_name), alpha = 0.5) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Run Number") +
  scale_color_viridis(discrete = T) + 
  scale_x_reverse() + 
  ggtitle("LPW FT-NIRS Spectra, Scans 1 & 2")
```

### PCA - compare runs

```{r, cache = T}
pca_IBM <- pca(dfmeta_IBM[25:ncol(dfmeta_IBM)], scale = T)
pcs <- as.data.frame(cbind(pc2 = pca_IBM$calres$scores[, 2], run_number = dfmeta_IBM$run_number)) # extract scores for PC2
pcs <- cbind(pc1 = pca_IBM$calres$scores[, 1], pcs) # extract scores for PC1
# plot PC1 & PC2, colored by run_number
ggplot(pcs) +
  geom_point(aes(x = pc1, y = pc2, color = as.factor(run_number)), size = 3) +
  labs(x = paste("PC1 (",
                 round(pca_IBM$calres$expvar[1], digits = 3), # variance explained by PC1
                 "% var. explained )"),
       y = paste("PC2 (", 
                 round(pca_IBM$calres$expvar[2], digits = 3), # variance explained by PC2
                 "% var. explained )"),
       color = "Run Number"
  ) +
  scale_color_viridis(discrete = T)
rm(pca_IBM, pcs)
```

Will only use new scans from cleaned otoliths (run 2)

## Preprocessing
```{r, cache = T}
# preprocess immediately right now
dfmeta_IBM_proc <- cbind(dfmeta_IBM[, c(1:24)],prospectr::savitzkyGolay(dfmeta_IBM[,25:ncol(dfmeta_IBM)],m = 1, p = 3, w = 17))
dfmeta_IBM_proc_long <- pivot_longer(dfmeta_IBM_proc, cols = `11472`:`4016`, names_to = "name", values_to = "value")
dfmeta_IBM_proc_long$name <- as.numeric(dfmeta_IBM_proc_long$name)
```

#### Spectra - Raw

```{r, cache = T}
ggplot(dfmeta_IBM_long[dfmeta_IBM_long$run_number == 2,]) + geom_path(aes(x = wavenumber, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse()
```

#### Spectra - Savitzky-Golay

```{r, cache = T}
plot1 <- ggplot(dfmeta_IBM_proc_long[dfmeta_IBM_proc_long$run_number == 2,]) + geom_path(aes(x = name, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SG")
plot1
```

This data looks really noisy after applying a Savitzky-Golay filter.  Going to look at using standard normal variate (SNV) to correct first.

## SNV

#### Spectra - SNV

```{r, cache = T}
testing1 <- prep.snv(as.matrix(dfmeta_IBM[,25:ncol(dfmeta_IBM)]))
testing1 <- as.data.frame(unlist(testing1))
testing1 <- cbind(dfmeta_IBM[,1:24], testing1)
testing1_long <- pivot_longer(testing1, cols = `11536`:`3952`, names_to = "name", values_to = "value")
testing1_long$name <- as.numeric(testing1_long$name)
plot2 <- ggplot(testing1_long[testing1_long$run_number == 2,]) + geom_path(aes(x = name, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SNV")
plot3
rm(testing1_long)
```

SNV normalization on it's own looks very strange.  Will trying combinations of filters.

#### Spectra - SNV + SG

```{r, cache = T}
testing2 <- cbind(testing1[, c(1:24)],prospectr::savitzkyGolay(testing1[,25:ncol(testing1)],m = 1, p = 3, w = 17))
testing2_long <- pivot_longer(testing2, cols = `11472`:`4016`, names_to = "name", values_to = "value")
testing2_long$name <- as.numeric(testing2_long$name)
plot3 <- ggplot(testing2_long[testing2_long$run_number == 2,]) + geom_path(aes(x = name, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SNV + SG")
grid.arrange(plot2,plot1,plot3,ncol = 1)
rm(testing2_long)
```

Using both SNV & SG doesn't appear to change much, may just use SG on its own.  Will remove wavenumbers >7500 like in LPW preprocessing now.

#### Spectra - SG + <7500

```{r, cache = T}
testing3 <- dfmeta_IBM[,-c(25:529)]
testing3 <- cbind(testing3[, c(1:24)],prospectr::savitzkyGolay(testing3[,25:ncol(testing3)],m = 1, p = 3, w = 17))
testing3_long <- pivot_longer(testing3, cols = `7432`:`4016`, names_to = "name", values_to = "value")
testing3_long$name <- as.numeric(testing3_long$name)
plot4 <- ggplot(testing3_long[testing3_long$run_number == 2,]) + geom_path(aes(x = name, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SG & <7500")
rm(testing3_long)
```

#### Spectra - SNV + SG + <7500

```{r, cache = T}
testing4 <- testing2[,-c(25:521)]
testing4_long <- pivot_longer(testing4, cols = `7496`:`4016`, names_to = "name", values_to = "value")
testing4_long$name <- as.numeric(testing4_long$name)
# <7500 for SNV -> SAVITZKY
plot5 <- ggplot(testing4_long[testing4_long$run_number == 2,]) + geom_path(aes(x = name, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SNV + SG & <7500")
rm(testing4_long)
grid.arrange(plot4,plot5,ncol = 1)
rm(plot1,plot2,plot3,plot4,plot5,testing1,testing2,testing2_long,testing3,testing4,dfmeta_IBM_long,dfmeta_IBM_proc,dfmeta_IBM_proc_long)
```

To make things simpler, I will do same preprocessing as LPW: remove wavenumbers >7500 cm-1 and apply a Savitzky-Golay filter (1st derivative, 3rd order, 17-point).



## Preprocess/Filter data

```{r, cache = T}
# preprocess, filter data
dfmeta_IBM_proc <- cbind(dfmeta_IBM[, c(1:24)],prospectr::savitzkyGolay(dfmeta_IBM[,25:ncol(dfmeta_IBM)],m = 1, p = 3, w = 17))
dfmeta_IBM_proc <- dfmeta_IBM_proc[,-c(25:521)]
rm(dfmeta_IBM)
# saveRDS(dfmeta_IBM_proc, "RDS_dataframes/IBM_proc_filter.RDS")
```

## PCA looking for outliers

```{r}
pca_temp <- pca(dfmeta_IBM_proc[, 25:ncol(dfmeta_IBM_proc), ])
dfmeta_IBM_proc$PC1 <- pca_temp$res$cal$scores[, 1]
dfmeta_IBM_proc$PC2 <- pca_temp$res$cal$scores[, 2]
rm(pca_temp)
library(plotly)
ggplotly(
  ggplot() +
    geom_point(data = dfmeta_IBM_proc,
               aes(x = PC1, y = PC2, color = length), size = 2
    ) + 
    scale_color_viridis()
)
rm(rm_specimens)
```

# Model Length - FT-NIRS

#### Load Data

```{r}
dfmeta_LPW_proc <- readRDS("RDS_dataframes/LPW_scan_avg_proc_filter.RDS")
dfmeta_LPW_proc <- dfmeta_LPW_proc[complete.cases(dfmeta_LPW_proc$structure_weight), ]
# only select scan 2, otoliths with weight > 0.0001
dfmeta_IBM_proc <- readRDS("RDS_dataframes/IBM_proc_filter.RDS")
dfmeta_IBM_proc <- dfmeta_IBM_proc %>% dplyr::filter(run_number == 2, structure_weight > 0.0001)
# store metrics
RMSE.weight <- list()
r2.weight <- list()
AIC.weight <- list()
AICc.weight <- list()
mod.summary.weight <- data.frame(
  model = c("lm", "gam", "pls","pls_vip", "lm.IBM", "gam.IBM", "pls.IBM", "vip.IBM"),
  r2 = 1:8,
  RMSE = 1:8,
  AIC = c(0,0,NA,NA,NA,NA,NA,NA),
  AICc = c(0,0,NA,NA,NA,NA,NA,NA))
preds.IBM <- list()
```

#### Pick PC's

```{r}
# 10-fold CV split for LPW models - create folds
set.seed(6)
splits <- caret::createFolds(dfmeta_LPW_proc$structure_weight, k = 10, list = TRUE, returnTrain = FALSE)
# extract PC's for each LPW calibration set, apply PC rotation to LPW test set and IBM dataset as well
cal <- list()
test <- list()
test.IBM <- list()
for (i in 1:10) {
  # calibration set and PC's
  pc.mod <- preProcess(dfmeta_LPW_proc[-splits[[i]], -c(1:20)], method = "pca", thresh = 0.95, pcaComp = 10)
  pc.cal <- predict(pc.mod, dfmeta_LPW_proc[-splits[[i]], -c(1:20)])
  pc.cal <- cbind(pc.cal, dfmeta_LPW_proc[-splits[[i]], ])
  cal[[i]] <- pc.cal
  rm(pc.cal)
  # test sets
  pc.test <- predict(pc.mod, dfmeta_LPW_proc[splits[[i]], -c(1:20)])
  pc.test <- cbind(pc.test, dfmeta_LPW_proc[splits[[i]], ])
  test[[i]] <- pc.test
  # PC's for IBM
  pc.IBM <- predict(pc.mod, dfmeta_IBM_proc[,-c(1:24)])
  pc.IBM <- cbind(pc.IBM, dfmeta_IBM_proc)
  test.IBM[[i]] <- pc.IBM
  rm(pc.mod, pc.test, pc.IBM)
}
# determine which PC's to include via step & AIC selection
```

```{r results = 'hide', message = F, warning = F}
mod.sel <- list()
invisible(
  for (i in 1:10) {
  pctest <- cal[[i]]
  temp <- step(lm(data = pctest[-splits[[i]], ], length ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6),trace = 0)
  mod.sel[[i]] <- rownames(summary(temp)$coef)
  rm(pctest, temp)
}
)
```

```{r}
knitr::kable(table(unlist(mod.sel)), align = "c")
```

Use PC's 1,3,4,6

## MLR

```{r}
mods.lm.weight <- list()
preds.temp <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  mod <- lm(data = calibrate, structure_weight ~ PC1 + PC3 + PC4 + PC6)
  preds <- predict(mod, newdata = testing)
  RMSE.weight$lm.test[i] <- caret::RMSE(pred = preds, obs = testing[, "structure_weight"])
  RSS <- sum((testing$structure_weight - preds)^2)
  TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
  r2.weight$lm.test[i] <- 1 - (RSS / TSS)
  AIC.weight$lm[i] <- AIC(mod)
  AICc.weight$lm[i] <- AICc(mod)
  mods.lm.weight[[i]] <- mod
  # IBM
  testing <- test.IBM[[i]]
  preds.temp[[i]] <- predict(mod, newdata = testing)
  RMSE.weight$lm.IBM[i] <- caret::RMSE(pred = preds.temp[[i]], obs = dfmeta_IBM_proc[, "structure_weight"])
  RSS <- sum((testing$structure_weight -  preds.temp[[i]])^2)
  TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
  r2.weight$lm.IBM[i] <- 1 - (RSS / TSS)
  rm(mod, preds, RSS, TSS)
}
preds.IBM$lm <- preds.temp
rm(preds.temp)
mod.summary.weight[1,2] <- round(mean(r2.weight$lm.test), 3)
mod.summary.weight[1,3] <- round(mean(RMSE.weight$lm.test), 3)
mod.summary.weight[1,4] <- round(mean(AIC.weight$lm) , 3)
mod.summary.weight[1,5] <- round(mean(AICc.weight$lm) , 3)
mod.summary.weight[5,2] <- round(mean(r2.weight$lm.IBM), 3)
mod.summary.weight[5,3] <- round(mean(RMSE.weight$lm.IBM), 3)
```

## GAM

```{r message=FALSE, cache=TRUE}
# GAM with 10 fold CV, select = T allows PCs to be penalized and effectively removed from model if appropriate.
mods.gam.weight <- list()
preds.temp <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  mod <- gam(data = calibrate, structure_weight ~ s(PC1) + s(PC2) + s(PC3) + s(PC4) + s(PC5) + s(PC6), method = "REML")
  preds <- predict(mod, newdata = testing)
  RSS <- sum((testing$structure_weight - preds)^2)
  TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
  RMSE.weight$gam.test[i] <- caret::RMSE(pred = preds, obs = testing[, "structure_weight"])
  r2.weight$gam.test[i] <- 1 - (RSS / TSS)
  AIC.weight$gam[i] <- AIC(mod)
  AICc.weight$gam[i] <- AICc(mod)
  mods.gam.weight[[i]] <- mod
  # IBM
  testing <- test.IBM[[i]]
  preds.temp[[i]] <- predict(mod, newdata = testing)
  RSS <- sum((testing$structure_weight - preds.temp[[i]])^2)
  TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
  RMSE.weight$gam.IBM[i] <- caret::RMSE(pred = preds.temp[[i]], obs = dfmeta_IBM_proc[, "structure_weight"])
  r2.weight$gam.IBM[i] <- 1 - (RSS / TSS)
  rm(mod, RSS, TSS, i, calibrate, testing)
}
preds.IBM$gam <- preds.temp
rm(preds.temp)
mod.summary.weight[2,2] <- round(mean(r2.weight$gam.test), 3)
mod.summary.weight[2,3] <- round(mean(RMSE.weight$gam.test), 3)
mod.summary.weight[2,4] <- round(mean(AIC.weight$gam), 3)
mod.summary.weight[2,5] <- round(mean(AICc.weight$gam), 3)
mod.summary.weight[6,2] <- round(mean(r2.weight$gam.IBM), 3)
mod.summary.weight[6,3] <- round(mean(RMSE.weight$gam.IBM), 3)
rm(AIC.weight, AICc.weight)
```

## PLS

```{r, cache = T}
mods.pls.weight <- list()
mods.vip.weight <- list()
preds.temp1 <- list()
preds.temp2 <- list()
# pls, no variable selection & VIP > 1
for (i in 1:10) {
  calibrate <- dfmeta_LPW_proc[-splits[[i]], ]
  testing <- dfmeta_LPW_proc[splits[[i]], ]
  mod <- pls(calibrate[, 21:ncol(calibrate)], calibrate[, "structure_weight"],
    ncomp.selcrit = "wold", scale = F, center = T, cv = 1,
    info = "Otolith Weight Prediction Model",
    x.test = testing[, 21:ncol(testing)],
    y.test = testing[, "structure_weight"]
  )
  ncomp <- mod$ncomp.selected
  RMSE.weight$pls.test[i] <- mod$testres$rmse[[ncomp]]
  r2.weight$pls.test[i] <- mod$testres$r2[[ncomp]]
  mods.pls.weight[[i]] <- mod
  # IBM
  testing <- test.IBM[[i]]
  preds.temp1[[i]] <- predict(
    mod,
    dfmeta_IBM_proc[, c(25:ncol(dfmeta_IBM_proc))],
    dfmeta_IBM_proc[, 3]
  )
  ncomp <- preds.temp1[[i]]$ncomp.selected
  RMSE.weight$pls.IBM[i] <- preds.temp1[[i]]$rmse[ncomp]
  r2.weight$pls.IBM[i] <- preds.temp1[[i]]$r2[ncomp]
  ### VIP > 1
  vip <- as.data.frame(vipscores(mod))
  testing <- dfmeta_LPW_proc[splits[[i]], ]
  mod <- pls(calibrate[, 21:ncol(calibrate)], calibrate[, "structure_weight"],
    ncomp.selcrit = "wold", scale = F, center = T, cv = 1,
    info = "Otolith Weight Prediction Model",
    x.test = testing[, 21:ncol(testing)],
    y.test = testing[, "structure_weight"],
    exclcols = vip$V1 < 1
  )
  ncomp <- mod$ncomp.selected
  ncomp <- mod$ncomp.selected
  RMSE.weight$vip.test[i] <- mod$testres$rmse[[ncomp]]
  r2.weight$vip.test[i] <- mod$testres$r2[[ncomp]]
  mods.vip.weight[[i]] <- mod
  # IBM
  testing <- test.IBM[[i]]
  preds.temp2[[i]] <- predict(
    mod,
    dfmeta_IBM_proc[, c(25:ncol(dfmeta_IBM_proc))],
    dfmeta_IBM_proc[, 3]
  )
  ncomp <- preds.temp2[[i]]$ncomp.selected
  RMSE.weight$vip.IBM[i] <- preds.temp2[[i]]$rmse[ncomp]
  r2.weight$vip.IBM[i] <- preds.temp2[[i]]$r2[ncomp]
  rm(calibrate, mod, ncomp, testing, i, vip)
}
preds.IBM$pls <- preds.temp1
preds.IBM$vip <- preds.temp2
rm(preds.temp1, preds.temp2)
mod.summary.weight[3, 2] <- round(mean(r2.weight$pls.test), 3)
mod.summary.weight[3, 3] <- round(mean(RMSE.weight$pls.test), 3)
mod.summary.weight[4, 2] <- round(mean(r2.weight$vip.test), 3)
mod.summary.weight[4, 3] <- round(mean(RMSE.weight$vip.test), 3)
mod.summary.weight[7, 2] <- round(mean(r2.weight$pls.IBM), 3)
mod.summary.weight[7, 3] <- round(mean(RMSE.weight$pls.IBM), 3)
mod.summary.weight[8, 2] <- round(mean(r2.weight$vip.IBM), 3)
mod.summary.weight[8, 3] <- round(mean(RMSE.weight$vip.IBM), 3)
knitr::kable(mod.summary.weight, align = "c")
rm(r2.weight,RMSE.weight)
```

# Results

### Plots

```{r}
rm(r2.weight, RMSE.weight)
preds.IBM$structure_weight <- dfmeta_IBM_proc$structure_weight
preds.IBM$pcs <- list()
for(i in 1:10){
  preds.IBM$pcs[[i]] <- test.IBM[[i]][1:10]
}
# lm predictions
ggplot() + 
  geom_point(aes(x = preds.IBM$lm[[1]], y = preds.IBM$y)) +
  geom_abline(aes(slope = 1, intercept = 0))
# GAM
ggplot() + 
  geom_point(aes(x = preds.IBM$gam[[1]], y = preds.IBM$y)) +
  geom_abline(aes(slope = 1, intercept = 0))
# PLS
ncomp <- preds.IBM$pls[[1]]$ncomp.selected

ggplot() +
  geom_point(aes(x = preds.IBM$pls[[1]]$y.pred[,ncomp,1], 
                 y = preds.IBM$pls[[1]]$y.ref), col = "red") + 
  scale_y_continuous(limits = c(0,0.004)) +
  scale_x_continuous(limits = c(-0.01, 0.005)) + 
  geom_abline(aes(slope = 1, intercept = 0))
  


ggplot() +
  geom_smooth(aes(x = preds.IBM$pls[[1]]$y.pred[,ncomp,1], 
                 y = preds.IBM$pls[[1]]$y.ref), 
              method = "lm", se = T,
              col = "red") + 
  scale_y_continuous(limits = c(0,0.004)) +
  scale_x_continuous(limits = c(-0.01, 0.005)) + 
  geom_abline(aes(slope = 1, intercept = 0))
# VIP
ggplot() + 
  geom_point(aes(x = preds.IBM$vip[[1]]$y.ref, y = preds.IBM$y)) +
  geom_abline(aes(slope = 1, intercept = 0))


  geom_smooth(method = "lm", formula = structure_weight ~ PC1 + PC3 + PC4 + PC6, mapping = aes(x = ))
plot_lm

mods.lm.weight[[1]]$terms
coef(mods.lm.weight[[1]])

preds.IBM$pls[[1]]$y.pred[,ncomp,1]

```
try log link function for GAM
```{r}
r2.weight <- list()
RMSE.weight <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  mod <- gam(data = calibrate, structure_weight ~ s(PC1) + s(PC2) + s(PC3) + s(PC4) + s(PC5) + s(PC6), method = "REML")
  preds <- predict(mod, newdata = testing)
  RSS <- sum((testing$structure_weight - preds)^2)
  TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
  RMSE.weight$gam.test[i] <- caret::RMSE(pred = preds, obs = testing[, "structure_weight"])
  r2.weight$gam.test[i] <- 1 - (RSS / TSS)
  AIC.weight$gam[i] <- AIC(mod)
  AICc.weight$gam[i] <- AICc(mod)
  mods.gam.weight[[i]] <- mod
  # IBM
  testing <- test.IBM[[i]]
  preds.temp[[i]] <- predict(mod, newdata = testing)
  RSS <- sum((testing$structure_weight - preds.temp[[i]])^2)
  TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
  RMSE.weight$gam.IBM[i] <- caret::RMSE(pred = preds.temp[[i]], obs = dfmeta_IBM_proc[, "structure_weight"])
  r2.weight$gam.IBM[i] <- 1 - (RSS / TSS)
  rm(mod, RSS, TSS, i, calibrate, testing)
}

```

