---
title: FT-NIRS Otolith Age Prediction - IBM
output: 
  html_document:
    theme: simplex
    toc: yes
    toc_float:
      collapsed: true
---

# Load Packages

```{r, results = 'hide', warning = F, message = F}
# Install pacman if not already installed
if (!require("pacman")) install.packages("pacman")
library(pacman)
# Install and load packages 
p_load(caret, corrplot, devtools, dplyr, gglm, ggplot2, plotly, gratia, gridExtra, 
       knitr, lubridate, mdatools, mgcv, MuMIn, opusreader2, prospectr, 
       splitstackshape, shiny, tidyr, viridis)
# Install and load GitHub packages
p_load_gh("spectral-cockpit/opusreader2")
# Uncomment the following line if you want to install simplerspec
# p_load_gh("philipp-baumann/simplerspec")
```

# Importing FT-NIRS Spectra

### Metadata + Read Ages

```{r eval=FALSE, include=FALSE}
ibm_meta<-read.csv("metadata/IBM_metadata.csv", header = TRUE)
ibm_meta<-ibm_meta[ibm_meta$file_name !="",] # 8 specimens removed (missing scans?)
ibm_meta$file_path<-paste0(getwd(), "/FT-NIRS_spectra/IBM_scans/", ibm_meta$file_name)
ibm_ages<-read.csv("metadata/IBM_otoreadages.csv", header = TRUE)
# Match and update read_age values in metadata
ibm_meta$read_age <- ifelse(ibm_meta$specimen %in% ibm_ages$specimen,
                           ibm_ages$read_age[match(ibm_meta$specimen, ibm_ages$specimen)],
                           ibm_meta$read_age)
# Replace zeros with NA in the read_age column
ibm_meta$read_age <- ifelse(ibm_meta$read_age == 0, NA, ibm_meta$read_age)


Opusfiles<-as.vector(ibm_meta$file_path)
exists<-as.vector(lapply(Opusfiles, file.exists))
ibm_meta$exists<-exists
meta1<-ibm_meta[ibm_meta$exists=="TRUE",]
Opusfiles<-as.vector(meta1$file_path)
rm(exists, meta1, ibm_ages)
```

### Opus files

```{r eval=FALSE, include=FALSE}
file <- Opusfiles[1]
data_list <- read_opus(dsn = file)
rm(data_list)
rm(file)
SPCfiles_nooffset<-lapply(Opusfiles,read_opus)
spectra<-lapply(SPCfiles_nooffset, function (x) x[[1]]$ab$data)
species<-lapply(SPCfiles_nooffset,function (x) x$lab_and_process_param_raw$parameters$FC2$parameter_value)
file_id<-lapply(SPCfiles_nooffset,function (x) x$lab_and_process_param_raw$parameters$FD1$parameter_value)
instrument<-lapply(SPCfiles_nooffset,function (x) x[[1]]$instrument_ref$parameters$INS$parameter_value)
wavenumber<-lapply(SPCfiles_nooffset,function (x) x[[1]]$ab$wavenumbers)
spectra<-lapply(spectra,as.data.frame)
for (i in 1:length(spectra)){
  colnames(spectra[[i]])<-wavenumber[[i]]
}
for (i in 1:length(spectra)){
  spectra[[i]]$species<-species[[i]]
}
for (i in 1:length(spectra)){
  spectra[[i]]$file_id<-file_id[[i]]
}
for (i in 1:length(spectra)){
  spectra[[i]]$instrument<-instrument[[i]]
}
for (i in 1:length(spectra)){
  spectra[[i]]$file_path<-Opusfiles[[i]]
}
file_name<-lapply(spectra, function (x) splitstackshape::cSplit(as.data.frame(x$file_path),sep="/",splitCols="x$file_path",type.convert=FALSE)%>%select(tail(names(.), 1)))
for (i in 1:length(spectra)){
  spectra[[i]]$file_name<-file_name[[i]][[1,1]]
}
```

### Combine dataframes

```{r eval=FALSE, include=FALSE}
df <- as.data.frame(do.call(rbind,spectra))
dfmeta_IBM<-dplyr::left_join(ibm_meta,df,by=c("file_name","file_path"))
dfmeta_IBM<-dfmeta_IBM%>%dplyr::select(.,-c(exists, instrument))
rm(ibm_meta, df, wavenumber, spectra, instrument, species, SPCfiles_nooffset,file_id)
colnames(dfmeta_IBM)<-as.character(colnames(dfmeta_IBM))
dfmeta_IBM$run_number <- as.factor(dfmeta_IBM$run_number)
# saveRDS(dfmeta_IBM, "RDS_dataframes/IBM_dfmeta.RDS")
dfmeta_IBM_long<-tidyr::pivot_longer(dfmeta_IBM, cols= -c(1:24)) #make it long format to plot it
dfmeta_IBM_long<-dfmeta_IBM_long%>%rename(.,"wavenumber"="name")
dfmeta_IBM_long$wavenumber<-as.numeric(as.character(dfmeta_IBM_long$wavenumber))
rm(Opusfiles, file_name, dfmeta_IBM, dfmeta_IBM_long, i)
```

# FUNCTION: Make Long Format

format to convert dataframe to long format for plotting

```{r}
make_long <- function(df) {
  df_long <- df %>%
    tidyr::pivot_longer(cols = -c(1:24)) %>%
    rename(wavenumber = name) %>%
    mutate(wavenumber = as.numeric(as.character(wavenumber)))
  
  return(df_long)
}
```

### Plot check

```{r, cache = T}
dfmeta_IBM_long <- make_long(readRDS("RDS_dataframes/IBM_dfmeta.RDS"))
ggplot()+
  geom_path(data=dfmeta_IBM_long[dfmeta_IBM_long$specimen %in% c(423801,423812,423818,423902, 423904,423914,423918), ] ,aes(x=wavenumber,y=value,group=file_name),linewidth=.5)+ 
  scale_x_reverse()+
  labs(y="Absorbance units",x= expression(paste("Wavenumber ", cm^-1)))+
  theme(axis.text =element_text(size=10),
        axis.title=element_text(size=12),
        strip.text = element_text(size=14),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))+
  facet_wrap(~specimen)
```

# Exploratory Analysis

### Spectra - Scans 1 vs 2

Concerns about dirty otoliths motivated sonicating and cleaning a subset of otoliths (run number 2).

```{r}
dfmeta_IBM <- readRDS("RDS_dataframes/IBM_dfmeta.RDS")
dfmeta_IBM_long <- make_long(readRDS("RDS_dataframes/IBM_dfmeta.RDS"))

ggplot(dfmeta_IBM_long) + geom_path(aes(x = wavenumber, y = value, color = run_number, group = file_name), alpha = 0.5) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Run Number") +
  scale_color_viridis(discrete = T) + 
  scale_x_reverse() + 
  ggtitle("IBM FT-NIRS Spectra, Scans 1 & 2")
```

### PCA - compare runs

```{r, cache = T}
pca_IBM <- pca(dfmeta_IBM[25:ncol(dfmeta_IBM)], scale = T)
pcs <- as.data.frame(cbind(pc2 = pca_IBM$calres$scores[, 2], run_number = dfmeta_IBM$run_number)) # extract scores for PC2
pcs <- cbind(pc1 = pca_IBM$calres$scores[, 1], pcs) # extract scores for PC1
# plot PC1 & PC2, colored by run_number
ggplot(pcs) +
  geom_point(aes(x = pc1, y = pc2, color = as.factor(run_number)), size = 3) +
  labs(x = paste("PC1 (",
                 round(pca_IBM$calres$expvar[1], digits = 3), # variance explained by PC1
                 "% var. explained )"),
       y = paste("PC2 (", 
                 round(pca_IBM$calres$expvar[2], digits = 3), # variance explained by PC2
                 "% var. explained )"),
       color = "Run Number"
  ) +
  scale_color_viridis(discrete = T)
rm(pca_IBM, pcs)
```

Will only use new scans from cleaned otoliths (run 2)

## Preprocessing

```{r, cache = T}
# preprocess immediately right now
dfmeta_IBM_proc <- cbind(dfmeta_IBM[, c(1:24)],prospectr::savitzkyGolay(dfmeta_IBM[,25:ncol(dfmeta_IBM)],m = 1, p = 3, w = 17))
dfmeta_IBM_proc_long <- make_long(dfmeta_IBM_proc)
```

#### Spectra - Raw

```{r, cache = T}
ggplot(dfmeta_IBM_long[dfmeta_IBM_long$run_number == 2,]) + geom_path(aes(x = wavenumber, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse()
```

#### Spectra - Savitzky-Golay

```{r, cache = T}
plot1 <- ggplot(dfmeta_IBM_proc_long[dfmeta_IBM_proc_long$run_number == 2,]) + geom_path(aes(x = wavenumber, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SG")
plot1
```

This data looks really noisy after applying a Savitzky-Golay filter. Going to look at using standard normal variate (SNV) to correct first.

## SNV

#### Spectra - SNV

```{r, cache = T}
testing1 <- prep.snv(as.matrix(dfmeta_IBM[,25:ncol(dfmeta_IBM)]))
testing1 <- as.data.frame(unlist(testing1))
testing1 <- cbind(dfmeta_IBM[,1:24], testing1)
testing1_long <- make_long(testing1)
plot2 <- ggplot(testing1_long[testing1_long$run_number == 2,]) + geom_path(aes(x = wavenumber, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SNV")
plot2
rm(testing1_long)
```

SNV normalization on it's own looks very strange. Will trying combinations of filters.

#### Spectra - SNV + SG

```{r, cache = T}
testing2 <- cbind(testing1[, c(1:24)],prospectr::savitzkyGolay(testing1[,25:ncol(testing1)],m = 1, p = 3, w = 17))
testing2_long <- make_long(testing2)
plot3 <- ggplot(testing2_long[testing2_long$run_number == 2,]) + geom_path(aes(x = wavenumber, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SNV + SG")
grid.arrange(plot2,plot1,plot3,ncol = 1)
```

Using both SNV & SG doesn't appear to change much, may just use SG on its own. Originally was removing wavenumbers \>7500 like in LPW preprocessing, but after LPW analysis I don't think I want to go that route.

#### Spectra - SG + \<7500

```{r, cache = T}
testing3 <- dfmeta_IBM[,-c(25:529)]
testing3 <- cbind(testing3[, c(1:24)],prospectr::savitzkyGolay(testing3[,25:ncol(testing3)],m = 1, p = 3, w = 17))
testing3_long <- make_long(testing3)
plot4 <- ggplot(testing3_long[testing3_long$run_number == 2,]) + geom_path(aes(x = wavenumber, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SG & <7500")
rm(testing3_long)
```

#### Spectra - SNV + SG + \<7500

```{r, cache = T}
testing4 <- testing2[,-c(25:521)]
testing4_long <- make_long(testing4)
# <7500 for SNV -> SAVITZKY
plot5 <- ggplot(testing4_long[testing4_long$run_number == 2,]) + geom_path(aes(x = wavenumber, y = value, color = length, group = file_name)) + 
  labs(y = "Absorbance units", x = expression(paste("Wavenumber ", cm^-1)), color = "Length (mm)") +
  scale_color_viridis() + 
  scale_x_reverse() + 
  ggtitle("SNV + SG & <7500")
rm(testing4_long)
grid.arrange(plot4,plot5,ncol = 1)
rm(plot1,plot2,plot3,plot4,plot5,testing1,testing2,testing2_long,testing3,testing4,dfmeta_IBM_long,dfmeta_IBM_proc,dfmeta_IBM_proc_long)
```

To make things simpler, I will do same preprocessing as LPW: Savitzky-Golay filter (1st derivative, 3rd order, 17-point). I may try some SNV analysis later

# Preprocess/Filter data

```{r, cache = T}
dfmeta_IBM <- readRDS("RDS_dataframes/IBM_dfmeta.RDS") %>% filter(run_number == 2)
# preprocess, filter data
df <- cbind(dfmeta_IBM[, c(1:24)],prospectr::savitzkyGolay(dfmeta_IBM[,25:ncol(dfmeta_IBM)],m = 1, p = 3, w = 17)) %>% filter(run_number == 2)
rm(dfmeta_IBM)
df <- df[complete.cases(df$read_age), ]
```

## PCA looking for outliers

Some really big outliers, but those are just large individuals, nothing to be concerned about.

```{r}
pca_temp <- pca(df[, 25:ncol(df), ])
PC1 <- pca_temp$res$cal$scores[, 1]
PC2 <- pca_temp$res$cal$scores[, 2]
pca_temp <- cbind(PC1, PC2, df)
ggplotly(
  ggplot() +
    geom_point(data = pca_temp,
               aes(x = PC1, y = PC2, color = length), size = 2
    ) + 
    scale_color_viridis()
)
rm(pca_temp, PC1, PC2)
```

# -------- Model Age - FT-NIRS --------

## Pick PC's

```{r}
df <- readRDS("RDS_dataframes/IBM_dfmeta.RDS") %>% filter(run_number == 2)
# preprocess, filter data
df <- cbind(df[, c(1:24)],prospectr::savitzkyGolay(df[,25:ncol(df)],m = 1, p = 3, w = 17)) %>% filter(run_number == 2)
df <- df[complete.cases(df$read_age), ]


pca_temp <- pca(df[, 25:ncol(df), ])
pc_df <- data.frame(PC1 = rep(0,nrow(df)))
for (i in 1:10) {
  pc_df[, paste0("PC", i)] <- pca_temp$res$cal$scores[, i]
  rm(i)
}
pc_df <- cbind(pc_df,df)
global_lm <- lm(data = pc_df, read_age ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10)
global_gam <- gam(data = pc_df, read_age ~ s(PC1,k = 4) + s(PC2,k = 4) + s(PC3,k = 4) + s(PC4,k = 4) + s(PC5,k = 4) + s(PC6,k = 4))

# dredge to find top 5 models
options(na.action = "na.fail")
dredge_lm <- dredge(global_lm)
top5_lm <- get.models((dredge_lm), subset = 1:5)
dredge_gam <- dredge(global_gam)
top5_gam <- get.models(dredge_gam, subset = 1:5)
terms_lm <- list()
terms_gam <- list()
for(i in 1:5){
  terms_lm[[i]] <- top5_lm[[i]]$terms
  terms_gam[[i]] <- top5_gam[[i]]$formula
  rm(i)
}
rm(global_gam, global_lm, top5_gam, top5_lm, pc_df, pca_temp, dredge_gam,dredge_lm)
```

# 10-fold split

```{r message=FALSE, echo = T}
# 10 fold CV split - create folds
set.seed(6)
splits <- caret::createFolds(df$read_age, k = 10, list = TRUE, returnTrain = FALSE)
# extract PC's for each calibration set, create test sets with ages and spectra
cal <- list()
test <- list()
for (i in 1:10) {
  # calibration set and PC's
  pc.mod <- preProcess(df[-splits[[i]], -c(1:24)], method = c("pca","center"), pcaComp = 10)
  pc.cal <- predict(pc.mod, df[-splits[[i]], -c(1:24)])
  pc.cal <- cbind(pc.cal, df[-splits[[i]], ])
  cal[[i]] <- pc.cal
  rm(pc.cal)
  # test sets
  pc.test <- predict(pc.mod, df[splits[[i]], -c(1:24)])
  pc.test <- cbind(pc.test, df[splits[[i]], ])
  test[[i]] <- pc.test
  rm(pc.test, pc.mod, i)
}
```

# LM - top 5 models

```{r message=FALSE}
mods.lm <- replicate(10, list())
r2_lm <- list()
rmse_lm <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  r2_temp <- rep(0, 5)
  rmse_temp <- rep(0, 5)
  for (j in 1:5) {
    mod <- lm(data = calibrate, terms_lm[[j]])
    preds <- predict(mod, newdata = testing)
    rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
    RSS <- sum((testing$read_age - preds)^2)
    TSS <- sum((testing$read_age - mean(testing$read_age))^2)
    r2_temp[j] <- 1 - (RSS / TSS)
    mods.lm[[i]][[j]] <- mod
    rm(RSS, TSS, preds)
  }
  r2_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), r2 = r2_temp)
  rmse_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), rmse = rmse_temp)
  rm(mod, i, j, r2_temp, rmse_temp, calibrate, testing)
}
# Unlist the r2 & rmse values and calc means for 10 splits, combine into dataframe
r2_lm <- sapply(r2_lm, "[[", "r2")
r2_lm <- rowMeans(r2_lm)
rmse_lm <- sapply(rmse_lm, "[[", "rmse")
rmse_lm <- rowMeans(rmse_lm)
(lm_result <- data.frame(
  Model = paste0("Linear ", 1:5),
  Mean_R2 = r2_lm,
  Mean_RMSE = rmse_lm
))
rm(r2_lm, rmse_lm)
```

# GAM - top 5 models

```{r}
mods.gam <- replicate(10, list())
r2_gam <- list()
rmse_gam <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  r2_temp <- rep(0, 5)
  rmse_temp <- rep(0, 5)
  for (j in 1:5) {
    mod <- gam(data = calibrate, terms_gam[[j]])
    preds <- predict(mod, newdata = testing)
    rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
    RSS <- sum((testing$read_age - preds)^2)
    TSS <- sum((testing$read_age - mean(testing$read_age))^2)
    r2_temp[j] <- 1 - (RSS / TSS)
    mods.gam[[i]][[j]] <- mod
    rm(RSS, TSS, preds)
  }
  r2_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), r2 = r2_temp)
  rmse_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), rmse = rmse_temp)
  rm(j, i, testing, calibrate, mod, r2_temp, rmse_temp)
}
r2_gam <- sapply(r2_gam, "[[", "r2")
r2_gam <- rowMeans(r2_gam)
rmse_gam <- sapply(rmse_gam, "[[", "rmse")
rmse_gam <- rowMeans(rmse_gam)
gam_result <- data.frame(
  Model = paste0("GAM ", 1:5),
  Mean_R2 = r2_gam,
  Mean_RMSE = rmse_gam
)
(results <- rbind(lm_result,gam_result))
rm(r2_gam, rmse_gam, lm_result, gam_result)
```

# PLS models

VIP scores to filter wavenumbers seems to always make model fit more poorly; likely will NOT be using.

```{r}
# # 10 fold CV split - create folds
# # set.seed(6)
# # splits <- caret::createFolds(df$read_age, k = 10, list = TRUE, returnTrain = FALSE)
# # extract PC's for each calibration set, create test sets with ages and spectra
# cal <- list()
# test <- list()
# for (i in 1:10) {
#   # calibration set and PC's
#   pc.mod <- preProcess(df[-splits[[i]], -c(1:24)], method = c("pca","center"), pcaComp = 10)
#   pc.cal <- predict(pc.mod, df[-splits[[i]], -c(1:24)])
#   pc.cal <- cbind(pc.cal, df[-splits[[i]], ])
#   cal[[i]] <- pc.cal
#   rm(pc.cal)
#   # test sets
#   pc.test <- predict(pc.mod, df[splits[[i]], -c(1:24)])
#   pc.test <- cbind(pc.test, df[splits[[i]], ])
#   test[[i]] <- pc.test
#   rm(pc.test, pc.mod, i)
# }
# removed LOO CV for test set for now, it was taking forever to run
mods.pls <- list()
mods.vip <- list()
r2.pls <- list()
rmse.pls <- list()
for (i in 1:1) {
  calibrate <- cal[[i]]
  testing <- test[[i]] 
  mod <- pls(calibrate[, 35:ncol(calibrate)], calibrate[, "read_age"],
    scale = F, center = T,
    info = "Length Prediction Model",
    x.test = testing[, 35:ncol(testing)],
    y.test = testing[, "read_age"]
  )
  ncomp <- mod$ncomp.selected
  rmse.pls$pls.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
  r2.pls$pls.test[i] <- mod$testres$r2[[ncomp]] # extract r2
  
  # VIP filter of wavenumbers
  mods.pls[[i]] <- mod
  mods.pls[[1]]$ncomp.selected
  vip <- as.data.frame(vipscores(mod))
  mod <- pls(calibrate[, 35:ncol(calibrate)], calibrate[, "read_age"],
    scale = F, center = T,
    info = "Length Prediction Model",
    x.test = testing[, 35:ncol(testing)],
    y.test = testing[, "read_age"],
    exclcols = vip$V1 < 0.5
  )
  ncomp <- mod$ncomp.selected
  r2.pls$vip.test[i] <- mod$testres$r2[[ncomp]] # extract r2
  mods.vip[[i]] <- mod
  rmse.pls$vip.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
  rm(calibrate, mod, ncomp, testing, i, vip)
}
(results <- rbind(results, 
                 c("PLS", mean(r2.pls$pls.test), mean(rmse.pls$pls.test)),
                 c("VIP", mean(r2.pls$vip.test), mean(rmse.pls$vip.test))
))
rm(rmse.pls, r2.pls)
```

# ------- Model Age - structure_weight included --------

```{r}
df <- readRDS("RDS_dataframes/IBM_dfmeta.RDS") %>% filter(run_number == 2)
# preprocess, filter data
df <- cbind(df[, c(1:24)],prospectr::savitzkyGolay(df[,25:ncol(df)],m = 1, p = 3, w = 17)) %>% filter(run_number == 2)
df <- df[complete.cases(df$read_age), ]


# 10 fold CV split - create folds
set.seed(6)
splits <- caret::createFolds(df$read_age, k = 10, list = TRUE, returnTrain = FALSE)
# extract PC's for each calibration set, create test sets with ages and spectra
cal <- list()
test <- list()
for (i in 1:10) {
  # calibration set and PC's
  pc.mod <- preProcess(df[-splits[[i]], -c(1:24)], method = c("pca","center"), pcaComp = 10)
  pc.cal <- predict(pc.mod, df[-splits[[i]], -c(1:24)])
  pc.cal <- cbind(pc.cal, df[-splits[[i]], ])
  cal[[i]] <- pc.cal
  rm(pc.cal)
  # test sets
  pc.test <- predict(pc.mod, df[splits[[i]], -c(1:24)])
  pc.test <- cbind(pc.test, df[splits[[i]], ])
  test[[i]] <- pc.test
  rm(pc.test, pc.mod, i)
}



# PLS models
mods.pls <- list()
mods.vip <- list()
r2.pls <- list()
rmse.pls <- list()
for (i in 1:1) {
  calibrate <- cal[[i]]
  testing <- test[[i]] 
  mod <- pls(x = calibrate[, c(13,35:ncol(calibrate))], 
             y = calibrate[, "read_age"],
    scale = F, center = T,
    info = "Length Prediction Model",
    x.test = testing[, c(13,35:ncol(testing))],
    y.test = testing[, "read_age"]
  )
  ncomp <- mod$ncomp.selected
  rmse.pls$pls.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
  r2.pls$pls.test[i] <- mod$testres$r2[[ncomp]] # extract r2
  
  # VIP filter of wavenumbers
  mods.pls[[i]] <- mod
  mods.pls[[1]]$ncomp.selected
  vip <- as.data.frame(vipscores(mod))
  mod <- pls(x = calibrate[, c(13,35:ncol(calibrate))], 
             y = calibrate[, "read_age"],
    scale = F, center = T,
    info = "Length Prediction Model",
    x.test = testing[, c(13,35:ncol(testing))],
    y.test = testing[, "read_age"],
    exclcols = vip$V1 < 0.5
  )
  ncomp <- mod$ncomp.selected
  r2.pls$vip.test[i] <- mod$testres$r2[[ncomp]] # extract r2
  mods.vip[[i]] <- mod
  rmse.pls$vip.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
  rm(calibrate, mod, ncomp, testing, i, vip)
}
(results <- rbind(c("PLS", mean(r2.pls$pls.test), mean(rmse.pls$pls.test)),
                 c("VIP", mean(r2.pls$vip.test), mean(rmse.pls$vip.test))
))
rm(rmse.pls, r2.pls)
```

# ------- only model with length and weight -------

```{r}
df <- readRDS("RDS_dataframes/IBM_dfmeta.RDS") %>% filter(run_number == 2)
# preprocess, filter data
df <- cbind(df[, c(1:24)],prospectr::savitzkyGolay(df[,25:ncol(df)],m = 1, p = 3, w = 17)) %>% filter(run_number == 2)
df <- df[complete.cases(df$read_age), ]


mod <- lm(data = df, read_age ~ length + structure_weight)
plot(mod)
summary(mod)

ggplot(df, aes(x = length, y = read_age)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red")

ggplot(df, aes(x = structure_weight, y = read_age)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red")

mod2 <- gam(data = df, read_age ~ s(length, k = 4) + s(structure_weight, k =4))
draw(mod2)
summary(mod2)
plot(mod2)
mod2$fitted.values
mod2$model$read_age
sqrt(mean((mod2$fitted.values -mod2$model$read_age)^2)) # RMSE


ggplot(df, aes(x = length, y = read_age)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red")

ggplot(df, aes(x = structure_weight, y = read_age)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red")

# 
# 
# # VIP filter of wavenumbers
# vip <- as.data.frame(vipscores(mod))
# mod <- pls(x = calibrate[, 12:13], 
#            y = calibrate[, "read_age"],
#   scale = F, center = T,
#   info = "Length Prediction Model",
#   x.test = testing[, 12:13],
#   y.test = testing[, "read_age"],
#   exclcols = vip$V1 < 0.5
# )
# ncomp <- mod$ncomp.selected
# r2.pls$vip.test[i] <- mod$testres$r2[[ncomp]] # extract r2
# mods.vip[[i]] <- mod
# rmse.pls$vip.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
# rm(calibrate, mod, ncomp, testing, i, vip)
# 
# (results <- rbind(c("PLS", mean(r2.pls$pls.test), mean(rmse.pls$pls.test)),
#                  c("VIP", mean(r2.pls$vip.test), mean(rmse.pls$vip.test))
# ))
# rm(rmse.pls, r2.pls)
```

# ----------- OLD CODE -----------

```{r results = 'hide', message = F, warning = F}
# mod.sel <- list()
# invisible(
#   for (i in 1:10) {
#   pctest <- cal[[i]]
#   temp <- step(lm(data = pctest[-splits[[i]], ], length ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6),trace = 0)
#   mod.sel[[i]] <- rownames(summary(temp)$coef)
#   rm(pctest, temp)
# }
# )
# 
# 
# 
# knitr::kable(table(unlist(mod.sel)), align = "c")
# 
# 
# # 
# # Use PC's 1,3,4,6
# # 
# # ## MLR
# 
# mods.lm.weight <- list()
# preds.temp <- list()
# for (i in 1:10) {
#   calibrate <- cal[[i]]
#   testing <- test[[i]]
#   mod <- lm(data = calibrate, structure_weight ~ PC1 + PC3 + PC4 + PC6)
#   preds <- predict(mod, newdata = testing)
#   RMSE.weight$lm.test[i] <- caret::RMSE(pred = preds, obs = testing[, "structure_weight"])
#   RSS <- sum((testing$structure_weight - preds)^2)
#   TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
#   r2.weight$lm.test[i] <- 1 - (RSS / TSS)
#   AIC.weight$lm[i] <- AIC(mod)
#   AICc.weight$lm[i] <- AICc(mod)
#   mods.lm.weight[[i]] <- mod
#   # IBM
#   testing <- test.IBM[[i]]
#   preds.temp[[i]] <- predict(mod, newdata = testing)
#   RMSE.weight$lm.IBM[i] <- caret::RMSE(pred = preds.temp[[i]], obs = dfmeta_IBM_proc[, "structure_weight"])
#   RSS <- sum((testing$structure_weight -  preds.temp[[i]])^2)
#   TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
#   r2.weight$lm.IBM[i] <- 1 - (RSS / TSS)
#   rm(mod, preds, RSS, TSS)
# }
# preds.IBM$lm <- preds.temp
# rm(preds.temp)
# mod.summary.weight[1,2] <- round(mean(r2.weight$lm.test), 3)
# mod.summary.weight[1,3] <- round(mean(RMSE.weight$lm.test), 3)
# mod.summary.weight[1,4] <- round(mean(AIC.weight$lm) , 3)
# mod.summary.weight[1,5] <- round(mean(AICc.weight$lm) , 3)
# mod.summary.weight[5,2] <- round(mean(r2.weight$lm.IBM), 3)
# mod.summary.weight[5,3] <- round(mean(RMSE.weight$lm.IBM), 3)
# 
# 
# ## GAM
# 
# # GAM with 10 fold CV, select = T allows PCs to be penalized and effectively removed from model if appropriate.
# mods.gam.weight <- list()
# preds.temp <- list()
# for (i in 1:10) {
#   calibrate <- cal[[i]]
#   testing <- test[[i]]
#   mod <- gam(data = calibrate, structure_weight ~ s(PC1) + s(PC2) + s(PC3) + s(PC4) + s(PC5) + s(PC6), method = "REML")
#   preds <- predict(mod, newdata = testing)
#   RSS <- sum((testing$structure_weight - preds)^2)
#   TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
#   RMSE.weight$gam.test[i] <- caret::RMSE(pred = preds, obs = testing[, "structure_weight"])
#   r2.weight$gam.test[i] <- 1 - (RSS / TSS)
#   AIC.weight$gam[i] <- AIC(mod)
#   AICc.weight$gam[i] <- AICc(mod)
#   mods.gam.weight[[i]] <- mod
#   # IBM
#   testing <- test.IBM[[i]]
#   preds.temp[[i]] <- predict(mod, newdata = testing)
#   RSS <- sum((testing$structure_weight - preds.temp[[i]])^2)
#   TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
#   RMSE.weight$gam.IBM[i] <- caret::RMSE(pred = preds.temp[[i]], obs = dfmeta_IBM_proc[, "structure_weight"])
#   r2.weight$gam.IBM[i] <- 1 - (RSS / TSS)
#   rm(mod, RSS, TSS, i, calibrate, testing)
# }
# preds.IBM$gam <- preds.temp
# rm(preds.temp)
# mod.summary.weight[2,2] <- round(mean(r2.weight$gam.test), 3)
# mod.summary.weight[2,3] <- round(mean(RMSE.weight$gam.test), 3)
# mod.summary.weight[2,4] <- round(mean(AIC.weight$gam), 3)
# mod.summary.weight[2,5] <- round(mean(AICc.weight$gam), 3)
# mod.summary.weight[6,2] <- round(mean(r2.weight$gam.IBM), 3)
# mod.summary.weight[6,3] <- round(mean(RMSE.weight$gam.IBM), 3)
# rm(AIC.weight, AICc.weight)
# 
# ## PLS
# 
# 
# mods.pls.weight <- list()
# mods.vip.weight <- list()
# preds.temp1 <- list()
# preds.temp2 <- list()
# # pls, no variable selection & VIP > 1
# for (i in 1:10) {
#   calibrate <- dfmeta_LPW_proc[-splits[[i]], ]
#   testing <- dfmeta_LPW_proc[splits[[i]], ]
#   mod <- pls(calibrate[, 21:ncol(calibrate)], calibrate[, "structure_weight"],
#     ncomp.selcrit = "wold", scale = F, center = T, cv = 1,
#     info = "Otolith Weight Prediction Model",
#     x.test = testing[, 21:ncol(testing)],
#     y.test = testing[, "structure_weight"]
#   )
#   ncomp <- mod$ncomp.selected
#   RMSE.weight$pls.test[i] <- mod$testres$rmse[[ncomp]]
#   r2.weight$pls.test[i] <- mod$testres$r2[[ncomp]]
#   mods.pls.weight[[i]] <- mod
#   # IBM
#   testing <- test.IBM[[i]]
#   preds.temp1[[i]] <- predict(
#     mod,
#     dfmeta_IBM_proc[, c(25:ncol(dfmeta_IBM_proc))],
#     dfmeta_IBM_proc[, 3]
#   )
#   ncomp <- preds.temp1[[i]]$ncomp.selected
#   RMSE.weight$pls.IBM[i] <- preds.temp1[[i]]$rmse[ncomp]
#   r2.weight$pls.IBM[i] <- preds.temp1[[i]]$r2[ncomp]
#   ### VIP > 1
#   vip <- as.data.frame(vipscores(mod))
#   testing <- dfmeta_LPW_proc[splits[[i]], ]
#   mod <- pls(calibrate[, 21:ncol(calibrate)], calibrate[, "structure_weight"],
#     ncomp.selcrit = "wold", scale = F, center = T, cv = 1,
#     info = "Otolith Weight Prediction Model",
#     x.test = testing[, 21:ncol(testing)],
#     y.test = testing[, "structure_weight"],
#     exclcols = vip$V1 < 1
#   )
#   ncomp <- mod$ncomp.selected
#   ncomp <- mod$ncomp.selected
#   RMSE.weight$vip.test[i] <- mod$testres$rmse[[ncomp]]
#   r2.weight$vip.test[i] <- mod$testres$r2[[ncomp]]
#   mods.vip.weight[[i]] <- mod
#   # IBM
#   testing <- test.IBM[[i]]
#   preds.temp2[[i]] <- predict(
#     mod,
#     dfmeta_IBM_proc[, c(25:ncol(dfmeta_IBM_proc))],
#     dfmeta_IBM_proc[, 3]
#   )
#   ncomp <- preds.temp2[[i]]$ncomp.selected
#   RMSE.weight$vip.IBM[i] <- preds.temp2[[i]]$rmse[ncomp]
#   r2.weight$vip.IBM[i] <- preds.temp2[[i]]$r2[ncomp]
#   rm(calibrate, mod, ncomp, testing, i, vip)
# }
# preds.IBM$pls <- preds.temp1
# preds.IBM$vip <- preds.temp2
# rm(preds.temp1, preds.temp2)
# mod.summary.weight[3, 2] <- round(mean(r2.weight$pls.test), 3)
# mod.summary.weight[3, 3] <- round(mean(RMSE.weight$pls.test), 3)
# mod.summary.weight[4, 2] <- round(mean(r2.weight$vip.test), 3)
# mod.summary.weight[4, 3] <- round(mean(RMSE.weight$vip.test), 3)
# mod.summary.weight[7, 2] <- round(mean(r2.weight$pls.IBM), 3)
# mod.summary.weight[7, 3] <- round(mean(RMSE.weight$pls.IBM), 3)
# mod.summary.weight[8, 2] <- round(mean(r2.weight$vip.IBM), 3)
# mod.summary.weight[8, 3] <- round(mean(RMSE.weight$vip.IBM), 3)
# knitr::kable(mod.summary.weight, align = "c")
# rm(r2.weight,RMSE.weight)
# 
# 
# 
# # Results
# 
# 
# ### Plots
# 
# 
# preds.IBM$structure_weight <- dfmeta_IBM_proc$structure_weight
# preds.IBM$pcs <- list()
# for(i in 1:10){
#   preds.IBM$pcs[[i]] <- test.IBM[[i]][1:10]
# }
# # lm predictions
# ggplot() + 
#   geom_point(aes(x = preds.IBM$lm[[1]], y = preds.IBM$y)) +
#   geom_abline(aes(slope = 1, intercept = 0))
# # GAM
# ggplot() + 
#   geom_point(aes(x = preds.IBM$gam[[1]], y = preds.IBM$y)) +
#   geom_abline(aes(slope = 1, intercept = 0))
# # PLS
# ncomp <- preds.IBM$pls[[1]]$ncomp.selected
# 
# ggplot() +
#   geom_point(aes(x = preds.IBM$pls[[1]]$y.pred[,ncomp,1], 
#                  y = preds.IBM$pls[[1]]$y.ref), col = "red") + 
#   scale_y_continuous(limits = c(0,0.004)) +
#   scale_x_continuous(limits = c(-0.01, 0.005)) + 
#   geom_abline(aes(slope = 1, intercept = 0))
#   
# 
# 
# ggplot() +
#   geom_smooth(aes(x = preds.IBM$pls[[1]]$y.pred[,ncomp,1], 
#                  y = preds.IBM$pls[[1]]$y.ref), 
#               method = "lm", se = T,
#               col = "red") + 
#   scale_y_continuous(limits = c(0,0.004)) +
#   scale_x_continuous(limits = c(-0.01, 0.005)) + 
#   geom_abline(aes(slope = 1, intercept = 0))
# # VIP
# ggplot() + 
#   geom_point(aes(x = preds.IBM$vip[[1]]$y.ref, y = preds.IBM$y)) +
#   geom_abline(aes(slope = 1, intercept = 0))
# 
# 
#   geom_smooth(method = "lm", formula = structure_weight ~ PC1 + PC3 + PC4 + PC6, mapping = aes(x = ))
# plot_lm
# 
# mods.lm.weight[[1]]$terms
# coef(mods.lm.weight[[1]])
# 
# preds.IBM$pls[[1]]$y.pred[,ncomp,1]
# 
# 
# # try log link function for GAM
# 
# r2.weight <- list()
# RMSE.weight <- list()
# for (i in 1:10) {
#   calibrate <- cal[[i]]
#   testing <- test[[i]]
#   mod <- gam(data = calibrate, structure_weight ~ s(PC1) + s(PC2) + s(PC3) + s(PC4) + s(PC5) + s(PC6), method = "REML")
#   preds <- predict(mod, newdata = testing)
#   RSS <- sum((testing$structure_weight - preds)^2)
#   TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
#   RMSE.weight$gam.test[i] <- caret::RMSE(pred = preds, obs = testing[, "structure_weight"])
#   r2.weight$gam.test[i] <- 1 - (RSS / TSS)
#   AIC.weight$gam[i] <- AIC(mod)
#   AICc.weight$gam[i] <- AICc(mod)
#   mods.gam.weight[[i]] <- mod
#   # IBM
#   testing <- test.IBM[[i]]
#   preds.temp[[i]] <- predict(mod, newdata = testing)
#   RSS <- sum((testing$structure_weight - preds.temp[[i]])^2)
#   TSS <- sum((testing$structure_weight - mean(testing$structure_weight))^2)
#   RMSE.weight$gam.IBM[i] <- caret::RMSE(pred = preds.temp[[i]], obs = dfmeta_IBM_proc[, "structure_weight"])
#   r2.weight$gam.IBM[i] <- 1 - (RSS / TSS)
#   rm(mod, RSS, TSS, i, calibrate, testing)
# }

```
