---
title: "ICP-MS Data Filtering & Analysis"
output: 
  html_document:
    code_folding: hide
---
# Packages
```{r fig.width=12, fig.height=8, results='asis', warning=FALSE, message=FALSE}
# Setup and Package Loading
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)

if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(
  broom,      # for tidying statistical test results
  corrplot,   # nice correlation plots
  DT,         # for interactive tables
  htmltools,  # for HTML widget manipulation
  kableExtra, # for formatting tables
  knitr,      # for knitting RMarkdown
  plotly,     # for interactive plots
  scales,     # for scale functions in ggplot2
  shiny,      # shiny apps
  tidyverse,  # for data manipulation and visualization
  vegan,      # various modeling tools
  viridis,    # for color palettes
  zoo         # for rolling functions
)

theme_set(theme_minimal(base_size = 14))
```
# Load Data
Load in data, filter to remove reference materials, assign otolith number, transect number, region variables
```{r}
# Create lookup table (slide number to specimen number)
lookup <- tribble(
  ~slide_num, ~specimen_number,
  1, 425957, 2, 425900, 3, 425925, 4, 425959, 5, 424584,
  6, 424289, 7, 424295, 8, 426162, 9, 424251, 10, 424333,
  11, 425940, 12, 424564, 13, 425999, 14, 426076, 15, 424568,
  16, 424177, 17, 424178, 18, 424182, 19, 426153, 20, 426216,
  21, 424195, 22, 424222, 23, 424224, 24, 424230, 25, 424232,
  26, 424595, 27, 424597, 28, 425915, 29, 425919, 30, 425990,
  31, 424184, 32, 424327, 33, 424339, 34, 426129, 35, 426165,
  36, 424527, 37, 424547, 38, 425907, 39, 425945, 40, 426007
)

# Load Data and Preprocess
df_ICP <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")
valid_indices <- setdiff(30:184, c(grep("oto10-2", names(df_ICP)), grep("oto30-", names(df_ICP))))

df_ICP <- df_ICP[valid_indices] %>% 
  bind_rows(.id = "transect_id") %>%
  mutate(
    otolith_id = str_extract(transect_id, "oto\\d+"),
    slide_num = as.integer(str_extract(otolith_id, "\\d+")),
    transect_num = as.integer(str_extract(transect_id, "\\d+$")),
    region = case_when(
      slide_num %in% c(6:10, 16:25, 31:35) ~ "Kodiak",
      TRUE ~ "Shumagin"
    )
  ) %>%
  left_join(lookup, by = "slide_num") %>%
  select(-slide_num)

# sort variable names
df_ICP <- df_ICP %>%
  select(
    otolith_id, specimen_number, transect_num, transect_id, region, `Elapsed Time`,
    sort(setdiff(names(df_ICP), c("otolith_id", "specimen_number", "transect_num", "transect_id", "region", "Elapsed Time"))) 
  ) %>%
  rename(time = `Elapsed Time`)

rm(lookup, valid_indices)
```
# Data Filtering
This section implements an improved method for detecting the boundaries of the otolith in each transect. It uses a rolling median to smooth the Calcium-43 data and then identifies the start and end points of the otolith based on a threshold relative to the baseline concentration.

### Filter df

```{r}
# After getting manual_adjustments.csv:
manual_adj <- read.csv("metadata/manual_adjustments.csv")

# Create the segmented dataset incorporating manual adjustments
df_ICP_filt <- df_ICP %>%
  group_by(transect_id) %>%
  mutate(
    row_num = row_number(),
    # Check if this transect has manual adjustments
    has_manual_adj = transect_id %in% manual_adj$transect_id,
    # Get start point (either manual or quantile-based)
    start = if_else(
      has_manual_adj,
      manual_adj$manual_start[match(transect_id, manual_adj$transect_id)],
      min(which(Ca43 >= quantile(Ca43, 0.6, na.rm = TRUE)))
    ),
    # Get end point (either manual or quantile-based)
    end = if_else(
      has_manual_adj,
      manual_adj$manual_end[match(transect_id, manual_adj$transect_id)],
      max(which(Ca43 >= quantile(Ca43, 0.6, na.rm = TRUE)))
    )
  ) %>%
  filter(row_num >= start & row_num <= end) %>%
  group_by(transect_id) %>%
  mutate(
    total_rows = n(),
    segment = case_when(
      row_number() <= 0.2 * total_rows ~ "core",
      row_number() > 0.8 * total_rows ~ "edge",
      TRUE ~ "middle"
    )
  ) %>%
  select(-has_manual_adj, -row_num, -start, -end, -total_rows) %>%
  ungroup()
rm(manual_adj)
```

### ++++ Shiny: mark cutoffs ++++
Manually mark cutoffs for problem specimens.  Displays current cutoffs in blue, allows you to mark new cutoffs in red and save in a csv.

```{r eval=FALSE, include=FALSE}
ui <- fluidPage(
  titlePanel("Otolith Transect Manual Adjustment"),

  sidebarLayout(
    sidebarPanel(
      # Select transect
      selectInput("transect", "Select Transect ID:",
                  choices = unique(df_ICP$transect_id)),

      # Sliders for start and end points
      sliderInput("start_point", "Start Point:",
                  min = 1, max = 100, value = 1),
      sliderInput("end_point", "End Point:",
                  min = 1, max = 100, value = 100),

      # Save button
      actionButton("save", "Save Cutoffs"),

      # Download all adjustments
      downloadButton("download", "Download All Adjustments")
    ),

    mainPanel(
      plotOutput("transect_plot", height = "600px"),
      verbatimTextOutput("saved_values")
    )
  )
)

server <- function(input, output, session) {
  # Reactive values to store manual adjustments
  manual_adjustments <- reactiveVal(data.frame(
    transect_id = character(),
    manual_start = numeric(),
    manual_end = numeric(),
    stringsAsFactors = FALSE
  ))

  # Update slider ranges when transect is selected
  observe({
    req(input$transect)
    transect_data <- df_ICP %>% filter(transect_id == input$transect)
    n_points <- nrow(transect_data)

    updateSliderInput(session, "start_point",
                      min = 1, max = n_points,
                      value = 1)
    updateSliderInput(session, "end_point",
                      min = 1, max = n_points,
                      value = n_points)
  })

  # Plot
  output$transect_plot <- renderPlot({
    req(input$transect)

    transect_data <- df_ICP %>% filter(transect_id == input$transect)
    
    # Get the current cutoffs from df_segmented
    current_cutoffs <- df_ICP_filt %>% 
      filter(transect_id == input$transect) %>% 
      summarise(start_time = min(time), end_time = max(time))

    ggplot(transect_data, aes(x = time, y = Ca43)) +
      geom_line() +
      geom_vline(xintercept = current_cutoffs$start_time, color = "blue", linetype = "dashed") +
      geom_vline(xintercept = current_cutoffs$end_time, color = "blue", linetype = "dashed") +
      geom_vline(xintercept = transect_data$time[input$start_point], color = "red", linetype = "dashed") +
      geom_vline(xintercept = transect_data$time[input$end_point], color = "red", linetype = "dashed") +
      labs(title = paste("Transect:", input$transect),
           x = "Elapsed Time",
           y = "Ca43") +
      theme_minimal()
  })

  # Save adjustments
  observeEvent(input$save, {
    current_adjustments <- manual_adjustments()
    new_adjustment <- data.frame(
      transect_id = input$transect,
      manual_start = transect_data$time[input$start_point],
      manual_end = transect_data$time[input$end_point],
      stringsAsFactors = FALSE
    )

    # Update or add new adjustment
    if(input$transect %in% current_adjustments$transect_id) {
      current_adjustments[current_adjustments$transect_id == input$transect,] <- new_adjustment
    } else {
      current_adjustments <- rbind(current_adjustments, new_adjustment)
    }

    manual_adjustments(current_adjustments)
  })

  # Display saved adjustments
  output$saved_values <- renderPrint({
    adjustments <- manual_adjustments()
    if(nrow(adjustments) == 0) {
      "No adjustments saved yet."
    } else {
      adjustments
    }
  })

  # Download handler
  output$download <- downloadHandler(
    filename = function() {
      "manual_adjustments.csv"
    },
    content = function(file) {
      write.csv(manual_adjustments(), file, row.names = FALSE)
    }
  )
}

shinyApp(ui = ui, server = server)
```

### Average per otolith "zone" (core, middle, edge)

```{r}

# List of elemental variables
elemental_vars <- names(df_ICP_filt)[which(names(df_ICP_filt) == "Ca43"):which(names(df_ICP_filt) == "Zn66_ppm")]

# Create the new data frame
df_ICP_avg <- df_ICP_filt %>%
  group_by(specimen_number, segment) %>%
  summarise(across(all_of(elemental_vars), mean, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(
    id_cols = specimen_number,
    names_from = segment,
    values_from = all_of(elemental_vars),
    names_glue = "{.value}_{segment}"
  )

# Add region information
df_ICP_avg <- df_ICP_filt %>%
  group_by(specimen_number) %>%
  summarise(region = first(region)) %>%
  right_join(df_ICP_avg, by = "specimen_number")

# Reorder columns to have specimen_number and region first
df_ICP_avg <- df_ICP_avg %>%
  select(specimen_number, region, everything())
rm(elemental_vars)
#saveRDS(df_ICP_avg,"RDS_dataframes/ICP-MS_avg.RDS")
```


# Combine with FT-NIRS Data
Import raw data, preprocess with SG, combine with elemental data.
```{r}
# Load and process FT-NIRS data
df_FTNIRS <- readRDS("RDS_dataframes/IBM_dfmeta.RDS") %>% filter(run_number == 2)
df_FTNIRS <- cbind(df_FTNIRS[, c(1:24)], prospectr::savitzkyGolay(df_FTNIRS[,25:ncol(df_FTNIRS)], m = 1, p = 3, w = 17))

# Step 1: Select only the wavenumber columns and the specimen number from FT-NIRS data
ft_nirs_wavenumbers <- df_FTNIRS %>%
  select(specimen, length, matches("^\\d+$"))  # This selects 'specimen' and all columns that are just numbers

# Step 2: Join df_specimen_averages with FT-NIRS data
df_combined <- df_ICP_avg %>%
  left_join(ft_nirs_wavenumbers, by = c("specimen_number" = "specimen"))

# saveRDS(df_combined, "RDS_dataframes/ICP-MS_combined")

rm(df_ICP,ft_nirs_wavenumbers)
```

# Code to run analysis on line transects?

```{r}
library(tidyverse)

# Function to calculate the position along a line
calculate_position_along_line <- function(point, start, end) {
  # Calculate vector from start to end
  line_vector <- c(end$x - start$x, end$y - start$y)
  
  # Calculate vector from start to point
  point_vector <- c(point$x - start$x, point$y - start$y)
  
  # Calculate projection of point vector onto line vector
  projection <- sum(point_vector * line_vector) / sum(line_vector^2)
  
  # Return the proportion along the line (0 = start, 1 = end)
  return(max(0, min(1, projection)))
}

# Function to process annotation data and return time ranges
process_annotations <- function(line_data, max_time) {
  start_point <- list(x = line_data$start_x, y = line_data$start_y)
  end_point <- list(x = line_data$end_x, y = line_data$end_y)
  
  # Create points for each annotation (handling NAs)
  result <- list()
  
  # Process standard annotations
  for (annotation_type in c("start", "core", "end")) {
    x_col <- paste0(annotation_type, "_annotation_x")
    y_col <- paste0(annotation_type, "_annotation_y")
    
    if (!is.na(line_data[[x_col]]) && !is.na(line_data[[y_col]])) {
      point <- list(x = line_data[[x_col]], y = line_data[[y_col]])
      pos <- calculate_position_along_line(point, start_point, end_point)
      result[[paste0(annotation_type, "_time")]] <- pos * max_time
    } else {
      result[[paste0(annotation_type, "_time")]] <- NA
    }
  }
  
  # Process data quality annotations
  for (annotation_type in c("data_start", "data_end")) {
    x_col <- paste0(annotation_type, "_annotation_x")
    y_col <- paste0(annotation_type, "_annotation_y")
    
    if (!is.na(line_data[[x_col]]) && !is.na(line_data[[y_col]])) {
      point <- list(x = line_data[[x_col]], y = line_data[[y_col]])
      pos <- calculate_position_along_line(point, start_point, end_point)
      result[[paste0(annotation_type, "_time")]] <- pos * max_time
    } else {
      # If data quality markers aren't set, use otolith start/end as defaults
      if (annotation_type == "data_start" && !is.na(result$start_time)) {
        result$data_start_time <- result$start_time
      } else if (annotation_type == "data_end" && !is.na(result$end_time)) {
        result$data_end_time <- result$end_time
      } else {
        result[[paste0(annotation_type, "_time")]] <- NA
      }
    }
  }
  
  # Add quality information
  result$quality <- line_data$quality
  result$notes <- line_data$notes
  
  return(result)
}

# Main function to filter and process your data
# Modified process_icp_data function
process_icp_data <- function(df_ICP, annotations) {
  # Check if annotations is a filename or dataframe
  if (is.character(annotations) && length(annotations) == 1) {
    # It's a filename
    annotations <- read.csv(annotations)
  } else if (!is.data.frame(annotations)) {
    stop("annotations must be either a filename or a data frame")
  }
  
  # Create a data frame to store the time ranges for each line
  time_ranges <- data.frame(
    line_name = character(),
    quality = character(),
    notes = character(),
    otolith_start_time = numeric(),
    core_time = numeric(),
    otolith_end_time = numeric(),
    data_start_time = numeric(),
    data_end_time = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Process each line
  for (i in 1:nrow(annotations)) {
    # Get the line data
    line_data <- annotations[i, ]
    
    # Determine the maximum time for this line
    line_id <- line_data$line_name
    
    # Skip if this line isn't in the ICP data
    if (!line_id %in% df_ICP$transect_id) {
      warning(paste("Line", line_id, "not found in ICP data, skipping"))
      next
    }
    
    max_time <- max(df_ICP$time[df_ICP$transect_id == line_id])
    
    # Process the annotations
    result <- process_annotations(line_data, max_time)
    
    # Add to the time_ranges data frame
    time_ranges <- rbind(time_ranges, data.frame(
      line_name = line_id,
      quality = result$quality,
      notes = result$notes,
      otolith_start_time = result$start_time,
      core_time = result$core_time,
      otolith_end_time = result$end_time,
      data_start_time = result$data_start_time,
      data_end_time = result$data_end_time,
      stringsAsFactors = FALSE
    ))
  }
  
  # Create a list to store filtered data frames
  filtered_data <- list()
  
  # Process each line
  for (i in 1:nrow(time_ranges)) {
    # Get the time range data
    range_data <- time_ranges[i, ]
    
    # Skip poor quality transects
    if (range_data$quality == "poor") {
      message(paste("Skipping poor quality line:", range_data$line_name))
      next
    }
    
    # Filter the data frame
    line_data <- df_ICP[df_ICP$transect_id == range_data$line_name, ]
    
    # Only use data between data_start and data_end
    if (!is.na(range_data$data_start_time) && !is.na(range_data$data_end_time)) {
      filtered_line_data <- line_data[
        line_data$time >= range_data$data_start_time & 
          line_data$time <= range_data$data_end_time, 
      ]
      
      # Skip if no data points remain after filtering
      if (nrow(filtered_line_data) == 0) {
        message(paste("No data points remain after filtering for line:", range_data$line_name))
        next
      }
      
      # Add columns with position information
      filtered_line_data$position_from_core <- filtered_line_data$time - range_data$core_time
      filtered_line_data$transect_quality <- range_data$quality
      
      # Store in the list
      filtered_data[[range_data$line_name]] <- filtered_line_data
    } else {
      message(paste("Missing data start/end markers for line:", range_data$line_name))
    }
  }
  
  # Combine all filtered data
  if (length(filtered_data) > 0) {
    all_filtered_data <- bind_rows(filtered_data)
    
    # Add normalized position information
    all_filtered_data <- all_filtered_data %>%
      group_by(transect_id) %>%
      mutate(
        # Calculate distances as percentages
        distance_from_core_pct = case_when(
          position_from_core >= 0 ~ 
            position_from_core / (
              time_ranges$otolith_end_time[match(transect_id, time_ranges$line_name)] - 
                time_ranges$core_time[match(transect_id, time_ranges$line_name)]
            ) * 100,
          TRUE ~ 
            position_from_core / (
              time_ranges$core_time[match(transect_id, time_ranges$line_name)] - 
                time_ranges$otolith_start_time[match(transect_id, time_ranges$line_name)]
            ) * -100
        ),
        # Create region labels
        otolith_region = case_when(
          distance_from_core_pct < -50 ~ "Inner Core",
          distance_from_core_pct < 0 ~ "Core",
          distance_from_core_pct < 50 ~ "Mid-Radius",
          TRUE ~ "Edge"
        )
      ) %>%
      ungroup()
    
    return(list(
      filtered_data = all_filtered_data,
      time_ranges = time_ranges
    ))
  } else {
    warning("No valid data remained after filtering")
    return(NULL)
  }
}

# Add transect_id to your dataframe
df$transect_id <- paste0(df$specimen, "-", df$transect)

# Process your data with the fixed function
results <- process_icp_data(df, allotos)
  
  specimens <- c("oto05", "oto10", "oto12", "oto14", "oto18", "oto19", "oto20",
                          "oto21", "oto22", "oto23", "oto24", "oto25",
                          "oto26", "oto27", "oto28", "oto29", "oto30",
                          "oto31", "oto32", "oto33", "oto34", "oto35",
                          "oto36", "oto37", "oto38", "oto39", "oto40"
)

temp <- results$filtered_data %>% filter(specimen == "oto13", transect_quality == "good")
temp <- temp %>% filter(time < 103)

ggplot(temp, aes(x = time, y = Mn55_ppm, color = transect_id)) + 
  geom_line()



# Create transect_id column to match line_name in annotations
df$transect_id <- paste0(df$specimen, "-", df$transect)


allotos <- read.csv("allotos.csv")
# Process your data with the fixed function
results <- process_icp_data(df, allotos)  # Note: passing the dataframe directly, not a filename

# Access the filtered data
filtered_data <- results$filtered_data

# Plot example element profile
ggplot(filtered_data, aes(x = distance_from_core_pct, y = Sr88_ppm, color = transect_id)) +
  geom_line() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~otolith_region, scales = "free") +
  theme_minimal() +
  labs(title = "Sr88 Concentration by Otolith Region", 
       x = "Distance from Core (%)", 
       y = "Sr88 (ppm)")
```



# Vegan & CCora 
ICP-MS and FT-NIRS Data combined

```{r}
# Select FT-NIRS data (wavenumbers)
ccora_y <- df_combined %>% 
  select(matches("^\\d+$")) %>% # Select columns that are just numbers
  as.matrix()

# Select ICP-MS data
ccora_x <- df_combined %>% 
  select(ends_with("_core"), ends_with("_edge"), ends_with("_middle")) %>%
  select(-starts_with("Ca43")) %>% # Exclude Ca43 if you don't want it in the analysis
  as.matrix()

# Perform PCA on FT-NIRS data
pca_ftnirs <- prcomp(ccora_y, scale. = TRUE)
# Determine the number of PCs to retain (explaining 95% of variance)
var_explained <- cumsum(pca_ftnirs$sdev^2 / sum(pca_ftnirs$sdev^2))

# Extract the PC scores for the selected number of components
ftnirs_pcs <- pca_ftnirs$x[, 1:10]
print(paste("Number of PCs retained:", 10))
print(paste("Variance explained:", round(var_explained[10] * 100, 2), "%"))

cca_result <- CCorA(X = ccora_x, Y = ftnirs_pcs)

# Print summary
print(summary(cca_result))

# Extract and print canonical correlations
can_corr <- cca_result$CanCorr
print("Canonical Correlations:")
print(can_corr)

# Plot canonical correlations
barplot(can_corr, 
        names.arg = 1:length(can_corr),
        xlab = "Canonical Dimension",
        ylab = "Canonical Correlation",
        main = "Canonical Correlations",
        col = "skyblue")
abline(h = 0.5, col = "red", lty = 2)

# Print top correlations for the first few canonical variates
print("Top 5 ICP-MS correlations with first 3 canonical variates:")
icp_ms_correlations <- cca_result$corr.X.Cx
print(head(icp_ms_correlations[order(-abs(icp_ms_correlations[,1])),1:3], 5))

print("Top 5 FT-NIRS PC correlations with first 3 canonical variates:")
ftnirs_pc_correlations <- cca_result$corr.Y.Cy
print(head(ftnirs_pc_correlations[order(-abs(ftnirs_pc_correlations[,1])),1:3], 5))
```

### Correlation heatmaps:

```{r}
library(reshape2)
library(pheatmap)


# Use the full correlation matrix
cor_matrix <- cor(ccora_x, ftnirs_pcs)

# Get element names and sort them alphabetically
element_names <- rownames(cor_matrix)
sorted_elements <- sort(element_names, method = "radix")

# Reorder the correlation matrix rows
cor_matrix_sorted <- cor_matrix[sorted_elements, ]

# Create heatmap of all correlations
pheatmap(cor_matrix_sorted,
         main = "Correlations: ICP-MS vs FT-NIRS PCs",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_row = 6,
         fontsize_col = 8,
         angle_col = 45)

```

### Correlation Network

```{r}
library(igraph)
library(ggraph)
library(tidygraph)

# Prepare the data
cor_df <- as.data.frame(as.table(cor_matrix))
names(cor_df) <- c("Element", "PC", "Correlation")

# Filter for strong correlations
cor_threshold <- 0.3  # Adjust this threshold as needed
cor_df_filtered <- cor_df[abs(cor_df$Correlation) > cor_threshold, ]

# Create a graph
g <- graph_from_data_frame(cor_df_filtered, directed = FALSE)

# Create a layout
layout <- create_layout(g, layout = "fr")

# Plot
ggraph(layout) +
  geom_edge_link(aes(edge_alpha = abs(Correlation), edge_width = abs(Correlation), color = Correlation)) +
  geom_node_point(aes(color = name %in% unique(cor_df_filtered$PC)), size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_edge_colour_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  scale_edge_alpha(range = c(0.1, 1)) +
  scale_edge_width(range = c(0.5, 2)) +
  scale_color_manual(values = c("orange", "purple"), guide = "none") +
  theme_void() +
  labs(title = "Correlation Network: ICP-MS Elements vs FT-NIRS PCs",
       edge_alpha = "Absolute Correlation",
       edge_width = "Absolute Correlation",
       edge_color = "Correlation")

# If you want to save the plot
# ggsave("correlation_network.png", width = 12, height = 10, dpi = 300)
```

### Dendrogram/Hierarchical Cluster

```{r}
library(dendextend)

# Perform hierarchical clustering
hc <- hclust(dist(cor_matrix))

# Create dendrogram
dend <- as.dendrogram(hc)

# Customize the dendrogram
dend <- set(dend, "labels_cex", 0.6)  # Reduce label size
dend <- set(dend, "branches_lwd", 1.5)  # Increase branch line width

# Color branches (this should work in most versions)
dend <- color_branches(dend, k = 5)  # Color branches for 5 main clusters (adjust as needed)

# Plot the dendrogram
par(mar = c(5, 3, 3, 20))  # Adjust margins to make room for labels
plot(dend, 
     main = "Hierarchical Clustering of Elements",
     horiz = TRUE,  # Horizontal orientation
     nodePar = list(cex = .007),
     axes = FALSE,  # Remove axes
     leaflab = "perpendicular")  # Make leaf labels perpendicular to the branches

# Add a scale bar
axis(side = 1, line = -1.5, cex.axis = 0.8)

```

### CIRCLE CORRELATION CHORD

```{r}
library(circlize)

# Prepare data for chord diagram
chord_data <- cor_matrix
chord_data[abs(chord_data) < 0.4] <- 0  # Set weak correlations to 0

# Get the actual names from the data and sort them
actual_vars <- union(rownames(chord_data), colnames(chord_data))
sorted_vars <- sort(actual_vars)

# Clear any existing circular plot
circos.clear()

# Create chord diagram with sorted variables
chordDiagram(chord_data, 
             transparency = 0.5,
             col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
             annotationTrack = "grid", # This removes the default labels
             preAllocateTracks = list(track.height = 0.1),
             order = sorted_vars) # Use the sorted order here

title("Correlation Chord Diagram")

# Add the links
circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
  xlim = get.cell.meta.data("xlim")
  ylim = get.cell.meta.data("ylim")
  sector.name = get.cell.meta.data("sector.index")
  circos.text(mean(xlim), ylim[1], sector.name, facing = "clockwise",
              niceFacing = TRUE, cex = 0.6, srt = 0)
}, bg.border = NA)

```



### random plotting code for CCora

```{r}
# Get the loadings for the first 10 PCs
pc_loadings <- pca_ftnirs$rotation[, 1:10]

# Function to get top contributing wavelengths
top_wavelengths <- function(loadings, n = 5) {
  top_indices <- order(abs(loadings), decreasing = TRUE)[1:n]
  return(data.frame(
    wavelength = as.numeric(names(loadings)[top_indices]),
    loading = loadings[top_indices]
  ))
}

# Print top contributing wavelengths for PC3, PC8, and PC9
print("Top 5 contributing wavelengths for PC3:")
print(top_wavelengths(pc_loadings[,3]))

print("Top 5 contributing wavelengths for PC8:")
print(top_wavelengths(pc_loadings[,8]))

print("Top 5 contributing wavelengths for PC9:")
print(top_wavelengths(pc_loadings[,9]))




# Extract scores
x_scores <- cca_result$Cx[, 1:6]
y_scores <- cca_result$Cy[, 1:6]

# Create a scatter plot
plot(x_scores[,5], y_scores[,5], 
     xlab = "First Canonical Variate (ICP-MS)", 
     ylab = "First Canonical Variate (FT-NIRS PCs)",
     main = "First vs Second Canonical Variate",
     pch = 19, col = "blue")

# Add labels for each point
text(x_scores[,1], y_scores[,1], labels = rownames(ccora_x), pos = 3, cex = 0.7)

# Add a diagonal line for reference
abline(0, 1, col = "red", lty = 2)



print("Top 5 ICP-MS correlations with second canonical variate:")
print(head(icp_ms_correlations[order(-abs(icp_ms_correlations[,2])),2, drop = FALSE], 5))

print("Top 5 FT-NIRS PC correlations with second canonical variate:")
print(head(ftnirs_pc_correlations[order(-abs(ftnirs_pc_correlations[,2])),2, drop = FALSE], 5))

print("Top 5 ICP-MS correlations with third canonical variate:")
print(head(icp_ms_correlations[order(-abs(icp_ms_correlations[,3])),3, drop = FALSE], 5))

print("Top 5 FT-NIRS PC correlations with third canonical variate:")
print(head(ftnirs_pc_correlations[order(-abs(ftnirs_pc_correlations[,3])),3, drop = FALSE], 5))



```

# ------MODELING --------
### Quick modelling elemental data

```{r}
library(dplyr)
library(prospectr)
library(mgcv)
library(MuMIn)
library(caret)
library(pls)

model_element <- function(df, element) {
  # Preprocess data
  df <- df %>% 
    select(all_of(element), everything()) %>%
    filter(complete.cases(.))
  spec_cols <- 70:ncol(df)
  # Split data
  set.seed(123)
  train_index <- createDataPartition(df[[element]], p = 0.6, list = FALSE)
  train_data <- df[train_index, ]
  test_data <- df[-train_index, ]
  
  # PCA
  pca_model <- preProcess(train_data[, spec_cols], method = c("pca", "center"), pcaComp = 10)
  train_pc <- predict(pca_model, train_data[, spec_cols])
  test_pc <- predict(pca_model, test_data[, spec_cols])
  
  train_pc <- cbind(train_pc, train_data[, 1:(spec_cols[1]-1)])
  test_pc <- cbind(test_pc, test_data[, 1:(spec_cols[1]-1)])
  
  # Linear Models
  global_lm <- lm(as.formula(paste(element, "~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10")), data = train_pc)
  dredge_lm <- dredge(global_lm)
  top5_lm <- get.models(dredge_lm, subset = 1:5)
  
  # GAM Models
  global_gam <- gam(as.formula(paste(element, "~ s(PC1,k=4) + s(PC2,k=4) + s(PC3,k=4) + s(PC4,k=4) + s(PC5,k=4) + s(PC6,k=4)")), data = train_pc)
  dredge_gam <- dredge(global_gam)
  top5_gam <- get.models(dredge_gam, subset = 1:5)
  
  # Evaluate models
  results <- data.frame(Model = character(), R2 = numeric(), RMSE = numeric())
  
  for (i in 1:5) {
    # Linear Models
    preds_lm <- predict(top5_lm[[i]], newdata = test_pc)
    r2_lm <- 1 - sum((test_pc[[element]] - preds_lm)^2) / sum((test_pc[[element]] - mean(test_pc[[element]]))^2)
    rmse_lm <- sqrt(mean((test_pc[[element]] - preds_lm)^2))
    results <- rbind(results, data.frame(Model = paste0("Linear ", i), R2 = r2_lm, RMSE = rmse_lm))
    
    # GAM Models
    preds_gam <- predict(top5_gam[[i]], newdata = test_pc)
    r2_gam <- 1 - sum((test_pc[[element]] - preds_gam)^2) / sum((test_pc[[element]] - mean(test_pc[[element]]))^2)
    rmse_gam <- sqrt(mean((test_pc[[element]] - preds_gam)^2))
    results <- rbind(results, data.frame(Model = paste0("GAM ", i), R2 = r2_gam, RMSE = rmse_gam))
  }
  
  # PLS Model
  pls_model <- pls(train_data[, spec_cols], train_data[[element]], scale = FALSE, center = TRUE, ncomp = 10, y.test = test_data[[element]],x.test = test_data[, spec_cols])
  ncomp <- pls_model$ncomp.selected
  preds_pls <- predict(pls_model, newdata = test_data[, spec_cols], ncomp = pls_model$ncomp)
  r2_pls <- pls_model$testres$r2[ncomp]
  rmse_pls <- pls_model$testres$rmse[ncomp]
  results <- rbind(results, data.frame(Model = "PLS", R2 = r2_pls, RMSE = rmse_pls))
  
  return(results)
}
```

### +++++ Shiny: elements/spectra +++++
### CURRENT NOT CALCULATING R2 AND RMSE CORRECTLY FOR PLS
```{r}
library(shiny)
library(DT)
library(dplyr)
library(mdatools)
library(mgcv)
library(MuMIn)
library(caret)
library(ggplot2)
library(gglm)
library(gratia)

# Assuming df_combined is already loaded in your environment

# Define the elements (adjust as needed)
elements <- names(df_combined)[3:68]  # Assuming elements are from column 3 to 68

model_element <- function(data, element) {
  # Preprocess data
  df <- data %>% 
    select(all_of(element), everything()) %>%
    na.omit()  # Remove rows with NA values
  
  spec_cols <- (which(names(df) == element) + 1):ncol(df)
  
  # Convert spectral columns to numeric and remove any that fail
  spec_data <- sapply(df[, spec_cols], function(x) as.numeric(as.character(x)))
  spec_data <- spec_data[, !is.na(colSums(spec_data))]
  
  # Recreate df with only numeric spectral columns
  df <- cbind(df[, 1:which(names(df) == element)], spec_data)
  spec_cols <- (which(names(df) == element) + 1):ncol(df)
  
  # Split data
  set.seed(123)
  train_index <- createDataPartition(df[[element]], p = 0.6, list = FALSE)
  train_data <- df[train_index, ]
  test_data <- df[-train_index, ]
  
  # PCA
  pca_model <- preProcess(train_data[, spec_cols], method = c("pca", "center"), pcaComp = 10)
  train_pc <- predict(pca_model, train_data[, spec_cols])
  test_pc <- predict(pca_model, test_data[, spec_cols])
  
  # Add the target element to the PC data
  train_pc <- cbind(Target = train_data[[element]], train_pc)
  test_pc <- cbind(Target = test_data[[element]], test_pc)
  
  # Linear Model
  lm_model <- lm(Target ~ ., data = train_pc)
  
  # GAM Model
  gam_model <- gam(Target ~ s(PC1, k=4) + s(PC2, k=4) + s(PC3, k=4) + s(PC4, k=4) + s(PC5, k=4), data = train_pc)
  
  # PLS Model using mdatools
  pls_model <- mdatools::pls(train_data[, spec_cols], train_data[[element]], ncomp = 10, cv = 1)
  
  # Evaluate models
  results <- data.frame(Model = character(), R2 = numeric(), RMSE = numeric())
  
  # Linear Model evaluation
  preds_lm <- predict(lm_model, newdata = test_pc)
  r2_lm <- 1 - sum((test_pc$Target - preds_lm)^2) / sum((test_pc$Target - mean(test_pc$Target))^2)
  rmse_lm <- sqrt(mean((test_pc$Target - preds_lm)^2))
  results <- rbind(results, data.frame(Model = "Linear", R2 = r2_lm, RMSE = rmse_lm))
  
  # GAM Model evaluation
  preds_gam <- predict(gam_model, newdata = test_pc)
  r2_gam <- 1 - sum((test_pc$Target - preds_gam)^2) / sum((test_pc$Target - mean(test_pc$Target))^2)
  rmse_gam <- sqrt(mean((test_pc$Target - preds_gam)^2))
  results <- rbind(results, data.frame(Model = "GAM", R2 = r2_gam, RMSE = rmse_gam))
  
  # PLS Model evaluation
  preds_pls <- predict(pls_model, test_data[, spec_cols])
  r2_pls <- pls_model$calres$r2[10]  # R2 for 10 components
  rmse_pls <- pls_model$calres$rmse[10]  # RMSE for 10 components
  results <- rbind(results, data.frame(Model = "PLS", R2 = r2_pls, RMSE = rmse_pls))
  
  return(list(results = results, models = list(Linear = lm_model, GAM = gam_model, PLS = pls_model)))
}

ui <- fluidPage(
  titlePanel("Element Modeling Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput("element", "Choose an element:", choices = elements),
      actionButton("process_all", "Process All Elements"),
      hr(),
      uiOutput("model_selection_ui"),
      conditionalPanel(
        condition = "input.selected_model == 'PLS'",
        sliderInput("n_components", "Number of PLS Components:", min = 1, max = 10, value = 1, step = 1)
      ),
      conditionalPanel(
        condition = "input.selected_model == 'GAM'",
        selectInput("gam_plot_type", "GAM Plot Type:", 
                    choices = c("Partial effects" = "partial", 
                                "Model checking" = "check"))
      )
    ),
    mainPanel(
      tabsetPanel(
        tabPanel("Results Table", DTOutput("resultsTable")),
        tabPanel("Model Visualization", 
                 plotOutput("model_plot", height = "600px"),
                 verbatimTextOutput("model_summary"))
      )
    )
  )
)

server <- function(input, output, session) {
  results <- reactiveVal(NULL)
  current_models <- reactiveVal(NULL)
  
  observeEvent(input$element, {
    model_output <- model_element(df_combined, input$element)
    results(model_output$results)
    current_models(model_output$models)
  })
  
  observeEvent(input$process_all, {
    withProgress(message = 'Processing elements', value = 0, {
      all_results <- lapply(elements, function(elem) {
        incProgress(1/length(elements), detail = paste("Processing", elem))
        res <- model_element(df_combined, elem)$results
        res$Element <- elem
        return(res)
      })
      results(do.call(rbind, all_results))
    })
  })
  
  output$resultsTable <- renderDT({
    req(results())
    datatable(results(), options = list(pageLength = 15, order = list(list(2, 'desc'))),
              filter = 'top', extensions = 'Buttons', selection = 'none') %>%
      formatRound(columns = c('R2', 'RMSE'), digits = 4)
  })
  
  output$model_selection_ui <- renderUI({
    req(results())
    selectInput("selected_model", "Select Model to Visualize:", 
                choices = c("Linear", "GAM", "PLS"))
  })
  
  output$model_plot <- renderPlot({
    req(input$selected_model, current_models())
    model <- current_models()[[input$selected_model]]
    
    switch(input$selected_model,
           "Linear" = gglm(model),
           "GAM" = {
             if(input$gam_plot_type == "partial") {
               gratia::draw(model)
             } else {
               gratia::appraise(model)
             }
           },
           "PLS" = {
             plot(model, ncomp = input$n_components)
           }
    )
  })
  
  output$model_summary <- renderPrint({
    req(input$selected_model, current_models())
    model <- current_models()[[input$selected_model]]
    
    switch(input$selected_model,
           "Linear" = summary(model,ncomp = input$n_components),
           "GAM" = summary(model),
           "PLS" = summary(model))
  })
}

shinyApp(ui = ui, server = server)
```




# Plotting
This section generates plots for each element of interest. For each element, it creates a boxplot showing the distribution across all otoliths, a histogram comparing the regional distributions, and a transect plot showing the concentration across the otolith for all transects. Outliers are removed for better plotting
```{r fig.width=12, fig.height=8, results='asis', warning=FALSE, message=FALSE}
remove_outliers <- function(x, na.rm = TRUE, ...) {
  qnt <- quantile(x, probs=c(.25, .75), na.rm = na.rm, ...)
  H <- 1.5 * IQR(x, na.rm = na.rm)
  y <- x
  y[x < (qnt[1] - H)] <- NA
  y[x > (qnt[2] + H)] <- NA
  y
}
```

### Boxplot - Element

```{r fig.width=12, fig.height=8, results='asis', warning=FALSE, message=FALSE}
plot_element_boxplot <- function(data, element_name, otolith_range) {
  data %>%
    filter(as.numeric(str_extract(otolith_id, "\\d+")) %in% otolith_range) %>%
    mutate(!!element_name := remove_outliers(.data[[element_name]])) %>%
    ggplot(aes(x = segment, y = .data[[element_name]], fill = segment)) +
    geom_boxplot(outlier.alpha = 0.5) +
    facet_wrap(~otolith_id, scales = "free_y") +
    scale_fill_viridis_d() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
    labs(title = paste("Distribution of", element_name, "by Segment"),
         subtitle = paste("Otoliths", min(otolith_range), "to", max(otolith_range)),
         x = "Segment", y = element_name)
}
```

### Histogram - Region

```{r fig.width=12, fig.height=8, results='asis', warning=FALSE, message=FALSE}
plot_region_histogram <- function(data, element_name) {
  data %>%
    mutate(!!element_name := remove_outliers(.data[[element_name]])) %>%
    ggplot(aes(x = .data[[element_name]], fill = region)) +
    geom_histogram(position = "dodge", bins = 30, alpha = 0.7) +
    facet_wrap(~segment, scales = "free_y") +
    scale_fill_viridis_d() +
    theme_minimal() +
    labs(title = paste("Distribution of", element_name, "by Region and Segment"),
         x = element_name, y = "Count")
}
```

### Heatmap - Elements

```{r fig.width=12, fig.height=8, results='asis', warning=FALSE, message=FALSE}
plot_element_heatmap <- function(data, element_name) {
  data %>%
    group_by(otolith_id, segment) %>%
    summarise(mean_concentration = mean(.data[[element_name]], na.rm = TRUE)) %>%
    ggplot(aes(x = segment, y = otolith_id, fill = mean_concentration)) +
    geom_tile() +
    scale_fill_viridis_c() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = paste("Heatmap of", element_name, "Concentration"),
         x = "Segment", y = "Otolith ID", fill = "Mean Concentration")
}
```

### Element Correlograms

```{r}
# Define the order of elements
element_order <- c("Ba138_ppm", "Ca43", "Mg24_ppm", "Mn55_ppm", "Sr86_ppm", "Sr87_ppm", "Sr88_ppm", "Zn66_ppm")

create_correlation_plot <- function(data, segment_name) {
  # Select only the numeric columns (elements) in the specified order
  element_data <- data %>%
    select(all_of(element_order))
  
  # Calculate correlation matrix
  cor_matrix <- cor(element_data, use = "pairwise.complete.obs")
  
  # Create correlation plot
  corrplot(cor_matrix, 
           method = "color", 
           type = "upper", 
           order = "original",  # Use the original order of the correlation matrix
           tl.col = "black",    # Set text label color to black for better visibility
           tl.srt = 45,         # Rotate text labels 45 degrees
           title = paste("Correlation Plot -", segment_name),
           mar = c(0,0,2,0))    # Increase top margin for title
}

```

### Data/Result Table

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE, results='asis'}
create_result_table <- function(data, caption) {
  columns_to_display <- intersect(
    c("otolith_id", "element", "segment", "n_transects", "p_value", "F_statistic", "df"),
    names(data)
  )

  formatted_data <- data %>%
    select(all_of(columns_to_display)) %>%
    mutate(
      p_value_original = p_value,
      p_value = ifelse(is.na(p_value), "NA", formatC(p_value, format = "e", digits = 2)),
      F_statistic = ifelse(is.na(F_statistic), "NA", round(F_statistic, 2)),
      Significant = case_when(
        is.na(p_value_original) ~ "NA",
        p_value_original < 0.05 ~ "Yes",
        TRUE ~ "No"
      ),
      p_value = cell_spec(p_value,
                          color = case_when(
                            is.na(p_value_original) ~ "black",
                            p_value_original < 0.05 ~ "white",
                            TRUE ~ "black"
                          ),
                          background = case_when(
                            is.na(p_value_original) ~ "lightgrey",
                            p_value_original < 0.05 ~ "#4CAF50",
                            TRUE ~ "white"
                          ),
                          bold = !is.na(p_value_original) & p_value_original < 0.05)
    ) %>%
    select(-p_value_original)

  DT::datatable(
    formatted_data,
    caption = caption,
    escape = FALSE,
    options = list(
      pageLength = 25,
      scrollX = TRUE,
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
    ),
    extensions = 'Buttons',
    filter = 'top',
    rownames = FALSE
  ) %>%
    formatStyle(
      'p_value',
      backgroundColor = styleEqual(c("TRUE", "FALSE", "NA"), c('#4CAF50', 'white', 'lightgrey'))
    ) %>%
    formatStyle(
      'Significant',
      backgroundColor = styleEqual(c("Yes", "No", "NA"), c('#4CAF50', 'white', 'lightgrey')),
      color = styleEqual(c("Yes", "No", "NA"), c('white', 'black', 'black'))
    )
}
```


#### Run Plots

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE, results='asis'}
elements_to_plot <- c("Sr86_ppm") 
                      # "Sr86_ppm", "Sr87_ppm", "Sr88_ppm", "Ba138_ppm", "Zn66_ppm", "Mg24_ppm", "Mn55_ppm")
)

for (element in elements_to_plot) {
  cat("##", element, "\n\n")

  cat("### Distribution Boxplot (Otoliths 1-20)\n\n")
  print(plot_element_boxplot(df_ICP_filt, element, 1:20))
  cat("\n\n")

  cat("### Distribution Boxplot (Otoliths 21-40)\n\n")
  print(plot_element_boxplot(df_ICP_filt, element, 21:40))
  cat("\n\n")

  cat("### Regional Distribution\n\n")
  print(plot_region_histogram(df_ICP_filt, element))
  cat("\n\n")

  cat("### Concentration Heatmap\n\n")
  print(plot_element_heatmap(df_ICP_filt, element))
  cat("\n\n")
}

# # Generate correlation plots for each segment
# cat("## Element Correlations\n\n")

# for (segment in unique(df_ICP_filt$segment)) {
#   cat(paste("### Correlation Plot -", segment, "\n\n"))
#   
#   segment_data <- df_ICP_filt %>%
#     filter(segment == !!segment)
#   
#   print(create_correlation_plot(segment_data, segment))
#   
#   cat("\n\n")
# }

# Generate overall correlation plot
# cat("### Overall Correlation Plot\n\n")
# print(create_correlation_plot(df_ICP_filt, "All Segments"))
```


```{r}
library(shiny)
library(DT)
library(dplyr)
library(mdatools)
library(mgcv)
library(MuMIn)
library(caret)
library(ggplot2)
library(gglm)
library(gratia)


model_element <- function(df, element) {
  print("Starting model_element function")
  
  # Preprocess data
  df <- df %>% 
    select(all_of(element), everything()) %>%
    filter(complete.cases(.))
  spec_cols <- 70:ncol(df)
  
  print("Data preprocessed")
  
  # PCA
  pca_temp <- pca(df[, spec_cols])
  pc_df <- data.frame(PC1 = rep(0, nrow(df)))
  for (i in 1:10) {
    pc_df[, paste0("PC", i)] <- pca_temp$res$cal$scores[, i]
  }
  pc_df <- cbind(pc_df, df)
  
  print("PCA completed")
  print(paste("Dimensions of pc_df:", dim(pc_df)[1], "x", dim(pc_df)[2]))
  
  # Linear Models
  print("Starting Linear Models")
  global_lm <- lm(as.formula(paste(element, "~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10")), data = pc_df)
  options(na.action = "na.fail")
  dredge_lm <- dredge(global_lm)
  top5_lm <- get.models(dredge_lm, subset = 1:5)
  
  print("Linear Models completed")
  
  # GAM Models
  print("Starting GAM Models")
  global_gam <- gam(as.formula(paste(element, "~ s(PC1,k=4) + s(PC2,k=4) + s(PC3,k=4) + s(PC4,k=4) + s(PC5,k=4) + s(PC6,k=4)")), data = pc_df)
  dredge_gam <- dredge(global_gam)
  top5_gam_temp <- dredge_gam[1:5,]
  
  reconstruct_formula <- function(model_row) {
    terms_used <- names(model_row)[!is.na(model_row)]
    terms_used <- terms_used[!terms_used %in% c("(Intercept)", "df", "logLik", "AICc", "delta", "weight")]
    if (length(terms_used) == 0) {
      return(as.formula(paste(element, "~ 1")))  # Return intercept-only model if no terms
    }
    as.formula(paste(element, "~", paste(terms_used, collapse = " + ")))
  }
  top5_gam <- lapply(1:5, function(i) {
    tryCatch({
      gam(reconstruct_formula(top5_gam_temp[i, ]), data = pc_df)
    }, error = function(e) {
      print(paste("Error in GAM model", i, ":", e$message))
      return(NULL)
    })
  })
  top5_gam <- top5_gam[!sapply(top5_gam, is.null)]  # Remove any NULL models
  
  print("GAM Models completed")
  
  options(na.action = "na.omit")  # Reset the option to default
  
  # Split data for evaluation
  set.seed(123)
  train_index <- createDataPartition(pc_df[[element]], p = 0.6, list = FALSE)
  train_pc <- pc_df[train_index, ]
  test_pc <- pc_df[-train_index, ]
  
  # Evaluate models
  results <- data.frame(Model = character(), R2 = numeric(), RMSE = numeric())
  
  for (i in 1:length(top5_lm)) {
    # Linear Models
    preds_lm <- predict(top5_lm[[i]], newdata = test_pc)
    r2_lm <- 1 - sum((test_pc[[element]] - preds_lm)^2) / sum((test_pc[[element]] - mean(test_pc[[element]]))^2)
    rmse_lm <- sqrt(mean((test_pc[[element]] - preds_lm)^2))
    results <- rbind(results, data.frame(Model = paste0("Linear ", i), R2 = r2_lm, RMSE = rmse_lm))
  }
  
  for (i in 1:length(top5_gam)) {
    # GAM Models
    preds_gam <- predict(top5_gam[[i]], newdata = test_pc)
    r2_gam <- 1 - sum((test_pc[[element]] - preds_gam)^2) / sum((test_pc[[element]] - mean(test_pc[[element]]))^2)
    rmse_gam <- sqrt(mean((test_pc[[element]] - preds_gam)^2))
    results <- rbind(results, data.frame(Model = paste0("GAM ", i), R2 = r2_gam, RMSE = rmse_gam))
  }
  
  print("Model evaluation completed")
  
  # PLS Model
  print("Starting PLS Model")
  tryCatch({
    # Prepare data for PLS
    X_train <- as.matrix(df[train_index, spec_cols])
    y_train <- as.vector(df[train_index, element])
    X_test <- as.matrix(df[-train_index, spec_cols])
    y_test <- as.vector(df[-train_index, element])
    
    # Regular PLS
    pls_model <- pls(X_train, y_train,
                     scale = FALSE, center = TRUE, cv = 1,
                     x.test = X_test,
                     y.test = y_test)
    
    ncomp <- pls_model$ncomp.selected
    r2_pls <- pls_model$testres$r2[[ncomp]]
    rmse_pls <- pls_model$testres$rmse[[ncomp]]
    
    # VIP-filtered PLS
    vip <- as.data.frame(vipscores(pls_model))
    vip_cols <- which(vip$V1 >= 0.5)
    pls_vip_model <- pls(X_train[, vip_cols], y_train,
                         scale = FALSE, center = TRUE, cv = 1,
                         x.test = X_test[, vip_cols],
                         y.test = y_test)
    
    ncomp_vip <- pls_vip_model$ncomp.selected
    r2_pls_vip <- pls_vip_model$testres$r2[[ncomp_vip]]
    rmse_pls_vip <- pls_vip_model$testres$rmse[[ncomp_vip]]
    
    # Add results to the summary table
    results <- rbind(results, 
                     data.frame(Model = "PLS", R2 = r2_pls, RMSE = rmse_pls),
                     data.frame(Model = "PLS-VIP", R2 = r2_pls_vip, RMSE = rmse_pls_vip))
    
    print("PLS Model completed")
    print(paste("PLS R2:", round(r2_pls, 4)))
    print(paste("PLS RMSE:", round(rmse_pls, 4)))
    print(paste("PLS-VIP R2:", round(r2_pls_vip, 4)))
    print(paste("PLS-VIP RMSE:", round(rmse_pls_vip, 4)))
  }, error = function(e) {
    print(paste("Error in PLS model:", e$message))
    results <- rbind(results, 
                     data.frame(Model = "PLS", R2 = NA, RMSE = NA),
                     data.frame(Model = "PLS-VIP", R2 = NA, RMSE = NA))
  })
  
  return(list(results = results, 
              models = list(Linear = top5_lm, 
                            GAM = top5_gam, 
                            PLS = if(exists("pls_model")) pls_model else NULL, 
                            PLS_VIP = if(exists("pls_vip_model")) pls_vip_model else NULL)))
}



ui <- fluidPage(
  titlePanel("Element Modeling Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput("element", "Choose an element:", choices = elements),
      actionButton("process_all", "Process All Elements"),
      hr(),
      uiOutput("model_selection_ui"),
      conditionalPanel(
        condition = "input.selected_model == 'PLS'",
        sliderInput("n_components", "Number of PLS Components:", min = 1, max = 10, value = 1, step = 1)
      ),
      conditionalPanel(
        condition = "input.selected_model == 'GAM' || input.selected_model == 'Linear'",
        selectInput("model_number", "Model Number:", choices = 1:5)
      ),
      conditionalPanel(
        condition = "input.selected_model == 'GAM'",
        selectInput("gam_plot_type", "GAM Plot Type:", 
                    choices = c("Partial effects" = "partial", 
                                "Model checking" = "check"))
      )
    ),
    mainPanel(
      tabsetPanel(
        tabPanel("Results Table", DTOutput("resultsTable")),
        tabPanel("Model Visualization", 
                 plotOutput("model_plot", height = "600px"),
                 verbatimTextOutput("model_summary"))
      )
    )
  )
)

server <- function(input, output, session) {
  results <- reactiveVal(NULL)
  current_models <- reactiveVal(NULL)
  
  observeEvent(input$element, {
    model_output <- model_element(df_combined, input$element)
    results(model_output$results)
    current_models(model_output$models)
  })
  
  observeEvent(input$process_all, {
    withProgress(message = 'Processing elements', value = 0, {
      all_results <- lapply(elements, function(elem) {
        incProgress(1/length(elements), detail = paste("Processing", elem))
        res <- model_element(df_combined, elem)$results
        res$Element <- elem
        return(res)
      })
      results(do.call(rbind, all_results))
    })
  })
  
  output$resultsTable <- renderDT({
    req(results())
    datatable(results(), options = list(pageLength = 15, order = list(list(2, 'desc'))),
              filter = 'top', extensions = 'Buttons', selection = 'none') %>%
      formatRound(columns = c('R2', 'RMSE'), digits = 4)
  })
  
  output$model_selection_ui <- renderUI({
    req(results())
    selectInput("selected_model", "Select Model to Visualize:", 
                choices = c("Linear", "GAM", "PLS"))
  })
  
  output$model_plot <- renderPlot({
    req(input$selected_model, current_models())
    models <- current_models()[[input$selected_model]]
    
    switch(input$selected_model,
           "Linear" = {
             model <- models[[as.numeric(input$model_number)]]
             gglm(model)
           },
           "GAM" = {
             model <- models[[as.numeric(input$model_number)]]
             if(input$gam_plot_type == "partial") {
               gratia::draw(model)
             } else {
               gratia::appraise(model)
             }
           },
           "PLS" = {
             plot(models, ncomp = input$n_components)
           }
    )
  })
  
  output$model_summary <- renderPrint({
    req(input$selected_model, current_models())
    models <- current_models()[[input$selected_model]]
    
    switch(input$selected_model,
           "Linear" = summary(models[[as.numeric(input$model_number)]]),
           "GAM" = summary(models[[as.numeric(input$model_number)]]),
           "PLS" = summary(models, ncomp = input$n_components))
  })
}

shinyApp(ui = ui, server = server)
```


# ------------ old code ---------------
in case I need to look at it again
```{r eval=FALSE, include=FALSE}

### Line Chart - Original Data vs Cutoffs
Trying to see if my cutoff values make sense
```{r eval=FALSE, include=FALSE}
library(ggplot2)
library(dplyr)

# Your detect_otolith_segments function here (as provided in your message)

# Function to plot original and filtered data for a single transect and multiple elements
plot_original_and_filtered <- function(data, transect_id, elements) {
  # Filter data for the selected transect
  plot_data <- data %>%
    filter(transect_id == !!transect_id)
  
  # Apply boundary detection
  boundaries <- detect_otolith_segments(plot_data$Ca43, plot_data$Cu63_ppm)
  start_time <- plot_data$time[boundaries$start]
  end_time <- plot_data$time[boundaries$end]
  
  # Function to create a plot for a single element
  create_element_plot <- function(element) {
    # Plot with all values and cutoffs
    plot_all <- ggplot(plot_data) +
      geom_line(aes(x = time, y = !!sym(element)), color = "blue") +
      geom_vline(xintercept = start_time, color = "red", linetype = "dashed") +
      geom_vline(xintercept = end_time, color = "red", linetype = "dashed") +
      labs(title = paste("All Values for", element, "in Transect", transect_id),
           subtitle = "Red dashed lines indicate cutoffs",
           x = "Time Elapsed",
           y = "Concentration") +
      theme_minimal()
    
    # Plot with filtered values
    plot_filtered <- plot_data %>%
      filter(time >= start_time & time <= end_time) %>%
      ggplot(aes(x = time, y = !!sym(element))) +
      geom_line(color = "blue") +
      labs(title = paste("Filtered Values for", element, "in Transect", transect_id),
           x = "Time Elapsed",
           y = "Concentration") +
      theme_minimal()
    
    # Combine plots vertically
    cowplot::plot_grid(plot_all, plot_filtered, ncol = 1)
  }
  
  # Create and print plots for each element
  for (element in elements) {
    plot <- create_element_plot(element)
    print(plot)
    # Add a prompt to continue to the next plot
    if (element != elements[length(elements)]) {
      readline(prompt = "Press [Enter] to see the next plot...")
    }
  }
}

# Example usage
transect_id_to_plot <- "oto31-3"  # Replace with the actual transect ID you want to plot
elements_to_plot <- c("Ca43", "Cu63_ppm")  # Add or remove elements as needed

# Create and print the plots
plot_original_and_filtered(df, transect_id_to_plot, elements_to_plot)


### Find Boundaries
```{r}
# Improved Boundary Detection and Segmentation using both Ca43 and Cu63_ppm
# detect_otolith_segments <- function(ca43_data, cu63_data, window_size = 7, ca_threshold_factor = 0.4, cu_threshold_factor = 1.5) {
#   # Smooth the data using rolling median
#   smoothed_ca43 <- zoo::rollmedian(ca43_data, k = window_size, fill = "extend")
#   smoothed_cu63 <- zoo::rollmedian(cu63_data, k = window_size, fill = "extend")
# 
#   n <- length(smoothed_ca43)
#   baseline_range <- round(n * 0.3):round(n * 0.7)
#   
#   # Calculate baseline and thresholds for Ca43
#   baseline_ca43 <- median(smoothed_ca43[baseline_range])
#   ca_threshold <- baseline_ca43 * ca_threshold_factor
# 
#   # Calculate baseline and thresholds for Cu63
#   baseline_cu63 <- median(smoothed_cu63[baseline_range])
#   cu_threshold <- baseline_cu63 * cu_threshold_factor
# 
#   # Determine start index based on Ca43 and Cu63 thresholds
#   start_idx_ca43 <- which(smoothed_ca43 > ca_threshold)[1]
#   start_idx_cu63 <- which(smoothed_cu63 < cu_threshold)[1]
#   start_idx <- max(1, start_idx_ca43, start_idx_cu63)
# 
#   # Determine end index based on Ca43 and Cu63 thresholds
#   end_idx_ca43 <- n - which(rev(smoothed_ca43) > ca_threshold)[1] + 1
#   end_idx_cu63 <- n - which(rev(smoothed_cu63) < cu_threshold)[1] + 1
#   end_idx <- min(n, end_idx_ca43, end_idx_cu63)
# 
#   return(list(start = start_idx, end = end_idx))
# }


# # Apply boundary detection to all transects using both Ca43 and Cu63_ppm
# df_segmented <- df %>%
#   group_by(transect_id) %>%
#   mutate(
#     # First smooth the Ca43 data using rolling median
#     smoothed_Ca43 = zoo::rollmedian(Ca43, k = 2, fill = "extend"),
#     # Apply quantile filtering to smoothed data
#     threshold = quantile(smoothed_Ca43, 0.6, na.rm = TRUE),
#     valid_points = smoothed_Ca43 >= threshold,
#     start = min(which(valid_points)),
#     end = max(which(valid_points)),
#     row_num = row_number(),
#     segment = case_when(
#       row_num < start + 0.2 * (end - start) ~ "core",
#       row_num > end - 0.2 * (end - start) ~ "edge",
#       TRUE ~ "middle"
#     )
#   ) %>%
#   filter(row_num >= start & row_num <= end) %>%
#   # Remove temporary columns
#   select(-threshold, -valid_points, -row_num, -start, -end, -smoothed_Ca43) %>%
#   ungroup()



# df_segmented <- df %>%
#   group_by(transect_id) %>%
#   mutate(
#     threshold = quantile(Ca43, 0.6, na.rm = TRUE),
#     valid_points = Ca43 >= threshold,
#     start = min(which(valid_points)),
#     end = max(which(valid_points)),
#     row_num = row_number(),
#     segment = case_when(
#       row_num < start + 0.2 * (end - start) ~ "core",
#       row_num > end - 0.2 * (end - start) ~ "edge",
#       TRUE ~ "middle"
#     )
#   ) %>%
#   filter(row_num >= start & row_num <= end) %>%
#   select(-threshold, -valid_points, -row_num, -start, -end) %>%
#   ungroup()


# 
# df_segmented <- df %>%
#   group_by(transect_id) %>%
#   mutate(
#     boundaries = list(detect_otolith_segments(Ca43, Cu63_ppm)),
#     start = map_dbl(boundaries, ~ .x$start),
#     end = map_dbl(boundaries, ~ .x$end),
#     row_num = row_number(),
#     total_rows = n(),
#     segment = case_when(
#       row_num < start + 0.2 * (end - start) ~ "core",
#       row_num > end - 0.2 * (end - start) ~ "edge",
#       TRUE ~ "middle"
#     )
#   ) %>%
#   filter(row_num >= start & row_num <= end) %>%
#   select(-boundaries, -row_num, -total_rows, -start, -end) %>%
#   ungroup()


# Load Data and Preprocess
load_and_clean_data <- function(file_path) {
  data <- readRDS(file_path)
  valid_indices <- setdiff(30:184, c(grep("oto10-2", names(data)), grep("oto30-", names(data))))
  
  cleaned_data <- data[valid_indices] %>%
    bind_rows(.id = "transect_id") %>%
    mutate(
      otolith_id = str_extract(transect_id, "oto\\d+"),
      transect_num = as.integer(str_extract(transect_id, "\\d+$")),
      region = case_when(
        as.integer(str_extract(otolith_id, "\\d+")) %in% c(6:10, 16:25, 31:35) ~ "Kodiak",
        TRUE ~ "Shumagin"
      )
    )
  
  return(cleaned_data)
}
df <- load_and_clean_data("RDS_dataframes/ICP-MS_clean.RDS")
```



#### Results tables

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE, results='asis'}
# Statistical Results
cat("## Statistical Results\n\n")

cat("### Transect Comparison Results\n\n")
create_result_table(analysis_results$transect_comparison, "Comparison of Element Concentrations Across Transects")
cat("\n\n")

cat("### Otolith Comparison Results\n\n")
create_result_table(analysis_results$otolith_comparison, "Comparison of Element Concentrations Across Otoliths")
cat("\n\n")

cat("### Region Comparison Results\n\n")
create_result_table(analysis_results$region_comparison, "Comparison of Element Concentrations Between Regions")
cat("\n\n")
```

### Statistical Analysis

```{r eval=FALSE, include=FALSE}
# Statistical Analysis
safe_test <- function(test_func, ...) {
  tryCatch(
    test_func(...),
    error = function(e) {
      warning(paste("Error in test:", e$message))
      return(data.frame(p.value = NA, statistic = NA, parameter = NA))
    }
  )
}
```

### Transects
```{r eval=FALSE, include=FALSE}
compare_transects <- function(data) {
  data %>%
    pivot_longer(cols = c(Ca43, ends_with("_ppm")),
                 names_to = "element",
                 values_to = "value") %>%
    group_by(otolith_id, element, segment) %>%
    summarise(
      n_transects = n_distinct(transect_id),
      test_result = if(n_transects > 1) {
        list(safe_test(aov, value ~ factor(transect_id)) %>% summary() %>% .[[1]])
      } else {
        list(data.frame(p.value = NA, statistic = NA, parameter = NA))
      },
      .groups = "drop"
    ) %>%
    mutate(
      p_value = map_dbl(test_result, ~ .x$`Pr(>F)`[1]),
      F_statistic = map_dbl(test_result, ~ .x$`F value`[1]),
      df = map_dbl(test_result, ~ .x$Df[1])
    ) %>%
    select(otolith_id, element, segment, n_transects, p_value, F_statistic, df)
}
```

### Otoliths

```{r eval=FALSE, include=FALSE}
compare_otoliths <- function(data) {
  data %>%
    pivot_longer(cols = c(Ca43, ends_with("_ppm")),
                 names_to = "element",
                 values_to = "value") %>%
    group_by(element, segment) %>%
    summarise(
      test_result = list(safe_test(aov, value ~ factor(otolith_id)) %>% summary() %>% .[[1]]),
      .groups = "drop"
    ) %>%
    mutate(
      p_value = map_dbl(test_result, ~ .x$`Pr(>F)`[1]),
      F_statistic = map_dbl(test_result, ~ .x$`F value`[1]),
      df = map_dbl(test_result, ~ .x$Df[1])
    ) %>%
    select(element, segment, p_value, F_statistic, df)
}
```

### Regions

```{r eval=FALSE, include=FALSE}
compare_regions <- function(data) {
  data %>%
    pivot_longer(cols = c(Ca43, ends_with("_ppm")),
                 names_to = "element",
                 values_to = "value") %>%
    group_by(element, segment) %>%
    summarise(
      test_result = list(safe_test(aov, value ~ factor(region)) %>% summary() %>% .[[1]]),
      .groups = "drop"
    ) %>%
    mutate(
      p_value = map_dbl(test_result, ~ .x$`Pr(>F)`[1]),
      F_statistic = map_dbl(test_result, ~ .x$`F value`[1]),
      df = map_dbl(test_result, ~ .x$Df[1])
    ) %>%
    select(element, segment, p_value, F_statistic, df)
}
```

#### Run Analysis
```{r eval=FALSE, include=FALSE}
analysis_results <- list(
  transect_comparison = compare_transects(df_ICP_filt),
  otolith_comparison = compare_otoliths(df_ICP_filt),
  region_comparison = compare_regions(df_ICP_filt)
)
```

