I'm redoing my modelling approach a bit per Franz's suggestions. The general approach will be: - start with the entire dataframe (no splits) and pick the number of PC's I would like to use. \~10 for LM and maybe less for GAM - fit LM and GAM with PC's from the full dataset (MuMIn, dredge function) - choose subset of best models, up to 5. - 10-fold CV and evaluate top 5 models from each approach - List all selected models and AIC from dredge function

```{r, results = 'hide', warning = F, message = F}
# Install packages not yet installed
packages <- c("caret", "corrplot", "dplyr","devtools", "DT", "dplyr", "gglm", "ggplot2", "gratia", "gridExtra", "knitr", "lubridate", "mdatools", "mgcv", "MuMIn", "opusreader2", "prospectr", "splitstackshape","shiny","shinyWidgets", "stringr","tidyr", "viridis")
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  utils::install.packages(pkgs = packages[!installed_packages])
}
invisible(lapply(packages, library, character.only = TRUE)) # load all packages in list
# install packages not on CRAN
if (!require("remotes")) install.packages("remotes")
if (!require("opusreader2")) {
  remotes::install_github("spectral-cockpit/opusreader2")
  library("opusreader2")
}
# if (!require("simplerspec")) {
#   remotes::install_github("philipp-baumann/simplerspec")
#   library("simplerspec")
# }
rm(installed_packages, packages) # remove objects from environment
```

# Load dataframe

I'm trying to decide whether to cutoff the wavenumbers \>7500 since this region is relatively "noisy". After some investigating I may opt to include this region as my models tend to fit a bit better. Preprocessing tweaks don't seem to be making a huge different so opting for Matta et al's 1,3,17 for SG filter.

```{r}
# cut off >7500 wavenumber
# df_filter <- readRDS("RDS_dataframes/LPW_scan_avg_proc_filter.RDS")
# df <- df_filter
# ALL wavenumbers
df <- readRDS("RDS_dataframes/LPW_scan_avg_proc.RDS")
df <- df[complete.cases(df$read_age), ]
df <- df %>% filter(specimen != 77)

# adjust preprocessing
# df <- readRDS("RDS_dataframes/LPW_scan_avg_unproc.RDS")
# df <- cbind(df[, c(1:20)], savitzkyGolay(df[, 21:length(df)], m = 1, p = 3, w = 17))
```

# 10 PCs, LM/GAM dredging with all samples

I've been trying to decide whether I want to select the best models based on PCA of spectra from ALL specimens (including those that are unaged), or only PCA's based on aged specimens. I've played around with both and it appears that model selecting with only aged specimens may be performing slightly better.

I had to reduce the number of PC's in my GAMs models to improve overall fit. R2 values with 10 PCs were all over the place (large and/or negative), and fit seems to be better around 6 overall PC's for the dredge function.

```{r}
# unfiltered spectra
# remove unaged specimens before calculating PCs


pca_temp <- pca(df[, 21:ncol(df), ])
pc_df <- data.frame(PC1 = rep(0,nrow(df)))
for (i in 1:10) {
  pc_df[, paste0("PC", i)] <- pca_temp$res$cal$scores[, i]
  rm(i)
}
pc_df <- cbind(pc_df,df)
global_lm <- lm(data = pc_df, read_age ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10)
global_gam <- gam(data = pc_df, read_age ~ s(PC1,k = 4) + s(PC2,k = 4) + s(PC3,k = 4) + s(PC4,k = 4) + s(PC5,k = 4) + s(PC6,k = 4)) # adding too many PC's makes GAM have horrible fit and R2



# dredge to find top 5 models
options(na.action = "na.fail")
dredge_lm <- dredge(global_lm)
top5_lm <- get.models((dredge_lm), subset = 1:5)
dredge_gam <- dredge(global_gam)
top5_gam <- get.models(dredge_gam, subset = 1:5)
terms_lm <- list()
terms_gam <- list()
for(i in 1:5){
  terms_lm[[i]] <- top5_lm[[i]]$terms
  terms_gam[[i]] <- top5_gam[[i]]$formula
  rm(i)
}


rm(global_gam, global_lm, top5_gam, top5_lm, pc_df, pca_temp, dredge_gam,dredge_lm)
# rm(global_gam_filter, global_lm_filter, pc_filter_df, pca_filter_temp, dredge_gam_filter, dredge_lm_filter)
```

# 10-fold split

```{r message=FALSE, echo = T, results = 'hide'}
# 10 fold CV split - create folds
set.seed(6)
splits <- caret::createFolds(df$read_age, k = 10, list = TRUE, returnTrain = FALSE)
# extract PC's for each calibration set, create test sets with ages and spectra
cal <- list()
test <- list()
mods.pca <- list()
for (i in 1:10) {
  # calibration set and PC's
  pc.mod <- preProcess(df[-splits[[i]], -c(1:20)], method = c("pca","center"), pcaComp = 10)
  pc.cal <- predict(pc.mod, df[-splits[[i]], -c(1:20)])
  pc.cal <- cbind(pc.cal, df[-splits[[i]], ])
  cal[[i]] <- pc.cal
  rm(pc.cal)
  # test sets
  pc.test <- predict(pc.mod, df[splits[[i]], -c(1:20)])
  pc.test <- cbind(pc.test, df[splits[[i]], ])
  test[[i]] <- pc.test
  mods.pca[[i]] <- pc.mod
  rm(pc.test, pc.mod, i)
}
```

# LM - top 5 models

```{r message=FALSE}
mods.lm <- replicate(10, list())
r2_lm <- list()
rmse_lm <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  r2_temp <- rep(0, 5)
  rmse_temp <- rep(0, 5)
  for (j in 1:5) {
    mod <- lm(data = calibrate, terms_lm[[j]])
    preds <- predict(mod, newdata = testing)
    rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
    RSS <- sum((testing$read_age - preds)^2)
    TSS <- sum((testing$read_age - mean(testing$read_age))^2)
    r2_temp[j] <- 1 - (RSS / TSS)
    mods.lm[[i]][[j]] <- mod
    rm(RSS, TSS, preds)
  }
  r2_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), r2 = r2_temp)
  rmse_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), rmse = rmse_temp)
  rm(mod, i, j, r2_temp, rmse_temp, calibrate, testing)
}
# Unlist the r2 & rmse values and calc means for 10 splits, combine into dataframe
r2_lm <- sapply(r2_lm, "[[", "r2")
r2_lm <- rowMeans(r2_lm)
rmse_lm <- sapply(rmse_lm, "[[", "rmse")
rmse_lm <- rowMeans(rmse_lm)
(lm_result <- data.frame(
  Model = paste0("Linear ", 1:5),
  Mean_R2 = r2_lm,
  Mean_RMSE = rmse_lm
))
rm(r2_lm, rmse_lm)
```

# GAM - top 5 models

```{r}
mods.gam <- replicate(10, list())
r2_gam <- list()
rmse_gam <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  r2_temp <- rep(0, 5)
  rmse_temp <- rep(0, 5)
  for (j in 1:5) {
    mod <- gam(data = calibrate, terms_gam[[j]])
    preds <- predict(mod, newdata = testing)
    rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
    RSS <- sum((testing$read_age - preds)^2)
    TSS <- sum((testing$read_age - mean(testing$read_age))^2)
    r2_temp[j] <- 1 - (RSS / TSS)
    mods.gam[[i]][[j]] <- mod
    rm(RSS, TSS, preds)
  }
  r2_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), r2 = r2_temp)
  rmse_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), rmse = rmse_temp)
  rm(j, i, testing, calibrate, mod, r2_temp, rmse_temp)
}
r2_gam <- sapply(r2_gam, "[[", "r2")
r2_gam <- rowMeans(r2_gam)
rmse_gam <- sapply(rmse_gam, "[[", "rmse")
rmse_gam <- rowMeans(rmse_gam)
gam_result <- data.frame(
  Model = paste0("GAM ", 1:5),
  Mean_R2 = r2_gam,
  Mean_RMSE = rmse_gam
)
(results <- rbind(lm_result,gam_result))
rm(r2_gam, rmse_gam, lm_result, gam_result)
```

# PLS models

VIP scores to filter wavenumbers seems to always make model fit more poorly; likely will NOT be using.

```{r}
mods.pls <- list()
mods.vip <- list()
r2.pls <- list()
rmse.pls <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  mod <- pls(calibrate[, 31:ncol(calibrate)], calibrate[, "read_age"],
    scale = F, center = T, ncomp.selcrit = "wold", 
    info = "Age Prediction Model",
    x.test = testing[, 31:ncol(testing)],
    y.test = testing[, "read_age"]
  )
  ncomp <- mod$ncomp.selected
  rmse.pls$pls.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
  r2.pls$pls.test[i] <- mod$testres$r2[[ncomp]] # extract r2
  
  # VIP filter of wavenumbers
  mods.pls[[i]] <- mod
  mods.pls[[1]]$ncomp.selected
  vip <- as.data.frame(vipscores(mod))
  mod <- pls(calibrate[, 31:ncol(calibrate)], calibrate[, "read_age"],
    ncomp.selcrit = "wold", 
    scale = F, center = T,
    info = "Age Prediction Model",
    x.test = testing[, 31:ncol(testing)],
    y.test = testing[, "read_age"],
    exclcols = vip$V1 < 0.25
  )
  ncomp <- mod$ncomp.selected
  r2.pls$vip.test[i] <- mod$testres$r2[[ncomp]] # extract r2
  mods.vip[[i]] <- mod
  ncomp <- mod$ncomp.selected
  rmse.pls$vip.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
  rm(calibrate, mod, ncomp, testing, i, vip)
}
(results <- rbind(results, 
                 c("PLS", mean(r2.pls$pls.test), mean(rmse.pls$pls.test)),
                 c("VIP", mean(r2.pls$vip.test), mean(rmse.pls$vip.test))
))
# rm(rmse.pls, r2.pls)
```

# Simple Length & Weight Models

```{r}
mods.simple <- list()
r2.simple <- list()
rmse.simple <- list()

for (i in 1:10) {
  calibrate <- cal[[i]]
  calibrate <- calibrate[complete.cases(calibrate$structure_weight), ] 
  testing <- test[[i]]
  testing <- testing[complete.cases(testing$structure_weight), ] 
  mod <- lm(data = calibrate, read_age ~ length + structure_weight, na.action = na.omit)
  preds <- predict(mod, newdata = testing)
  mods.simple$lm[[i]] <- mod
  RSS <- sum((testing$read_age - preds)^2)
  TSS <- sum((testing$read_age - mean(testing$read_age))^2)
  r2.simple$lm[[i]] <- 1 - (RSS / TSS)
  rmse.simple$lm[[i]] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
###### GAM #####
  mod <- gam(data = calibrate, read_age ~ s(length, k = 4) + s(structure_weight, k = 4))
  preds <- predict(mod, newdata = testing)
  mods.simple$gam[[i]] <- mod
  rmse.simple$gam[[i]] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
  RSS <- sum((testing$read_age - preds)^2)
  TSS <- sum((testing$read_age - mean(testing$read_age))^2)
  r2.simple$gam[[i]] <- 1 - (RSS / TSS)
  rm(mod, i, calibrate, testing)
  rm(RSS, TSS, preds)
}

# Calculate mean R2 for simple lm and gam
mean_r2_lm <- mean(unlist(r2.simple$lm))
mean_r2_gam <- mean(unlist(r2.simple$gam))
# Calculate mean RMSE for simple lm and gam
mean_rmse_lm <- mean(unlist(rmse.simple$lm))
mean_rmse_gam <- mean(unlist(rmse.simple$gam))
# Create new rows to add to results
new_rows <- data.frame(
  Model = c("Simple lm", "Simple gam"),
  Mean_R2 = c(mean_r2_lm, mean_r2_gam),
  Mean_RMSE = c(mean_rmse_lm, mean_rmse_gam)
)
# Append the new rows to the results dataframe
(results <- rbind(results, new_rows))
rm(r2.pls,r2.simple,rmse.pls, rmse.simple, mean_r2_gam, mean_r2_lm, mean_rmse_gam, mean_rmse_lm, new_rows)


```

# ()()()()()() --- AI approach to above --- ()()()()()()

```{r eval=FALSE, include=FALSE}
# Load required libraries
library(caret)
library(mgcv)
library(MuMIn)
library(mdatools)
library(dplyr)
library(pls)

# Load and prepare data
df <- readRDS("RDS_dataframes/LPW_scan_avg_proc.RDS")
df <- df[complete.cases(df$read_age), ] %>% 
  filter(specimen != 77)

# 10-fold CV split - create folds
set.seed(6)
splits <- caret::createFolds(df$read_age, k = 10, list = TRUE, returnTrain = FALSE)

# Extract PC's for each calibration set, create test sets with ages and spectra
cal <- list()
test <- list()
mods.pca <- list()

for (i in 1:10) {
  # Calibration set and PC's
  pc.mod <- preProcess(df[-splits[[i]], -c(1:20)], method = c("pca","center"), pcaComp = 10)
  pc.cal <- predict(pc.mod, df[-splits[[i]], -c(1:20)])
  pc.cal <- cbind(pc.cal, df[-splits[[i]], ])
  cal[[i]] <- pc.cal
  
  # Test sets
  pc.test <- predict(pc.mod, df[splits[[i]], -c(1:20)])
  pc.test <- cbind(pc.test, df[splits[[i]], ])
  test[[i]] <- pc.test
  mods.pca[[i]] <- pc.mod
}

# Extract PCs from all data for model selection
pca_result <- pca(df[, 21:ncol(df)])
pc_scores <- pca_result$res$cal$scores[, 1:10]
pc_df <- as.data.frame(pc_scores)
colnames(pc_df) <- paste0("PC", 1:10)
pc_df <- cbind(pc_df, df)

# Define global models for dredging
global_lm <- lm(data = pc_df, read_age ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10)
global_gam <- gam(data = pc_df, read_age ~ s(PC1, k=4) + s(PC2, k=4) + s(PC3, k=4) + 
                    s(PC4, k=4) + s(PC5, k=4) + s(PC6, k=4))

# Dredge to find top 5 models
options(na.action = "na.fail")
dredge_lm <- dredge(global_lm)
top5_lm <- get.models((dredge_lm), subset = 1:5)
dredge_gam <- dredge(global_gam)
top5_gam <- get.models(dredge_gam, subset = 1:5)

# Extract model formulas
terms_lm <- lapply(top5_lm, function(x) x$terms)
terms_gam <- lapply(top5_gam, function(x) x$formula)

# Function to evaluate a model with CV
evaluate_cv_model <- function(model_type, terms, df, folds, cal, test) {
  results <- data.frame(fold = 1:10, r2 = NA, rmse = NA)
  
  if (model_type %in% c("lm", "gam")) {
    for (i in 1:10) {
      calibrate <- cal[[i]]
      testing <- test[[i]]
      
      # Fit model
      if (model_type == "lm") {
        mod <- lm(terms, data = calibrate)
        preds <- predict(mod, newdata = testing)
      } else {
        mod <- gam(terms, data = calibrate)
        preds <- predict(mod, newdata = testing)
      }
      
      # Calculate metrics
      rmse_val <- RMSE(preds, testing$read_age)
      RSS <- sum((testing$read_age - preds)^2)
      TSS <- sum((testing$read_age - mean(testing$read_age))^2)
      r2_val <- 1 - (RSS / TSS)
      
      results$r2[i] <- r2_val
      results$rmse[i] <- rmse_val
    }
  } else if (model_type == "pls") {
    r2_vals <- numeric(10)
    rmse_vals <- numeric(10)
    
    for (i in 1:10) {
      calibrate <- cal[[i]]
      testing <- test[[i]]
      
      mod <- pls(calibrate[, 31:ncol(calibrate)], calibrate[, "read_age"],
                 scale = FALSE, center = TRUE, ncomp.selcrit = "wold", 
                 info = "Age Prediction Model",
                 x.test = testing[, 31:ncol(testing)],
                 y.test = testing[, "read_age"])
      
      ncomp <- mod$ncomp.selected
      rmse_vals[i] <- mod$testres$rmse[[ncomp]]
      r2_vals[i] <- mod$testres$r2[[ncomp]]
    }
    
    results$r2 <- r2_vals
    results$rmse <- rmse_vals
    
  } else if (model_type == "vip") {
    r2_vals <- numeric(10)
    rmse_vals <- numeric(10)
    
    for (i in 1:10) {
      calibrate <- cal[[i]]
      testing <- test[[i]]
      
      # First build regular PLS model to get VIP scores
      mod <- pls(calibrate[, 31:ncol(calibrate)], calibrate[, "read_age"],
                 scale = FALSE, center = TRUE, ncomp.selcrit = "wold", 
                 info = "Age Prediction Model",
                 x.test = testing[, 31:ncol(testing)],
                 y.test = testing[, "read_age"])
      
      # Calculate VIP scores
      vip <- as.data.frame(vipscores(mod))
      
      # Build model with VIP filtering
      mod <- pls(calibrate[, 31:ncol(calibrate)], calibrate[, "read_age"],
                 ncomp.selcrit = "wold", 
                 scale = FALSE, center = TRUE,
                 info = "Age Prediction Model",
                 x.test = testing[, 31:ncol(testing)],
                 y.test = testing[, "read_age"],
                 exclcols = vip$V1 < 0.25)
      
      ncomp <- mod$ncomp.selected
      rmse_vals[i] <- mod$testres$rmse[[ncomp]]
      r2_vals[i] <- mod$testres$r2[[ncomp]]
    }
    
    results$r2 <- r2_vals
    results$rmse <- rmse_vals
    
  } else if (model_type == "simple_lm") {
    for (i in 1:10) {
      calibrate <- cal[[i]]
      calibrate <- calibrate[complete.cases(calibrate$structure_weight), ]
      testing <- test[[i]]
      testing <- testing[complete.cases(testing$structure_weight), ]
      
      mod <- lm(read_age ~ length + structure_weight, data = calibrate)
      preds <- predict(mod, newdata = testing)
      
      # Calculate metrics
      rmse_val <- RMSE(preds, testing$read_age)
      RSS <- sum((testing$read_age - preds)^2)
      TSS <- sum((testing$read_age - mean(testing$read_age))^2)
      r2_val <- 1 - (RSS / TSS)
      
      results$r2[i] <- r2_val
      results$rmse[i] <- rmse_val
    }
  } else if (model_type == "simple_gam") {
    for (i in 1:10) {
      calibrate <- cal[[i]]
      calibrate <- calibrate[complete.cases(calibrate$structure_weight), ]
      testing <- test[[i]]
      testing <- testing[complete.cases(testing$structure_weight), ]
      
      mod <- gam(read_age ~ s(length, k=4) + s(structure_weight, k=4), data = calibrate)
      preds <- predict(mod, newdata = testing)
      
      # Calculate metrics
      rmse_val <- RMSE(preds, testing$read_age)
      RSS <- sum((testing$read_age - preds)^2)
      TSS <- sum((testing$read_age - mean(testing$read_age))^2)
      r2_val <- 1 - (RSS / TSS)
      
      results$r2[i] <- r2_val
      results$rmse[i] <- rmse_val
    }
  }
  
  return(list(
    mean_r2 = mean(results$r2, na.rm = TRUE),
    mean_rmse = mean(results$rmse, na.rm = TRUE),
    fold_results = results
  ))
}

# Evaluate all models
results_list <- list()

# Evaluate top 5 linear models
for (i in 1:5) {
  results_list[[paste0("Linear ", i)]] <- evaluate_cv_model("lm", terms_lm[[i]], df, splits, cal, test)
}

# Evaluate top 5 GAM models
for (i in 1:5) {
  results_list[[paste0("GAM ", i)]] <- evaluate_cv_model("gam", terms_gam[[i]], df, splits, cal, test)
}

# Evaluate PLS model
results_list[["PLS"]] <- evaluate_cv_model("pls", NULL, df, splits, cal, test)

# Evaluate VIP model
results_list[["VIP"]] <- evaluate_cv_model("vip", NULL, df, splits, cal, test)

# Evaluate simple models
results_list[["Simple lm"]] <- evaluate_cv_model("simple_lm", NULL, df, splits, cal, test)
results_list[["Simple gam"]] <- evaluate_cv_model("simple_gam", NULL, df, splits, cal, test)

# Compile final results
final_results <- data.frame(
  Model = names(results_list),
  Mean_R2 = sapply(results_list, function(x) x$mean_r2),
  Mean_RMSE = sapply(results_list, function(x) x$mean_rmse)
)

# Print the final results
print(final_results)

# Optional: Save the results
# saveRDS(final_results, "model_comparison_results.rds")
# saveRDS(results_list, "detailed_model_results.rds")
```



# Model Performance Plot

```{r eval=FALSE, include=FALSE}
# Figure 4 - model performance plots
actual <- list()
pred_lm <- list()
pred_gam <- list()
pred_pls <- list()
# create dataframe with predictions vs actual age
for (i in 1:10) {
   for (j in 1:5){
  ncomp <- mods.pls[[i]]$ncomp.selected
  pred_pls[[i]] <- mods.pls[[i]]$testres$y.pred[,ncomp,]
  actual[[i]] <- test[[i]]$read_age
  pred_lm[[i]][[j]] <- predict(mods.lm[[i]][[j]], test[[i]])
  pred_gam[[i]][[j]] <- predict(mods.gam[[i]][[j]], test[[i]])
  }
}


summary(mods.lm[[1]][[6]])
predictions <- data.frame(
  actual = rep(0, 46),
  pred_lm = rep(0, 46),
  pred_gam = rep(0, 46)
)
predictions$actual <- unlist(actual)
predictions$pred_lm <- unlist(pred_lm)
predictions$pred_gam <- unlist(pred_gam)
predictions$pred_pls <- unlist(pred_pls)
rm(actual, pred_lm, pred_gam)
unlist(pred_pls)
# LM model performance
RMSE.age$lm.test <- mean(RMSE.age$lm.test)
r2lmlab <- paste("r2 =", round(mod.summary[2, 2], 3))
rmselmlab <- paste("RMSE = ", round(RMSE.age$lm.test, 3))

plot4 <- ggplot() +
  theme_bw() +
  geom_point(data = predictions, aes(x = actual, y = pred_lm), size = 3) +
  geom_abline(slope = 1, linewidth = 1.25, col = "red", linetype = "longdash") +
  xlab("Age (days)") +
  ylab("Predicted age (days)") +
  geom_text(aes(x = 184, y = 130), size = 4, label = r2lmlab) +
  geom_text(aes(x = 180, y = 124), size = 4, label = rmselmlab) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, size = 20)
  ) +
  ggtitle("MLR")

# GAM model performance
r2gamlab <- paste("r2 =", round(mod.summary[4, 2], 3))
rmsegamlab <- paste("RMSE = ", round(RMSE.age$GAM.test, 3))

plot5 <- ggplot() +
  theme_bw() +
  geom_point(data = predictions, aes(x = actual, y = pred_gam), size = 3) +
  geom_abline(slope = 1, linewidth = 1.25, col = "red", linetype = "longdash") +
  xlab("Age (days)") +
  ylab("Predicted age (days)") +
  geom_text(aes(x = 184, y = 125), size = 4, label = r2gamlab) +
  geom_text(aes(x = 180, y = 119), size = 4, label = rmsegamlab) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, size = 20)
  ) +
  ggtitle("GAM")

plot4 + plot5
```



# FUNCTION: Dredge 3-10 PCs

Run GAM and LM dredge with 3-10 PCs, then run models with 10-fold split, store results for comparisons. Save top 5 models for each approach, regardless of number of PCs used for dredge

```{r eval=FALSE, include=FALSE}
df <- readRDS("RDS_dataframes/LPW_scan_avg_proc.RDS")
df <- df[complete.cases(df$read_age), ] 
df <- df %>% filter(specimen != 77)
# Create a list to store all results
all_results <- list()

# Loop through different numbers of PCs (3 to 10)
for(num_pcs in 3:10) {
  # Create PC dataframe
  pca_temp <- pca(df[, 25:ncol(df), ])
  pc_df <- data.frame(PC1 = rep(0,nrow(df)))
  for (i in 1:num_pcs) {
    pc_df[, paste0("PC", i)] <- pca_temp$res$cal$scores[, i]
  }
  pc_df <- cbind(pc_df,df)
  
  # Create formula strings dynamically based on num_pcs
  lm_formula_terms <- paste0("PC", 1:num_pcs, collapse = " + ")
  gam_formula_terms <- paste0("s(PC", 1:num_pcs, ",k = 4)", collapse = " + ")
  
  # Create global models
  global_lm <- lm(as.formula(paste("read_age ~", lm_formula_terms)), data = pc_df)
  global_gam <- gam(as.formula(paste("read_age ~", gam_formula_terms)), data = pc_df)
  
  # Dredge to find top 5 models
  options(na.action = "na.fail")
  dredge_lm <- dredge(global_lm)
  top5_lm <- get.models((dredge_lm), subset = 1:5)
  dredge_gam <- dredge(global_gam)
  top5_gam <- get.models(dredge_gam, subset = 1:5)
  
  terms_lm <- list()
  terms_gam <- list()
  for(i in 1:5){
    terms_lm[[i]] <- top5_lm[[i]]$terms
    terms_gam[[i]] <- top5_gam[[i]]$formula
  }
  
  # 10 fold CV split
  set.seed(6)
  splits <- caret::createFolds(df$read_age, k = 10, list = TRUE, returnTrain = FALSE)
  
  # Create calibration and test sets
  cal <- list()
  test <- list()
  for (i in 1:10) {
    pc.mod <- preProcess(df[-splits[[i]], -c(1:24)], method = c("pca","center"), pcaComp = num_pcs)
    pc.cal <- predict(pc.mod, df[-splits[[i]], -c(1:24)])
    pc.cal <- cbind(pc.cal, df[-splits[[i]], ])
    cal[[i]] <- pc.cal
    
    pc.test <- predict(pc.mod, df[splits[[i]], -c(1:24)])
    pc.test <- cbind(pc.test, df[splits[[i]], ])
    test[[i]] <- pc.test
  }
  
  # LM - top 5 models
  mods.lm <- replicate(10, list())
  r2_lm <- list()
  rmse_lm <- list()
  for (i in 1:10) {
    calibrate <- cal[[i]]
    testing <- test[[i]]
    r2_temp <- rep(0, 5)
    rmse_temp <- rep(0, 5)
    for (j in 1:5) {
      mod <- lm(data = calibrate, terms_lm[[j]])
      preds <- predict(mod, newdata = testing)
      rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
      RSS <- sum((testing$read_age - preds)^2)
      TSS <- sum((testing$read_age - mean(testing$read_age))^2)
      r2_temp[j] <- 1 - (RSS / TSS)
      mods.lm[[i]][[j]] <- mod
    }
    r2_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), r2 = r2_temp)
    rmse_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), rmse = rmse_temp)
  }
  
  # GAM - top 5 models
  mods.gam <- replicate(10, list())
  r2_gam <- list()
  rmse_gam <- list()
  for (i in 1:10) {
    calibrate <- cal[[i]]
    testing <- test[[i]]
    r2_temp <- rep(0, 5)
    rmse_temp <- rep(0, 5)
    for (j in 1:5) {
      mod <- gam(data = calibrate, terms_gam[[j]])
      preds <- predict(mod, newdata = testing)
      rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
      RSS <- sum((testing$read_age - preds)^2)
      TSS <- sum((testing$read_age - mean(testing$read_age))^2)
      r2_temp[j] <- 1 - (RSS / TSS)
      mods.gam[[i]][[j]] <- mod
    }
    r2_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), r2 = r2_temp)
    rmse_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), rmse = rmse_temp)
  }
  
  # Calculate final results
  r2_lm <- sapply(r2_lm, "[[", "r2")
  r2_lm <- rowMeans(r2_lm)
  rmse_lm <- sapply(rmse_lm, "[[", "rmse")
  rmse_lm <- rowMeans(rmse_lm)
  # Create lm_result with formulas
  lm_result <- data.frame(
    Model = paste0("Linear ", 1:5),
    Mean_R2 = r2_lm,
    Mean_RMSE = rmse_lm,
    Formula = sapply(top5_lm, function(x) Reduce(paste, deparse(formula(x))))  # Extract formula
  )
  # GAM results
  r2_gam <- sapply(r2_gam, "[[", "r2")
  r2_gam <- rowMeans(r2_gam)
  rmse_gam <- sapply(rmse_gam, "[[", "rmse")
  rmse_gam <- rowMeans(rmse_gam)
  # Create gam_result with formulas
  gam_result <- data.frame(
    Model = paste0("GAM ", 1:5),
    Mean_R2 = r2_gam,
    Mean_RMSE = rmse_gam,
    Formula = sapply(top5_gam, function(x) Reduce(paste, deparse(formula(x))))  # Extract formula
  )
  # Combine results and add number of PCs
  results <- rbind(lm_result, gam_result)
  results$Num_PCs <- num_pcs
  # Store results for this iteration
  all_results[[paste0("PCs_", num_pcs)]] <- results
}
# Combine all results into one dataframe
final_results <- do.call(rbind, all_results)
rownames(final_results) <- NULL
final_results <-  dplyr::distinct(final_results, Mean_RMSE, .keep_all = TRUE) # remove duplicates

rm(cal, calibrate, dredge_gam, dredge_lm, gam_result, global_gam, global_lm, lm_result, mod, mods.gam, mods.lm, pc_df, pc.cal, pc.mod, pc.test, pca_temp, splits, terms_gam, test, testing, gam_formula_terms, i, j, lm_formula_terms, num_pcs, preds, r2_gam, r2_lm, r2_temp, rmse_gam, rmse_lm, rmse_temp, RSS, TSS)

# list top 5 for GAM and LM
top_linear_models <- final_results %>%
  dplyr::filter(str_detect(Model, "Linear")) %>%
  dplyr::arrange(Mean_RMSE) %>%
  dplyr::slice_head(n = 5)
top_gam_models <- final_results %>%
  dplyr::filter(str_detect(Model, "GAM")) %>%
  dplyr::arrange(Mean_RMSE) %>%
  dplyr::slice_head(n = 5)
# Create formula objects (as they would be used in modeling functions)
linear_formulas <- lapply(top_linear_models$Formula, as.formula)
gam_formulas <- lapply(top_gam_models$Formula, as.formula)

# Create named lists with additional metadata
top5_lm <- list(
  formulas = linear_formulas,
  performance = top_linear_models[, c("Model", "Mean_R2", "Mean_RMSE")]
)
top5_gam <- list(
  formulas = gam_formulas,
  performance = top_gam_models[, c("Model", "Mean_R2", "Mean_RMSE")]
)
rm(top_linear_models, top_gam_models, linear_formulas, gam_formulas)
```

# Function - run model with df

For above code to run all modelling with a given data frame (playing around with removing specimens to see performance) are in this section

```{r eval=FALSE, include=FALSE}
run_models <- function(mydf){
df <- {{mydf}}
df <- df[complete.cases(df$read_age), ] 
pca_temp <- pca(df[, 21:ncol(df), ])
pc_df <- data.frame(PC1 = rep(0,nrow(df)))
for (i in 1:10) {
  pc_df[, paste0("PC", i)] <- pca_temp$res$cal$scores[, i]
  rm(i)
}
pc_df <- cbind(pc_df,df) 
global_lm <- lm(data = pc_df, read_age ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10)
global_gam <- gam(data = pc_df, read_age ~ s(PC1,k=4) + s(PC2,k=4) + s(PC3,k=4) + s(PC4,k=4) + s(PC5,k=4) + s(PC6,k=4)) # adding too many PC's makes GAM have horrible fit and R2

# dredge to find top 5 models
options(na.action = "na.fail")
dredge_lm <- dredge(global_lm)
top5_lm <- get.models(dredge_lm, subset = 1:5)

# get.models was not working for gam in a function format so had to adapt this insane code to extract the top models....
dredge_gam <- dredge(global_gam)
top5_gam_temp <- dredge_gam[1:5,] 
reconstruct_formula <- function(model_row) {
  terms_used <- names(model_row)[!is.na(model_row)]   # Extract the terms used in the model
  terms_used <- terms_used[!terms_used %in% c("(Intercept)", "df", "logLik", "AICc", "delta", "weight")]   # Remove non-predictor columns (e.g., AIC, delta, weight)
  as.formula(paste("read_age ~", paste(terms_used, collapse = " + ")))   # Construct the formula
}
top5_gam <- lapply(1:5, function(i) {reconstruct_formula(top5_gam_temp[i, ])})
rm(top5_gam_temp, reconstruct_formula)

# create lists for top 5 models of lm and gam
terms_lm <- list()
terms_gam <- list()
for(i in 1:5){
  terms_lm[[i]] <- top5_lm[[i]]$terms
  terms_gam[[i]] <- top5_gam[[i]]
  rm(i)
}
rm(global_gam, global_lm, top5_gam, top5_lm, pc_df, pca_temp, dredge_gam,dredge_lm)

# 10 fold CV split - create folds
set.seed(6)
splits <- caret::createFolds(df$read_age, k = 10, list = TRUE, returnTrain = FALSE)
# extract PC's for each calibration set, create test sets with ages and spectra
cal <- list()
test <- list()
for (i in 1:10) {
  # calibration set and PC's
  pc.mod <- preProcess(df[-splits[[i]], -c(1:20)], method = c("pca","center"), pcaComp = 10)
  pc.cal <- predict(pc.mod, df[-splits[[i]], -c(1:20)])
  pc.cal <- cbind(pc.cal, df[-splits[[i]], ])
  cal[[i]] <- pc.cal
  rm(pc.cal)
  # test sets
  pc.test <- predict(pc.mod, df[splits[[i]], -c(1:20)])
  pc.test <- cbind(pc.test, df[splits[[i]], ])
  test[[i]] <- pc.test
  rm(pc.test, pc.mod, i)
}

mods.lm <- replicate(10, list())
r2_lm <- list()
rmse_lm <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  r2_temp <- rep(0, 5)
  rmse_temp <- rep(0, 5)
  for (j in 1:5) {
    mod <- lm(data = calibrate, terms_lm[[j]])
    preds <- predict(mod, newdata = testing)
    rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
    RSS <- sum((testing$read_age - preds)^2)
    TSS <- sum((testing$read_age - mean(testing$read_age))^2)
    r2_temp[j] <- 1 - (RSS / TSS)
    mods.lm[[i]][[j]] <- mod
    rm(RSS, TSS, preds)
  }
  r2_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), r2 = r2_temp)
  rmse_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), rmse = rmse_temp)
  rm(mod, i, j, r2_temp, rmse_temp, calibrate, testing)
}
# Unlist the r2 & rmse values and calc means for 10 splits, combine into dataframe
r2_lm <- sapply(r2_lm, "[[", "r2")
r2_lm <- rowMeans(r2_lm)
rmse_lm <- sapply(rmse_lm, "[[", "rmse")
rmse_lm <- rowMeans(rmse_lm)
(lm_result <- data.frame(
  Model = paste0("Linear ", 1:5),
  Mean_R2 = r2_lm,
  Mean_RMSE = rmse_lm
))
rm(r2_lm, rmse_lm)

mods.gam <- replicate(10, list())
r2_gam <- list()
rmse_gam <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  r2_temp <- rep(0, 5)
  rmse_temp <- rep(0, 5)
  for (j in 1:5) {
    mod <- gam(data = calibrate, terms_gam[[j]])
    preds <- predict(mod, newdata = testing)
    rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
    RSS <- sum((testing$read_age - preds)^2)
    TSS <- sum((testing$read_age - mean(testing$read_age))^2)
    r2_temp[j] <- 1 - (RSS / TSS)
    mods.gam[[i]][[j]] <- mod
    rm(RSS, TSS, preds)
  }
  r2_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), r2 = r2_temp)
  rmse_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), rmse = rmse_temp)
  rm(j, i, testing, calibrate, mod, r2_temp, rmse_temp)
}
r2_gam <- sapply(r2_gam, "[[", "r2")
r2_gam <- rowMeans(r2_gam)
rmse_gam <- sapply(rmse_gam, "[[", "rmse")
rmse_gam <- rowMeans(rmse_gam)
gam_result <- data.frame(
  Model = paste0("GAM ", 1:5),
  Mean_R2 = r2_gam,
  Mean_RMSE = rmse_gam
)
(results <- rbind(lm_result,gam_result))
rm(r2_gam, rmse_gam, lm_result, gam_result)

mods.pls <- list()
mods.vip <- list()
r2.pls <- list()
rmse.pls <- list()
for (i in 1:10) {
  calibrate <- cal[i]
  testing <- test[i]
  mod <- pls(calibrate[, 31:ncol(calibrate)], calibrate[, "read_age"],
             scale = F, center = T, cv = 1, #ncomp.selcrit = "wold", removing for now, seems too conservative with number of selected components
             info = "Age Prediction Model",
             x.test = testing[, 31:ncol(testing)],
             y.test = testing[, "read_age"]
  )
  ncomp <- mod$ncomp.selected
  rmse.pls$pls.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
  r2.pls$pls.test[i] <- mod$testres$r2[[ncomp]] # extract r2
  
  # VIP filter of wavenumbers
  mods.pls[[i]] <- mod
  mods.pls[[1]]$ncomp.selected
  vip <- as.data.frame(vipscores(mod))
  mod <- pls(calibrate[, 31:ncol(calibrate)], calibrate[, "read_age"],
             # ncomp.selcrit = "wold", 
             scale = F, center = T, cv = 1,
             info = "Age Prediction Model",
             x.test = testing[, 31:ncol(testing)],
             y.test = testing[, "read_age"],
             exclcols = vip$V1 < 0.5
  )
  ncomp <- mod$ncomp.selected
  r2.pls$vip.test[i] <- mod$testres$r2[[ncomp]] # extract r2
  mods.vip[[i]] <- mod
  rmse.pls$vip.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
  rm(calibrate, mod, ncomp, testing, i, vip)
}
(results <- rbind(results, 
                  c("PLS", mean(r2.pls$pls.test), mean(rmse.pls$pls.test)),
                  c("VIP", mean(r2.pls$vip.test), mean(rmse.pls$vip.test))
))
rm(rmse.pls, r2.pls)
mods.lm <<- mods.lm
mods.gam <<- mods.gam
mods.pls <<- mods.pls
mods.vip <<- mods.vip
return(results)
}

### Store models from output

get.my.models <- function() {
  model.list <- list(lm = list(), gam = list(), pls = list(), vip = list())
  model.list$lm <- mods.lm
  model.list$gam <- mods.gam
  model.list$pls <- mods.pls
  model.list$vip <- mods.vip
  rm(mods.lm, mods.gam, mods.pls, mods.vip, pos = .GlobalEnv)
  return(model.list)
}

```

### All waves, no outliers removed

```{r}
# ALL waves
# df <- readRDS("RDS_dataframes/LPW_scan_avg_proc.RDS")
# results_allwaves <- run_models(df)
# mods_allwaves <- get.my.models()
# saveRDS(results_allwaves, "results_allwaves.RDS")
# saveRDS(mods_allwaves, "mods_allwaves.RDS")
```

### Filter waves \> 7500, no outliers removed

```{r}
# cut off >7500 wavenumber
#df <- readRDS("RDS_dataframes/LPW_scan_avg_proc_filter.RDS")
#results_filtered <- run_models(df)
#mods_filtered <- get.my.models()
#saveRDS(results_filtered, "results_filtered.RDS")
#saveRDS(mods_filtered, "mods_filtered.RDS")
```

### All wavenumbers, remove specimen #77 (very large)

```{r}
# remove specimen 77, all waves
#df <- readRDS("RDS_dataframes/LPW_scan_avg_proc.RDS")
#df <- df %>% filter(specimen != 77)
#results_allwaves_no77 <- run_models(df)
#mods_allwaves_no77 <- get.my.models()
#saveRDS(results_allwaves_no77, "results_allwaves_no77.RDS")
#saveRDS(mods_allwaves_no77, "mods_allwaves_no77.RDS")
```

# Outliers

Below are a couple approaches to find outliers I'm looking at without nit-picking via spectra and PCA. Many of the specimens are borderline whether they should be removed or not, however I have previously seen that my largest specimen (#77, 182 cm) seems to throw off estimates and may be excluded from analysis.

```{r}
df <- readRDS("RDS_dataframes/LPW_scan_avg_proc.RDS")
df <- df[complete.cases(df$read_age), ] 
# find outliers based on length and age - >2 SD
model <- lm(read_age ~ length, data = df)
df.temp <- df %>%
    mutate(residuals = resid(model))
residual_threshold <- 2 * sd(df.temp$residuals)
df.temp <- df.temp %>%
    mutate(outlier = ifelse(abs(residuals) > residual_threshold, TRUE, FALSE))
df.temp %>%
    ggplot(aes(x = length, y = read_age)) +
    geom_smooth(color = "#003d5b", 
                method = "lm", 
                se = FALSE,
                linewidth = 3,
                alpha = 0.2) +
    geom_point(aes(color = outlier), 
               size = 4, 
               alpha = 0.6,
               ) + 
    scale_color_manual(values = c("FALSE" = "#00798c", 
                                  "TRUE" = "#c1121f")) +
    ggrepel::geom_label_repel(
      data = filter(df.temp, outlier),
      aes(label = specimen),
      size = 3, 
      nudge_x = 0.1, 
      nudge_y = 0.1, 
      color = "#c1121f"
    ) +
    theme_minimal() + 
    labs(title = "length vs age",
         x = "length",
         y = "read age") +
    guides(color = guide_legend(override.aes = list(size = 4))) + 
    theme(text = element_text(size = 12),  # Default text size for all text elements
          plot.title = element_text(size = 20, face="bold"),  # Plot title
          axis.title = element_text(size = 16),  # Axis titles (both x and y)
          axis.text = element_text(size = 14),  # Axis text (both x and y)
          legend.title = element_text(size = 14),  # Legend title
          legend.text = element_text(size = 12))  # Legend items
# outs <- c(31,67,74,95)


# find outliers based on otolith weight and age - >2 SD
df.temp <- df[complete.cases(df$structure_weight), ] # filter only aged specimens
model <- lm(read_age ~ structure_weight, data = df.temp)
df.temp <- df.temp %>%
    mutate(residuals = resid(model))
residual_threshold <- 2 * sd(df.temp$residuals)
df.temp <- df.temp %>%
    mutate(outlier = ifelse(abs(residuals) > residual_threshold, TRUE, FALSE))
df.temp %>%
    ggplot(aes(x = structure_weight, y = read_age)) +
    geom_smooth(color = "#003d5b", 
                method = "lm", 
                se = FALSE,
                linewidth = 3,
                alpha = 0.2) +
    geom_point(aes(color = outlier), 
               size = 4, 
               alpha = 0.6,
               ) + 
    scale_color_manual(values = c("FALSE" = "#00798c", 
                                  "TRUE" = "#c1121f")) +
    ggrepel::geom_label_repel(
      data = filter(df.temp, outlier),
      aes(label = specimen),
      size = 3, 
      nudge_x = 0.1, 
      nudge_y = 0.1, 
      color = "#c1121f"
    ) +
    theme_minimal() + 
    labs(title = "structure_weight vs age",
         x = "structure_weight",
         y = "read age") +
    guides(color = guide_legend(override.aes = list(size = 4))) + 
    theme(text = element_text(size = 12),  # Default text size for all text elements
          plot.title = element_text(size = 20, face="bold"),  # Plot title
          axis.title = element_text(size = 16),  # Axis titles (both x and y)
          axis.text = element_text(size = 14),  # Axis text (both x and y)
          legend.title = element_text(size = 14),  # Legend title
          legend.text = element_text(size = 12))  # Legend items
# outs <- c(67,74,77,95)
```
