Load Packages & Data
```{r}
library(pacman)
p_load(mdatools, prospectr, tidyverse, stringr, patchwork, gridExtra, cowplot, purrr, viridis, readxl, MuMIn, mgcv, tidyr, future, furrr)


df <- readRDS("RDS_dataframes/combined_IBM_LPW_raw.RDS") %>% filter(complete.cases(read_age))
# combine "Shumagins" and "Kodiak" for region
df <- df %>%
  mutate(
    region = case_when(
      region %in% c("E Shumagins", "W Shumagins") ~ "Shumagin",
      region %in% c("N Kodiak", "S Kodiak") ~ "Kodiak",
      TRUE ~ region
    )
  )

kodiak <- df %>% filter(region == "Kodiak")
shumagin <- df %>% filter(region == "Shumagin")
lpw <- df %>% filter(region == "LPW")
```

# Functions
Things to speed up writing some code....!
```{r}
# -------------------------------------------------------------------------- #
# Data processing
# -------------------------------------------------------------------------- #
# Evenly split dataframe based on read_age
generate_multiple_splits <- function(data, n_splits) {
  purrr::map(1:n_splits, ~caret::createFolds(data[["read_age"]], k = 10, list = TRUE, returnTrain = FALSE))
}

# get FT-NIRS wavenumber columns by search for only numerical column names
find_cols <- function(df){
  names(df)[grepl("^\\d", names(df))]
}

# take FT-NIRS dataframe, make it long for plotting
make_long <- function(df) {
  speccols <- find_cols(df)
  df_long <- df %>%
    tidyr::pivot_longer(cols = speccols) %>%
    # rename(wavenumber = name) %>%
    # mutate(wavenumber = as.numeric(as.character(wavenumber)))
    mutate(name = as.numeric(as.character(name)))
  return(df_long)
}


# -------------------------------------------------------------------------- #
# FT-NIRS preprocessing functions: Savitzky-Golay & SNV
# -------------------------------------------------------------------------- #
# apply SG filtering to dataframe
sg_filter <- function(df, m, p, w) {
  speccols <- find_cols(df)
  metacols <- setdiff(names(df), speccols)
  sg_matrix <- savitzkyGolay(df[, speccols], m = m, p = p, w = w)
  new_speccols <- find_cols(sg_matrix)
  names(sg_matrix) <- new_speccols
  return(dplyr::bind_cols(df[, metacols], as.data.frame(sg_matrix)))
}

# apply SNV to dataframe
snv_filter <- function(df){
  speccols <- find_cols(df)
  metacols <- setdiff(names(df), speccols)
  snv_matrix <- prep.snv(as.matrix(df[,speccols]))
  return(dplyr::bind_cols(df[, metacols], as.data.frame(snv_matrix)))
}

wave_filter <- function(df, wave){
  speccols <- find_cols(df)
  metacols <- setdiff(names(df), speccols)
  keep_me <- as.numeric(speccols) <= wave
  speccols_keep <- speccols[keep_me]
  return(df[, c(metacols, speccols_keep)])
}

# -------------------------------------------------------------------------- #
# Model Performance
# -------------------------------------------------------------------------- #
# get RPD from model
calculate_rpd <- function(observed, predicted) { sd(observed) / sqrt(mean((observed - predicted)^2)) }
# get bias from model
calculate_bias <- function(observed, predicted) { mean(predicted - observed) }

# -------------------------------------------------------------------------- #
# Plotting
# -------------------------------------------------------------------------- #
spec.fig <- function(df, color) {
  mydf <- make_long(df)
  # 1. Quoting to allow column name use later
  color_col <- rlang::enquo(color) 

  # 2. Sort the data by the grouping variable (specimen) and the x-axis (name)
  mydf_sorted <- mydf %>% 
    dplyr::arrange(specimen, name) 
  
  # 3. Plot the sorted data
  ggplot(mydf_sorted) +
    geom_path(aes(x = name, y = value, color = !!color_col, group = specimen)) +
    scale_x_reverse() +
    scale_color_viridis() +
    labs(y = "Absorbance", x = expression(paste("Wavenumber ", cm^-1))) + 
    theme_bw(base_size = 16)
}
```

# Overall goals


I am looking to explore variations in region-specific models.  I have specimens from LPW (eastern GOA), Kodiak (~ central), and Shumagin Islands (~ central but more western) to test various combinations of train/test sets.  

1) First will look at using one region to predict another, with expectations that Kodiak and Shumagin should predict each other fairly well.  LPW specimens are far older than any found in Shumagin/Kodiak so those are a crap shoot.  There are three iterations of this:

- Train: Kodiak, Test: Shumagin & LPW seperately
- Train: Shumagin, Test: Kodiak * LPW seperately
- Train: LPW, Test: Kodiak & Shumagin seperately

No 10-fold CV possible here, unless I do an additional internal 10-fold CV test?  Might be worth confirming the models predict themselves well, just like LPW.

2) Explore "combined"/"generalized" models that use specimens from multiple regions.

- Train: Shumagin & Kodiak combined (Wild)
- Train: Shumagin, Kodiak & LPW combined (ALL Combined)

I hope to determine whether models are best built with only regional specimens or if we can combine multiple regions and still predict fairly well.

# Preprocessing questions

I need to decide how to deal with preprocessing my spectral data.  LPW specimens are 109-205 days old; Shumagin 56-102 days; Kodiak 64-127.  Attempting to predict LPW specimens with the wild is extrapolation and I imagine is going to suck.  The biggest issue is that the spectral signal for wild fish seems "noisier" possibly due to dirty specimens and poorer quality scans, and far lower magnitude (peaks just aren't as large as LPW).  My LPW models worked best using ALL wavenumber and a first order, 3rd polymonial with a windows size of 17 Savitzky-Golay filter.  I need to quickly test a handful of different filters on these regions to see which works best when combining regions.  

I'll start with using a single simple PLS model (and LOOCV for selecting ncomp) to determine which filter works best for each region individually.  Then I'll expand to using 10-fold CV on the same sort of simple LOOCV PLS model with the wild and ALL combined dataframes.  I will collect RPD, Bias, RMSE and R2 from all these models and compare them.

Things to test: 

1) Various SG filters, starting with 1,3,17.  I'll consider smaller/larger windows, as well as 2nd order polynomials (2,3,17?)
2) SNV normalization (via prep.snv(as.matrix(df)) )
3) Wavenumber filtering (seems to be best when removing >7500 cm-1 based on previous papers).


# Individual region model testing with preprocessing
```{r}
# # Function to run a 10-fold CV PLS model from a single region, collecting R2, RMSE, RPD, & Bias from the splits.
# run_pls <- function(df, num_rounds){
#   # identify FT-NIRS columns
#   speccols <- find_cols(df)
#   
#   # split dataframe into test and cal dataframes:
#   test.idx <- generate_multiple_splits(df, num_rounds) # generate test indices for the 10-fold CV with number of rounds we want
#   cal.idx <- purrr::map(test.idx, ~ {
#   purrr::map(., ~ setdiff(1:nrow(df), .)) # . is one fold's indices
# })
# 
#   # Run each 10-fold CV:
#   rounds_results <- purrr::map2_dfr(cal.idx, test.idx, ~{
#     # .x is a list of 10 cal folds (for this one round)
#     # .y is a list of 10 test folds (for this one round)
#     
#     # 4. Iterate over the folds *within* this round
#     # We add .id = "fold" to track the fold (1-10)
#     fold_results <- purrr::map2_dfr(.x, .y, ~{
#       # .x is now ONE vector of cal indices
#       # .y is now ONE vector of test indices
#       
#       # --- This is your inner model logic, same as before ---
#       cal_data <- df[.x, ]
#       test_data <- df[.y, ]
#       
#       Xc <- as.matrix(cal_data[, speccols])
#       Yc <- cal_data$read_age
#       
#       Xt <- as.matrix(test_data[, speccols])
#       Yt <- test_data$read_age
#       
#       # Run PLS model with LOOCV to find ncomp
#       pls_model <- mdatools::pls(x = Xc, y = Yc, cv = 1, x.test = Xt, y.test = Yt, scale = F, center = T)
#       
#       # Extract ncomp from LOOCV
#       ncomp_selected <- pls_model$ncomp.selected
#       
# # 1. Create the prediction data frame first
#       pred_df <- data.frame(
#         specimen = test_data$specimen, 
#         obs_age = Yt,
#         pred_age = pls_model$testres$y.pred[, ncomp_selected, 1]
#       )
#       
#       # 2. Return ONE data.frame with metrics AND the predictions
#       #    This is now the *last* expression, so it gets returned
#       data.frame(
#         ncomp = ncomp_selected,
#         R2 = pls_model$testres$r2[ncomp_selected],
#         RMSE = pls_model$testres$rmse[ncomp_selected],
#         RPD = pls_model$testres$rpd[ncomp_selected],
#         Bias = pls_model$testres$bias[ncomp_selected],
#         predictions = I(list(pred_df)) # Nest pred_df in a list-column
#       )
#       
#     }, .id = "fold") # This .id gives you fold 1-10
#     
#     return(fold_results) # Return the data frame of 10 folds
#     
#   }, .id = "round") # This .id gives you round 1-10
#   
#   # Return the final data frame with 100 rows
#   return(rounds_results)
# }

```


# testing
```{r}
# Sys.time()
# # kodiak
# results_kodiak <- run_pls(sg_filter(kodiak, 1,3,17), 2)
# Sys.time()
# predictions_kodiak <- results_kodiak %>% 
#   tidyr::unnest(cols = predictions) %>% select(round, fold, specimen, obs_age, pred_age)
# results_kodiak <- results_kodiak %>% select(-predictions)
# # summarize across all folds for a given round, average for metrics R2:Bias
# summary_kodiak <- results_kodiak %>% group_by(round) %>%
#   summarize(
#     mean_ncomp = mean(ncomp),
#     mean_R2 = mean(R2),
#     mean_RMSE = mean(RMSE),
#     mean_RPD = mean(RPD),
#     mean_Bias = mean(Bias)
#   )

```

# now making it parallel for faster shit

```{r}
run_pls_parallel <- function(df, num_rounds = 10) {
  
  # Identify FT-NIRS columns
  speccols <- find_cols(df)
  
  # 1. Generate all indices (this step is still sequential and fast)
  test.idx_nested <- generate_multiple_splits(df, num_rounds) 
  cal.idx_nested <- purrr::map(test.idx_nested, ~ {
    purrr::map(., ~ setdiff(1:nrow(df), .)) 
  })

  # 2. "Flatten" the nested lists
  # This creates two simple lists, each (num_rounds * 10) elements long
  all_cal_idx <- purrr::flatten(cal.idx_nested)
  all_test_idx <- purrr::flatten(test.idx_nested)
  
  # --- Add 'round' and 'fold' IDs ---
  # This creates a data frame to track which job is which
  job_ids <- tidyr::expand_grid(round = 1:num_rounds, fold = 1:10)

  # 3. Run all models in parallel
  # We use future_map2_dfr to iterate over the two flattened lists
  all_results <- furrr::future_map2_dfr(
    all_cal_idx, 
    all_test_idx, 
    ~{
      # .x is one vector of cal indices
      # .y is one vector of test indices
      
      # --- This is your inner model logic, same as before ---
      cal_data <- df[.x, ]
      test_data <- df[.y, ]
      
      Xc <- as.matrix(cal_data[, speccols])
      Yc <- cal_data$read_age
      
      Xt <- as.matrix(test_data[, speccols])
      Yt <- test_data$read_age
      
      # Run PLS model 
      pls_model <- mdatools::pls(
        x = Xc, y = Yc, 
        cv = 1, 
        x.test = Xt, y.test = Yt, 
        scale = FALSE, center = TRUE,
      )
      
      # Extract ncomp from LOOCV
      ncomp_selected <- pls_model$ncomp.selected
      
      # Create the prediction data frame
      pred_df <- data.frame(
        specimen = test_data$specimen, 
        obs_age = Yt,
        pred_age = pls_model$testres$y.pred[, ncomp_selected, 1]
      )
      
      # Return ONE data.frame with metrics AND the predictions
      data.frame(
        ncomp = ncomp_selected,
        R2 = pls_model$testres$r2[ncomp_selected],
        RMSE = pls_model$testres$rmse[ncomp_selected],
        RPD = pls_model$testres$rpd[ncomp_selected],
        Bias = pls_model$testres$bias[ncomp_selected],
        predictions = I(list(pred_df)) # Our list-column
      )
      
    }, 
    # This is critical for reproducible results in parallel
    .options = furrr_options(seed = TRUE) 
  )
  
  # Combine the job IDs (round, fold) with the results
  final_results <- dplyr::bind_cols(job_ids, all_results)
  
  return(final_results)
}
```

# test

```{r}
# # USE 8 CORES I GUESS, SEEMS TO WORK FAIRLY WELL
# plan(multisession, workers = 8)
# 
# Sys.time()
# # results_kodiak_parallel <- run_pls_parallel(sg_filter(kodiak, m=1, p=3, w=17), num_rounds = 2)
# Sys.time()
# plan(sequential)
# 
# predictions_kodiak <- results_kodiak_parallel %>% 
#   tidyr::unnest(cols = predictions) %>% select(round, fold, specimen, obs_age, pred_age)
# results_kodiak_parallel <- results_kodiak_parallel %>% select(-predictions)
# # summarize across all folds for a given round, average for metrics R2:Bias
# 
# 
# summary_kodiak <- results_kodiak_parallel %>%
#   group_by(round) %>%
#   summarize(
#     across(
#       c(ncomp, R2, RMSE, RPD, Bias), # The columns you want to summarize
#       mean,                          # The function to apply
#       .names = "mean_{.col}"         # How to name the new columns
#     ),
#     comp_range = paste0(min(ncomp), " - ", max(ncomp))
#   )
# 
# summary_kodiak_expanded <- results_kodiak_parallel %>%
#   group_by(round) %>%
#   summarize(
#     across(
#       c(ncomp, R2, RMSE, RPD, Bias),
#       list(mean = mean, sd = sd), # Provide a named list of functions
#       .names = "{.fn}_{.col}"     # '{.fn}' uses the names from your list
#     ),
#     comp_range = paste0(min(ncomp), " - ", max(ncomp))
#   )




```
parallel works WAY better, going to use that to identify WHICH version of filtering works best.  



Okay now run models for each region with the following filters: 

1) Only SG: 1,3,17
2) Only SNV
3) SNV FIRST, then SG: 1,3,17
4) SG: 1,3,17 FIRST, then SNV
5) remove waves > 7500, then SG 1,3,17

# we'll also try a seperate chunk of models with various SG filters only

1) 1,3,17
2) 2,3,17,
3) 1,3,19
4) 1,3,21


# Code to allow me to run a seperate test and calibration dataframe:

```{r}
# =============================================================================
#
# FT-NIRS MODELING FRAMEWORK - SIMPLE TRAIN/TEST SCRIPT
#
# This script is designed to run all model types on a *single*
# calibration (cal_df) and testing (test_df) dataset.
# It does *not* perform cross-validation and is used to test the
# core logic, metrics, and outputs before building a larger CV loop.
#
# =============================================================================

# --- 1. Load Required Libraries ---
# Make sure all are loaded in your session
# library(pacman)
# p_load(mdatools, prospectr, tidyverse, stringr, purrr, viridis,
#        mgcv, MuMIn, caret, ranger, xgboost, dplyr)

# =============================================================================
# --- 2. HELPER FUNCTIONS (Adapted from your Chapter 1 script) ---
# =============================================================================

# --- A. Metric Calculators ---

# --- C. PCA Importance Helpers ---
# Gets the PC numbers (e.g., 1, 3, 5) from an lm() model term
get_pc_indices_lm <- function(term) {
  as.numeric(stringr::str_extract(attr(term, "term.labels"), "\\d+"))
}

# Gets the PC numbers from a gam() model formula
get_pc_indices_gam <- function(formula) {
  pc_matches <- stringr::str_extract_all(as.character(formula)[3], "PC\\d+")[[1]]
  as.numeric(stringr::str_extract(pc_matches, "\\d+"))
}

# Calculates the vector length (importance) for a set of PCs
calculate_pca_vector_length <- function(pca_rotation, pc_indices) {
  if (length(pc_indices) == 0 || is.null(pc_indices)) {
    return(rep(0, nrow(pca_rotation)))
  }
  valid_indices <- pc_indices[!is.na(pc_indices)]
  if (length(valid_indices) == 0) {
    return(rep(0, nrow(pca_rotation)))
  }
  loadings_subset <- pca_rotation[, valid_indices, drop = FALSE]
  sqrt(rowSums(loadings_subset^2))
}


# =============================================================================
# --- 3. INDIVIDUAL MODEL RUNNER FUNCTIONS ---
# =============================================================================
# Each function takes cal_df and test_df, runs one model type,
# and returns a list with 3 data frames:
# $metrics, $predictions, $importance

# --- A. PCA-LM and PCA-GAM Runner ---
# --- A. PCA-LM and PCA-GAM Runner ---
run_pca_models <- function(cal_df, test_df, terms_lm, terms_gam) {
  
  cli::cli_h1("Running PCA-LM & PCA-GAM Models")
  spec_cols <- find_cols(cal_df)
  
  # 1. Perform PCA on cal_df
  cli::cli_alert_info("Calculating PCA...")
  pc.mod <- caret::preProcess(cal_df[, spec_cols],
                              method = c("center", "pca"),
                              pcaComp = 10
  )
  
  # Add PC scores to cal and test data
  cal_pca <- cbind(predict(pc.mod, cal_df[, spec_cols]), cal_df)
  test_pca <- cbind(predict(pc.mod, test_df[, spec_cols]), test_df)
  
  # 2. Run Dredge to find top 10 models for LM and GAM
  # --- THIS SECTION HAS BEEN REMOVED ---
  # The dredge logic is now run *before* calling this function.
  # We are now using the 'terms_lm' and 'terms_gam' arguments passed in.
  
  # --- 3. Run LM Models & Get Metrics ---
  cli::cli_alert_info("Running 10 LM models...")
  metrics_lm_list <- list()
  preds_lm_list <- list()
  importance_lm_list <- list()
  
  for (j in 1:10) {
    mod_name <- paste("PCA-LM", j)
    mod <- lm(data = cal_pca, terms_lm[[j]])
    preds <- predict(mod, newdata = test_pca)
    
    metrics_lm_list[[j]] <- data.frame(
      model_type = "PCA-LM",
      model_name = mod_name,
      R2 = cor(preds, test_pca$read_age)^2,
      RMSE = caret::RMSE(pred = preds, obs = test_pca$read_age),
      RPD = calculate_rpd(test_pca$read_age, preds),
      Bias = calculate_bias(test_pca$read_age, preds)
    )
    
    preds_lm_list[[j]] <- data.frame(
      specimen = test_pca$specimen,
      obs_age = test_pca$read_age,
      pred_age = preds,
      model_type = "PCA-LM",
      model_name = mod_name
    )
    
    pc_indices <- get_pc_indices_lm(terms_lm[[j]])
    importance_lm_list[[j]] <- data.frame(
      model_type = "PCA-LM",
      model_name = mod_name,
      wavenumber = as.numeric(rownames(pc.mod$rotation)),
      importance = calculate_pca_vector_length(pc.mod$rotation, pc_indices)
    )
  }
  
  # --- 4. Run GAM Models & Get Metrics ---
  cli::cli_alert_info("Running 10 GAM models...")
  metrics_gam_list <- list()
  preds_gam_list <- list()
  importance_gam_list <- list()
  
  for (j in 1:10) {
    mod_name <- paste("PCA-GAM", j)
    mod <- gam(data = cal_pca, terms_gam[[j]], method = "REML")
    preds <- predict(mod, newdata = test_pca)
    
    metrics_gam_list[[j]] <- data.frame(
      model_type = "PCA-GAM",
      model_name = mod_name,
      R2 = cor(preds, test_pca$read_age)^2,
      RMSE = caret::RMSE(pred = preds, obs = test_pca$read_age),
      RPD = calculate_rpd(test_pca$read_age, preds),
      Bias = calculate_bias(test_pca$read_age, preds)
    )
    
    preds_gam_list[[j]] <- data.frame(
      specimen = test_pca$specimen,
      obs_age = test_pca$read_age,
      pred_age = preds,
      model_type = "PCA-GAM",
      model_name = mod_name
    )
    
    pc_indices <- get_pc_indices_gam(terms_gam[[j]])
    importance_gam_list[[j]] <- data.frame(
      model_type = "PCA-GAM",
      model_name = mod_name,
      wavenumber = as.numeric(rownames(pc.mod$rotation)),
      importance = calculate_pca_vector_length(pc.mod$rotation, pc_indices)
    )
  }
  
  # 5. Combine and Return
  return(list(
    metrics = bind_rows(metrics_lm_list, metrics_gam_list),
    predictions = bind_rows(preds_lm_list, preds_gam_list),
    importance = bind_rows(importance_lm_list, importance_gam_list)
  ))
}

# --- B. PLS and PLS-VIP Runner ---
run_pls_models <- function(cal_df, test_df, vip_threshold = 1.0) {
  
  cli::cli_h1("Running PLS & PLS-VIP Models")
  spec_cols <- find_cols(cal_df)
  
  Xc <- as.matrix(cal_df[, spec_cols])
  Yc <- as.matrix(cal_df$read_age) # <-- FIX: Convert vector to matrix
  Xt <- as.matrix(test_df[, spec_cols])
  Yt <- as.matrix(test_df$read_age) # <-- FIX: Convert vector to matrix
  
  # --- Model 1: Standard PLS ---
  cli::cli_alert_info("Running standard PLS with LOOCV...")
  mod_pls <- mdatools::pls(Xc, Yc,
                           cv = 1, # LOOCV to find ncomp
                           scale = FALSE, center = TRUE,
                           x.test = Xt, y.test = Yt
  )
  ncomp_pls <- mod_pls$ncomp.selected
  
  # Get VIP scores from this first model
  vip_scores <- mdatools::vipscores(mod_pls, ncomp = ncomp_pls)
  
  # --- Model 2: PLS-VIP ---
  cli::cli_alert_info(paste("Running PLS-VIP (threshold >", vip_threshold, ")..."))
  
  # Find columns to exclude
  excl_cols_vip <- which(vip_scores < vip_threshold)
  
  mod_vip <- mdatools::pls(Xc, Yc,
                           cv = 1, # LOOCV to find ncomp
                           scale = FALSE, center = TRUE,
                           x.test = Xt, y.test = Yt,
                           exclcols = excl_cols_vip
  )
  ncomp_vip <- mod_vip$ncomp.selected
  
  # --- 3. Combine Metrics ---
  metrics <- bind_rows(
    data.frame(
      model_type = "PLS",
      model_name = "PLS",
      R2 = mod_pls$testres$r2[ncomp_pls],
      RMSE = mod_pls$testres$rmse[ncomp_pls],
      RPD = mod_pls$testres$rpd[ncomp_pls],
      Bias = mod_pls$testres$bias[ncomp_pls]
    ),
    data.frame(
      model_type = "PLS-VIP",
      model_name = paste0("PLS-VIP (>", vip_threshold, ")"),
      R2 = mod_vip$testres$r2[ncomp_vip],
      RMSE = mod_vip$testres$rmse[ncomp_vip],
      RPD = mod_vip$testres$rpd[ncomp_vip],
      Bias = mod_vip$testres$bias[ncomp_vip]
    )
  )
  
  # --- 4. Combine Predictions ---
  predictions <- bind_rows(
    data.frame(
      specimen = test_df$specimen,
      obs_age = Yt,
      pred_age = mod_pls$testres$y.pred[, ncomp_pls, 1],
      model_type = "PLS",
      model_name = "PLS"
    ),
    data.frame(
      specimen = test_df$specimen,
      obs_age = Yt,
      pred_age = mod_vip$testres$y.pred[, ncomp_vip, 1],
      model_type = "PLS-VIP",
      model_name = paste0("PLS-VIP (>", vip_threshold, ")")
    )
  )
  
  # --- 5. Combine Importance ---
  importance <- data.frame(
    model_type = "PLS-VIP",
    model_name = "PLS-VIP", # VIP scores are from the main PLS model
    wavenumber = as.numeric(colnames(Xc)),
    importance = as.vector(vip_scores)
  )
  
  return(list(
    metrics = metrics,
    predictions = predictions,
    importance = importance
  ))
}

# --- C. XGBoost Runner ---
run_xgb_model <- function(cal_df, test_df, best_params_xgb) {
  
  cli::cli_h1("Running XGBoost Model")
  spec_cols <- find_cols(cal_df)
  
  # 1. Prepare DMatrix
  dtrain <- xgb.DMatrix(data = as.matrix(cal_df[, spec_cols]), label = cal_df$read_age)
  dtest <- xgb.DMatrix(data = as.matrix(test_df[, spec_cols]), label = test_df$read_age)
  
  # 2. Set Parameters from your tuning
  params <- list(
    objective = "reg:squarederror",
    booster = "gbtree",
    nthread = 1,
    eta = best_params_xgb$eta,
    max_depth = best_params_xgb$max_depth,
    min_child_weight = best_params_xgb$min_child_weight,
    subsample = best_params_xgb$subsample,
    gamma = best_params_xgb$gamma,
    colsample_bytree = best_params_xgb$colsample_bytree
  )
  best_nrounds <- round(best_params_xgb$mean_nrounds)
  
  # 3. Train Model
  cli::cli_alert_info(paste("Training for", best_nrounds, "rounds..."))
  mod_xgb <- xgb.train(
    params = params,
    data = dtrain,
    nrounds = best_nrounds,
    verbose = 0
  )
  
  # 4. Get Predictions and Metrics
  preds <- predict(mod_xgb, dtest)
  obs <- test_df$read_age
  
  metrics <- data.frame(
    model_type = "XGB",
    model_name = "XGBoost",
    R2 = cor(preds, obs)^2,
    RMSE = caret::RMSE(pred = preds, obs = obs),
    RPD = calculate_rpd(obs, preds),
    Bias = calculate_bias(obs, preds)
  )
  
  predictions <- data.frame(
    specimen = test_df$specimen,
    obs_age = obs,
    pred_age = preds,
    model_type = "XGB",
    model_name = "XGBoost"
  )
  
  # 5. Get Importance
  imp_matrix <- xgb.importance(model = mod_xgb)
  imp_df <- data.frame(Feature = colnames(cal_df[, spec_cols])) %>%
    left_join(imp_matrix, by = "Feature") %>%
    mutate(
      model_type = "XGB",
      model_name = "XGBoost",
      wavenumber = as.numeric(Feature),
      importance = ifelse(is.na(Gain), 0, Gain)
    ) %>%
    select(model_type, model_name, wavenumber, importance)
  
  return(list(
    metrics = metrics,
    predictions = predictions,
    importance = imp_df
  ))
}

# --- D. Random Forest Runner ---
run_rf_model <- function(cal_df, test_df, best_params_rf) {
  
  cli::cli_h1("Running Random Forest Model")
  spec_cols <- find_cols(cal_df)
  
  # 1. Train Model
  cli::cli_alert_info("Training with ranger...")
  mod_rf <- ranger(
    x = cal_df[, spec_cols],
    y = cal_df$read_age,
    mtry = best_params_rf$mtry,
    min.node.size = best_params_rf$min.node.size,
    splitrule = "variance",
    importance = 'permutation',
    num.trees = 1000,
    num.threads = 1 # Keep single-threaded for simple test
  )
  
  # 2. Get Predictions and Metrics
  preds <- predict(mod_rf, data = test_df)$predictions
  obs <- test_df$read_age
  
  metrics <- data.frame(
    model_type = "RF",
    model_name = "Random Forest",
    R2 = cor(preds, obs)^2,
    RMSE = caret::RMSE(pred = preds, obs = obs),
    RPD = calculate_rpd(obs, preds),
    Bias = calculate_bias(obs, preds)
  )
  
  predictions <- data.frame(
    specimen = test_df$specimen,
    obs_age = obs,
    pred_age = preds,
    model_type = "RF",
    model_name = "Random Forest"
  )
  
  # 3. Get Importance
  importance <- data.frame(
    model_type = "RF",
    model_name = "Random Forest",
    wavenumber = as.numeric(names(mod_rf$variable.importance)),
    importance = as.vector(mod_rf$variable.importance)
  )
  
  return(list(
    metrics = metrics,
    predictions = predictions,
    importance = importance
  ))
}

# --- E. Simple Models (Bio-only) Runner ---
run_simple_models <- function(cal_df, test_df) {
  
  cli::cli_h1("Running Simple (Bio-Only) Models")
  
  # --- CRITICAL: Filter for complete cases for *these* models ---
  bio_vars <- c("read_age", "length", "structure_weight")
  
  cal_simple <- cal_df[complete.cases(cal_df[, bio_vars]), ]
  test_simple <- test_df[complete.cases(test_df[, bio_vars]), ]
  
  if (nrow(cal_simple) == 0 || nrow(test_simple) == 0) {
    cli::cli_alert_warning("Skipping simple models: no complete cases for bio-vars.")
    return(NULL)
  }
  
  cli::cli_alert_info(paste(
    "Using", nrow(cal_simple), "cal samples and",
    nrow(test_simple), "test samples."
  ))
  
  # --- Model 1: Simple LM ---
  mod_lm <- lm(read_age ~ length + structure_weight, data = cal_simple)
  preds_lm <- predict(mod_lm, newdata = test_simple)
  obs <- test_simple$read_age
  
  metrics_lm <- data.frame(
    model_type = "Simple",
    model_name = "Simple-LM (L+SW)",
    R2 = cor(preds_lm, obs)^2,
    RMSE = caret::RMSE(pred = preds_lm, obs = obs),
    RPD = calculate_rpd(obs, preds_lm),
    Bias = calculate_bias(obs, preds_lm)
  )
  
  preds_lm_df <- data.frame(
    specimen = test_simple$specimen,
    obs_age = obs,
    pred_age = preds_lm,
    model_type = "Simple",
    model_name = "Simple-LM (L+SW)"
  )
  
  # --- Model 2: Simple GAM ---
  mod_gam <- mgcv::gam(read_age ~ s(length, k = 4) + s(structure_weight, k = 4),
                       data = cal_simple,
                       method = "REML"
  )
  preds_gam <- predict(mod_gam, newdata = test_simple)
  
  metrics_gam <- data.frame(
    model_type = "Simple",
    model_name = "Simple-GAM (L+SW)",
    R2 = cor(preds_gam, obs)^2,
    RMSE = caret::RMSE(pred = preds_gam, obs = obs),
    RPD = calculate_rpd(obs, preds_gam),
    Bias = calculate_bias(obs, preds_gam)
  )
  
  preds_gam_df <- data.frame(
    specimen = test_simple$specimen,
    obs_age = obs,
    pred_age = preds_gam,
    model_type = "Simple",
    model_name = "Simple-GAM (L+SW)"
  )
  
  # --- 3. Combine and Return ---
  # No importance for these models
  return(list(
    metrics = bind_rows(metrics_lm, metrics_gam),
    predictions = bind_rows(preds_lm_df, preds_gam_df),
    importance = data.frame() # Return empty data frame
  ))
}


# =============================================================================
# --- 4. MAIN WRAPPER FUNCTION ---
# =============================================================================

#' Run All Models on a Single Train/Test Split
#'
#' This function runs all 5 model types (PCA-LM/GAM, PLS/PLS-VIP,
#' XGB, RF, and Simple-LM/GAM) on a given calibration and
#' testing dataset.
#'
#' @param cal_df The dataframe for model calibration (training).
#' @param test_df The dataframe for model testing.
#' @param best_params_xgb A 1-row data frame with tuned XGBoost params.
#' @param best_params_rf A 1-row data frame with tuned Random Forest params.
#' @param vip_threshold The VIP score threshold for the PLS-VIP model.
#'
#' @return A list containing three data frames:
#'         $all_metrics, $all_predictions, $all_importance
#'
run_all_models_simple <- function(cal_df, test_df,
                                  best_params_xgb, best_params_rf,
                                  terms_lm, terms_gam, # <-- ADDED ARGUMENTS
                                  vip_threshold = 1.0) {
  
  # --- Initialize containers ---
  metrics_list <- list()
  predictions_list <- list()
  importance_list <- list()
  
  # --- Run Model Groups ---
  # Use tryCatch to allow the script to continue if one model fails
  
  tryCatch({
    # Pass the dredge results into the function
    pca_results <- run_pca_models(cal_df, test_df, terms_lm, terms_gam) # <-- UPDATED CALL
    metrics_list$pca <- pca_results$metrics
    predictions_list$pca <- pca_results$predictions
    importance_list$pca <- pca_results$importance
  }, error = function(e) { cli::cli_alert_danger("PCA Models FAILED: {e$message}") })
  
  tryCatch({
    pls_results <- run_pls_models(cal_df, test_df, vip_threshold)
    metrics_list$pls <- pls_results$metrics
    predictions_list$pls <- pls_results$predictions
    importance_list$pls <- pls_results$importance
  }, error = function(e) { cli::cli_alert_danger("PLS Models FAILED: {e$message}") })
  
  tryCatch({
    xgb_results <- run_xgb_model(cal_df, test_df, best_params_xgb)
    metrics_list$xgb <- xgb_results$metrics
    predictions_list$xgb <- xgb_results$predictions
    importance_list$xgb <- xgb_results$importance
  }, error = function(e) { cli::cli_alert_danger("XGBoost Model FAILED: {e$message}") })
  
  tryCatch({
    rf_results <- run_rf_model(cal_df, test_df, best_params_rf)
    metrics_list$rf <- rf_results$metrics
    predictions_list$rf <- rf_results$predictions
    importance_list$rf <- rf_results$importance
  }, error = function(e) { cli::cli_alert_danger("Random Forest Model FAILED: {e$message}") })
  
  tryCatch({
    simple_results <- run_simple_models(cal_df, test_df)
    if (!is.null(simple_results)) {
      metrics_list$simple <- simple_results$metrics
      predictions_list$simple <- simple_results$predictions
      # importance_list$simple is not needed (it's empty)
    }
  }, error = function(e) { cli::cli_alert_danger("Simple Models FAILED: {e$message}") })
  
  
  # --- Combine All Results ---
  cli::cli_h1("Combining All Results")
  
  all_metrics <- bind_rows(metrics_list)
  all_predictions <- bind_rows(predictions_list)
  all_importance <- bind_rows(importance_list)
  
  return(list(
    all_metrics = all_metrics,
    all_predictions = all_predictions,
    all_importance = all_importance
  ))
}
```


```{r}
# =============================================================================
# --- 5. EXAMPLE USAGE (How to run this script) ---
# =============================================================================

cal_df <- NULL
test_df <- NULL



# --- NEW: Run Dredge Pre-computation (moved from inside the function) ---
cli::cli_h1("Running Dredge Pre-computation")

# 1. Perform PCA on the *calibration* data
pca_spec_cols <- find_cols(cal_df)
dredge_pc.mod <- caret::preProcess(cal_df[, pca_spec_cols],
                                   method = c("center", "pca"),
                                   pcaComp = 10
)
dredge_cal_pca <- cbind(predict(dredge_pc.mod, cal_df[, pca_spec_cols]), cal_df)

# 2. Create data for dredging
dredge_model_data <- dredge_cal_pca[, c("read_age", paste0("PC", 1:10))]

# 3. Build global models (with model = TRUE)
dredge_global_lm <- lm(read_age ~ ., data = dredge_model_data, na.action = "na.fail", model = TRUE)
dredge_global_gam <- mgcv::gam(
  data = dredge_model_data,
  read_age ~ s(PC1, k = 4) + s(PC2, k = 4) + s(PC3, k = 4) + s(PC4, k = 4) + s(PC5, k = 4) +
    s(PC6, k = 4) + s(PC7, k = 4) + s(PC8, k = 4) + s(PC9, k = 4) + s(PC10, k = 4),
  na.action = "na.fail",
  model = TRUE
)

# 4. Run dredge
cli::cli_alert_info("Dredging LM... (this may take a moment)")
suppressMessages({
  top10_lm_models <- get.models(dredge(dredge_global_lm, trace = FALSE), subset = 1:10)
})
cli::cli_alert_info("Dredging GAM... (this may take a moment)")
suppressMessages({
  top10_gam_models <- get.models(dredge(dredge_global_gam, trace = FALSE), subset = 1:10)
})

# 5. Extract terms/formulas to pass to the function
terms_lm <- lapply(top10_lm_models, function(m) m$terms)
terms_gam <- lapply(top10_gam_models, function(m) m$formula)


# 5. Run the main function
cli::cli_h1("Running Main Analysis")

simple_test_results <- run_all_models_simple(
  cal_df = cal_df,
  test_df = test_df,
  best_params_xgb = best_xgb_all,
  best_params_rf = best_rf_all,
  terms_lm = terms_lm,         # <-- PASS IN DREDGE RESULTS
  terms_gam = terms_gam,       # <-- PASS IN DREDGE RESULTS
  vip_threshold = 1.0
)

# 6. Inspect your results
cli::cli_h1("Analysis Complete!")

cat("\n--- METRICS ---\n")
simple_test_results$all_metrics %>% arrange(RMSE)

cat("\n--- PREDICTIONS (head) ---\n")
print(head(simple_test_results$all_predictions))

cat("\n--- IMPORTANCE (head) ---\n")
print(head(simple_test_results$all_importance))



saveRDS(simple_test_results, "Chapter 2/train_kod_test_shu_results.RDS")
```



Plot results from above

```{r}
# PLOT RESULTS QUICKLY

# =============================================================================
#
# SCRIPT TO PLOT MODEL RESULTS
#
# Assumes an object named 'simple_test_results' exists in your
# environment, which is the list output from 'run_all_models_simple()'.
#
# =============================================================================

library(ggplot2)
library(dplyr)
library(patchwork) # For combining plots

# --- 1. Get the data from your results object ---
importance_df <- simple_test_results$all_importance
predictions_df <- simple_test_results$all_predictions


# =============================================================================
# --- 2. PLOT 1: VARIABLE IMPORTANCE ---
# =============================================================================

cli::cli_h1("Generating Importance Plot")

# --- A. Select and normalize data ---

# Select the best PCA-GAM model from your table
# (PCA-GAM 2 was your #2 best overall)
models_to_plot_imp <- c(
  "PLS-VIP",
  "XGBoost",
  "Random Forest",
  "PCA-GAM 2"
)

# Normalize importance from 0-1 for each method to make them comparable
importance_to_plot <- importance_df %>%
  filter(model_name %in% models_to_plot_imp) %>%
  # Ensure wavenumber is numeric
  mutate(wavenumber = as.numeric(wavenumber)) %>%
  # Group by model to scale importance (0-1) *within* each model
  group_by(model_name) %>%
  mutate(importance_scaled = (importance - min(importance)) / (max(importance) - min(importance))) %>%
  ungroup() %>%
  # Re-order the facets for a logical layout
  mutate(model_name = factor(model_name, levels = models_to_plot_imp))


# --- B. Create the plot ---
p_importance <- ggplot(importance_to_plot, aes(x = wavenumber, y = importance_scaled, color = model_name)) +
  geom_line(linewidth = 1) +
  scale_x_reverse(name = expression(paste("Wavenumber ", cm^-1))) +
  scale_y_continuous(name = "Normalized Importance (0-1)") +
  facet_wrap(~model_name, ncol = 1, scales = "free_y") +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    strip.background = element_blank()
  ) +
  ggtitle("Model Variable Importance")

print(p_importance)


# =============================================================================
# --- 3. PLOT 2: PREDICTIONS (OBSERVED VS. PREDICTED) ---
# =============================================================================

cli::cli_h1("Generating Prediction Plot")

# --- A. Select models to plot ---
# (Based on your list and the best performers)
models_to_plot_pred <- c(
  "PCA-GAM 2",        # Best GAM
  "PCA-LM 6",         # Best LM
  "PLS",
  "PLS-VIP (>1)",
  "XGBoost",
  "Random Forest",
  "Simple-LM (L+SW)",
  "Simple-GAM (L+SW)"
)

# Filter and re-order data
predictions_to_plot <- predictions_df %>%
  filter(model_name %in% models_to_plot_pred) %>%
  mutate(model_name = factor(model_name, levels = models_to_plot_pred))

# --- B. Create the plot ---
p_predictions <- ggplot(predictions_to_plot, aes(x = obs_age, y = pred_age)) +
  # 1:1 line
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  # Model trend line
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 0.75) +
  # Data points
  geom_point(alpha = 0.6) +
  # Facet by model
  facet_wrap(~model_name, ncol = 4) +
  labs(
    x = "Observed Age (days)",
    y = "Predicted Age (days)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    strip.background = element_blank()
  ) +
  ggtitle("Model Predictions: Observed vs. Predicted")

print(p_predictions)





# 1. Get the list of specimens that the simple models were *actually* tested on
simple_specimens <- predictions_df %>%
  filter(model_type == "Simple") %>%
  distinct(specimen)

# 2. Filter ALL model predictions to *only* include those specimens
fair_comparison_preds <- predictions_df %>%
  filter(specimen %in% simple_specimens$specimen)

# 3. Now, re-calculate the metrics
fair_metrics_table <- fair_comparison_preds %>%
  group_by(model_type, model_name) %>%
  summarize(
    RMSE = caret::RMSE(pred = pred_age, obs = obs_age),
    RPD = calculate_rpd(obs_age, pred_age),
    Bias = calculate_bias(obs_age, pred_age)
  ) %>%
  arrange(RMSE)

# Print the new, fair metrics
print(fair_metrics_table)

```

