# --- 1. Load All Libraries ---
```{r}
library(dplyr)
library(tidyr)
library(stringr) # For str_remove_all
library(pls)       # For plsr
library(pheatmap)  # For heatmap
library(vegan)     # For rda
library(ggplot2)
library(ggrepel)
library(broom)     # For model summaries
library(purrr)     # For map_dfr
library(prospectr)

ICPMS <- readRDS(file = "C:/Users/poopm/Thesis Writing/Github/LPW_FT-NIRS/RDS_dataframes/ICPMS_filtered_GOOD_USEME.RDS")

FTNIRS <- readRDS("C:/Users/poopm/Thesis Writing/Github/LPW_FT-NIRS/RDS_dataframes/combined_IBM_LPW_raw.RDS")
FTNIRS <- FTNIRS %>% filter(region != "LPW")# select only IBM specimensof combined dataframe

# apply savitzky golay 1,5,17 filter
speccols <- names(FTNIRS)[grepl("^\\d", names(FTNIRS))]
metacols <- setdiff(names(FTNIRS), speccols)
FTNIRS <- cbind(FTNIRS[,metacols],
             savitzkyGolay(as.matrix(FTNIRS[speccols]), m = 1, p = 5, w = 17)
             )
speccols <- names(FTNIRS)[grepl("^\\d", names(FTNIRS))] # reselect the FT-NIRS columns
# perform PCA on all FT-NIRS columns, store alongside specimen
pca_FTNIRS <- mdatools::pca(FTNIRS[speccols], center = TRUE, scale = TRUE)
# pca dataframe with first 20 PC's
pca_df <- data.frame(specimen = FTNIRS$specimen, 
                     pca_FTNIRS$calres$scores[,1:20])
# clean up column names
colnames(pca_df)[2:21] <- paste0("PC", 1:20)
```

# --- 2. Data Preparation  ---
```{r}

# Define the *original* column names you want to select
elements_of_interest <- c(
  "Mg24_ppm", 
  "Mn55_ppm", 
  "Li7_ppm", 
  "Zn66_ppm", 
  "Sr88_ppm",  # The one Sr we want
  "Ba138_ppm"
)

# Define the *clean* names for the pivot step later
elements_clean <- c("Mg", "Mn", "Li", "Zn", "Sr", "Ba")

# Create the summary, averaging transects
ICPMS_region_avg <- ICPMS %>%
  group_by(specimen, oto_region) %>%
  summarize(
    across(all_of(elements_of_interest), \(x) median(x, na.rm = TRUE))
  ) %>%
  ungroup() # Always good to ungroup after summarize

# *** THE KEY FIX: Rename columns *before* pivoting ***
ICPMS_region_avg <- ICPMS_region_avg %>%
  rename_with(~str_remove_all(.x, "[0-9]+|_ppm"), .cols = all_of(elements_of_interest))

# Pivot wider using the new, clean names
ICPMS_wide_region <- ICPMS_region_avg %>%
  pivot_wider(
    names_from = oto_region,
    values_from = all_of(elements_clean), # Use the clean names
    names_sep = "_"                      # Makes "Mg_core", "Mg_middle", etc.
  )

# Join with PCA data (pca_df)
# We use na.omit() to ensure all models run
pca_df_clean <- inner_join(pca_df, ICPMS_wide_region, by = "specimen") %>%
  na.omit()

# --- 3. Create Final X and Y Variables ---
X_vars <- pca_df_clean %>% 
  dplyr::select(starts_with("PC"))

Y_vars <- pca_df_clean %>% 
  dplyr::select(-specimen, -starts_with("PC")) # Now has clean names

# Create the scaled Y matrix (CRITICAL for PLS, RDA, Heatmap)
Y_vars_scaled <- scale(Y_vars)
```

# investigate regions by element

If my main question is whether FT-NIRS is related to ICP-MS data, this isn't the way to do it....

```{r}
# add in `region` information from FT-NIRS df, match by specimen
ICPMS_region_avg <- ICPMS_region_avg %>% left_join(FTNIRS %>% dplyr::select(specimen, region), by = "specimen")
# make ICPMS long for elements for plotting
ICPMS_long <- ICPMS_region_avg %>% pivot_longer(cols = Mg:Ba, 
                                 names_to = "element", 
                                 values_to = "concentration") %>%
  mutate(element = str_remove_all(element, "[0-9]+|_ppm"))


ggplot(ICPMS_long, aes(x = oto_region, y = concentration)) + 
  geom_boxplot() + 
  facet_grid(element ~ region , scales = "free_y") +
  theme_bw()
```



Analysis 1: Linear Model Summary

Same with this. Franz said you'll find spurious relationships by just looking for LM models that match
```{r}
# Get the new, clean Y names
Y_names <- colnames(Y_vars)
pc_names <- colnames(X_vars)
formula_predictors <- paste(pc_names, collapse = " + ")

# Run the loop
model_list <- lapply(Y_names, function(chem_name) {
  full_formula <- as.formula(paste(chem_name, "~", formula_predictors))
  lm(full_formula, data = pca_df_clean)
})
names(model_list) <- Y_names

# Create the clean summary table
model_summary_table <- map_dfr(model_list, glance, .id = "element_region") %>%
  dplyr::select(element_region, r.squared, adj.r.squared, sigma, p.value) %>%
  rename(RSE = sigma) %>%
  arrange(desc(adj.r.squared))

# Print the clean summary
print(head(model_summary_table))
```

Analysis 2: PLSR (PCs vs. Chemistry)
```{r}
X_vars_pls <- as.matrix(X_vars)

# Run the PLSR model on scaled Y data
pls_model <- plsr(Y_vars_scaled ~ X_vars_pls, ncomp = 10, validation = "CV")

# --- Create Loadings DF ---
y_loadings_df <- as.data.frame(Yloadings(pls_model)[, 1:2]) %>%
  mutate(variable = rownames(.), type = "Element (Y)") %>%
  rename(LV1 = `Comp 1`, LV2 = `Comp 2`)

x_loadings_df <- as.data.frame(loadings(pls_model)[, 1:2]) %>%
  mutate(variable = rownames(.), type = "PC (X)") %>%
  rename(LV1 = `Comp 1`, LV2 = `Comp 2`)

loadings_df <- rbind(x_loadings_df, y_loadings_df)

# --- Create Biplot (with clean labels) ---
ggplot(loadings_df, aes(x = LV1, y = LV2, label = variable, color = type)) +
  geom_point() +
  geom_text_repel() +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(title = "PLSR Loadings Biplot (Scaled Data)") +
  theme_minimal()
```

Analysis 3: Correlation Heatmap
```{r}
# Calculate correlation matrix (use scaled Y for better color range)
cor_matrix <- cor(X_vars, Y_vars_scaled) 

# Create heatmap (with clean labels)
pheatmap(
  cor_matrix,
  main = "Correlation Heatmap (PCs vs. Elements)",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
)
```

Analysis 4: RDA (PCs vs. Chemistry)
```{r}
# Run RDA on scaled Y data
rda_model <- rda(Y_vars_scaled ~ ., data = X_vars)

# Plot RDA (with clean labels)
plot(rda_model, scaling = 2, main = "RDA Triplot (Scaled Data)")
```

5. Analysis 5: PLSR (Full Spectra vs. Chemistry)
```{r}
# Join full FTNIRS data with *clean* ICPMS data
pls_data_full <- inner_join(FTNIRS, ICPMS_wide_region, by = "specimen")

# Select spectral columns (safer than using indices)
X_vars_full <- pls_data_full %>% dplyr::select(all_of(speccols))
X_vars_full <- as.matrix(X_vars_full)
# Select chemical columns (safer than using indices)
Y_vars_full <- pls_data_full %>% dplyr::select(all_of(colnames(Y_vars)))
Y_vars_full_scaled <- scale(Y_vars_full)

# Run the full PLSR model
pls_model_full <- plsr(Y_vars_full_scaled ~ X_vars_full, 
                       ncomp = 10, 
                       validation = "CV")
# --- Create Loadings DF ---
y_loadings_df <- as.data.frame(Yloadings(pls_model_full)[, 1:2]) %>%
  mutate(variable = rownames(.), type = "Element (Y)") %>%
  rename(LV1 = `Comp 1`, LV2 = `Comp 2`)

x_loadings_df <- as.data.frame(loadings(pls_model_full)[, 1:2]) %>%
  mutate(variable = rownames(.), type = "PC (X)") %>%
  rename(LV1 = `Comp 1`, LV2 = `Comp 2`)

loadings_df <- rbind(x_loadings_df, y_loadings_df)

# --- Create Biplot (with clean labels) ---
ggplot(loadings_df, aes(x = LV1, y = LV2, label = variable, color = type)) +
  geom_point() +
  geom_text_repel() +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(title = "PLSR Loadings Biplot (Scaled Data)") +
  theme_minimal()
```

# CCA that works

```{r}
# --- 2a. Load and Prep FT-NIRS Data ---
# Using the path from your uploaded 'ICP_MS_dataexploration_1.1.Rmd'
# ‼️ UPDATE THIS PATH if it's incorrect
ftnirs_path <- "C:/Users/poopm/Thesis Writing/Github/LPW_FT-NIRS/RDS_dataframes/combined_IBM_LPW_raw.RDS"
if (!file.exists(ftnirs_path)) {
  stop(paste("FT-NIRS file not found at:", ftnirs_path))
}
FTNIRS_raw <- readRDS(ftnirs_path)

# Prepare FTNIRS data (based on your script)
FTNIRS <- FTNIRS_raw %>% 
  filter(region != "LPW") # Select only IBM specimens

# Apply Savitzky-Golay filter
speccols <- names(FTNIRS)[grepl("^\\d", names(FTNIRS))]
metacols <- setdiff(names(FTNIRS), speccols)
FTNIRS_processed <- cbind(
  FTNIRS[, metacols],
  savitzkyGolay(as.matrix(FTNIRS[speccols]), m = 1, p = 3, w = 17)
)
# Ensure specimen is a factor for joining
FTNIRS_processed$specimen <- as.factor(FTNIRS_processed$specimen)
speccols <- names(FTNIRS_processed)[grepl("^\\d", names(FTNIRS_processed))]
cat(paste("Loaded and processed FT-NIRS data for", nrow(FTNIRS_processed), "specimens.\n"))


# --- 2b. Load and Prep ICP-MS Data ---
# We load your 'final_analysis_data' and pivot it to a wide format.
# This assumes 'final_analysis_data' is in your environment.
# If not, you must run Part 3 of 'Final_Otolith_Workflow.Rmd' first.
ICPMS <- readRDS(file = "C:/Users/poopm/Thesis Writing/Github/LPW_FT-NIRS/RDS_dataframes/ICPMS_filtered_GOOD_USEME.RDS")

# Define the *original* column names you want to select
elements_of_interest <- c(
  "Mg24_ppm", 
  "Mn55_ppm", 
  "Li7_ppm", 
  "Zn66_ppm", 
  "Sr88_ppm",  # The one Sr we want
  "Ba138_ppm"
)

# Define the *clean* names for the pivot step later
elements_clean <- c("Mg", "Mn", "Li", "Zn", "Sr", "Ba")

# Create the summary, averaging transects
ICPMS_region_avg <- ICPMS %>%
  group_by(specimen, oto_region) %>%
  summarize(
    across(all_of(elements_of_interest), \(x) median(x, na.rm = TRUE))
  ) %>%
  ungroup() # Always good to ungroup after summarize

# *** THE KEY FIX: Rename columns *before* pivoting ***
ICPMS_region_avg <- ICPMS_region_avg %>%
  rename_with(~str_remove_all(.x, "[0-9]+|_ppm"), .cols = all_of(elements_of_interest))

# Pivot wider using the new, clean names
ICPMS_wide <- ICPMS_region_avg %>%
  pivot_wider(
    names_from = oto_region,
    values_from = all_of(elements_clean), # Use the clean names
    names_sep = "_"                      # Makes "Mg_core", "Mg_middle", etc.
  )


FTNIRS_processed$specimen <- as.factor(FTNIRS_processed$specimen)
ICPMS_wide$specimen <- as.factor(ICPMS_wide$specimen)
# --- 2c. Join Datasets and Create X/Y Matrices ---
# Join the two datasets by specimen ID
analysis_data <- inner_join(
  FTNIRS_processed,
  ICPMS_wide,
  by = "specimen"
)

cat(paste("Joined data: Found", nrow(analysis_data), "matching specimens.\n"))

# Create the two matrices for analysis
# X = FT-NIRS Spectra (Predictors)
X_matrix <- analysis_data %>%
  select(all_of(speccols)) %>%
  as.matrix()

# Y = ICP-MS Elemental Data (Responses)
Y_matrix <- analysis_data %>%
  select(ends_with("_core"), ends_with("_middle"), ends_with("_edge")) %>%
  as.matrix()

# Scale the Y matrix (highly recommended for both methods)
Y_matrix_scaled <- scale(Y_matrix)

cat("Created X (FT-NIRS) and Y (ICP-MS) matrices.\n")

# --- 3a. Run PCA on FT-NIRS Data (X) ---
# Center and scale the spectra
X_scaled <- scale(X_matrix)
pca_result <- prcomp(X_scaled)

# Extract the first 10 Principal Components (PCs)
# This is our new, low-dimension X matrix
X_pca <- pca_result$x[, 1:10]

# --- 3b. Run Canonical Correlation Analysis (CCA) ---
# Use the built-in stats::cancor() function
# It requires Y (ICP-MS) and X (FT-NIRS PCs)
# It also requires n > p and n > q
# Our n = ~40, p = 10 (good!), q = ~24 (good!)
cca_result <- cancor(X_pca, Y_matrix_scaled)

# --- 3c. View Results ---
cat("--- Canonical Correlation (PCA + CCA) Results ---\n")
# The 'cor' values are the correlations for each canonical variate.
# The first value is the most important.
print(cca_result$cor)

# We can test the significance of these correlations using the 'CCP' package
# install.packages("CCP")
library(CCP)
p.asym(cca_result$cor, 
       N = nrow(X_pca), 
       p = ncol(X_pca), 
       q = ncol(Y_matrix_scaled), 
       tstat = "Wilks")


# --- 3c. View Results ---
cat("--- Canonical Correlation (PCA + CCA) Results ---\n")
print(cca_result$cor)

# ‼️ ADD THIS: View the weights for the Y-variables (Elements)
# This shows which elements are most important for the first (most significant) pair
cat("\n--- Elemental Loadings for CV1 ---\n")
cv1_y_loadings <- cca_result$ycoef[, 1]
print(sort(abs(cv1_y_loadings), decreasing = TRUE))

```

