# Load packages
Install and load necessary packages
```{r}
packages <- c("tidyverse")
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages) > 0) {
  install.packages(new_packages)
}
invisible(lapply(packages, library, character.only = TRUE))
rm(packages, new_packages)
```


# Data Import
Read and prepare initial dataframe for otolith elemental analysis from saved RDS file

```{r}
# Load ICP-MS data from RDS file
df <- readRDS("RDS_dataframes/ICP-MS_clean.RDS") # I think this is just the raw data without any filtering yet
# Select only otolith specimens (rows 30-184)
df <- df[c(30:184)]
```

### create single dataframe (remove lists)
```{r}
# no specimen 14
# specimen 24 missing a third transect
for(i in 1:155){
  spec <- names(df[i])
  df[[i]]$specimen <- as.factor(str_sub(spec, start= 0,end = 5))
  df[[i]]$transect <- as.factor(as.numeric(str_sub(spec, start= -1)) + 1)
  rm(spec,i)
}
df <- do.call("rbind", df)
df <- rename(df, "time" = `Elapsed Time`)
# saveRDS(df, "RDS_dataframes/ICP-MS_df.RDS")
# rm(df)
```

### filter out the junk first
```{r}
# unusable = missed core
unusable <- paste0("oto", c(05, 10, 12, 14, 30, 37, 38, 40))
# bad = nearly missed our basically outside core
bad <- paste0("oto", c(16, 18, 20, 28, 35))
# good = good enough, close to or thru core
good <- as.character(unique(df$specimen))
good <- good[!good %in% c(unusable, bad)]
```


# Best Approach



### find_otolith_boundaries()
Detect Otolith Boundaries using Ca43 and Cu65 signals
ca43_data: Numeric vector of Ca43 measurements
cu65_data: Numeric vector of Cu65 measurements
window_size:  Size of moving average window (default: 7)
```{r}
find_otolith_boundaries <- function(ca43_data, cu65_data, window_size = 7) {
  # Data preparation
  n_points <- length(ca43_data)
  # Calculate rolling medians with larger window for stability
  ma_ca43 <- stats::runmed(ca43_data, k = window_size)
  ma_cu65 <- stats::runmed(cu65_data, k = window_size)
  # Calculate percentiles using middle portion of data for better baseline
  mid_start <- ceiling(n_points * 0.3)
  mid_end <- floor(n_points * 0.7)
  mid_data_ca43 <- ma_ca43[mid_start:mid_end]
  
  # Get reference levels
  # ‼️ Added na.rm = TRUE, just in case
  ca43_median <- median(mid_data_ca43, na.rm = TRUE) 
  ca43_start_threshold <- ca43_median * 0.4  # More lenient threshold
  
  # Start detection with minimum signal requirements
  min_stable_points <- 5
  start_idx <- NULL
  
  # Look for sustained rise in Ca43
  for(i in window_size:(n_points-min_stable_points)) {
    window <- ma_ca43[i:(i+min_stable_points)]
    # ‼️ Added na.rm = TRUE
    if(all(window > ca43_start_threshold, na.rm = TRUE) &&  
       mean(window, na.rm = TRUE) > mean(ma_ca43[(i-window_size):i], na.rm = TRUE)) {
      start_idx <- i
      break
    }
  }
  
  # Fallback if no clear start found
  if(is.null(start_idx)) {
    # Use first major rise in Ca43
    for(i in 2:n_points) {
      if(ma_ca43[i] > ca43_start_threshold && 
         ma_ca43[i] > mean(ma_ca43[max(1, i-5):i], na.rm = TRUE) * 1.5) {
        start_idx <- i
        break
      }
    }
  }
  
  # If still no start found, use conservative approach
  if(is.null(start_idx)) {
    start_idx <- ceiling(n_points * 0.1)  # Use first 10% as default
  }

  # --- ‼️ REVISED END DETECTION LOGIC ---
  
  # Default to the end of the file
  end_idx <- n_points 
  
  # Only look for end if we have enough points
  if(n_points - start_idx > window_size * 3) {
    
    # Start our search from the middle (where we know signal is good)
    # and search *forwards* to the end.
    search_start_for_end <- mid_start
    
    # Loop from the middle to the end
    for(i in search_start_for_end:(n_points - min_stable_points)) {
      
      # Get a small window of points
      window <- ma_ca43[i:(i + min_stable_points - 1)]
      
      # Check if *all* points in this window have dropped below the threshold
      # (Using the same threshold as the start)
      if(all(window < ca43_start_threshold, na.rm = TRUE)) {
        end_idx <- i  # This is the first point of the drop
        break        # We found it, so stop searching
      }
    }
  }
  
  # --- END REVISED LOGIC ---

  # Validate and adjust boundaries
  if(end_idx - start_idx < 100) {  # Minimum reasonable length
    end_idx <- n_points
    start_idx <- min(start_idx, ceiling(n_points * 0.15))  # Max 15% from start
  }
  
  return(list(
    start = start_idx,
    end = end_idx,
    filtered_data = ca43_data[start_idx:end_idx],
    diagnostics = list(
      ca43_median = ca43_median,
      start_threshold = ca43_start_threshold
    )
  ))
}

```


```{r}
#' Find Otolith Boundaries using Cu65 "Bowl"
#'
#' Detects otolith boundaries using the inverse hyperbolic sine (asinh)
#' transformed Cu65 signal, which typically forms a "bowl" shape.
#'
#' @param ca43_data (Unused, for compatibility with process_otoliths)
#' @param cu65_data Numeric vector of Cu65 measurements
#' @param window_size Size of moving average window (default: 7)
#' @param threshold_offset How far above the baseline to set the threshold
#'                         (on asinh scale). Tunable.
#' @param buffer How many points to include on the slope (default: 5)
#'
find_otolith_boundaries_cu <- function(ca43_data, cu65_data, 
                                     window_size = 7, 
                                     threshold_offset = 3.0, 
                                     buffer = 2) {
  
  # --- 1. Preparation ---
  n_points <- length(cu65_data)
  min_stable_points <- 5
  
  # Apply asinh transform and smoothing (same as in plots)
  # Using asinh() because it handles 0 and negative values
  ma_cu <- asinh(stats::runmed(cu65_data, k = window_size))

  # --- 2. Find Baseline & Threshold ---
  # Find the "bottom of the bowl" baseline
  mid_start <- ceiling(n_points * 0.3)
  mid_end <- floor(n_points * 0.7)
  cu_baseline <- median(ma_cu[mid_start:mid_end], na.rm = TRUE)
  
  # Set the threshold slightly *above* the bottom of the bowl
  cu_threshold <- cu_baseline + threshold_offset 
  
  # --- 3. Start Detection (Find the drop) ---
  start_idx <- NULL
  # Search from the beginning for the first stable *low* period
  for(i in window_size:(mid_end)) {
    window <- ma_cu[i:(i + min_stable_points - 1)]
    
    # if all points in the window are *below* the threshold...
    if(all(window < cu_threshold, na.rm = TRUE)) {
      # ...we found the start of the bowl bottom.
      # Back up by 'buffer' to capture the slope.
      start_idx <- max(1, i - buffer) 
      break
    }
  }
  
  # Fallback if no clear start found
  if(is.null(start_idx)) {
    start_idx <- ceiling(n_points * 0.1) # Conservative default
  }

  # --- 4. End Detection (Find the rise) ---
  end_idx <- n_points # Default to end
  # Search from the middle for the first *rise*
  for(i in mid_start:(n_points - min_stable_points)) {
    
    # Check a single point (more responsive to the *start* of the rise)
    if(is.na(ma_cu[i])) next # Skip NAs
    
    if(ma_cu[i] > cu_threshold) {
      # We found the start of the rise.
      # Go *forward* by 'buffer' to capture the slope.
      end_idx <- min(n_points, i + buffer) 
      break
    }
  }

  # --- 5. Validate and Return ---
  if(end_idx - start_idx < 100) { # Minimum reasonable length
    start_idx <- max(1, start_idx - buffer*2) # Be more generous if it's too short
    end_idx <- min(n_points, end_idx + buffer*2)
  }
  
  # Final safety check
  if (end_idx <= start_idx) {
      start_idx <- ceiling(n_points * 0.1)
      end_idx <- n_points
  }

  return(list(
    start = start_idx,
    end = end_idx,
    # Note: filtered_data is now irrelevant as we're just finding indices
    filtered_data = ca43_data[start_idx:end_idx], 
    diagnostics = list(
      cu_baseline = cu_baseline,
      cu_threshold = cu_threshold
    )
  ))
}
```



### process_otoliths()
Process all otolith transects
```{r}
#'  otolith_list List of dataframes containing Ca43 and Cu65_ppm measurements
#'  Summary dataframe of processing results
process_otoliths <- function(otolith_list) {
  results <- list()
  failed_specimens <- character()
  
  for(name in names(otolith_list)) {
    tryCatch({
      bounds <- find_otolith_boundaries_cu(
        otolith_list[[name]]$Ca43,
        otolith_list[[name]]$Cu65_ppm
      )
      results[[name]] <- bounds
      cat("Processed", name, 
          "- Start:", bounds$start, 
          "End:", bounds$end, "\n")
    }, error = function(e) {
      cat("Error processing", name, ":", e$message, "\n")
      failed_specimens <- c(failed_specimens, name)
    })
  }
  summary_df <- data.frame(
    transect = names(results),
    start_index = sapply(results, `[[`, "start"),
    end_index = sapply(results, `[[`, "end"),
    original_length = sapply(otolith_list[names(results)], nrow),
    filtered_length = sapply(results, function(x) length(x$filtered_data))
  )
  summary_df$points_removed <- summary_df$original_length - summary_df$filtered_length
  attr(summary_df, "failed_specimens") <- failed_specimens
  return(summary_df)
}
```

#### outlier from analysis (this is not current as of right now, 11/04/2025)
```{r}
# # Documentation for anomalous specimen oto10-2
# ANOMALY_NOTE <- list(
#   specimen_id = "oto10-2",
#   issue = "Mid-transect elemental data disruption",
#   details = list(
#     observation = "All elemental data show aberrant values in middle section",
#     impact = "Creates artificial truncation in data processing",
#     location = "Around point 288 in transect",
#     comparison = "Other oto10 series (0,1,3) show normal patterns"
#   ),
#   recommendation = "Should be flagged for separate analysis or potential exclusion from bulk processing"
# )

```

### Filter DF

```{r}
# remove the junk first
df_filtered <- df %>% filter(specimen %in% good)


# ‼️ FIX 1: Split the dataframe back into a list
# Your process_otoliths() function expects a *list* of dataframes,
# not one single big dataframe. We'll split the big df by each
# unique combination of specimen and transect.
df_list <- df_filtered %>%
  group_split(specimen, transect)

# ‼️ FIX 2: Name the list
# The function also expects a *named* list to identify the transects.
# We'll create names by pasting the specimen and transect IDs together
# (e.g., "oto01-1", "oto01-2", etc.)
list_names <- purrr::map_chr(df_list, function(transect_df) {
  # Get the first (and only) specimen/transect value from this df
  spec <- as.character(transect_df$specimen[1])
  tran <- as.character(transect_df$transect[1])
  paste(spec, tran, sep = "-")
})

# Apply the new names to the list
names(df_list) <- list_names
```





```{r}
# ---
# Now, pass the correctly formatted *list* to your function
# ---
results <- process_otoliths(df_list)

# You can now view the summary dataframe
print(results)
```


# generate plots of all transects, excluding the filtered junk
```{r}
library(tidyr)
# 1. Add a row-number index *within each transect* to df_filtered
df_with_index <- df_filtered %>%
  group_by(specimen, transect) %>%
  mutate(index = row_number()) %>%
  ungroup()

# 2. Split the 'results' df to make its columns match df_filtered
results_split <- results %>%
  # Split "oto01-1" into "oto01" and "1"
  tidyr::separate(transect, into = c("specimen_char", "transect_char"), sep = "-", remove = FALSE) %>%
  # Convert new columns to factors to match the types in df_filtered
  mutate(
    specimen = factor(specimen_char),
    transect = factor(transect_char)
  ) %>%
  select(specimen, transect, start_index, end_index)

# 3. Find the 'time' value for the start_index
start_times <- df_with_index %>%
  # Join where specimen, transect, and index match
  inner_join(results_split, by = c("specimen", "transect", "index" = "start_index")) %>%
  select(specimen, transect, start_time = time)

# 4. Find the 'time' value for the end_index
end_times <- df_with_index %>%
  inner_join(results_split, by = c("specimen", "transect", "index" = "end_index")) %>%
  select(specimen, transect, end_time = time)

# 5. Join start and end times into one cutoff dataframe
cutoff_data <- start_times %>%
  inner_join(end_times, by = c("specimen", "transect"))

# 6. Pivot to a "long" format, which ggplot prefers for plotting
cutoff_data_long <- cutoff_data %>%
  tidyr::pivot_longer(
    cols = c(start_time, end_time),
    names_to = "cutoff_type",
    values_to = "time_value"
  )

cat("Cutoff time lookup table created successfully.\n")

# Make sure all necessary packages are loaded
library(ggplot2)
library(dplyr)
library(tidyr) # We need this for pivot_longer

# --- 1. Data Preparation ---
# (Assumes you've already run the code to create 'df_filtered' and 'cutoff_data_long')

# ‼️ NEW: Pivot the main data to a "long" format
# This is so we can facet by element (Ca43 vs Cu65_ppm)
df_long <- df_filtered %>%
  # Select just the columns we need for plotting
  select(specimen, transect, time, Ca43, Cu65_ppm) %>%
  pivot_longer(
    cols = c(Ca43, Cu65_ppm), # Pivot these two columns
    names_to = "element",    #...into a new column named "element"
    values_to = "value"      #...and their values into a new column named "value"
  )

cat("Pivoted data to long format for plotting.\n")


# 2. ‼️ NEW: Apply transformation
#    We will apply the asinh() transform to Cu65_ppm values
#    and update the 'element' label to show this in the plot.
df_long_transformed <- df_long %>%
  mutate(
    # Apply asinh (pseudo-log) transformation ONLY to Cu65_ppm values
    value = ifelse(element == "Cu65_ppm", asinh(value), value),
    
    # Rename the 'element' factor to reflect the transformation in the plot label
    element = factor(element, 
                     levels = c("Ca43", "Cu65_ppm"), 
                     labels = c("Ca43", "Cu65_ppm (asinh)"))
  )

cat("Applied asinh() transformation to Cu65_ppm data.\n")


# --- 2. PDF Generation Loop ---

pdf_output_file <- "Ca43_Cu65_transect_plots_LOG_SCALES_thresh2_buffer2.pdf"
specimen_list <- unique(df_filtered$specimen)

# Set a wider page width
pdf(pdf_output_file, width = 11, height = 8.5) # Landscape

for (spec_id in specimen_list) {
  
  # Filter the *transformed* data
  spec_data_transformed <- df_long_transformed %>% 
    filter(specimen == spec_id)
  
  # Filter the cutoff data (this is the same as before)
  spec_cutoffs <- cutoff_data_long %>%
    filter(specimen == spec_id)
  
  # Create the plot
  p <- ggplot(spec_data_transformed, aes(x = time, y = value)) +
    
    # Draw the line, colored by the (new) element factor
    geom_line(aes(color = element), linewidth = 0.7) +
    
    # Add vertical lines (same as before)
    geom_vline(
      data = spec_cutoffs,
      aes(xintercept = time_value),
      color = "red",
      linetype = "dashed",
      linewidth = 0.7
    ) +
    
    # Use facet_wrap (same as before)
    facet_wrap(transect ~ element, scales = "free", ncol = 2) +
    
    # ‼️ UPDATED: Use the new factor label in the color scale
    scale_color_manual(values = c("Ca43" = "#0060A0", "Cu65_ppm (asinh)" = "#D55E00")) +
    
    labs(
      title = paste("Specimen:", spec_id),
      # ‼️ UPDATED: New subtitle
      subtitle = "Ca43 (raw) and Cu65 (asinh-transformed) Signals with Boundaries",
      x = "Time (Elapsed)",
      y = "Signal Value / asinh(Value)"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(size = 10, face = "bold"),
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5),
      legend.position = "none"
    )
  
  print(p)
  cat("Plotted specimen:", as.character(spec_id), "\n")
}

dev.off()

cat("\nDone! Successfully created PDF:", pdf_output_file, "\n")

```


```{r}
# # Process all specimens
# results <- process_otoliths(df)
# # Create filtered version of original dataset
# filtered_df <- lapply(names(df), function(specimen) {
#   # Get boundaries for this specimen
#   bounds <- results[results$transect == specimen, ]
#   
#   # Skip if specimen is the documented anomaly
#   if(specimen == "oto10-2") {
#     warning(paste("Skipping known anomalous specimen:", specimen))
#     return(NULL)
#   }
#   
#   # Extract relevant portion of data
#   df[[specimen]][bounds$start_index:bounds$end_index, ]
# })
# # Name the list elements
# names(filtered_df) <- names(df)
# # Remove NULL entries (anomalous specimens)
# filtered_df <- filtered_df[!sapply(filtered_df, is.null)]
# # saveRDS(filtered_df, "RDS_dataframes/ICP-MS_cutoffs_df.RDS")
```

### Validation checks
```{r validation}
# # Check filtered dataset
# cat("Original specimens:", length(df), "\n")
# cat("Filtered specimens:", length(filtered_df), "\n")
# cat("Removed specimens:", setdiff(names(df), names(filtered_df)), "\n")
# 
# # Check lengths match results
# all_match <- all(sapply(names(filtered_df), function(specimen) {
#   nrow(filtered_df[[specimen]]) == 
#     results$filtered_length[results$transect == specimen]
# }))
# cat("All lengths match results:", all_match, "\n")
```


# shiny for marking ends; shows both Ca and Cu, filtering unusable specimens, showing only best 2 transects from each

```{r}
# --- 0. Load Packages ---
# Install if needed
# install.packages(c("shiny", "tidyverse", "readr", "ggplot2", "purrr"))
library(shiny)
library(ggplot2)
library(tidyverse)
library(readr)
library(purrr)
library(readxl)

# --- 1. Load Raw ICP-MS Data ---
cat("Loading raw ICP-MS data...\n")
df_raw <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")
df_raw <- df_raw[c(30:184)] # Select otoliths

# Bind all transects into one dataframe
df_list_raw <- list()
for(i in 1:length(df_raw)){
  spec <- names(df_raw[i])
  df_list_raw[[i]] <- df_raw[[i]]
  df_list_raw[[i]]$specimen <- as.factor(str_sub(spec, start= 0,end = 5))
  df_list_raw[[i]]$transect <- as.factor(as.numeric(str_sub(spec, start= -1)) + 1)
}
df <- do.call("rbind", df_list_raw)
df <- rename(df, "time" = `Elapsed Time`)

# --- 2. Load Transect Grades & Filter ---
cat("Loading transect grades...\n")
grades_file <- "transect grades.xlsx"
if (!file.exists(grades_file)) {
  stop(paste("Error:", grades_file, "not found. Please upload it."))
}

grades_wide <- read_xlsx(grades_file)

# Convert grades to long format
grades_long <- grades_wide %>%
  rename(specimen_num = `Specimen Number`) %>%
  pivot_longer(
    cols = starts_with("Transect"),
    names_to = "transect_name",
    values_to = "grade"
  ) %>%
  # Filter out transects that weren't graded (NA)
  filter(!is.na(grade)) %>%
  mutate(
    # Format specimen name to match ICP-MS data (e.g., 1 -> oto01)
    specimen = sprintf("oto%02d", specimen_num),
    # Format transect name to match (e.g., "Transect 1" -> "1")
    transect = str_extract(transect_name, "\\d+"),
  ) %>%
  select(specimen, transect, grade)

# Get the "Top 2" transects for each specimen
top_2_transects <- grades_long %>%
  group_by(specimen) %>%
  # Keep only the 2 transects with the lowest grade
  slice_min(order_by = grade, n = 2, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    # Convert to factor to match 'df' for joining
    specimen = as.factor(specimen),
    transect = as.factor(transect)
  )

cat("Successfully identified Top 2 transects per specimen.\n")

# --- 3. Filter Main Data ---
# Define "unusable" (red) list.
unusable <- paste0("oto", c(05, 10, 12, 14, 30, 37, 38, 40))

# 1. Remove unusable specimens
df_semi_filtered <- df %>%
  filter(!specimen %in% unusable)

# 2. Keep ONLY the top 2 transects
# This correctly INCLUDES the "bad" (yellow) list specimens
df_to_process <- df_semi_filtered %>%
  inner_join(top_2_transects, by = c("specimen", "transect"))

cat("Main dataframe filtered. Ready for pre-processing.\n")

# --- 4. Automated Pre-Filter (Functions) ---

# Copy of the Cu-bowl finder function
find_otolith_boundaries_cu <- function(ca43_data, cu65_data, 
                                       window_size = 7, 
                                       threshold_offset = 1.0, # Using 1.0 as you found it worked better
                                       buffer = 3) { # Using a smaller buffer
  
  n_points <- length(cu65_data)
  min_stable_points <- 5
  ma_cu <- asinh(stats::runmed(cu65_data, k = window_size))
  
  mid_start <- ceiling(n_points * 0.3)
  mid_end <- floor(n_points * 0.7)
  cu_baseline <- median(ma_cu[mid_start:mid_end], na.rm = TRUE)
  cu_threshold <- cu_baseline + threshold_offset 
  
  start_idx <- NULL
  for(i in window_size:(mid_end)) {
    window <- ma_cu[i:(i + min_stable_points - 1)]
    if(all(window < cu_threshold, na.rm = TRUE)) {
      start_idx <- max(1, i - buffer) 
      break
    }
  }
  if(is.null(start_idx)) { start_idx <- ceiling(n_points * 0.1) }
  
  end_idx <- n_points 
  for(i in mid_start:(n_points - min_stable_points)) {
    if(is.na(ma_cu[i])) next 
    if(ma_cu[i] > cu_threshold) {
      end_idx <- min(n_points, i + buffer) 
      break
    }
  }
  
  if (end_idx <= start_idx) {
    start_idx <- ceiling(n_points * 0.1)
    end_idx <- n_points
  }
  
  return(list(
    start = start_idx,
    end = end_idx,
    filtered_data = ca43_data[start_idx:end_idx], 
    diagnostics = list(cu_baseline = cu_baseline, cu_threshold = cu_threshold)
  ))
}

# Copy of the processing function
process_otoliths <- function(otolith_list) {
  results <- list()
  for(name in names(otolith_list)) {
    tryCatch({
      bounds <- find_otolith_boundaries_cu(
        otolith_list[[name]]$Ca43,
        otolith_list[[name]]$Cu65_ppm
      )
      results[[name]] <- bounds
    }, error = function(e) {
      cat("Error processing", name, ":", e$message, "\n")
    })
  }
  
  summary_df <- data.frame(
    transect = names(results),
    start_index = sapply(results, `[[`, "start"),
    end_index = sapply(results, `[[`, "end")
  )
  return(summary_df)
}

# --- 5. Run Pre-Filter & Prepare for App ---
cat("Running automated pre-filter...\n")

# Split the dataframe into a list
df_list_to_process <- df_to_process %>%
  group_split(specimen, transect)

# Create names for the list
list_names <- purrr::map_chr(df_list_to_process, function(transect_df) {
  paste(transect_df$specimen[1], transect_df$transect[1], sep = "-")
})
names(df_list_to_process) <- list_names

# Run the automated function
initial_results_indexed <- process_otoliths(df_list_to_process)

# Add a row-number index *within each transect* to the data
df_app_data <- df_to_process %>%
  group_by(specimen, transect) %>%
  mutate(index = row_number()) %>%
  ungroup()

# Convert start/end *indices* to *time values*
initial_cutoffs_df <- initial_results_indexed %>%
  tidyr::separate(transect, into = c("specimen", "transect"), sep = "-", remove = FALSE) %>%
  mutate(specimen = as.factor(specimen), transect = as.factor(transect)) %>%
  # Join to find start_time
  left_join(
    df_app_data %>% select(specimen, transect, index, time),
    by = c("specimen", "transect", "start_index" = "index")
  ) %>%
  rename(start_time = time) %>%
  # Join to find end_time
  left_join(
    df_app_data %>% select(specimen, transect, index, time),
    by = c("specimen", "transect", "end_index" = "index")
  ) %>%
  rename(end_time = time) %>%
  select(specimen, transect, start_time, end_time)

# Handle any NAs (e.g., if index was out of bounds) by setting to min/max
initial_cutoffs_df <- initial_cutoffs_df %>%
  group_by(specimen, transect) %>%
  mutate(
    min_time = min(df_app_data$time[df_app_data$specimen == .data$specimen & df_app_data$transect == .data$transect]),
    max_time = max(df_app_data$time[df_app_data$specimen == .data$specimen & df_app_data$transect == .data$transect])
  ) %>%
  ungroup() %>%
  mutate(
    start_time = ifelse(is.na(start_time), min_time, start_time),
    end_time = ifelse(is.na(end_time), max_time, end_time)
  ) %>%
  select(specimen, transect, start_time, end_time)

cat("Pre-filter complete. Launching Shiny app...\n")

# --- 6. Manual Review Shiny App ---

# Make data globally available to the app
app_data <- df_app_data
initial_cutoffs <- initial_cutoffs_df

# --- App UI ---
ui <- fluidPage(
  titlePanel("Otolith Cutoff Review (Top 2 Transects)"),
  sidebarLayout(
    sidebarPanel(
      selectInput("specimen", "Select Specimen:",
                  choices = sort(unique(app_data$specimen))),
      uiOutput("transect_selector"),
      
      h4("Selected Cutoffs:"),
      textOutput("start_time_display"),
      textOutput("end_time_display"),
      
      br(),
      actionButton("set_start", "Set Start from Click"),
      actionButton("set_end", "Set End from Click"),
      br(),
      br(),
      actionButton("save_cutoffs", "Save All Changes"),
      downloadButton("download_cutoffs", "Download Cutoffs as CSV"),
      
      hr(),
      h4("Click Info:"),
      verbatimTextOutput("click_info")
    ),
    mainPanel(
      plotOutput("cutoff_plot",
                 height = "600px",
                 click = "plot_click",
                 brush = brushOpts(id = "plot_brush", direction = "x")
      )
    )
  )
)

# --- App Server ---
server <- function(input, output, session) {
  
  # Reactive values to store all cutoffs
  all_cutoffs <- reactiveVal(initial_cutoffs)
  
  # Reactive value for the last click
  last_click <- reactiveVal(NULL)
  
  # Dynamic UI for transect selection
  output$transect_selector <- renderUI({
    req(input$specimen)
    transects <- app_data %>%
      filter(specimen == input$specimen) %>%
      pull(transect) %>%
      unique()
    
    selectInput("transect", "Select Transect:",
                choices = sort(transects))
  })
  
  # Get data for the selected transect
  current_transect_data <- reactive({
    req(input$specimen, input$transect)
    app_data %>%
      filter(specimen == input$specimen, transect == input$transect)
  })
  
  # Get the current start/end times for the selected transect
  current_times <- reactive({
    req(input$specimen, input$transect)
    all_cutoffs() %>%
      filter(specimen == input$specimen, transect == input$transect)
  })
  
  # Store click info
  observeEvent(input$plot_click, {
    last_click(input$plot_click$x)
  })
  
  # --- Update Buttons ---
  observeEvent(input$set_start, {
    req(last_click())
    new_start <- last_click()
    current_end <- current_times()$end_time
    
    if (new_start >= current_end) {
      showNotification("Start time must be before end time.", type = "error")
      return()
    }
    
    # Update the reactiveVal
    df <- all_cutoffs()
    df$start_time[df$specimen == input$specimen & df$transect == input$transect] <- new_start
    all_cutoffs(df)
  })
  
  observeEvent(input$set_end, {
    req(last_click())
    new_end <- last_click()
    current_start <- current_times()$start_time
    
    if (new_end <= current_start) {
      showNotification("End time must be after start time.", type = "error")
      return()
    }
    
    # Update the reactiveVal
    df <- all_cutoffs()
    df$end_time[df$specimen == input$specimen & df$transect == input$transect] <- new_end
    all_cutoffs(df)
  })
  
  # --- Text Displays ---
  output$start_time_display <- renderText({
    paste("Start Time:", round(current_times()$start_time, 2))
  })
  
  output$end_time_display <- renderText({
    paste("End Time:", round(current_times()$end_time, 2))
  })
  
  output$click_info <- renderPrint({
    req(input$plot_click)
    cat(paste("Time:", round(input$plot_click$x, 2)))
  })
  
  # --- Main Plot ---
  output$cutoff_plot <- renderPlot({
    req(current_transect_data(), current_times())
    
    data <- current_transect_data()
    times <- current_times()
    
    ggplot(data, aes(x = time)) +
      geom_line(aes(y = Ca43), color = "blue", alpha = 0.7) +
      geom_line(aes(y = asinh(Cu65_ppm) * 1e5), color = "darkorange", alpha = 0.7) + # Scale for visibility
      
      geom_vline(xintercept = times$start_time, color = "green", linetype = "dashed", linewidth = 1) +
      geom_vline(xintercept = times$end_time, color = "red", linetype = "dashed", linewidth = 1) +
      
      scale_y_continuous(
        name = "Ca43 (Blue)",
        sec.axis = sec_axis(~. / 1e5, name = "asinh(Cu65) (Orange)")
      ) +
      labs(title = paste("Specimen:", input$specimen, "- Transect:", input$transect),
           x = "Time (Elapsed)", y = "Signal") +
      theme_bw() +
      theme(
        plot.title = element_text(size = 16, face = "bold"),
        axis.title.y.left = element_text(color = "blue"),
        axis.title.y.right = element_text(color = "darkorange")
      )
  })
  
  # --- Save/Download ---
  observeEvent(input$save_cutoffs, {
    # This is more for a real app; for this one, download is the save
    showNotification("Changes saved in app session. Click Download to get the CSV.", type = "message")
  })
  
  output$download_cutoffs <- downloadHandler(
    filename = function() {
      "manual_cutoffs.csv"
    },
    content = function(file) {
      write.csv(all_cutoffs(), file, row.names = FALSE)
    }
  )
}

# --- Run the App ---
shinyApp(ui = ui, server = server)
```

# filter dataframe after above shiny

```{r}
# --- 1. Load Packages ---
library(tidyverse)
library(readr) # ‼️ Added
library(purrr) # ‼️ Added
library(reshape2)

# --- 2. Load and Filter Data (Mirrors App Logic) ---

# ‼️ --- THIS ENTIRE SECTION IS UPDATED --- ‼️

cat("Loading raw ICP-MS data...\n")
df_raw <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")
df_raw <- df_raw[c(30:184)] # Select otoliths

# Bind all transects into one dataframe
df_list_raw <- list()
for(i in 1:length(df_raw)){
  spec <- names(df_raw[i])
  df_list_raw[[i]] <- df_raw[[i]]
  df_list_raw[[i]]$specimen <- as.factor(str_sub(spec, start= 0,end = 5))
  df_list_raw[[i]]$transect <- as.factor(as.numeric(str_sub(spec, start= -1)) + 1)
}
df <- do.call("rbind", df_list_raw)
df <- rename(df, "time" = `Elapsed Time`)

# --- 2. Load Transect Grades & Filter ---
cat("Loading transect grades...\n")
grades_file <- "transect grades.xlsx"
if (!file.exists(grades_file)) {
  stop(paste("Error:", grades_file, "not found. Please upload it."))
}

grades_wide <- read_xlsx(grades_file)


# Convert grades to long format
grades_long <- grades_wide %>%
  rename(specimen_num = `Specimen Number`) %>%
  pivot_longer(
    cols = starts_with("Transect"),
    names_to = "transect_name",
    values_to = "grade"
  ) %>%
  filter(!is.na(grade)) %>%
  mutate(
    specimen = sprintf("oto%02d", specimen_num),
    transect = str_extract(transect_name, "\\d+"),
  ) %>%
  select(specimen, transect, grade)

# Get the "Top 2" transects for each specimen
top_2_transects <- grades_long %>%
  group_by(specimen) %>%
  slice_min(order_by = grade, n = 2, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    specimen = as.factor(specimen),
    transect = as.factor(transect)
  )

cat("Successfully identified Top 2 transects per specimen.\n")

# Define "unusable" (red) list.
unusable <- paste0("oto", c(05, 10, 12, 14, 30, 37, 38, 40))

# 1. Remove unusable specimens
df_semi_filtered <- df %>%
  filter(!specimen %in% unusable)

# 2. Keep ONLY the top 2 transects
# This is the base dataframe we will apply cutoffs to
df_to_filter <- df_semi_filtered %>%
  inner_join(top_2_transects, by = c("specimen", "transect"))

cat("Main dataframe filtered. Ready for manual cutoffs.\n")

# ‼️ --- END OF UPDATED SECTION --- ‼️


# --- 3. Load Your Manual Cutoffs ---
# Make sure the CSV file is in your working directory
cutoff_file <- "manual_cutoffs_new.csv"

if (!file.exists(cutoff_file)) {
  stop("Error: 'manual_cutoffs.csv' not found. 
       Please make sure you saved the file from the Shiny app.")
}

# Read the cutoffs file.
# We must ensure the column types match 'df_filtered' for the join.
manual_cutoffs <- read_csv(cutoff_file, 
                           col_types = cols(
                             specimen = col_factor(),
                             transect = col_factor(),
                             start_time = col_double(),
                             end_time = col_double()
                           ))

cat("Successfully loaded 'manual_cutoffs.csv'.\n")

# --- 4. Join and Filter Data ---

# Join the cutoffs to the main dataframe
# ‼️ UPDATED: use df_to_filter
df_joined <- df_to_filter %>%
  inner_join(manual_cutoffs, by = c("specimen", "transect"))

# Filter the data based on the manual start and end times
final_filtered_data <- df_joined %>%
  filter(time >= start_time & time <= end_time) %>%
  # Remove the extra time columns
  select(-start_time, -end_time)

cat("Successfully filtered data based on manual cutoffs.\n")

# --- 5. Save and Inspect ---
# You can now use 'final_filtered_data' for your analysis
cat("Final dataframe 'final_filtered_data' is ready.\n")
head(final_filtered_data)

# Optional: Save this final, clean dataframe to a new file
saveRDS(final_filtered_data, "RDS_dataframes/ICP-MS_manual_filtered_11042025.RDS")
write.csv(final_filtered_data, "ICPMS_final_filtered_data_11042025.csv", row.names = FALSE)

```


# identify core, mark middle as point after, until ~10% of end which = edge



```{r}
# --- 0. Load Packages ---
library(shiny)
library(ggplot2)
library(tidyverse)
library(readr)
library(purrr)
library(readxl) # ‼️ Added for read_xlsx

# --- 1. Load and Prepare Data ---
# This section runs ONCE when the app starts.
# It loads all necessary data and prepares the 'final_filtered_data'.

cat("Loading and preparing data for app...\n")

# --- Load Raw Data ---
df_raw <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")
df_raw <- df_raw[c(30:184)] # Select otoliths
df_list_raw <- list()
for(i in 1:length(df_raw)){
  spec <- names(df_raw[i])
  df_list_raw[[i]] <- df_raw[[i]]
  df_list_raw[[i]]$specimen <- as.factor(str_sub(spec, start= 0,end = 5))
  df_list_raw[[i]]$transect <- as.factor(as.numeric(str_sub(spec, start= -1)) + 1)
}
df <- do.call("rbind", df_list_raw)
df <- rename(df, "time" = `Elapsed Time`)

# --- Load Grades and Filter to Top 2 ---
cat("Loading transect grades...\n")
# ‼️ UPDATED based on your correction
grades_file <- "transect grades.xlsx"
if (!file.exists(grades_file)) {
  stop(paste("Error:", grades_file, "not found. Please upload it."))
}

grades_wide <- read_xlsx(grades_file)
# ‼️ END UPDATED BLOCK

grades_long <- grades_wide %>%
  rename(specimen_num = `Specimen Number`) %>%
  pivot_longer(
    cols = starts_with("Transect"),
    names_to = "transect_name",
    values_to = "grade"
  ) %>%
  filter(!is.na(grade)) %>%
  mutate(
    specimen = sprintf("oto%02d", specimen_num),
    transect = str_extract(transect_name, "\\d+"),
  ) %>%
  select(specimen, transect, grade)

top_2_transects <- grades_long %>%
  group_by(specimen) %>%
  slice_min(order_by = grade, n = 2, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    specimen = as.factor(specimen),
    transect = as.factor(transect)
  )

# --- Filter Main DF ---
unusable <- paste0("oto", c(05, 10, 12, 14, 30, 37, 38, 40))
df_semi_filtered <- df %>%
  filter(!specimen %in% unusable)

df_to_filter <- df_semi_filtered %>%
  inner_join(top_2_transects, by = c("specimen", "transect"))

# --- Load Manual TRANSECT Cutoffs ---
cutoff_file <- "manual_cutoffs.csv"
if (!file.exists(cutoff_file)) {
  stop("Error: 'manual_cutoffs.csv' not found. 
       Please make sure you saved this file from the previous app.")
}
manual_cutoffs <- read_csv(cutoff_file, 
                           col_types = cols(
                             specimen = col_factor(),
                             transect = col_factor(),
                             start_time = col_double(),
                             end_time = col_double()
                           ))

# --- Create the Final, Filtered Data ---
df_joined <- df_to_filter %>%
  inner_join(manual_cutoffs, by = c("specimen", "transect"))

# This is the dataframe the app will use
final_app_data <- df_joined %>%
  filter(time >= start_time & time <= end_time) %>%
  select(-start_time, -end_time) %>%
  # Add a row index *within each transect*
  group_by(specimen, transect) %>%
  mutate(index = row_number()) %>%
  ungroup()

# --- Prepare App Data ---
# Get list of elements for the dropdown
# ‼️ UPDATED: Restrict choices to only these three elements
element_choices <- c("Ba138_ppm", "Mg24_ppm", "Mn55_ppm")

# Create a dataframe to store the core boundaries
initial_core_df <- final_app_data %>%
  group_by(specimen, transect) %>%
  summarise() %>% # Get unique specimen-transect pairs
  ungroup() %>%
  mutate(
    core_start_index = 1,     # Default start
    core_end_index = 20     # Default end
  )

cat("Data processing complete. Launching app...\n")

# --- 2. Shiny App UI ---
ui <- fluidPage(
  titlePanel("Otolith Core Definition"),
  sidebarLayout(
    sidebarPanel(
      selectInput("specimen", "Select Specimen:",
                  choices = sort(unique(final_app_data$specimen))),
      
      # ‼️ NEW: Transect selector
      uiOutput("transect_selector_ui"),
      
      selectInput("element", "Select Element to Plot:",
                  choices = element_choices,
                  selected = "Ba138_ppm"),
      
      hr(),
      h4("Edit Core Boundaries:"),
      actionButton("set_start", "Set Core Start from Click"),
      actionButton("set_end", "Set Core End from Click"),
      
      hr(),
      h4("Click Info:"),
      verbatimTextOutput("click_info"),
      
      h4("Current Core Boundaries:"),
      verbatimTextOutput("core_info_display"),
      
      hr(),
      downloadButton("download_core_defs", "Download Core Definitions")
    ),
    mainPanel(
      # ‼️ UPDATED: Now a single plot, not faceted
      plotOutput("core_plot",
                 height = "600px", # Can be shorter now
                 click = "plot_click"
      )
    )
  )
)

# --- 3. Shiny App Server ---
server <- function(input, output, session) {
  
  # Reactive value to store all core boundaries
  all_core_data <- reactiveVal(initial_core_df)
  
  # Reactive value to store the last click info
  last_click_info <- reactiveVal(NULL)
  
  # ‼️ NEW: Dynamic UI for transect selection
  output$transect_selector_ui <- renderUI({
    req(input$specimen)
    transect_choices <- final_app_data %>%
      filter(specimen == input$specimen) %>%
      pull(transect) %>%
      unique() %>%
      sort()
    
    selectInput("transect", "Select Transect:",
                choices = transect_choices)
  })
  
  # ‼️ UPDATED: Get data for the selected TRANSECT
  current_transect_data <- reactive({
    req(input$specimen, input$transect)
    final_app_data %>%
      filter(specimen == input$specimen, transect == input$transect)
  })
  
  # ‼️ UPDATED: Get the core boundaries for the selected TRANSECT
  current_core_bounds <- reactive({
    req(input$specimen, input$transect)
    all_core_data() %>%
      filter(specimen == input$specimen, transect == input$transect)
  })
  
  # Store click info
  observeEvent(input$plot_click, {
    # ‼️ UPDATED: Simpler click logic
    last_click_info(list(
      index = round(input$plot_click$x)
    ))
  })
  
  # --- Update Buttons ---
  observeEvent(input$set_start, {
    # ‼️ UPDATED: Simpler logic
    req(last_click_info(), input$specimen, input$transect)
    
    clicked_index <- last_click_info()$index
    
    # Get current end time for validation
    current_end <- current_core_bounds()$core_end_index
    
    if (clicked_index >= current_end) {
      showNotification("Core start must be before core end.", type = "error")
      return()
    }
    
    # Update the reactiveVal
    df <- all_core_data()
    df$core_start_index[df$specimen == input$specimen & df$transect == input$transect] <- clicked_index
    all_core_data(df)
  })
  
  observeEvent(input$set_end, {
    # ‼️ UPDATED: Simpler logic
    req(last_click_info(), input$specimen, input$transect)
    
    clicked_index <- last_click_info()$index
    
    # Get current start time for validation
    current_start <- current_core_bounds()$core_start_index
    
    if (clicked_index <= current_start) {
      showNotification("Core end must be after core start.", type = "error")
      return()
    }
    
    # Update the reactiveVal
    df <- all_core_data()
    df$core_end_index[df$specimen == input$specimen & df$transect == input$transect] <- clicked_index
    all_core_data(df)
  })
  
  # --- Text Displays ---
  output$click_info <- renderPrint({
    info <- last_click_info()
    if (is.null(info)) {
      cat("Click on a plot to see info.")
    } else {
      cat(paste("Clicked Index:", info$index))
    }
  })
  
  output$core_info_display <- renderPrint({
    bounds <- current_core_bounds()
    print(bounds[, c("transect", "core_start_index", "core_end_index")], row.names = FALSE)
  })
  
  # --- Main Plot ---
  output$core_plot <- renderPlot({
    # ‼️ UPDATED: Simpler plot definition
    req(current_transect_data(), input$element, current_core_bounds())
    
    plot_data <- current_transect_data()
    plot_element <- input$element
    core_lines <- current_core_bounds()
    
    ggplot(plot_data, aes(x = index, y = .data[[plot_element]])) +
      geom_point(color = "grey70", alpha = 0.6) +
      geom_smooth(color = "blue", linewidth = 1.2, se = FALSE, method = "loess", span = 0.2) +
      
      # Add vertical lines for core
      geom_vline(aes(xintercept = core_lines$core_start_index), 
                 color = "green", linetype = "dashed", linewidth = 1) +
      geom_vline(aes(xintercept = core_lines$core_end_index), 
                 color = "red", linetype = "dashed", linewidth = 1) +
      
      # ‼️ REMOVED: facet_wrap()
      
      labs(title = paste("Specimen:", input$specimen, "- Transect:", input$transect),
           subtitle = "Grey: Raw Data, Blue: Smoothed Trend, Green/Red: Core Boundaries",
           x = "Row Index (from filtered start point)",
           y = plot_element) +
      theme_bw() +
      theme(
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)
      )
  })
  
  # --- Download Handler ---
  output$download_core_defs <- downloadHandler(
    filename = function() {
      "manual_core_definitions.csv"
    },
    content = function(file) {
      write.csv(all_core_data(), file, row.names = FALSE)
    }
  )
}

# --- 4. Run the App ---

options = list(launch.browser = T)
shinyApp(ui = ui, server = server, options = list(launch.browser = TRUE))



```












#### ANOTHER FUCKING SHINY APP, THIS TIME WITH IMAGES

```{r}
# --- 0. Set Upload Limit & Load Packages ---
# ‼️ THIS IS THE FIX: Increase max file upload size to 300MB
options(shiny.maxRequestSize = 300 * 1024^2)

# install.packages(c("shiny", "tidyverse", "readxl", "magick", "viridis", "purrr"))
library(shiny)
library(tidyverse)
library(readxl)      # For read_xlsx
library(magick)      # For image processing
library(viridis)     # For viridis color scale
library(purrr)

# --- 1. Load & Prepare Data (Global) ---
# This code runs once when the app starts.

# --- 1a. Load Transect Grades ---
cat("Loading transect grades...\n")
grades_file <- "transect grades.xlsx" # Corrected file name
if (!file.exists(grades_file)) {
  stop(paste("Error:", grades_file, "not found. Please upload it."))
}

grades_wide <- read_xlsx(grades_file) # Corrected function

# Convert grades to long format
grades_long <- grades_wide %>%
  rename(specimen_num = `Specimen Number`) %>%
  pivot_longer(
    cols = starts_with("Transect"),
    names_to = "transect_name",
    values_to = "grade"
  ) %>%
  filter(!is.na(grade)) %>%
  mutate(
    specimen = sprintf("oto%02d", specimen_num),
    transect = str_extract(transect_name, "\\d+"),
  ) %>%
  select(specimen, transect, grade)

# Get the "Top 2" transects for each specimen
top_2_transects <- grades_long %>%
  group_by(specimen) %>%
  slice_min(order_by = grade, n = 2, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    specimen = as.factor(specimen),
    transect = as.factor(transect)
  )

# --- 1b. Load Raw ICP-MS Data ---
cat("Loading raw ICP-MS data...\n")
df_raw <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")
df_raw <- df_raw[c(30:184)] # Select otoliths

# Bind all transects into one dataframe
df_list_raw <- list()
for(i in 1:length(df_raw)){
  spec <- names(df_raw[i])
  df_list_raw[[i]] <- df_raw[i][[1]]
  df_list_raw[[i]]$specimen <- as.factor(str_sub(spec, start= 0,end = 5))
  df_list_raw[[i]]$transect <- as.factor(as.numeric(str_sub(spec, start= -1)) + 1)
}
df <- do.call("rbind", df_list_raw)
df <- rename(df, "time" = `Elapsed Time`)

# --- 1c. Filter Main Data ---
unusable <- paste0("oto", c(05, 10, 12, 14, 30, 37, 38, 40))
df_semi_filtered <- df %>%
  filter(!specimen %in% unusable)

# Keep ONLY the top 2 transects
df_to_process <- df_semi_filtered %>%
  inner_join(top_2_transects, by = c("specimen", "transect"))

# --- 1d. Load Manual Cutoffs ---
cat("Loading manual cutoffs...\n")
cutoff_file <- "manual_cutoffs.csv"
if (!file.exists(cutoff_file)) {
  stop(paste("Error:", cutoff_file, "not found. Run previous app first."))
}
manual_cutoffs <- read_csv(cutoff_file, 
                           col_types = cols(
                             specimen = col_factor(),
                             transect = col_factor(),
                             start_time = col_double(),
                             end_time = col_double()
                           ))

# --- 1e. Create Final Filtered Dataframe ---
df_joined <- df_to_process %>%
  inner_join(manual_cutoffs, by = c("specimen", "transect"))

final_filtered_data <- df_joined %>%
  filter(time >= start_time & time <= end_time) %>%
  select(-start_time, -end_time) %>%
  group_by(specimen, transect) %>%
  mutate(index = row_number()) %>% # Add the row index
  ungroup()

cat("Data processing complete. Final 'final_filtered_data' ready.\n")

# --- 1f. Prepare for App ---
# This is the data the app will use
app_data <- final_filtered_data

# Define the elements for the dropdown
element_choices <- c("Ba138_ppm", "Mg24_ppm", "Mn55_ppm")

# Create a dataframe to store core cutoffs
# Initialize with NA values
core_cutoff_data <- app_data %>%
  distinct(specimen, transect) %>%
  mutate(
    core_start_index = NA_integer_,
    core_end_index = NA_integer_
  )

# --- 2. Shiny App UI ---
ui <- fluidPage(
  titlePanel("Otolith Core Validator: 1D Plot + 2D Image"),
  
  # --- 2a. Main Row ---
  fluidRow(
    # --- 2b. Left Column (1D Plot & Controls) ---
    column(6,
      wellPanel(
        fluidRow(
          column(4, selectInput("specimen", "Select Specimen:", 
                                choices = sort(unique(app_data$specimen)))),
          column(4, uiOutput("transect_selector")),
          column(4, selectInput("element", "Select Element:",
                                choices = element_choices, selected = "Ba138_ppm"))
        )
      ),
      plotOutput("element_plot", 
                 height = "400px", 
                 click = "plot_click"),
      
      h4("Core Selection:"),
      textOutput("core_start_display"),
      textOutput("core_end_display"),
      actionButton("set_core_start", "Set Core Start from Click"),
      actionButton("set_core_end", "Set Core End from Click"),
      downloadButton("download_cores", "Download Core Cutoffs (CSV)"),
      
      h4("Click Info (1D Plot):"),
      verbatimTextOutput("click_info_1d")
    ),
    
    # --- 2c. Right Column (2D Image) ---
    column(6,
      wellPanel(
        fileInput("upload_image", "Upload Otolith Image (.tif, .jpg, .png)",
                  accept = c("image/tiff", "image/jpeg", "image/png", ".tif", ".jpg", ".tiff"))
      ),
      plotOutput("image_plot", 
                 height = "600px", 
                 click = "image_click"),
      
      h4("Click Info (Image):"),
      verbatimTextOutput("click_info_2d")
    )
  )
)

# --- 3. Shiny App Server ---
server <- function(input, output, session) {
  
  # --- 3a. Reactive Values ---
  rv <- reactiveValues(
    core_cutoffs = core_cutoff_data, # Holds all core start/end indices
    last_plot_click = NULL,          # Last click on 1D plot
    image_clicks = list(),           # Stores the two clicks for the transect line
    transect_image = NULL            # Stores the uploaded image
  )
  
  # --- 3b. Dynamic Transect UI ---
  output$transect_selector <- renderUI({
    req(input$specimen)
    transects <- app_data %>%
      filter(specimen == input$specimen) %>%
      pull(transect) %>%
      unique()
    
    selectInput("transect", "Select Transect:",
                choices = sort(transects))
  })
  
  # --- 3c. Current Selections ---
  current_transect_data <- reactive({
    req(input$specimen, input$transect)
    app_data %>%
      filter(specimen == input$specimen, transect == input$transect)
  })
  
  current_core_times <- reactive({
    req(input$specimen, input$transect)
    rv$core_cutoffs %>%
      filter(specimen == input$specimen, transect == input$transect)
  })
  
  current_element_data <- reactive({
    req(current_transect_data(), input$element)
    current_transect_data() %>%
      select(index, value = all_of(input$element))
  })
  
  # --- 3d. 1D Plot Logic (Left Column) ---
  
  # Store 1D plot click
  observeEvent(input$plot_click, {
    rv$last_plot_click <- input$plot_click$x
  })
  
  # Set core start/end
  observeEvent(input$set_core_start, {
    req(rv$last_plot_click)
    new_start <- round(rv$last_plot_click)
    current_end <- current_core_times()$core_end_index
    
    if (!is.na(current_end) && new_start >= current_end) {
      showNotification("Core start must be before core end.", type = "error")
      return()
    }
    
    df <- rv$core_cutoffs
    df$core_start_index[df$specimen == input$specimen & df$transect == input$transect] <- new_start
    rv$core_cutoffs <- df
  })
  
  observeEvent(input$set_core_end, {
    req(rv$last_plot_click)
    new_end <- round(rv$last_plot_click)
    current_start <- current_core_times()$core_start_index
    
    if (!is.na(current_start) && new_end <= current_start) {
      showNotification("Core end must be after core start.", type = "error")
      return()
    }
    
    df <- rv$core_cutoffs
    df$core_end_index[df$specimen == input$specimen & df$transect == input$transect] <- new_end
    rv$core_cutoffs <- df
  })
  
  # 1D Plot Output
  output$element_plot <- renderPlot({
    req(current_element_data())
    
    data <- current_element_data()
    times <- current_core_times()
    
    ggplot(data, aes(x = index, y = value)) +
      geom_point(color = "grey80", alpha = 0.5) +
      geom_smooth(color = "blue", fill = "blue", alpha = 0.2, method = "loess", span = 0.2) +
      
      geom_vline(xintercept = times$core_start_index, color = "green", linetype = "dashed", linewidth = 1) +
      geom_vline(xintercept = times$core_end_index, color = "red", linetype = "dashed", linewidth = 1) +
      
      labs(title = paste("1D Elemental Plot:", input$element),
           x = "Row Index (from filtered start point)",
           y = "Element Value") +
      theme_bw(base_size = 14)
  })
  
  # 1D Click Info Output
  output$click_info_1d <- renderPrint({
    if (!is.null(rv$last_plot_click)) {
      cat(paste("Index:", round(rv$last_plot_click)))
    }
  })
  
  # Core text displays
  output$core_start_display <- renderText({
    paste("Core Start Index:", current_core_times()$core_start_index)
  })
  output$core_end_display <- renderText({
    paste("Core End Index:", current_core_times()$core_end_index)
  })
  
  # --- 3e. 2D Image Plot Logic (Right Column) ---
  
  # Load image
  observeEvent(input$upload_image, {
    req(input$upload_image)
    rv$image_clicks <- list() # Reset clicks on new image
    tryCatch({
      rv$transect_image <- image_read(input$upload_image$datapath)
    }, error = function(e) {
      showNotification(paste("Error reading image:", e$message), type = "error")
      rv$transect_image <- NULL
    })
  })
  
  # Store image clicks
  observeEvent(input$image_click, {
    if (length(rv$image_clicks) < 2) {
      rv$image_clicks <- c(rv$image_clicks, list(c(input$image_click$x, input$image_click$y)))
    } else {
      rv$image_clicks <- list(c(input$image_click$x, input$image_click$y)) # Start over
    }
  })
  
  # 2D Image Plot Output
  output$image_plot <- renderPlot({
    if (is.null(rv$transect_image)) {
      plot(1, type="n", axes=FALSE, xlab="", ylab="")
      text(1, 1, "Please upload a .tif or .jpg image", cex=1.5)
      return()
    }
    
    # Plot the base image
    par(mar = c(0, 0, 0, 0)) # Remove plot margins
    plot(rv$transect_image)
    
    # If we have two clicks, draw the colored line
    if (length(rv$image_clicks) == 2) {
      p1 <- rv$image_clicks[[1]]
      p2 <- rv$image_clicks[[2]]
      
      # Get the 1D elemental data to map
      elem_data <- current_element_data()$value
      n_points <- length(elem_data)
      
      # --- Robust Color Scaling (as requested) ---
      q_low <- quantile(elem_data, 0.05, na.rm = TRUE)
      q_high <- quantile(elem_data, 0.95, na.rm = TRUE)
      
      if (is.na(q_low) || is.na(q_high) || q_low == q_high) {
        colors <- rep("#808080", n_points) # Fallback to grey
      } else {
        elem_data_capped <- pmax(q_low, pmin(q_high, elem_data))
        elem_data_norm <- (elem_data_capped - q_low) / (q_high - q_low)
        colors <- viridis(100)[cut(elem_data_norm, breaks = 100, labels = FALSE)]
        colors[is.na(colors)] <- "#808080" # Grey for NA
      }

      # Create x and y coordinates for the line segments
      x_coords <- seq(p1[1], p2[1], length.out = n_points)
      y_coords <- seq(p1[2], p2[2], length.out = n_points)
      
      # Draw the colored line
      segments(
        x0 = x_coords[-n_points], y0 = y_coords[-n_points],
        x1 = x_coords[-1], y1 = y_coords[-1],
        col = colors[-n_points],
        lwd = 3 # Line width
      )
      
      # --- ‼️ NEW: ADD CORE ARROWS ‼️ ---
      times <- current_core_times()
      
      # --- Calculate perpendicular offset for arrows ---
      line_vec_x <- p2[1] - p1[1]
      line_vec_y <- p2[2] - p1[2]
      line_length <- sqrt(line_vec_x^2 + line_vec_y^2)
      
      # Perpendicular vector (normalized)
      perp_vec_x <- -line_vec_y / line_length
      perp_vec_y <- line_vec_x / line_length
      
      # Offset amount (e.g., 5% of image height)
      offset_amount <- image_info(rv$transect_image)$height * 0.05
      
      # Draw CORE START arrow
      if (!is.na(times$core_start_index) && times$core_start_index <= n_points) {
        idx <- times$core_start_index
        # Point on the line
        line_x <- x_coords[idx]
        line_y <- y_coords[idx]
        
        # Arrow start/end points
        arrow_start_x <- line_x + (perp_vec_x * offset_amount)
        arrow_start_y <- line_y + (perp_vec_y * offset_amount)
        arrow_end_x <- line_x + (perp_vec_x * offset_amount * 0.1) # Point closer to line
        arrow_end_y <- line_y + (perp_vec_y * offset_amount * 0.1)
        
        arrows(arrow_start_x, arrow_start_y, arrow_end_x, arrow_end_y,
               col = "red", lwd = 3, length = 0.15)
      }
      
      # Draw CORE END arrow
      if (!is.na(times$core_end_index) && times$core_end_index <= n_points) {
        idx <- times$core_end_index
        # Point on the line
        line_x <- x_coords[idx]
        line_y <- y_coords[idx]
        
        # Arrow start/end points
        arrow_start_x <- line_x + (perp_vec_x * offset_amount)
        arrow_start_y <- line_y + (perp_vec_y * offset_amount)
        arrow_end_x <- line_x + (perp_vec_x * offset_amount * 0.1)
        arrow_end_y <- line_y + (perp_vec_y * offset_amount * 0.1)
        
        arrows(arrow_start_x, arrow_start_y, arrow_end_x, arrow_end_y,
               col = "red", lwd = 3, length = 0.15)
      }
      # --- ‼️ END NEW CODE ‼️ ---
      
    } 
    # Draw the click points
    else if (length(rv$image_clicks) == 1) {
      points(rv$image_clicks[[1]][1], rv$image_clicks[[1]][2], col = "red", pch = 3, cex = 2, lwd = 3)
    }
  })
  
  # 2D Click Info Output
  output$click_info_2d <- renderPrint({
    if (is.null(rv$transect_image)) {
        cat("To zoom: Hold Ctrl and use mouse wheel.\n\n")
        cat("Waiting for image upload...")
    } else if (length(rv$image_clicks) == 0) {
      cat("To zoom: Hold Ctrl and use mouse wheel.\n\n")
      cat("Click on the image to set the transect START point.")
    } else if (length(rv$image_clicks) == 1) {
      cat("To zoom: Hold Ctrl and use mouse wheel.\n\n")
      cat("Click on the image to set the transect END point.")
    } else {
      cat("To zoom: Hold Ctrl and use mouse wheel.\n\n")
      cat("Transect line drawn. Click again to restart.")
    }
  })
  
  # --- 3f. Download Logic ---
  output$download_cores <- downloadHandler(
    filename = function() {
      "core_cutoffs.csv"
    },
    content = function(file) {
      # Save only the rows that have been marked
      final_cutoffs <- rv$core_cutoffs %>%
        filter(!is.na(core_start_index) & !is.na(core_end_index))
      write.csv(final_cutoffs, file, row.names = FALSE)
    }
  )
}

# --- 4. Run the App ---
shinyApp(ui = ui, server = server)



# THIS WORKS WELL, CUTOFFS ARE FOUND IN : ICPMS_corevalues.csv
```





# apply new filter to my otolith dataframe.

```{r}
# --- 1. Load Packages ---
# install.packages(c("tidyverse", "readxl"))
library(tidyverse)
library(readxl)

# --- 2. Re-create the 'final_filtered_data' (from previous script) ---
# This ensures this script is runnable on its own and uses the correct data.

# 2a. Load Grades
cat("Loading transect grades...\n")
grades_file <- "transect grades.xlsx"
if (!file.exists(grades_file)) {
  stop(paste("Error:", grades_file, "not found."))
}
grades_wide <- read_xlsx(grades_file)

grades_long <- grades_wide %>%
  rename(specimen_num = `Specimen Number`) %>%
  pivot_longer(
    cols = starts_with("Transect"),
    names_to = "transect_name",
    values_to = "grade"
  ) %>%
  filter(!is.na(grade)) %>%
  mutate(
    specimen = sprintf("oto%02d", specimen_num),
    transect = str_extract(transect_name, "\\d+"),
  ) %>%
  select(specimen, transect, grade)

top_2_transects <- grades_long %>%
  group_by(specimen) %>%
  slice_min(order_by = grade, n = 2, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    specimen = as.factor(specimen),
    transect = as.factor(transect)
  )

# 2b. Load Raw ICP-MS Data
cat("Loading raw ICP-MS data...\n")
df_raw <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")
df_raw <- df_raw[c(30:184)]
df_list_raw <- list()
for(i in 1:length(df_raw)){
  spec <- names(df_raw[i])
  df_list_raw[[i]] <- df_raw[i][[1]]
  df_list_raw[[i]]$specimen <- as.factor(str_sub(spec, start= 0,end = 5))
  df_list_raw[[i]]$transect <- as.factor(as.numeric(str_sub(spec, start= -1)) + 1)
}
df <- do.call("rbind", df_list_raw)
df <- rename(df, "time" = `Elapsed Time`)

# 2c. Filter by "Top 2" and "Unusable"
unusable <- paste0("oto", c(05, 10, 12, 14, 30, 37, 38, 40))
df_semi_filtered <- df %>%
  filter(!specimen %in% unusable)
df_to_process <- df_semi_filtered %>%
  inner_join(top_2_transects, by = c("specimen", "transect"))

# 2d. Load Manual Transect Cutoffs
cat("Loading manual transect cutoffs (manual_cutoffs.csv)...\n")
cutoff_file <- "manual_cutoffs.csv"
if (!file.exists(cutoff_file)) {
  stop(paste("Error:", cutoff_file, "not found."))
}
manual_cutoffs <- read_csv(cutoff_file, 
                           col_types = cols(
                             specimen = col_factor(),
                             transect = col_factor(),
                             start_time = col_double(),
                             end_time = col_double()
                           ))

# 2e. Create the 'final_filtered_data'
df_joined <- df_to_process %>%
  inner_join(manual_cutoffs, by = c("specimen", "transect"))

final_filtered_data <- df_joined %>%
  filter(time >= start_time & time <= end_time) %>%
  select(-start_time, -end_time) %>%
  group_by(specimen, transect) %>%
  mutate(index = row_number()) %>% # The index *within* the filtered transect
  ungroup()

cat("Created 'final_filtered_data'.\n")

# --- 3. Load Your Manually-Defined Core Cutoffs ---
cat("Loading core cutoffs (ICPMS_corevalues.csv)...\n")
core_cutoff_file <- "ICPMS_corevalues.csv"
if (!file.exists(core_cutoff_file)) {
  stop(paste("Error:", core_cutoff_file, "not found."))
}
core_cutoffs <- read_csv(core_cutoff_file,
                         col_types = cols(
                           specimen = col_factor(),
                           transect = col_factor(),
                           core_start_index = col_integer(),
                           core_end_index = col_integer()
                         ))

# --- 4. Join Core Cutoffs and Define Regions ---
cat("Joining core cutoffs and defining regions...\n")

# Join the core definitions to the filtered data
data_with_cores <- final_filtered_data %>%
  left_join(core_cutoffs, by = c("specimen", "transect"))

# Define regions based on your rules
data_with_regions_raw <- data_with_cores %>%
  # Handle any transects you didn't mark in the app
  filter(!is.na(core_start_index)) %>% 
  group_by(specimen, transect) %>%
  # Get total length *after* core start (as per your Rule 2)
  # This step is complex. Let's do it simpler.
  
  # --- Simpler & Better Logic ---
  # 1. Get total length of the *entire* transect (after time filtering)
  mutate(total_length = n()) %>%
  # 2. Define the edge_start_index based on your 10% rule
  mutate(edge_start_index = floor(total_length * 0.90)) %>%
  
  # 3. Assign regions using case_when()
  mutate(
    oto_region = case_when(
      # Rule 2: All points *before* core are NULL
      index < core_start_index ~ NA_character_,
      
      # Rule 1: "core"
      index >= core_start_index & index <= core_end_index ~ "core",
      
      # Rule 4: "edge" (Last 10%)
      index >= edge_start_index ~ "edge",
      
      # Rule 3: "middle" (everything else)
      index > core_end_index & index < edge_start_index ~ "middle"
    )
  ) %>% 
  ungroup()

# --- 5. Final Filtering and Formatting ---
cat("Cleaning final dataframe...\n")

final_analysis_data <- data_with_regions_raw %>%
  # Rule 2: Remove all points BEFORE the core
  filter(!is.na(oto_region)) %>%
  
  # Re-level the factor so "core" is the baseline
  mutate(
    oto_region = factor(oto_region, levels = c("core", "middle", "edge"))
  )

# --- 6. Inspect Your New Data ---
cat("Done! Your final, analysis-ready dataframe is 'final_analysis_data'.\n\n")


saveRDS(final_analysis_data,"RDS_dataframes/ICPMS_filteredregionotos.RDS")
cat("Head of the new data:\n")
print(head(final_analysis_data))

cat("\n\nSummary of new 'oto_region' variable:\n")
print(summary(final_analysis_data$oto_region))

cat("\n\nCheck counts by specimen and region:\n")
print(
  final_analysis_data %>%
    group_by(specimen, oto_region) %>%
    summarise(n_points = n(), .groups = "drop"))
```
