---
output: html_document
editor_options: 
  chunk_output_type: console
---


# All old code no longer being used - archiving to see where mistakes were made and keep a general record of my coding.

# Variable/Wavelength Selection & PLS on __length__ as a test
Using temp_proc df produced from quickproc() function above.  VIP score to select wavelengths that are most informative to model

```{r}
quickproc(scan_avg,1,3,17) # SG of scan_avg DF to genereate temp_proc df for models below

test_pls_length <- pls(temp_proc[, 21:ncol(temp_proc)], temp_proc[, 2],
  scale = T, center = F, info = "Length Prediction Model",
  cv = 1
)
saveRDS(test_pls_length, "RDS_dataframes/PLS_Length")

# pls only with wavenumebrs VIP > 1
length_VIP <- vipscores(test_pls_length) # extract VIP values for wavenumbers of above pls model
plotVIPScores(test_pls_length)
test_waves <- names(subset(length_VIP[,1], length_VIP[,1] > 1)) 
all_waves <- names(temp_proc[,21:length(temp_proc)])
bad_waves <- all_waves[!(all_waves %in% test_waves)]
test_pls_length_VIP <- pls(temp_proc[,21:ncol(temp_proc)], temp_proc[,2],
  scale = T, center = F, info = "Length Prediction Model",
  exclcols = bad_waves,
  cv = 1
)
saveRDS(test_pls_length_VIP, "RDS_dataframes/PLS_Length_VIP")

# unprocessed df PLS
test_pls_length_unproc <- pls(scan_avg[, 21:ncol(scan_avg)], scan_avg[, 2],
  scale = T, center = F, info = "Length Prediction Model",
  cv = 1
)
saveRDS(test_pls_length_unproc, "RDS_dataframes/PLS_Length_unproc")

# 90% train, 10% test PLS, no VIP
set.seed(1)
idx <- floor(nrow(temp_proc) * 0.9)
train_idx <- sample(seq_len(nrow(temp_proc)), size = idx)
train <- temp_proc[train_idx, -c(1,3:20)]
test <- temp_proc[-train_idx, -c(1,3:20)]
test_pls_length_split <- pls(train[,-c(1)], train[,1],
  scale = T, center = F,
  info = "Length Prediction Model", cv = 1,
  x.test = test[,-c(1)], y.test = test[,1]
)
saveRDS(test_pls_length_split, "RDS_dataframes/PLS_Length_split")

# 90% train, 10% test PLS, VIP
test_pls_length_split_VIP <- pls(train[,-c(1)], train[,1],
  scale = T, center = F,
  info = "Length Prediction Model", cv = 1,
  x.test = test[,-c(1)], y.test = test[,1],
  exclcols = bad_waves
)
saveRDS(test_pls_length_split_VIP, "RDS_dataframes/PLS_Length_split_VIP")

plot(test_pls_length)
plot(test_pls_length_VIP)
plot(test_pls_length_unproc)
plot(test_pls_length_split)
plot(test_pls_length_split_VIP)

test_pls_length$res$cal$rmse[2]
test_pls_length_VIP$res$cal$rmse[3]
test_pls_length_unproc$res$cal$rmse[3]
test_pls_length_split$res$test$rmse[2]
test_pls_length_split_VIP$res$test$rmse[3]

rm(test_pls_length, test_pls_length_VIP, test_pls_length_unproc, test_pls_length_split, test_pls_length_split_VIP)
rm(test,train,length_VIP, all_waves, bad_waves, idx, test_waves, train_idx)
```

# PCR for age

```{r}
quickproc(age_only,1,3,17)
pca_age_LPW <- pca(temp_proc[21:ncol(temp_proc)], scale = T)
plot(pca_age_LPW)
mlr_age_pcs <- pca_age_LPW$calres$scores[,1:10]
mlr_age_pcs <- cbind(mlr_age_pcs, temp_proc[,1:11])
mlr_age <- lm(data = mlr_age_pcs, read_age ~ `Comp 1` + `Comp 2` + `Comp 3` + `Comp 4` + `Comp 5`)
summary(mlr_age)

# MLR with split
set.seed(1)
idx <- floor(nrow(age_only) * 0.9) # 90% of indices needed for training set
train_idx <- sample(seq_len(nrow(age_only)), size = idx) # generate indices for training set
train <- mlr_age_pcs[train_idx, -c(11:20)] # wavenumber measurement columns only
test <- mlr_age_pcs[-train_idx, -c(11:20)]
mlr_age_split <- lm(data = train, read_age ~`Comp 1` + `Comp 2` + `Comp 3` + `Comp 4` + `Comp 5`)
summary(mlr_age_split)
mlr_age_split_pred <- predict(mlr_age_split,test)
ggplot() +  # fitted values vs actual, out of sample prediction in red
  geom_point(aes(train$read_age,mlr_age_split$fitted.values)) + 
  geom_point(aes(test$read_age,mlr_age_split_pred),col = "red") + 
  geom_abline(slope=1, intercept = 0, col = "red")

# Split for PCR
set.seed(1)
idx <- floor(nrow(temp_proc) * 0.9)
train_idx <- sample(seq_len(nrow(temp_proc)), size = idx)
train <- age_only[train_idx, -c(1:10,12:20)] 
test <- age_only[-train_idx, -c(1:10,12:20)] 

pcr_age_model <- pcr(read_age ~ .,
                 data = train, 
                 scale = T,
                 validation = "CV",
                 ncomp = 10)
selectNcomp(pcr_age_model, "onesigma", plot = TRUE) # determine appropriate number of comps
scoreplot(pcr_age_model) # plot of first 2 PCs

pcr_age_pred <- predict(pcr_age_model, newdata = test, ncomp = 4) # fit model to test data, store fitted values
ggplot() +
  geom_point(aes(x = train$read_age, y = pcr_age_model$fitted.values[, , 4])) + # extract fitted values of model with 2 PCs.
  geom_point(aes(x = test$read_age, y = pcr_age_pred), col = "red") +
  geom_abline(slope = 1, col = "red")

RMSEP(pcr_age_model)


```



# PCR & PC GAMs using length

This code is no longer relevant; I was incorrectly applying the PCA prior to splitting dataset.
```{r}

# extract PC's
pca_LPW <- pca(scan_avg[21:ncol(scan_avg)], scale = T) # PCA for scan_avg
plot(pca_LPW)
pca_LPW <- pca_LPW$calres$scores[, 1:20] # extract PCs
pca_LPW <- cbind(pca_LPW[, 1:10], scan_avg[, 1:20]) # only store first 10 PC's
colnames(pca_LPW)[c(1:10)] <- c("PC1", "PC2", "PC3", "PC4", "PC5","PC6", "PC7", "PC8", "PC9", "PC10")

# PCR
PCR_length <- lm(data = pca_LPW, length ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10)
summary(PCR_length) # some PC's seem less informative than others
PCR_length2 <- update(PCR_length, length ~ PC1 + PC2 + PC5 + PC6 + PC8 + PC10)
summary(PCR_length2)
plot(PCR_length)
plot(PCR_length2)


PCR_length3 <- pcr(length ~ ., data = scan_avg[,c(2,21:ncol(scan_avg))], ncomp = 20,
                   validation = "CV")
selectNcomp(PCR_length3, "onesigma", plot = TRUE) # determine appropriate number of comps, suggested 4
scoreplot(PCR_length3) # plot of first 2 PCs


PCR_length3$fitted.values[,,4]


set.seed(1)
inTraining <- createDataPartition(scan_avg$length, p = .9, list = F)
training <- scan_avg[inTraining,-c(1,3:20)]
testing  <- scan_avg[-inTraining,-c(1,3:20)]
fitControl <- trainControl(## 10-fold CV
                           method = "repeatedcv",
                           number = 10,
                           ## repeated ten times
                           repeats = 10,
                           preProcOptions = list(pcaComp = 10))
PCR_length4 <- train(length ~ ., data = training, 
                 method = "lm", 
                 trControl = fitControl,
                 preProcess = "pca")
PCR_length4$
scan_avg[inTraining,-c(1,3:20)]

pca <- prcomp(scan_avg[inTraining,-c(1:20)],center=F,scale=T)
pca$rotation
pca$sdev
pca_LPW$PC1
dim(pca_LPW1$x)
pca_LPW1$loadings
testing1 <- preProcess(scan_avg[inTraining,-c(1:20)], method = "pca", pcaComp = 10)
predict(testing1, scan_avg[inTraining,-c(1,3:ncol(scan_a))])


postResampleSpectro(predict(PCR_length4,testing),testing$length)
PCR_length4
pcr1
PCR_length4$finalModel






# PCR with 10-fold split for RMSE
set.seed(1)
splits <- caret::createFolds(pca_LPW$length, k = 10, list = TRUE, returnTrain = FALSE)
PCR_errors <- vector()
# GAM
for(i in 1:10){
  mod <- lm(data = pca_LPW[-splits[[i]],], length ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10)
  preds <- predict(mod, newdata = pca_LPW[splits[[i]],])
  PCR_errors[i] <- RMSE(pca_LPW[splits[[i]],]$length, preds)
}
# RMSE
mean(PCR_errors)

# PCR with only significant PC's from lm() summary
for(i in 1:10){
  mod <- lm(data = pca_LPW[-splits[[i]],], length ~ PC1 + PC2 + PC5 + PC6 + PC8 + PC10)
  preds <- predict(mod, newdata = pca_LPW[splits[[i]],])
  PCR_errors2[i] <- RMSE(pca_LPW[splits[[i]],]$length, preds)
}
#RMSE
mean(PCR_errors2)

# choosing the best PC's for length results in lowest RMSE

pred <- predict(PCR_length2,pca_LPW[splits[[1]],])
ggplot() +  # fitted values vs actual, out of sample prediction in red
  geom_point(aes(pca_LPW[splits[[1]],]$length,PCR_length2$fitted.values)) + 
  geom_point(aes(test$length,PCR_length_split_pred),col = "red") + 
  geom_abline(slope=1, intercept = 0, col = "red")

# Split for PCR
set.seed(1)
idx <- floor(nrow(scan_avg) * 0.9) # 90% of indices needed for training set
train_idx <- sample(seq_len(nrow(scan_avg)), size = idx) # generate indices for training set
train <- scan_avg[train_idx, -c(1, 3:20)] # length and wavenumber measurement columns only
test <- scan_avg[-train_idx, -c(1, 3:20)]

# PCR model 
pcr_length_model <- pls::pcr(length ~ .,
  data = train,
  scale = T,
  validation = "CV",
  ncomp = 10
) # look at 10 components max
selectNcomp(pcr_length_model, "onesigma", plot = TRUE) # determine appropriate number of comps
scoreplot(pcr_length_model) # plot of first 2 PCs
loadingplot(pcr_length_model)

pcr_length_pred <- predict(pcr_length_model, newdata = test, ncomp = 2) # fit model to test data, store fitted values
ggplot() +
  geom_point(aes(x = train$length, y = pcr_length_model$fitted.values[, , 2])) + # extract fitted values of model with 2 PCs.
  geom_point(aes(x = test$length, y = pcr_length_pred), col = "red") +
  geom_abline(slope = 1, col = "red")

```


# PLS for Age

```{r}
# df with only aged specimens
age_only <- scan_avg[complete.cases(scan_avg$read_age),]

# test pls, no variable selection
test_pls_age <- pls(age_only[, 21:ncol(age_only)], age_only[, 11],
  scale = F, center = T, info = "Age Prediction Model",
  cv = 1
)
plot(test_pls_age)
# pls with VIP > 1
test_vip_age <- vipscores(test_pls_age)
plotVIPScores(test_pls_age)
good_waves <- as.numeric(names(subset(test_vip_age[,1], test_vip_age[,1] > 1)))
all_waves <- names(age_only[,21:length(age_only)])
bad_waves <- all_waves[!(all_waves %in% good_waves)]

test_pls_age_VIP <- pls(age_only[, 21:ncol(age_only)], age_only[, 11],
 scale = F, center = T, info = "Age Prediction Model",
 cv = 1, exclcols = bad_waves
)

# 90/10 split for PLS, no VIP
set.seed(1)
idx <- floor(nrow(age_only) * 0.9)
train_idx <- sample(seq_len(nrow(age_only)), size = idx)
train <- age_only[train_idx, -c(1:10,12:20)]
test <- age_only[-train_idx, -c(1:10,12:20)]

test_pls_age_split <- pls(train[,-1], train[,1],
  scale = F, center = T,
  info = "Age Prediction Model", cv = 1,
  x.test = test[,-1], y.test = test[,1]
)

# 90/10 split for PLS, VIP > 1
test_pls_age_split_VIP <- pls(train[,-1], train[,1],
  scale = F, center = T,
  info = "Age Prediction Model", cv = 1,
  x.test = test[,-c(1)], y.test = test[,1],
  exclcols = bad_waves
)

plot(test_pls_age)
plot(test_pls_age_VIP)
plot(test_pls_age_split)
plot(test_pls_age_split_VIP)

test_pls_age$res$cal$rmse[3]
test_pls_age_VIP$res$cal$rmse[2]
test_pls_age_split$res$test$rmse[3]
test_pls_age_split_VIP$res$cal$rmse[2]
```


# Trying some things with other kinds of preprocessing.  Doing savitzky-golay and then a follow-up SNV
a
```{r}
scan_2 <- dfmeta_LPW %>% dplyr::filter(run_number == 2)
# scan_2_long <- pivot_longer(scan_2, cols = `11536`:`3952`, names_to = "name", values_to = "value")
# scan_2_long$name <- as.numeric(scan_2_long$name)
scan_3 <- dfmeta_LPW %>% dplyr::filter(run_number == 3)
# scan_3_long <- pivot_longer(scan_3, cols = `11536`:`3952`, names_to = "name", values_to = "value")
# scan_3_long$name <- as.numeric(scan_3_long$name)
testing1 <- bind_cols(NULL, scan_2[, 1:20])
testing1 <- bind_cols(testing1, (scan_2[, 21:ncol(scan_2)] + scan_3[, 21:ncol(scan_3)]) / 2)
rm(scan_2, scan_3)
testing2 <- cbind(testing1[, c(1:20)],savitzkyGolay(testing1[,21:ncol(testing1)],m = 1, p = 3, w = 17))
testing3 <- prep.snv(as.matrix(testing2[,21:ncol(testing2)]))
testing3 <- as.data.frame(unlist(testing3))
testing3 <- cbind(testing2[,1:20], testing3)

testing_long <- pivot_longer(testing1, cols = `11536`:`3952`, names_to = "name", values_to = "value")
testing_long$name <- as.numeric(testing_long$name)

testing_long2 <- pivot_longer(testing2, cols = `11472`:`4016`, names_to = "name", values_to = "value")
testing_long2$name <- as.numeric(testing_long2$name)

testing_long3 <- pivot_longer(testing3, cols = `11472`:`4016`, names_to = "name", values_to = "value")
testing_long3$name <- as.numeric(testing_long3$name)

ggplot(testing_long[complete.cases(testing_long$read_age), ], aes(x = name, y = value, group = specimen, color = read_age)) + 
  geom_path() + 
  scale_x_reverse() + 
  scale_color_viridis() + 
  labs(color = "Age (days)") + 
  labs(y = "Preprocessed absorbance", x = expression(paste("Wavenumber ", cm^-1))) + 
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 15),
        legend.text = element_text(size = 10, vjust = .2),
        legend.title = element_text(size = 15, vjust = 1))

ggplot(testing_long2[complete.cases(testing_long2$read_age), ], aes(x = name, y = value, group = specimen, color = read_age)) + 
  geom_path() + 
  scale_x_reverse() + 
  scale_color_viridis() + 
  labs(color = "Age (days)") + 
  labs(y = "Preprocessed absorbance", x = expression(paste("Wavenumber ", cm^-1))) + 
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 15),
        legend.text = element_text(size = 10, vjust = .2),
        legend.title = element_text(size = 15, vjust = 1))

ggplot(testing_long3[complete.cases(testing_long3$read_age), ], aes(x = name, y = value, group = specimen, color = read_age)) + 
  geom_path() + 
  scale_x_reverse() + 
  scale_color_viridis() + 
  labs(color = "Age (days)") + 
  labs(y = "Preprocessed absorbance", x = expression(paste("Wavenumber ", cm^-1))) + 
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 15),
        legend.text = element_text(size = 10, vjust = .2),
        legend.title = element_text(size = 15, vjust = 1))
unproc <- cbind(mod.summary,AIC.summary)
```

```{r, echo = F, results=F}
# Create dataframe only with ages contained in metadata
age_only <- testing2[complete.cases(testing2$read_age), ]

age_only <- age_only[,-c(21:517)]

# outliers identified with plotly and PCA
# SG
age_only <- age_only %>%
dplyr::filter(read_age != 188, read_age != 181, read_age != 175, read_age != 161, read_age != 148, read_age != 147, read_age != 135)

# store metrics from each fold and each model type
RMSE.age <- list()
r2.age <- list()
AIC.age <- list()
AICc.age <- list()

# 10 fold CV split - create folds
set.seed(6)
splits <- caret::createFolds(age_only$read_age, k = 10, list = TRUE, returnTrain = FALSE)

# extract PC's for each calibration set, create test sets with ages and spectra
cal <- list()
test <- list()
for (i in 1:10) {
  # calibration set and PC's
  pc.mod <- preProcess(age_only[-splits[[i]], -c(1:20)], method = "pca", thresh = 0.95, pcaComp = 10)
  pc.cal <- predict(pc.mod, age_only[-splits[[i]], -c(1:20)])
  pc.cal <- cbind(pc.cal, age_only[-splits[[i]], ])
  cal[[i]] <- pc.cal
  # test sets
  pc.test <- predict(pc.mod, age_only[splits[[i]], -c(1:20)])
  pc.test <- cbind(pc.test, age_only[splits[[i]], ])
  test[[i]] <- pc.test
}
rm(pc.cal,pc.test)

# determine which PC's to include via step & AIC selection
mod.sel <- list()
for (i in 1:10) {
  pctest <- cal[[i]]
  temp <- step(lm(data = pctest[-splits[[i]], ], read_age ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10))
  mod.sel[[i]] <- rownames(summary(temp)$coef)
}

# PICK MODEL BASED ON TOP 5 AIC - worse performance than just using 4 PC's

# lm() with 10-fold CV
lm.mods <- list()
for (i in 1:10) {
  PC_5 <- sort(table(mod.sel[i]),decreasing = T)
  PC_5 <- rownames(PC_5)[2:6] # top 5 PC's
  PC_5 <- PC_5[complete.cases(PC_5)] # make sure not NA's
  calibrate <- cal[[i]]
  testing <- test[[i]]
  mod <- lm(data = calibrate, reformulate(PC_5, "read_age"))
  # RMSE.age$lm.cal[i] <- caret::RMSE(pred = mod$fitted.values, obs = calibrate[, 21])
  preds <- predict(mod, newdata = testing)
  RMSE.age$lm.test[i] <- caret::RMSE(pred = preds, obs = testing[, 21])
  # r2.age$lm.cal[i] <- summary(mod)$r.squared
  RSS <- sum((testing$read_age - preds)^2)
  TSS <- sum((testing$read_age - mean(testing$read_age))^2)
  r2.age$lm.test[i] <- 1 - (RSS / TSS)
  AIC.age$lm[i] <- AIC(mod)
  AICc.age$lm[i] <- AICc(mod)
  lm.mods[[i]] <- mod
}
# RMSE.age$lm.cal <- mean(RMSE.age$lm.cal)
RMSE.age$lm.test <- mean(RMSE.age$lm.test)
# r2.age$lm.cal <- mean(r2.age$lm.cal)
r2.age$lm.test <- mean(r2.age$lm.test)

# GAM with 10 fold CV, select = T allows PCs to be penalized and effectively removed from model if appropriate. 
GAM.mods <- list()
for (i in 1:10) {
  calibrate <- cal[[i]]
  testing <- test[[i]]
  mod <- gam(data = calibrate, read_age ~ s(PC1) + s(PC2) + s(PC3) + s(PC4), method = "REML", select = T)
  # Extract AIC, AICc, RMSE (cal & test) & r2 (cal & test)
  # RMSE.age$GAM.cal[i] <- caret::RMSE(pred = mod$fitted.values, obs = calibrate[, 21])
  preds <- predict(mod, newdata = testing)
  RMSE.age$GAM.test[i] <- caret::RMSE(pred = preds, obs = testing[, 21])
  # r2.age$gam.cal[i] <- summary(mod)$r.sq
  RSS <- sum((testing$read_age - preds)^2)
  TSS <- sum((testing$read_age - mean(testing$read_age))^2)
  r2.age$gam.test[i] <- 1 - (RSS / TSS)
  AIC.age$gam[i] <- AIC(mod)
  AICc.age$gam[i] <- AICc(mod)
  GAM.mods[[i]] <- mod
}
# r2.age$gam.cal <- mean(r2.age$gam.cal)
r2.age$gam.test <- mean(r2.age$gam.test)
# RMSE.age$GAM.cal <- mean(RMSE.age$GAM.cal)
RMSE.age$GAM.test <- mean(RMSE.age$GAM.test)

AIC.summary <- data.frame(
  model = c(rep("lm", 10), rep("gam", 10)),
  AIC = 1:20,
  AICc = 1:20
)
AIC.summary$AIC <- unlist(AIC.age)
AIC.summary$AICc <- unlist(AICc.age)

# LM
mean(AIC.summary[1:10,2]) # AIC
mean(AIC.summary[1:10,3]) # AICc

# GAM
mean(AIC.summary[11:20,2]) # AIC
mean(AIC.summary[11:20,3]) # AICc


```

# PCA functions for plotting and other stuff from old code 
```{r}
pca.var <- function(mydf, mypca) {
  PC1perc <<- round(mypca$eigenvals[1] / sum(mypca$eigenvals), digits = 4) * 100
  PC2perc <<- round(mypca$eigenvals[2] / sum(mypca$eigenvals), digits = 4) * 100
  varsc0 <- as.data.frame(mypca$loadings[, 1:3])
  objsc0 <- as.data.frame(mypca$calres$scores[, 1:3])
  Spectrascores_all <- as.data.frame(cbind(objsc0$`Comp 1`, objsc0$`Comp 2`, objsc0$`Comp 3`))
  colnames(Spectrascores_all) <- c("PCs1", "PCs2", "PCs3")
  SpecVec <- as.data.frame(cbind(varsc0$`Comp 1`, varsc0$`Comp 2`, varsc0$`Comp 3`))
  head(SpecVec)
  colnames(SpecVec) <- c("PC1", "PC2", "PC3")
  Spectrascores_all <- cbind(Spectrascores_all, mydf[, 1:8])
}


PCA.plot <- function(mydf, mypca, color.by) {
  temp <- pca.var(mydf = mydf, mypca = mypca)
  ggplot() +
    geom_hline(yintercept = 0, color = "gray") +
    geom_vline(xintercept = 0, color = "gray") +
    geom_point(data = temp, aes(x = PCs1, y = PCs2, color = {{ color.by }}), size = 3) +
    labs(x = paste("PC1 (", PC1perc, "%)", sep = ""), y = paste("PC2 (", PC2perc, "%)", sep = "")) +
    scale_color_viridis_c() +
    theme(
      panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text = element_text(size = 16),
      axis.title = element_text(size = 16, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 10.5),
      plot.title = element_text(size = 16, face = "bold"),
      strip.text.x = element_text(size = 14)
    )
}


PCA.plot.bigger <- function(mydf, mypca, color.by) {
  temp <- pca.var(mydf = mydf, mypca = mypca)
  ggplot() +
    geom_hline(yintercept = 0, color = "gray") +
    geom_vline(xintercept = 0, color = "gray") +
    geom_point(data = temp, aes(x = PCs1, y = PCs2, color = {{ color.by }}), size = 5) +
    labs(
      x = paste("PC1 (", PC1perc, "%)", sep = ""),
      y = paste("PC2 (", PC2perc, "%)", sep = "")
    ) +
    scale_color_viridis_c() +
    theme(
      panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text = element_text(size = 16),
      axis.title = element_text(size = 16, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 10.5),
      plot.title = element_text(size = 16, face = "bold"),
      strip.text.x = element_text(size = 14)
    )
}

mypca <- pca(age_only[,21:ncol(age_only)], scale = F, center = T)
plotVariance(mypca)
plot(mypca)
PC1perc <<- round(mypca$eigenvals[1] / sum(mypca$eigenvals), digits = 4) * 100
PC2perc <<- round(mypca$eigenvals[2] / sum(mypca$eigenvals), digits = 4) * 100
varsc0 <- as.data.frame(mypca$loadings[, 1:3])
objsc0 <- as.data.frame(mypca$calres$scores[, 1:3])
Spectrascores_all <- as.data.frame(cbind(objsc0$`Comp 1`, objsc0$`Comp 2`, objsc0$`Comp 3`))
colnames(Spectrascores_all) <- c("PCs1", "PCs2", "PCs3")
SpecVec <- as.data.frame(cbind(varsc0$`Comp 1`, varsc0$`Comp 2`, varsc0$`Comp 3`))
head(SpecVec)
colnames(SpecVec) <- c("PC1", "PC2", "PC3")
Spectrascores_all <- cbind(Spectrascores_all, read_age = age_only[,11])
fig <- ggplot() +
  geom_hline(yintercept = 0, color = "gray") +
  geom_vline(xintercept = 0, color = "gray") +
  geom_point(data = Spectrascores_all, aes(x = PCs1, y = PCs2, color = read_age), size = 3) +
  labs(x = paste("PC1 (", PC1perc, "%)", sep = ""), y = paste("PC2 (", PC2perc, "%)", sep = "")) +
  scale_color_viridis_c() +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text = element_text(size = 16),
    axis.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 10.5),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text.x = element_text(size = 14)
  )
#library(plotly)
ggplotly(fig)
```

# IBM extract specimen numbers from rescans
```{r, cache = T}
testing1 <- cat(strings, "\n") # concatenates all filenames in a directory and adds a space delimiter
list.files(path = "IBM_scans")
strings <- list.files(path = "get_filename", )
library(stringr)
# Define the pattern to match
pattern <- "(?<=JJ202101202_)\\d+(?=_.+)"
# Use str_extract_all to extract matches from each string
matches <- str_extract_all(strings, pattern)
testing <- unlist(matches)
# Convert the matches to a data frame for easier inspection
df <- data.frame(numbers = unlist(matches))
print(df)
testing
```

# LPW functions to run_models and get_my_models given a df

```{r}
run_models <- function(mydf){
  df <- {{mydf}}
  df <- df[complete.cases(df$read_age), ] 
  pca_temp <- pca(df[, 21:ncol(df), ])
  pc_df <- data.frame(PC1 = rep(0,nrow(df)))
  for (i in 1:10) {
    pc_df[, paste0("PC", i)] <- pca_temp$res$cal$scores[, i]
    rm(i)
  }
  pc_df <- cbind(pc_df,df) 
  global_lm <- lm(data = pc_df, read_age ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10)
  global_gam <- gam(data = pc_df, read_age ~ s(PC1) + s(PC2) + s(PC3) + s(PC4) + s(PC5) + s(PC6)) # adding too many PC's makes GAM have horrible fit and R2
  
  # dredge to find top 5 models
  options(na.action = "na.fail")
  dredge_lm <- dredge(global_lm)
  top5_lm <- get.models(dredge_lm, subset = 1:5)
  
  # get.models was not working for gam in a function format so had to adapt this insane code to extract the top models....
  dredge_gam <- dredge(global_gam)
  top5_gam_temp <- dredge_gam[1:5,] 
  reconstruct_formula <- function(model_row) {
    terms_used <- names(model_row)[!is.na(model_row)]   # Extract the terms used in the model
    terms_used <- terms_used[!terms_used %in% c("(Intercept)", "df", "logLik", "AICc", "delta", "weight")]   # Remove non-predictor columns (e.g., AIC, delta, weight)
    as.formula(paste("read_age ~", paste(terms_used, collapse = " + ")))   # Construct the formula
  }
  top5_gam <- lapply(1:5, function(i) {reconstruct_formula(top5_gam_temp[i, ])})
  rm(top5_gam_temp, reconstruct_formula)
  
  # create lists for top 5 models of lm and gam
  terms_lm <- list()
  terms_gam <- list()
  for(i in 1:5){
    terms_lm[[i]] <- top5_lm[[i]]$terms
    terms_gam[[i]] <- top5_gam[[i]]
    rm(i)
  }
  rm(global_gam, global_lm, top5_gam, top5_lm, pc_df, pca_temp, dredge_gam,dredge_lm)
  
  # 10 fold CV split - create folds
  set.seed(6)
  splits <- caret::createFolds(df$read_age, k = 10, list = TRUE, returnTrain = FALSE)
  # extract PC's for each calibration set, create test sets with ages and spectra
  cal <- list()
  test <- list()
  for (i in 1:10) {
    # calibration set and PC's
    pc.mod <- preProcess(df[-splits[[i]], -c(1:20)], method = c("pca","center"), pcaComp = 10)
    pc.cal <- predict(pc.mod, df[-splits[[i]], -c(1:20)])
    pc.cal <- cbind(pc.cal, df[-splits[[i]], ])
    cal[[i]] <- pc.cal
    rm(pc.cal)
    # test sets
    pc.test <- predict(pc.mod, df[splits[[i]], -c(1:20)])
    pc.test <- cbind(pc.test, df[splits[[i]], ])
    test[[i]] <- pc.test
    rm(pc.test, pc.mod, i)
  }
  
  mods.lm <- replicate(10, list())
  r2_lm <- list()
  rmse_lm <- list()
  for (i in 1:10) {
    calibrate <- cal[[i]]
    testing <- test[[i]]
    r2_temp <- rep(0, 5)
    rmse_temp <- rep(0, 5)
    for (j in 1:5) {
      mod <- lm(data = calibrate, terms_lm[[j]])
      preds <- predict(mod, newdata = testing)
      rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
      RSS <- sum((testing$read_age - preds)^2)
      TSS <- sum((testing$read_age - mean(testing$read_age))^2)
      r2_temp[j] <- 1 - (RSS / TSS)
      mods.lm[[i]][[j]] <- mod
      rm(RSS, TSS, preds)
    }
    r2_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), r2 = r2_temp)
    rmse_lm[[i]] <- data.frame(Model = paste0("Linear ", 1:5), rmse = rmse_temp)
    rm(mod, i, j, r2_temp, rmse_temp, calibrate, testing)
  }
  # Unlist the r2 & rmse values and calc means for 10 splits, combine into dataframe
  r2_lm <- sapply(r2_lm, "[[", "r2")
  r2_lm <- rowMeans(r2_lm)
  rmse_lm <- sapply(rmse_lm, "[[", "rmse")
  rmse_lm <- rowMeans(rmse_lm)
  (lm_result <- data.frame(
    Model = paste0("Linear ", 1:5),
    Mean_R2 = r2_lm,
    Mean_RMSE = rmse_lm
  ))
  rm(r2_lm, rmse_lm)
  
  mods.gam <- replicate(10, list())
  r2_gam <- list()
  rmse_gam <- list()
  for (i in 1:10) {
    calibrate <- cal[[i]]
    testing <- test[[i]]
    r2_temp <- rep(0, 5)
    rmse_temp <- rep(0, 5)
    for (j in 1:5) {
      mod <- gam(data = calibrate, terms_gam[[j]])
      preds <- predict(mod, newdata = testing)
      rmse_temp[j] <- caret::RMSE(pred = preds, obs = testing[, "read_age"])
      RSS <- sum((testing$read_age - preds)^2)
      TSS <- sum((testing$read_age - mean(testing$read_age))^2)
      r2_temp[j] <- 1 - (RSS / TSS)
      mods.gam[[i]][[j]] <- mod
      rm(RSS, TSS, preds)
    }
    r2_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), r2 = r2_temp)
    rmse_gam[[i]] <- data.frame(Model = paste0("GAM ", 1:5), rmse = rmse_temp)
    rm(j, i, testing, calibrate, mod, r2_temp, rmse_temp)
  }
  r2_gam <- sapply(r2_gam, "[[", "r2")
  r2_gam <- rowMeans(r2_gam)
  rmse_gam <- sapply(rmse_gam, "[[", "rmse")
  rmse_gam <- rowMeans(rmse_gam)
  gam_result <- data.frame(
    Model = paste0("GAM ", 1:5),
    Mean_R2 = r2_gam,
    Mean_RMSE = rmse_gam
  )
  (results <- rbind(lm_result,gam_result))
  rm(r2_gam, rmse_gam, lm_result, gam_result)
  
  mods.pls <- list()
  mods.vip <- list()
  r2.pls <- list()
  rmse.pls <- list()
  for (i in 1:10) {
    calibrate <- df[-splits[[i]], ]
    testing <- df[splits[[i]], ]
    mod <- pls(calibrate[, 31:ncol(calibrate)], calibrate[, "read_age"],
               scale = F, center = T, cv = 1, #ncomp.selcrit = "wold", removing for now, seems too conservative with number of selected components
               info = "Age Prediction Model",
               x.test = testing[, 31:ncol(testing)],
               y.test = testing[, "read_age"]
    )
    ncomp <- mod$ncomp.selected
    rmse.pls$pls.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
    r2.pls$pls.test[i] <- mod$testres$r2[[ncomp]] # extract r2
    
    # VIP filter of wavenumbers
    mods.pls[[i]] <- mod
    mods.pls[[1]]$ncomp.selected
    vip <- as.data.frame(vipscores(mod))
    mod <- pls(calibrate[, 31:ncol(calibrate)], calibrate[, "read_age"],
               # ncomp.selcrit = "wold", 
               scale = F, center = T, cv = 1,
               info = "Age Prediction Model",
               x.test = testing[, 31:ncol(testing)],
               y.test = testing[, "read_age"],
               exclcols = vip$V1 < 0.5
    )
    ncomp <- mod$ncomp.selected
    r2.pls$vip.test[i] <- mod$testres$r2[[ncomp]] # extract r2
    mods.vip[[i]] <- mod
    rmse.pls$vip.test[i] <- mod$testres$rmse[[ncomp]] # extract rmse
    rm(calibrate, mod, ncomp, testing, i, vip)
  }
  (results <- rbind(results, 
                    c("PLS", mean(r2.pls$pls.test), mean(rmse.pls$pls.test)),
                    c("VIP", mean(r2.pls$vip.test), mean(rmse.pls$vip.test))
  ))
  rm(rmse.pls, r2.pls)
  mods.lm <<- mods.lm
  mods.gam <<- mods.gam
  mods.pls <<- mods.pls
  mods.vip <<- mods.vip
  return(results)
}

get.my.models <- function() {
  model.list <- list(lm = list(), gam = list(), pls = list(), vip = list())
  model.list$lm <- mods.lm
  model.list$gam <- mods.gam
  model.list$pls <- mods.pls
  model.list$vip <- mods.vip
  rm(mods.lm, mods.gam, mods.pls, mods.vip, pos = .GlobalEnv)
  return(model.list)
}
```

# ICP-MS Cutoff value attempts (based on mean)

```{r}
# Cutoff Data
```{r}
ICP_MS <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")
df <- readRDS("RDS_dataframes/ICP-MS_df.RDS")
df <- df %>% replace(is.na(.), 0)
cutoffs <- df %>% group_by(specimen,transect) %>% summarize(cutoff = mean(Ca43) * 0.9)

# Merge the two data frames
merged_df <- merge(df, cutoffs, by = c("specimen", "transect"))

# Filter the merged data frame using subset
filtered_df <- subset(merged_df, Ca43 > cutoff)
# saveRDS(filtered_df, "RDS_dataframes/ICP-MS_cutoffs_df.RDS")

# Mg
filtered_df %>% filter(specimen == "oto01") %>% ggplot() + 
  geom_line(aes(x = time, y = Mg24_ppm)) + 
  facet_wrap(~transect)
# Mn
filtered_df %>% filter(specimen == "oto01") %>% ggplot() + 
  geom_line(aes(x = time, y = Mn55_ppm)) + 
  facet_wrap(~transect)
# Sr
filtered_df %>% filter(specimen == "oto01") %>% ggplot() + 
  geom_line(aes(x = time, y = Sr86_ppm)) + 
  facet_wrap(~transect)
filtered_df %>% filter(specimen == "oto01") %>% ggplot() + 
  geom_line(aes(x = time, y = Sr87_ppm)) + 
  facet_wrap(~transect)
filtered_df %>% filter(specimen == "oto01") %>% ggplot() + 
  geom_line(aes(x = time, y = Sr88_ppm)) + 
  facet_wrap(~transect)
```


# ICP-MS 

### elementR attempts

### Write specimen sheets to folder for elementR

```{r}
# ICP_MS <- readRDS("RDS_dataframes/ICP-MS_rawlist.RDS")
library(writexl)
# specimen 1-23, starts at index 30
idx <- seq(from = 1, to = 88, by = 4)
for (i in idx){
  i = i + 29
  path <- names(ICP_MS[i])
  path <- str_sub(path, 1, str_length(path)-2)
  write_xlsx(ICP_MS[c(i:(i+3))],paste0("ICP-MS/elementR/samples/",path,".xlsx"))
}
# specimen 24 missing a third scan
write_xlsx(ICP_MS[c(118:120)],"ICP-MS/elementR/samples/oto24.xlsx")
# 25-40
idx <- seq(from = 92, to = 155, by = 4)
for (i in idx){
  i = i + 29
  path <- names(ICP_MS[i])
  path <- str_sub(path, 1, str_length(path)-2)
  write_xlsx(ICP_MS[c(i:(i+3))],paste0("ICP-MS/elementR/samples/",path,".xlsx"))
}

```

### Write reference materials to folder

```{r}
write_xlsx(ICP_MS[c(4:8)],"ICP-MS/elementR/standards/ATHO.xlsx")
write_xlsx(ICP_MS[c(9:13)],"ICP-MS/elementR/standards/BM90-21.xlsx")
write_xlsx(ICP_MS[c(14:18)],"ICP-MS/elementR/standards/G_NIST612.xlsx")
write_xlsx(ICP_MS[c(19:20)],"ICP-MS/elementR/standards/KL2.xlsx")
write_xlsx(ICP_MS[c(21:25)],"ICP-MS/elementR/standards/NIST612.xlsx")
write_xlsx(ICP_MS[c(26:29)],"ICP-MS/elementR/standards/T1.xlsx")
```

### Reference Data
```{r}
ICP_MS <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")
#

testing <- ICP_MS[[4]]
testing <- retype(testing)
testing$Specimen <- as.factor(testing$Specimen)
testing$Otolith <- as.factor(testing$Otolith)
# testing <- testing[37:268,]
testing %>% filter(Otolith == "oto01")

ggplot() +
  geom_histogram(data = testing %>% filter(Otolith == "oto07"),aes(Li7_ppm_LOD_Longerich)) + 
  geom_histogram(data = testing %>% filter(Otolith == "oto08"),aes(Li7_ppm_LOD_Longerich), fill = "red") + 
  geom_histogram(data = testing %>% filter(Otolith == "oto09"),aes(Li7_ppm_LOD_Longerich), fill = "blue")


ggplot(testing) +
  geom_histogram(aes(Li7_ppm_LOD_Longerich, fill = Otolith))

```

### some visualization stuff

```{r}
library(tidyverse)
ICP_MS <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")
testing <- ICP_MS[30:33]
testing$`oto01-0`$transect = as.factor(1)
testing$`oto01-1`$transect = as.factor(2)
testing$`oto01-2`$transect = as.factor(3)
testing$`oto01-3`$transect = as.factor(4)
testing3 <- bind_rows(testing$`oto01-0`,testing$`oto01-1`,testing$`oto01-2`,testing$`oto01-3`)

testing3 <- testing3 %>% rename(time = `Elapsed Time`)
naming <- names(testing3)
colnames(testing3) <- as.character(naming)
names(testing3)
ggplot(testing3) +
  geom_path(aes(x = time, y = Ca43))+
  facet_wrap(~transect)
```


### specimens data exploration

```{r}
ICP_MS <- readRDS("RDS_dataframes/ICP-MS_df.RDS")
ICP_MS <- ICP_MS %>% replace(is.na(.), 0)

test <- ICP_MS[[30]]

testing <- ICP_MS %>% filter(Ca43 > 1000000)
testing %>% group_by(transect) %>% filter(specimen == "oto01") %>% ggplot(aes(x = time, y = Mg24_ppm)) + 
  geom_line() + 
  facet_wrap(~transect, scale = "free")

testing_longer <- testing %>% pivot_longer(-c(time, specimen, transect), names_to = "element", values_to = "value")
testing_longer$element <- as.factor(testing_longer$element)
testing_longer$specimen <- as.factor(testing_longer$specimen)

testing_longer %>% group_by(transect) %>% filter(specimen == "oto01") %>% 
  ggplot() + 
  geom_line(aes(x = time, y = value)) + 
  facet_wrap(~element,scales = "free")


testing_longer %>% group_by(transect) %>% filter(specimen == "oto01") %>% ggplot() + 
  geom_tile(aes(x = time, y = transect, color = value)) + 
  scale_color_viridis_b()


testing <- ICP_MS %>% filter(specimen == "oto01")
ggplot(testing) + 
  geom_line(aes(x = time, y = Ca43)) + 
  geom_hline(yintercept = 1102300) + 
  facet_grid(~transect, scale = "free")

testing1 <- testing %>% filter(transect == 1)
mean(testing1$Ca43)



means <- testing %>% group_by(transect) %>% summarise(Ca43 = mean(Ca43))
themean <- as.numeric(means[1,2])
beginning <- testing %>% filter(time < 10)
end <- testing %>% filter(time > 100)
mod.beg <- lm(data = beginning, Ca43 ~ time)
data.beg <- data.frame(time = seq(1,10,0.01))
data.beg$preds <- round(predict(mod.beg,data.beg))
idx.beg <- which.min(abs(data.beg$preds - themean))
idx.beg <- data.beg[idx.beg,1]
mod.end <- lm(data = end, Ca43 ~ time)
data.end <- data.frame(time = seq(80,120,0.01))
data.end$preds <- round(predict(mod.end,data.end))
idx.end <- which.min(abs(data.end$preds - themean))
idx.end <- data.end[idx.end,1]

testing2 <- testing %>% filter(transect == 1, 
                               time > idx.beg,
                               time < idx.end)
ggplot(testing2) + 
  geom_line(aes(x = time, y = Ca43))

ggplot(testing2) + 
  geom_line(aes(x = time, y = Mn55_ppm))

testing2_longer <- testing2 %>% pivot_longer(-c(time, specimen, transect), names_to = "element", values_to = "value")

ggplot(testing2_longer) + 
  geom_line(aes(x = time, y = value)) + 
  facet_wrap(~element,scales = "free")


testing1 <- testing %>% filter(time < 10)

mod <- lm(data = testing1, Ca43 ~ time)


ggplot(testing1,aes(x = time, y = Ca43)) + 
  geom_line() + 
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 1102300)
```

# filter roughly using Cu65, create determine mean and medium values for each transect/specimen

```{r}
df <- readRDS("RDS_dataframes/ICP-MS_df.RDS")
df <- df %>% replace(is.na(.), 0)
df <- df %>% filter(Cu65_ppm <= 5, Cu65_ppm > 0) 
# split each transect into thirds
df <- df %>% 
  group_by(specimen, transect) %>%
  mutate(portion = cut(time, breaks = 3, labels = c("Start", "Middle", "End"), include.lowest = TRUE))
# mean and median for all transects/thirds
spec_stats <- df %>% group_by(specimen, transect, portion) %>% summarise(across(c(Ca43:Th232_ppm), list(mean = ~mean(.x, na.rm = TRUE), median = ~median(.x, na.rm = TRUE))))
# combine transects, summary statistics for each specimen by third
spec_stats <- spec_stats %>% group_by(specimen, portion) %>% 
  summarise(across(c(Ca43_mean:Th232_ppm_median), ~mean(.x)))

# ggplot look at splits for all specimens; focus on Mg, Mn, Zn, 
ggplot(spec_stats) +
  geom_point(aes(y = specimen, x = Mg24_ppm_median, color = portion))

ggplot(spec_stats) +
  geom_point(aes(y = specimen, x = Mn55_ppm_median, color = portion))

ggplot(spec_stats) +
  geom_point(aes(y = specimen, x = Zn66_ppm_median, color = portion))

ggplot(spec_stats) +
  geom_point(aes(y = specimen, x = Sr86_ppm_median, color = portion))

ggplot(spec_stats) +
  geom_point(aes(y = specimen, x = Sr87_ppm_median, color = portion))

ggplot(spec_stats) +
  geom_point(aes(y = specimen, x = Sr88_ppm_median, color = portion))

# tukey to test differences
anova_result <- aov(Mg24_ppm_median ~ portion, data = spec_stats)
TukeyHSD(anova_result)




```



# Load packages
Install and load necessary packages
```{r}
packages <- "tidyverse"
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages) > 0) {
  install.packages(new_packages)
}
invisible(lapply(packages, library, character.only = TRUE))
rm(packages, new_packages)
```

# ----------------------
# CORE SETUP
# ----------------------

```{r}
`%||%` <- function(a, b) if (!is.null(a)) a else b  # Null-coalescing operator
```

# ----------------------
# DATA PREPARATION
# ----------------------
```{r}
# Load and filter specimens
df <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")[30:184] |> 
  (\(x) x[!grepl("oto10-2|oto30-", names(x))])()

# Parse specimen metadata
df <- Map(\(specimen_df, name) {
  # Extract otolith number and transect ID (e.g., oto1-0 → 1 and 0)
  matches <- regmatches(name, regexec("oto(\\d+)-(\\d+)", name))[[1]]
  specimen_df$otolith_id <- as.integer(matches[2])
  specimen_df$transect_id <- as.integer(matches[3])
  
  # Assign regions based on established groupings
  specimen_df$region <- ifelse(
    specimen_df$otolith_id %in% c(6:10, 16:25, 31:35),
    "Kodiak", "Shumagin"
  )
  
  specimen_df
}, df, names(df)) |> setNames(names(df))
```

# ----------------------
# BOUNDARY DETECTION 
# ----------------------

```{r}
detect_otolith_segments <- function(ca43_data, window_size = 7) {
  if(length(ca43_data) < 100) return(list(start = 1, end = length(ca43_data)))
  
  smoothed <- tryCatch(
    stats::runmed(ca43_data, window_size),
    error = function(e) ca43_data
  )
  n <- length(smoothed)
  
  baseline_median <- median(smoothed[round(n*0.3):round(n*0.7)], na.rm = TRUE)
  
  # Improved start detection
  start_idx <- detect_primary_start(smoothed, baseline_median) %||%
               detect_fallback_start(smoothed) %||% 
               round(n*0.1)
  
  # Robust end detection with fallback
  end_idx <- detect_terminal_end(smoothed, start_idx) %||% 
             detect_terminal_end_fallback(smoothed) %||%
             n
  
  # Ensure valid indices
  start_idx <- max(1, as.integer(start_idx))
  end_idx <- min(n, as.integer(end_idx))
  
  list(start = start_idx, end = end_idx)
}

# Helper functions for boundary detection
detect_primary_start <- function(smoothed, baseline) {
  threshold <- baseline * 0.4
  for(i in seq(7, length(smoothed)-5)) {
    if(all(smoothed[i:(i+4)] > threshold) && 
       mean(smoothed[i:(i+4)]) > mean(smoothed[pmax(1, i-10):i])) {
      return(i)
    }
  }
  NULL
}

detect_fallback_start <- function(smoothed) {
  which(smoothed > median(smoothed) * 0.6)[1]
}

detect_terminal_end <- function(smoothed, start) {
  n <- length(smoothed)
  tail_window <- max(start+50, n-100):n
  drop_point <- which(smoothed[tail_window] < median(smoothed)*0.3)[1]
  if(!is.na(drop_point)) tail_window[drop_point] else n
}

detect_terminal_end_fallback <- function(smoothed) {
  # Find last point above 25th percentile
  cutoff <- quantile(smoothed, 0.25)
  tail(which(smoothed > cutoff), 1)
}
```

# ----------------------
# STATISTICAL COMPARISONS
# ----------------------
### 1. Transect-level Comparisons (Within Otolith)

```{r}
compare_transects <- function(processed_data) {
  # Get all element columns (ending with _ppm)
  all_names <- names(processed_data[[1]])
  elements <- grep("_ppm$", all_names, value = TRUE)
  
  # Remove metadata columns explicitly
  elements <- setdiff(elements, c("otolith_id", "transect_id", "region"))
  
  # Rest of the function remains the same...
  otolith_ids <- unique(sapply(processed_data, \(df) df$otolith_id[1]))
  
  p_matrix <- matrix(nrow = length(otolith_ids), 
                     ncol = length(elements),
                     dimnames = list(otolith_ids, elements))
  
  for(elem in elements) {
    for(oto in otolith_ids) {
      transect_list <- processed_data[grep(paste0("^oto", oto, "-"), names(processed_data))]
      
      if(length(transect_list) < 2) next
      
      transect_data <- bind_rows(lapply(transect_list, \(df) {
        df |> select(transect_id, value = all_of(elem))
      }))
      
      if(n_distinct(transect_data$transect_id) > 1) {
        p_matrix[as.character(oto), elem] <- kruskal.test(
          value ~ transect_id, data = transect_data
        )$p.value
      }
    }
  }
  
  list(
    p_values = p_matrix,
    significant = apply(p_matrix, 2, \(col) {
      rownames(p_matrix)[which(col < 0.05 & !is.na(col))]
    })
  )
}
```
### 2. Otolith-level Comparisons (Between Otoliths)  
```{r}
compare_otoliths <- function(processed_data) {
  elements <- c("Mg24_ppm", "Sr86_ppm", "Sr87_ppm", "Sr88_ppm",
                "Ba138_ppm", "Mn55_ppm", "Li7_ppm")
  
  lapply(elements, function(elem) {
    # Aggregate data by otolith
    oto_data <- bind_rows(lapply(processed_data, \(df) {
      df |> group_by(otolith_id) |> 
        summarise(median_value = median(.data[[elem]], na.rm = TRUE))
    }))
    
    # Kruskal-Wallis test across otoliths
    list(
      element = elem,
      test = kruskal.test(median_value ~ as.factor(otolith_id), data = oto_data),
      otolith_medians = oto_data
    )
  }) |> setNames(elements)
}
```
### 3. Region-level Comparisons (Shumagin vs Kodiak)
```{r}

compare_regions <- function(processed_data) {
  elements <- c("Mg24_ppm", "Sr86_ppm", "Sr87_ppm", "Sr88_ppm",
                "Ba138_ppm", "Mn55_ppm", "Li7_ppm")
  
  lapply(elements, function(elem) {
    # Aggregate data by region
    region_data <- bind_rows(lapply(processed_data, \(df) {
      df |> group_by(region) |> 
        summarise(median_value = median(.data[[elem]], na.rm = TRUE))
    }))
    
    # Mann-Whitney U test between regions
    list(
      element = elem,
      test = wilcox.test(median_value ~ region, data = region_data),
      region_medians = region_data
    )
  }) |> setNames(elements)
}
```

# ----------------------
# VISUALIZATION
# ----------------------
### 3. Compare transects
```{r}
plot_transect_variability <- function(processed_data, element) {
  library(ggplot2)
  
  bind_rows(lapply(processed_data, \(df) {
    df |> select(otolith_id, transect_id, value = all_of(element))
  })) |>
    ggplot(aes(x = factor(transect_id), y = value)) +
    geom_boxplot() +
    facet_wrap(~otolith_id, scales = "free_x") +
    labs(title = paste("Transect Variability -", element),
         x = "Transect ID", y = "Concentration (ppm)") +
    theme_minimal()
}
```
### 2. Compare otoliths
```{r}
plot_otolith_comparison <- function(stats, element) {
  library(ggplot2)
  
  stats$otolith_comparison[[element]]$otolith_medians |>
    ggplot(aes(x = factor(otolith_id), y = median_value)) +
    geom_boxplot() +
    labs(title = paste("Otolith Comparison -", element),
         x = "Otolith ID", y = "Median Concentration (ppm)") +
    theme_minimal()
}
```
### 3. Compare regions
```{r}
plot_regional_differences <- function(stats, element) {
  library(ggplot2)
  
  stats$region_comparison[[element]]$region_medians |>
    ggplot(aes(x = region, y = median_value, fill = region)) +
    geom_boxplot() +
    labs(title = paste("Regional Differences -", element),
         x = "", y = "Median Concentration (ppm)") +
    theme_minimal() +
    theme(legend.position = "none")
}

```

# ----------------------
# ANALYSIS WORKFLOW
# ----------------------
```{r}
analyze_otoliths <- function(raw_data) {
  # 1. Process boundaries
  processed <- lapply(raw_data, \(df) {
    bounds <- detect_otolith_segments(df$Ca43)
    segmented <- df[bounds$start:bounds$end, ]
    attr(segmented, "boundaries") <- bounds  # Store boundaries as attribute
    segmented
  })
  
  # 2. Perform comparisons
  stats <- list(
    transect_comparison = compare_transects(processed),
    otolith_comparison = compare_otoliths(processed),
    region_comparison = compare_regions(processed)
  )
  
  # 3. Generate summary statistics
  summary_stats <- lapply(names(processed[[1]]), function(col) {
    if(col %in% c("Mg24_ppm", "Sr86_ppm", "Sr87_ppm", "Sr88_ppm",
                  "Ba138_ppm", "Mn55_ppm", "Li7_ppm")) {
      vals <- na.omit(unlist(lapply(processed, \(df) df[[col]])))
      list(
        mean = mean(vals),
        median = median(vals),
        sd = sd(vals),
        n = length(vals)
      )
    }
  }) |> setNames(names(processed[[1]]))
  
  list(
    processed_data = processed,
    statistics = stats,
    summary_stats = summary_stats
  )
}

```

# ----------------------
# EXECUTION & REPORTING
# ----------------------

```{r}
results <- analyze_otoliths(df)

# View regional results for Magnesium
print(results$statistics$region_comparison$Mg24_ppm$test)
plot_regional_differences(results$statistics, "Mg24_ppm")

# Check transect variability in Strontium-86
sig_otos <- names(which(!sapply(results$statistics$transect_comparison$Sr86_ppm$transect_tests, is.null)))
print(paste("Otoliths with significant transect differences (Sr86):", paste(sig_otos, collapse = ", ")))
plot_transect_variability(results$processed_data, "Sr86_ppm")

# Compare individual otoliths for Barium-138
print(results$statistics$otolith_comparison$Ba138_ppm$test)
plot_otolith_comparison(results$statistics, "Ba138_ppm")

# Save full results
saveRDS(results, "comprehensive_analysis_results.rds")
```


```{r}
# ----------------------
# ANNOTATED CODE BREAKDOWN
# ----------------------

### 1. CORE SETUP ###
# Null coalescing operator for safe default values
`%||%` <- function(a, b) if (!is.null(a)) a else b

### 2. DATA PREPARATION ###
# - Loads filtered specimen data
# - Extracts metadata from specimen names
# - Assigns regions based on otolith ID ranges

### 3. BOUNDARY DETECTION ###
# Identifies meaningful transect segments using:
# - Running median smoothing
# - Baseline threshold calculation
# - Fallback detection mechanisms

### 4. STATISTICAL COMPARISONS ###
# Three comparison levels:
# 1. Within-otolith transects (Kruskal-Wallis)
# 2. Between-otoliths (Kruskal-Wallis)
# 3. Regional differences (Mann-Whitney U)

### 5. VISUALIZATION ###
# Generates diagnostic plots for each comparison level

### 6. ANALYSIS WORKFLOW ###
# Coordinates data processing -> analysis -> visualization

# ----------------------
# VALIDATION TEST SUITE
# ----------------------

validate_boundaries <- function(processed_data) {
  lapply(processed_data, function(df) {
    bounds <- attr(df, "boundaries")
    data.frame(
      start = bounds$start,
      end = bounds$end,
      length = bounds$end - bounds$start,
      valid = bounds$start < bounds$end
    )
  }) |> bind_rows(.id = "specimen")
}



validate_results <- function(results) {
  # Test 1: Structure Check
  required_components <- c("processed_data", "statistics", "summary_stats")
  if(!all(required_components %in% names(results))) {
    stop("Missing critical result components")
  }
  
  # Test 2: Transect Comparison Sanity Check
  transect_pvals <- results$statistics$transect_comparison$p_values
  if(any(transect_pvals < 0, na.rm = TRUE) | any(transect_pvals > 1, na.rm = TRUE)) {
    warning("Invalid p-values in transect comparisons")
  }
  
  # Test 3: Metadata Preservation
  first_specimen <- results$processed_data[[1]]
  if(!all(c("otolith_id", "transect_id", "region") %in% names(first_specimen))) {
    stop("Metadata not preserved in processed data")
  }
  
  # Test 4: Visualization Capability
  tryCatch({
    print(plot_regional_differences(results$statistics, "Mg24_ppm"))
    print(plot_transect_variability(results$processed_data, "Sr86_ppm"))
  }, error = function(e) {
    warning("Visualization generation failed: ", e$message)
  })
  
  message("Validation complete. No critical errors found.")
}

element_sanity_check <- function(results, element = "Mg24_ppm") {
  # Check all analysis levels for one element
  list_safe <- function(x) tryCatch(x, error = function(e) NULL)
  
  list(
    transect_pvals = list_safe(results$statistics$transect_comparison$p_values[,element]),
    otolith_test = list_safe(results$statistics$otolith_comparison[[element]]$test),
    region_test = list_safe(results$statistics$region_comparison[[element]]$test),
    summary_stats = list_safe(results$summary_stats[[element]])
  )
}

# ----------------------
# EXECUTION VALIDATION
# ----------------------
# Run after executing results <- analyze_otoliths(df)

# 1. Structural validation
validate_results(results)

# 2. Element-level validation
mg_check <- element_sanity_check(results, "Mg24_ppm")
str(mg_check)

# 3. Statistical output check
cat("\nRegional Comparison (Mg24):\n")
print(results$statistics$region_comparison$Mg24_ppm$test)

# 4. Data completeness check
cat("\nProcessed data dimensions:\n")
print(lapply(results$processed_data[1:3], dim))

# 5. Boundary detection verification
cat("\nBoundary detection examples:\n")
lapply(results$processed_data[1:3], function(df) {
  c(start = attr(df, "boundaries")$start, 
    end = attr(df, "boundaries")$end)
}) |> str()

# ----------------------
# INTERPRETATION GUIDE
# ----------------------

### Transect Comparisons ###
# - Significant p-values (<0.05) indicate differences between transects
# - Expect most non-significant results (biological assumption)

### Otolith Comparisons ###
# - Significant p-values suggest elemental composition differences
# - Could indicate environmental/habitat variations

### Regional Comparisons ###
# - Significant differences support geographic population structure
# - Consider effect sizes along with p-values

# ----------------------
# TROUBLESHOOTING CHECKS
# ----------------------

# 1. Check element names match your data
cat("\nAvailable elements in first specimen:\n")
names(results$processed_data[[1]]) |> 
  grep("_ppm", value = TRUE, x = _) |> 
  print()

# 2. Verify region assignments
cat("\nRegion distribution:\n")
table(sapply(results$processed_data, \(df) unique(df$region)))

# 3. Check transect counts per otolith
transect_counts <- sapply(
  names(results$processed_data),
  \(x) sub(".*-(\\d+)", "\\1", x)
) |> table()
cat("\nTransects per otolith:\n")
print(transect_counts)

plot_segment_comparison <- function(specimen) {
  # Get data
  original <- df[[specimen]]$Ca43
  processed <- results$processed_data[[specimen]]$Ca43
  bounds <- attr(results$processed_data[[specimen]], "boundaries")
  
  if(is.null(bounds)) {
    warning("No boundaries found for ", specimen)
    return()
  }
  
  # Create plot layout
  par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
  
  # 1. Top Plot: Full Original Signal
  plot(original, type = "l", col = "gray60",
       main = paste("Original Signal -", specimen),
       xlab = "", ylab = "Ca43",
       xlim = c(1, length(original)))
  abline(v = bounds$start, col = "blue", lwd = 1.5)
  abline(v = bounds$end, col = "red", lwd = 1.5)
  
  # 2. Bottom Plot: Focused Processed Segment
  plot(bounds$start:bounds$end, processed, type = "l",
       col = "darkgreen", lwd = 1.5,
       main = paste("Processed Segment -", specimen),
       xlab = "Original Measurement Index", ylab = "Ca43",
       xlim = c(bounds$start, bounds$end))  # Critical adjustment
  
  # Add boundary markers
  points(x = bounds$start, y = processed[1], 
         pch = 17, col = "blue", cex = 1.2)
  points(x = bounds$end, y = tail(processed, 1), 
         pch = 17, col = "red", cex = 1.2)
  
  par(mfrow = c(1, 1))  # Reset layout
}

# Plot first 3 specimens
lapply(names(results$processed_data)[1:150], plot_segment_comparison)

# Check boundary storage for first specimen
cat("\nBoundary attributes for first specimen:\n")
str(attr(results$processed_data[[1]], "boundaries"))

# Verify Ca43 column exists
cat("\nColumns in processed data:\n")
names(results$processed_data[[1]])

# Test plotting with error handling
try_plot <- function(specimen) {
  tryCatch({
    plot_segment(specimen)
    TRUE
  }, error = function(e) {
    message("Plot failed for ", specimen, ": ", e$message)
    FALSE
  })
}

# Test first 3 specimens
sapply(1:3, try_plot)





# 1. Re-run analysis
results <- analyze_otoliths(df)

# 2. Validate all boundaries
boundary_report <- validate_boundaries(results$processed_data)
print(boundary_report)

# 3. Plot problematic specimens
problem_specimens <- boundary_report$specimen[!boundary_report$valid]
if(length(problem_specimens) > 0) {
  cat("\nPlotting problematic specimens:\n")
  lapply(problem_specimens, plot_segment)
}

# 4. Plot first 3 specimens with boundaries
lapply(names(results$processed_data)[1:3], plot_segment)





```



NEW CODE AGAIN

```{r}
# Improved ICP-MS Data Analysis Script
# This script processes and analyzes ICP-MS laser ablation trace element data
# from juvenile fish otolith specimens.

# Load required libraries
library(tidyverse)
library(plotly)
library(stats)

# 1. Data Import and Preprocessing
# --------------------------------

# Load the RDS file, excluding reference materials and known outliers
load_and_clean_data <- function(file_path) {
  data <- readRDS(file_path)
  
  # Select indices 30-184, excluding outliers
  valid_indices <- setdiff(30:184, c(grep("oto10-2", names(data)), grep("oto30-", names(data))))
  
  cleaned_data <- data[valid_indices] %>%
    bind_rows(.id = "transect_id") %>%
    mutate(
      otolith_id = str_extract(transect_id, "oto\\d+"),
      transect_num = as.integer(str_extract(transect_id, "\\d+$")),
      region = case_when(
        as.integer(str_extract(otolith_id, "\\d+")) %in% c(6:10, 16:25, 31:35) ~ "Kodiak",
        TRUE ~ "Shumagin"
      )
    )
  
  return(cleaned_data)
}

# Load the data
df <- load_and_clean_data("/Users/zachstamplis/Desktop/Thesis and Otoliths/Github/LPW_FT-NIRS/RDS_dataframes/ICP-MS_clean.RDS")

# 2. Improved Boundary Detection
# ------------------------------

detect_otolith_segments <- function(ca43_data, window_size = 7, threshold_factor = 0.4) {
  # Apply rolling median to smooth the data
  smoothed <- zoo::rollmedian(ca43_data, k = window_size, fill = "extend")
  
  # Calculate baseline using the middle 40% of the data
  n <- length(smoothed)
  baseline_range <- round(n * 0.3):round(n * 0.7)
  baseline <- median(smoothed[baseline_range])
  
  # Detect start and end points
  threshold <- baseline * threshold_factor
  start_idx <- which(smoothed > threshold)[1]
  end_idx <- n - which(rev(smoothed) > threshold)[1] + 1
  
  # Ensure we have valid indices
  start_idx <- max(1, start_idx)
  end_idx <- min(n, end_idx)
  
  return(list(start = start_idx, end = end_idx))
}

# Apply boundary detection to all transects
df_segmented <- df %>%
  group_by(transect_id) %>%
  mutate(
    boundaries = list(detect_otolith_segments(Ca43)),
    start = map_dbl(boundaries, ~ .x$start),
    end = map_dbl(boundaries, ~ .x$end),
    row_num = row_number()
  ) %>%
  filter(row_num >= start & row_num <= end) %>%
  select(-boundaries, -row_num) %>%
  ungroup()

# 3. Statistical Analysis
# -----------------------
library(tidyr)
library(dplyr)
library(purrr)

# Helper function to safely run tests
safe_test <- function(test_func, ...) {
  tryCatch(
    test_func(...),
    error = function(e) {
      warning(paste("Error in test:", e$message))
      return(data.frame(p.value = NA, statistic = NA, parameter = NA))
    }
  )
}

# Compare transects within otoliths for all elements
compare_transects <- function(data) {
  id_cols <- c("otolith_id", "transect_id", "region", "Elapsed Time", "start", "end", "transect_num")
  
  data %>%
    pivot_longer(cols = -all_of(id_cols),
                 names_to = "element",
                 values_to = "value") %>%
    group_by(otolith_id, element) %>%
    summarise(
      n_transects = n_distinct(transect_id),
      test_result = if(n_transects > 1) {
        list(safe_test(kruskal.test, value ~ factor(transect_id)))
      } else {
        list(data.frame(p.value = NA, statistic = NA, parameter = NA))
      },
      .groups = "drop"
    ) %>%
    mutate(
      p_value = map_dbl(test_result, ~ .x$p.value),
      significant = !is.na(p_value) & p_value < 0.05
    ) %>%
    select(otolith_id, element, n_transects, p_value, significant)
}

# Compare otoliths for all elements
compare_otoliths <- function(data) {
  id_cols <- c("otolith_id", "transect_id", "region", "Elapsed Time", "start", "end", "transect_num")
  
  data %>%
    pivot_longer(cols = -all_of(id_cols),
                 names_to = "element",
                 values_to = "value") %>%
    group_by(element) %>%
    summarise(
      test_result = list(safe_test(kruskal.test, value ~ factor(otolith_id))),
      .groups = "drop"
    ) %>%
    mutate(
      p_value = map_dbl(test_result, ~ .x$p.value),
      statistic = map_dbl(test_result, ~ .x$statistic),
      df = map_dbl(test_result, ~ .x$parameter)
    ) %>%
    select(element, p_value, statistic, df)
}

# Compare regions for all elements
compare_regions <- function(data) {
  id_cols <- c("otolith_id", "transect_id", "region", "Elapsed Time", "start", "end", "transect_num")
  
  data %>%
    pivot_longer(cols = -all_of(id_cols),
                 names_to = "element",
                 values_to = "value") %>%
    group_by(element) %>%
    summarise(
      test_result = list(safe_test(wilcox.test, value ~ factor(region))),
      .groups = "drop"
    ) %>%
    mutate(
      p_value = map_dbl(test_result, ~ .x$p.value),
      statistic = map_dbl(test_result, ~ .x$statistic),
      method = map_chr(test_result, ~ .x$method)
    ) %>%
    select(element, p_value, statistic, method)
}

# Perform all analyses
analysis_results <- list(
  transect_comparison = compare_transects(df_segmented),
  otolith_comparison = compare_otoliths(df_segmented),
  region_comparison = compare_regions(df_segmented)
)
# 4. Visualization
# ----------------

library(ggplot2)

plot_transect_comparison <- function(data) {
  ggplot(data, aes(x = element, y = -log10(p_value), color = significant)) +
    geom_point() +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
    facet_wrap(~ otolith_id, scales = "free_y") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = "Transect Comparison within Otoliths",
         x = "Element",
         y = "-log10(p-value)",
         color = "Significant")
}

transect_plot <- plot_transect_comparison(analysis_results$transect_comparison)
print(transect_plot)


plot_otolith_comparison <- function(data) {
  ggplot(data, aes(x = element, y = -log10(p_value))) +
    geom_col() +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = "Otolith Comparison",
         x = "Element",
         y = "-log10(p-value)")
}

otolith_plot <- plot_otolith_comparison(analysis_results$otolith_comparison)
print(otolith_plot)

plot_region_comparison <- function(data) {
  ggplot(data, aes(x = element, y = -log10(p_value))) +
    geom_col() +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = "Region Comparison",
         x = "Element",
         y = "-log10(p-value)")
}

region_plot <- plot_region_comparison(analysis_results$region_comparison)
print(region_plot)

transect_plot <- plot_transect_comparison(analysis_results$transect_comparison)
otolith_plot <- plot_otolith_comparison(analysis_results$otolith_comparison)
region_plot <- plot_region_comparison(analysis_results$region_comparison)

# Print or save the plots as needed
print(transect_plot)
print(otolith_plot)
print(region_plot)

# If you want to save the plots
ggsave("transect_comparison.png", transect_plot, width = 12, height = 8)
ggsave("otolith_comparison.png", otolith_plot, width = 10, height = 6)
ggsave("region_comparison.png", region_plot, width = 10, height = 6)

# 5. Results and Output
# ---------------------

# Print summary of analysis results
print_analysis_summary <- function(results) {
  for (element in names(results)) {
    cat("\nResults for", element, ":\n")
    cat("Transects with significant differences:\n")
    print(results[[element]]$transect_comparison %>% filter(significant))
    cat("\nOtolith comparison p-value:", results[[element]]$otolith_comparison$p.value, "\n")
    cat("Region comparison p-value:", results[[element]]$region_comparison$p.value, "\n")
  }
}

print_analysis_summary(analysis_results)

# Save plots
for (element in elements_of_interest) {
  ggsave(paste0(element, "_distribution.png"), element_plots[[element]]$distribution, width = 12, height = 8)
  ggsave(paste0(element, "_regional.png"), element_plots[[element]]$regional, width = 8, height = 6)
}




library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(viridis)

# Function to remove outliers for plotting purposes
remove_outliers <- function(x, na.rm = TRUE, ...) {
  qnt <- quantile(x, probs=c(.25, .75), na.rm = na.rm, ...)
  H <- 1.5 * IQR(x, na.rm = na.rm)
  y <- x
  y[x < (qnt[1] - H)] <- NA
  y[x > (qnt[2] + H)] <- NA
  y
}

# Function to plot boxplots for a specific element with outlier handling
plot_element_boxplot <- function(data, element_name) {
  p <- data %>%
    mutate(!!element_name := remove_outliers(.data[[element_name]])) %>%
    ggplot(aes(x = otolith_id, y = .data[[element_name]], fill = otolith_id)) +
    geom_boxplot(outlier.alpha = 0.5) +
    scale_fill_viridis_d() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.position = "none") +
    labs(title = paste("Distribution of", element_name), x = "Otolith ID", y = element_name)
  
  print(p)
}

# Function to plot stacked histograms for regional comparisons
plot_region_histogram <- function(data, element_name) {
  p <- data %>%
    mutate(!!element_name := remove_outliers(.data[[element_name]])) %>%
    ggplot(aes(x = .data[[element_name]], fill = region)) +
    geom_histogram(position = "stack", bins = 30, alpha = 0.7) +
    scale_fill_viridis_d() +
    theme_minimal() +
    labs(title = paste("Distribution of", element_name, "by Region"), x = element_name, y = "Count")
  
  print(p)
}

# Function to plot element concentrations across the otolith transect
plot_element_transect <- function(data, element_name) {
  p <- data %>%
    mutate(!!element_name := remove_outliers(.data[[element_name]])) %>%
    ggplot(aes(x = `Elapsed Time`, y = .data[[element_name]], color = transect_id)) +
    geom_line(alpha = 0.7) +
    facet_wrap(~otolith_id, scales = "free_y") +
    scale_color_viridis_d() +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = paste(element_name, "Concentration Across Transects"),
         x = "Elapsed Time", y = element_name)
  
  print(p)
}

# List of elements to plot (adjust based on your data)
elements_to_plot <- c("Sr88_ppm", "Ba138_ppm", "Mg24_ppm", "Mn55_ppm")

# Create plots
for (element in elements_to_plot) {
  # Boxplot
  plot_element_boxplot(df_segmented, element)
  
  # Stacked histogram by region
  plot_region_histogram(df_segmented, element)
  
  # Transect plot
  plot_element_transect(df_segmented, element)
}
```




