Load Packages/Data

```{r}
# =============================================================================
#
# FT-NIRS AGE PREDICTION - Results
#
# This script plots and analyzes results from my FT-NIRS models.  I have 5 distinct sets of models:
#   1. All regions (LPW, Shumagin & Kodiak)
#   2. Wild Regions (Shumagin & Kodiak)
#   3. LPW only
#   4. Shumagin only
#   5. Kodiak only 
#
# Each set of models uses a repeat 10-fold CV approach.  If models are using multiple regions of specimens,
# 10-fold CV splits are calculated for each region seperately with createFolds() in caret and `read_age`
# as the variable to try and make fair splits for.  Each of the sets uses the following model types:
#
#   1. PCA-based models (best 10 Linear & best 10 GAM, using the first 10 PC's)
#   2. Partial Least Squares & PLs models with VIP filtering < 0.5
#   3. XGBoost
#   4. Random Forest
#   5. Simple biological models (Length + Structure Weight)
#
#
# =============================================================================

# =============================================================================
# SECTION 1: SETUP & DATA LOADING
# =============================================================================

# --- Load Required Packages ---
library(pacman)
p_load(
  arrow,
  caret,        # Machine learning utilities & preprocessing
  cowplot,      # Plot arrangements
  data.table,
  doParallel,   # Parallel backend
  doSNOW,       # Parallel progress tracking
  furrr,        # Parallel mapping with future
  future,       # Parallel computing framework
  gridExtra,    # Additional plot utilities
  mdatools,     # PLS modeling
  mgcv,         # GAM modeling
  MuMIn,        # Model selection (dredge)
  parallel,     # Core parallel utilities
  patchwork,    # Plot composition
  prospectr,    # Spectral preprocessing (Savitzky-Golay)
  purrr,        # Functional programming tools
  ranger,       # Random Forest implementation
  readxl,       # Excel file reading
  stringr,      # String manipulation
  tidyr,        # Data tidying
  tidyverse,    # Core tidyverse packages
  viridis,      # Color scales
  xgboost       # XGBoost modeling
)

# --- Load and Prepare Raw Data ---
df_raw <- readRDS("RDS_dataframes/combined_IBM_LPW_raw.RDS") %>% 
  filter(complete.cases(read_age))

# Consolidate regional groupings
df_raw <- df_raw %>%
  mutate(
    region = case_when(
      region %in% c("E Shumagins", "W Shumagins") ~ "Shumagin",
      region %in% c("N Kodiak", "S Kodiak") ~ "Kodiak",
      TRUE ~ region
    )
  )

# Create region-specific datasets
kodiak_raw <- df_raw %>% filter(region == "Kodiak")
shumagin_raw <- df_raw %>% filter(region == "Shumagin")
lpw_raw <- df_raw %>% filter(region == "LPW")
```



```{r}

# =============================================================================
# SECTION 2: UTILITY FUNCTIONS
# =============================================================================

# --- 2.1: Data Manipulation Functions ---

#' Find spectral column names
#' 
#' Identifies columns containing wavenumber data (numeric column names)
#' @param df Dataframe containing spectral data
#' @return Character vector of spectral column names
find_cols <- function(df) {
  names(df)[grepl("^\\d", names(df))]
}

#' Convert spectral data to long format
#' 
#' Transforms wide-format spectral data into long format for visualization
#' @param df Dataframe with spectral columns
#' @return Long-format dataframe with 'name' (wavenumber) and 'value' columns
make_long <- function(df) {
  speccols <- find_cols(df)
  df_long <- df %>%
    tidyr::pivot_longer(cols = speccols) %>%
    mutate(name = as.numeric(as.character(name)))
  return(df_long)
}

# --- 2.3: Spectral Preprocessing Functions ---

#' Apply Savitzky-Golay smoothing filter
#' 
#' @param df Dataframe with spectral columns
#' @param m Derivative order (0 = smoothing only, 1 = first derivative, etc.)
#' @param p Polynomial order
#' @param w Window size (must be odd)
#' @return Dataframe with filtered spectral data
sg_filter <- function(df, m, p, w) {
  speccols <- find_cols(df)
  metacols <- setdiff(names(df), speccols)
  sg_matrix <- prospectr::savitzkyGolay(df[, speccols], m = m, p = p, w = w)
  new_speccols <- find_cols(sg_matrix)
  names(sg_matrix) <- new_speccols
  return(dplyr::bind_cols(df[, metacols], as.data.frame(sg_matrix)))
}

#' Apply Standard Normal Variate (SNV) transformation
#' 
#' Centers and scales each spectrum to correct for scatter effects
#' @param df Dataframe with spectral columns
#' @return Dataframe with SNV-transformed spectral data
snv_filter <- function(df) {
  speccols <- find_cols(df)
  metacols <- setdiff(names(df), speccols)
  snv_matrix <- prospectr::prep.snv(as.matrix(df[, speccols]))
  return(dplyr::bind_cols(df[, metacols], as.data.frame(snv_matrix)))
}

#' Filter spectral data by wavelength threshold
#' 
#' Removes wavenumbers above a specified threshold
#' @param df Dataframe with spectral columns
#' @param wave Maximum wavenumber to retain
#' @return Dataframe with filtered wavelength range
wave_filter <- function(df, wave) {
  speccols <- find_cols(df)
  metacols <- setdiff(names(df), speccols)
  keep_me <- as.numeric(speccols) <= wave
  speccols_keep <- speccols[keep_me]
  return(df[, c(metacols, speccols_keep)])
}

# --- 2.7: Visualization Functions ---

#' Create spectral plot
#' 
#' Generates a line plot of absorbance vs wavenumber for multiple specimens
#' @param df Dataframe with spectral data
#' @param color Column name (unquoted) to use for coloring lines
#' @return ggplot object
spec.fig <- function(df, color) {
  mydf <- make_long(df)
  color_col <- rlang::enquo(color) 
  
  mydf_sorted <- mydf %>% 
    dplyr::arrange(specimen, name) 
  
  ggplot(mydf_sorted) +
    geom_path(aes(x = name, y = value, color = !!color_col, group = specimen)) +
    scale_x_reverse() +
    scale_color_viridis() +
    labs(y = "Absorbance", x = expression(paste("Wavenumber ", cm^-1))) + 
    theme_bw(base_size = 16)
}
```

```{r}
# --- : Preprocess Data ---
cli::cli_h1("Preprocessing Data")

kodiak_proc <- sg_filter(kodiak_raw, 1, 3, 19)
shumagin_proc <- sg_filter(shumagin_raw, 1, 3, 19)
lpw_proc <- sg_filter(lpw_raw, 1, 3, 19)
wild_proc <- bind_rows(kodiak_proc, shumagin_proc)
all_proc <- bind_rows(kodiak_proc, shumagin_proc, lpw_proc)

```

# density plot - ages for each region

Trying to demonstrate that specimens are wildly different ages so combined models likely don't make sense.

```{r}
density_plot <- ggplot(all_proc, aes(x = read_age, fill = region)) + 
  geom_density(alpha = 0.6) +
  scale_fill_viridis(discrete = T) + 
  theme_bw(base_size = 16) + 
    labs(
    x = "Thin Section Age (Days)",
    y = "Proportion",
    fill = "Region"
  )
density_plot
# ggsave("Chapter 2/age_comparisons_density.png", density_plot, width = 9, height = 6, dpi = 300)
```

I want to make age prediction scatterplots for each region from region-only, ALL and wild (for kod/shu)

```{r}
# =============================================================================
#
# THESIS FIGURE: MODEL RUN COMPARISON (BEST MODELS ONLY)
#
# =============================================================================

library(arrow)
library(dplyr)
library(ggplot2)
library(stringr)

# --- 1. CONFIGURATION ---

# Set the base path to your results
BASE_PATH <- "C:/Users/poopm/Thesis Writing/Github/LPW_FT-NIRS/Chapter 2/temp_arrow_files"

# Define the paths for each of your 5 runs
# (Paste the full folder names from your list)
RUN_PATHS <- list(
  All = file.path(BASE_PATH, "all_20251112_083140"),
  Wild = file.path(BASE_PATH, "wild_20251112_124214"),
  Kodiak = file.path(BASE_PATH, "kod_20251112_145212"),
  Shumagin = file.path(BASE_PATH, "shu_20251112_161318"),
  LPW = file.path(BASE_PATH, "LPW_20251113_144252")
)

# Define the same color palette from your other plots
COLOR_PALETTE <- c(
  "PCA" = "#4477AA",
  "PLS" = "#AA3377",
  "ML" = "#228833",
  "Simple" = "#CCBB44"
)

# --- 2. HELPER FUNCTION (MEMORY-SAFE LOADER) ---

#' Loads and summarizes predictions from a single model run
#'
#' This function opens the Arrow dataset *lazily*, calculates the
#' median prediction for each specimen/model, and *then* collects
#' the small, summarized dataframe into memory.
#'
#' @param run_path Path to the model run's root folder (e.g., ".../all_...")
#' @param run_name A short name to identify the run (e.g., "All")
#' @return A summarized tibble with median predictions
#'
load_and_summarize_preds <- function(run_path, run_name) {
  message(paste("Processing run:", run_name))
  
  pred_path <- file.path(run_path, "predictions")
  
  # 1. Open dataset lazily (uses 0 RAM)
  lazy_preds <- open_dataset(pred_path, format = "feather")
  
  # 2. Define the lazy summary
  summarized_data_lazy <- lazy_preds %>%
    group_by(specimen, model_name, model_type, obs_age) %>%
    summarise(median_pred_age = median(pred_age, na.rm = TRUE), .groups = "drop")
  
  # 3. Collect() the SMALL summarized data (low RAM)
  summarized_data <- collect(summarized_data_lazy)
  
  # 4. Add the 'model_run' label and return
  summarized_data %>%
    mutate(model_run = run_name)
}

# --- 3. GET SPECIMEN-REGION LOOKUP ---

# We need to get the 'region' for each 'specimen'.
if (!exists("all_proc")) {
  message("Warning: 'all_proc' dataframe not found. Loading from RDS...")
  # Fallback to load the raw data
  df_raw <- readRDS("RDS_dataframes/combined_IBM_LPW_raw.RDS") %>% 
    filter(complete.cases(read_age))
  
  df_raw <- df_raw %>%
    mutate(
      region = case_when(
        region %in% c("E Shumagins", "W Shumagins") ~ "Shumagin",
        region %in% c("N Kodiak", "S Kodiak") ~ "Kodiak",
        TRUE ~ region
      )
    )
  all_proc <- df_raw # Use the raw df for the lookup
}

specimen_region_lookup <- all_proc %>%
  select(specimen, region) %>%
  distinct()


# --- 4. LOAD AND COMBINE ALL RUNS ---

# Loop over our path list and run the loader function for each
all_summarized_preds_list <- purrr::imap(RUN_PATHS, load_and_summarize_preds)

# Combine the 5 tibbles into one big one
all_summarized_preds <- bind_rows(all_summarized_preds_list)

# Join with region data
final_data_for_plotting <- all_summarized_preds %>%
  left_join(specimen_region_lookup, by = "specimen")

# --- 5. LOAD SUMMARY & FIND BEST MODELS ---

results_ALL <- read.csv("C:/Users/poopm/Thesis Writing/Github/LPW_FT-NIRS/Chapter 2/results_unfiltered.csv")

# We must filter for finite RMSE values *before* slicing,
# otherwise NA/Inf values will cause groups to be dropped.
best_models_unstandardized <- results_ALL %>%
  filter(is.finite(RMSE)) %>%
  group_by(model_run) %>%
  slice_min(order_by = RMSE, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(model_run, model_name, RMSE)

message("Found the following best models (unstandardized):")
print(best_models_unstandardized)


# --- 6. STANDARDIZE & FILTER PREDICTIONS ---
#
# This section standardizes the names from BOTH the Arrow data and the CSV
# data to a common format *before* joining them.
#
# 1. Standardize names from the ARROW data (final_data_for_plotting)
preds_standardized <- final_data_for_plotting %>%
  mutate(
    # Create the new, simple 'Model' column
    Model = case_when(
      str_detect(model_name, "PCA-LM") ~ model_name, # Keep full PCA-LM names
      str_detect(model_name, "PCA-GAM") ~ model_name, # Keep full PCA-GAM names
      model_name == "PLS" ~ "PLS",
      str_detect(model_name, "PLS-VIP") ~ "PLS-VIP", # Standardize to "PLS-VIP"
      model_name == "XGBoost" ~ "XGB",             # Standardize to "XGB"
      model_name == "Random Forest" ~ "RF",         # Standardize to "RF"
      model_name == "Simple-LM (L+SW)" ~ "Simple-LM",   # Standardize to "Simple-LM"
      model_name == "Simple-GAM (L+SW)" ~ "Simple-GAM", # Standardize to "Simple-GAM"
      TRUE ~ model_name
    ),
    ModelType = case_when(
      str_detect(Model, "PCA-LM") | str_detect(Model, "PCA-GAM") ~ "PCA",
      Model %in% c("PLS", "PLS-VIP") ~ "PLS",
      Model %in% c("XGB", "RF") ~ "ML",
      Model %in% c("Simple-LM", "Simple-GAM") ~ "Simple",
      TRUE ~ model_type
    )
  ) %>%
  filter(!is.na(ModelType)) %>%
  # Select only the columns we need for the join
  select(specimen, obs_age, median_pred_age, model_run, region, Model, ModelType)

# 2. Standardize names from the CSV data (best_models)
best_models_standardized <- best_models_unstandardized %>%
  mutate(
    # Create the new, simple 'Model' column
    Model = case_when(
      model_name == "Simple LM" ~ "Simple-LM",
      model_name == "Simple GAM" ~ "Simple-GAM",
      # All others (XGB, RF, PLS, PLS-VIP, PCA-GAM 1, etc.)
      # already match the simple names from the CSV.
      TRUE ~ model_name
    )
  ) %>%
  # We join on the NEW 'Model' column, not the old 'model_name'
  select(model_run, Model, RMSE)

message("Standardized best models:")
print(best_models_standardized)

# 3. Join the two standardized tables
# This will join `(model_run="All", Model="XGB")`
# with `(model_run="All", Model="XGB")`
final_data_cleaned <- preds_standardized %>%
  inner_join(best_models_standardized, by = c("model_run", "Model"))

message(paste(
  "Join successful. Final filtered data has",
  nrow(final_data_cleaned), "rows."
))
print(unique(final_data_cleaned$model_run))


# 4. Set factor levels for plotting
final_type_order <- c("PCA", "PLS", "ML", "Simple")

final_data_cleaned <- final_data_cleaned %>%
  mutate(
    ModelType = factor(ModelType, levels = final_type_order)
  )
```


```{r}
# --- 7. METRIC & PLOTTING FUNCTIONS ---

# --- Helper functions for metric calculation ---
calculate_rmse <- function(obs, pred) {
  sqrt(mean((obs - pred)^2, na.rm = TRUE))
}

calculate_r2 <- function(obs, pred) {
  TSS <- sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)
  RSS <- sum((obs - pred)^2, na.rm = TRUE)
  1 - (RSS / TSS)
}

calculate_bias <- function(obs, pred) {
  mean(pred - obs, na.rm = TRUE)
}


#' Creates a comparison scatter plot grid
#'
#' @param plot_data The data to plot (already filtered)
#' @param title_text The main title for the plot
#' @param run_order A vector of strings for the row order (e.g., c("All", "Wild"))
#'

create_comparison_plot <- function(plot_data, title_text, run_order) {
  
  # Create a new label for the facet strips that includes the model name
  plot_data_labeled <- plot_data %>%
    # Note: 'Model' is the simplified name, e.g., "XGB"
    mutate(
      facet_label = paste0("Trained with: ", model_run, " (", Model, ")"),
      facet_label = factor(facet_label, levels = unique(facet_label[order(factor(model_run, levels = run_order))]))
    )
  
  # --- NEW: Calculate metrics for text labels ---
  metric_labels <- plot_data_labeled %>%
    group_by(facet_label) %>%
    summarise(
      plot_rmse = calculate_rmse(obs_age, median_pred_age),
      plot_r2 = calculate_r2(obs_age, median_pred_age),
      plot_bias = calculate_bias(obs_age, median_pred_age)
    ) %>%
    mutate(
      # Create the formatted text string for the label
      label_string = sprintf(
        "RMSE: %.2f\nR²: %.2f\nBias: %.2f",
        plot_rmse, plot_r2, plot_bias
      )
    )

  # --- Main Plot ---
  ggplot(plot_data_labeled, aes(x = obs_age, y = median_pred_age)) +
    # 1:1 line
    geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
    # Points
    geom_point(color = "black", alpha = 0.7, size = 3) +
    # Line of best fit
    geom_smooth(method = "lm", se = F, color = "red", linetype = "dotted", linewidth = 0.8) +
    
    # --- NEW: Add the text labels ---
    geom_text(
      data = metric_labels,
      aes(label = label_string),
      x = -Inf, y = Inf, # Position in top-left corner
      hjust = -0.1,       # Nudge right
      vjust = 1.2,        # Nudge down
      size = 5,
      lineheight = 0.9
    ) +
    # Facet by the new combined label, in a single column
    facet_wrap(~ facet_label, ncol = 1) +
    labs(
      title = title_text,
      x = "Thin Section Age (Days)",
      y = "Median Predicted Age (Days)"
    ) +
    theme_bw(base_size = 17) +
    theme(
      legend.position = "bottom",
      strip.text = element_text(face = "bold", size = 15),
      strip.background = element_rect(fill = "gray90")
    )
}
```


```{r}
# --- 8. GENERATE AND SAVE PLOTS ---
#
# PLOT 1: KODIAK SPECIMENS
#
plot_data_kod <- final_data_cleaned %>%
  filter(
    region == "Kodiak",
    model_run %in% c("All", "Wild", "Kodiak")
  )

# Define the row order for the plot
kod_run_order <- c("All", "Wild", "Kodiak")

plot_kod <- create_comparison_plot(
  plot_data_kod,
  "Kodiak Specimen Predictions",
  kod_run_order
)

print(plot_kod)
# ggsave(
#   "Chapter 2/comparison_plot_kodiak_best.png",
#   plot_kod,
#   width = 8, height = 15, dpi = 300 # Narrower, taller
# )


#
# PLOT 2: SHUMAGIN SPECIMENS
#
plot_data_shu <- final_data_cleaned %>%
  filter(
    region == "Shumagin",
    model_run %in% c("All", "Wild", "Shumagin")
  )

shu_run_order <- c("All", "Wild", "Shumagin") # <-- Corrected typo

plot_shu <- create_comparison_plot(
  plot_data_shu,
  "Shumagin Specimen Predictions",
  shu_run_order
)

print(plot_shu)
# ggsave(
#   "Chapter 2/comparison_plot_shumagin_best.png",
#   plot_shu,
#   width = 8, height = 15, dpi = 300 # Narrower, taller
# )


#
# PLOT 3: LPW SPECIMENS
#
plot_data_lpw <- final_data_cleaned %>%
  filter(
    region == "LPW",
    model_run %in% c("All", "LPW")
  )

lpw_run_order <- c("All", "LPW")

plot_lpw <- create_comparison_plot(
  plot_data_lpw,
  "LPW Specimen Predictions",
  lpw_run_order
)

print(plot_lpw)
# ggsave(
#   "Chapter 2/comparison_plot_lpw_best.png",
#   plot_lpw,
#   width = 8, height = 10, dpi = 300 # Narrower, shorter (2 rows)
# )

message("All comparison plots generated and saved.")
```


I'm thinking of using a table that shows model metrics for the best of all types that includes, model type, which set of models it came from.


```{r}

```

