# Load packages
Install and load necessary packages
```{r}
packages <- c("tidyverse")
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages) > 0) {
  install.packages(new_packages)
}
invisible(lapply(packages, library, character.only = TRUE))
rm(packages, new_packages)
```


# Data Import
Read and prepare initial dataframe for otolith elemental analysis from saved RDS file

```{r}
# Load ICP-MS data from RDS file
df <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")

# Select only otolith specimens (rows 30-184)
df <- df[c(30:184)]
```

### create single dataframe (remove lists)
```{r}
# no specimen 14
# specimen 24 missing a third transect
for(i in 1:155){
  spec <- names(df[i])
  df[[i]]$specimen <- as.factor(str_sub(spec, start= 0,end = 5))
  df[[i]]$transect <- as.factor(as.numeric(str_sub(spec, start= -1)) + 1)
  rm(spec,i)
}
df <- do.call("rbind", df)
df <- rename(df, "time" = `Elapsed Time`)
# saveRDS(df, "RDS_dataframes/ICP-MS_df.RDS")
rm(ICP_MS, df)
```

### outliers
```{r}
bad <- c(10,14,30,32,34,35)
good <- 1:40
good <- good[!good %in% c(bad)]
```


# Best Approach



### find_otolith_boundaries()
Detect Otolith Boundaries using Ca43 and Cu65 signals
ca43_data: Numeric vector of Ca43 measurements
cu65_data: Numeric vector of Cu65 measurements
window_size:  Size of moving average window (default: 7)
```{r}
find_otolith_boundaries <- function(ca43_data, cu65_data, window_size = 7) {
  # Data preparation
  n_points <- length(ca43_data)
  # Calculate rolling medians with larger window for stability
  ma_ca43 <- stats::runmed(ca43_data, k = window_size)
  ma_cu65 <- stats::runmed(cu65_data, k = window_size)
  # Calculate percentiles using middle portion of data for better baseline
  mid_start <- ceiling(n_points * 0.3)
  mid_end <- floor(n_points * 0.7)
  mid_data_ca43 <- ma_ca43[mid_start:mid_end]
  # Get reference levels
  ca43_median <- median(mid_data_ca43, na.rm = TRUE)
  ca43_start_threshold <- ca43_median * 0.4  # More lenient threshold
  # Start detection with minimum signal requirements
  min_stable_points <- 5
  start_idx <- NULL
  # Look for sustained rise in Ca43
  for(i in window_size:(n_points-min_stable_points)) {
    window <- ma_ca43[i:(i+min_stable_points)]
    if(all(window > ca43_start_threshold) && 
       mean(window) > mean(ma_ca43[(i-window_size):i])) {
      start_idx <- i
      break
    }
  }
  # Fallback if no clear start found
  if(is.null(start_idx)) {
    # Use first major rise in Ca43
    for(i in 2:n_points) {
      if(ma_ca43[i] > ca43_start_threshold && 
         ma_ca43[i] > mean(ma_ca43[max(1, i-5):i]) * 1.5) {
        start_idx <- i
        break
      }
    }
  }
  # If still no start found, use conservative approach
  if(is.null(start_idx)) {
    start_idx <- ceiling(n_points * 0.1)  # Use first 10% as default
  }
  # End detection
  end_idx <- n_points
  tail_length <- min(50, ceiling(n_points * 0.1))
  # Only look for end if we have enough points
  if(n_points - start_idx > window_size * 3) {
    tail_median <- median(ma_ca43[(n_points-tail_length):n_points])
    for(i in (n_points-tail_length):start_idx) {
      if(mean(ma_ca43[i:(i+min(4, n_points-i))]) < ca43_median * 0.3) {
        end_idx <- i
        break
      }
    }
  }
  # Validate and adjust boundaries
  if(end_idx - start_idx < 100) {  # Minimum reasonable length
    end_idx <- n_points
    start_idx <- min(start_idx, ceiling(n_points * 0.15))  # Max 15% from start
  }
  return(list(
    start = start_idx,
    end = end_idx,
    filtered_data = ca43_data[start_idx:end_idx],
    diagnostics = list(
      ca43_median = ca43_median,
      start_threshold = ca43_start_threshold
    )
  ))
}
```
### process_otoliths()
Process all otolith transects
```{r}
#'  otolith_list List of dataframes containing Ca43 and Cu65_ppm measurements
#'  Summary dataframe of processing results
process_otoliths <- function(otolith_list) {
  results <- list()
  failed_specimens <- character()
  
  for(name in names(otolith_list)) {
    tryCatch({
      bounds <- find_otolith_boundaries(
        otolith_list[[name]]$Ca43,
        otolith_list[[name]]$Cu65_ppm
      )
      results[[name]] <- bounds
      cat("Processed", name, 
          "- Start:", bounds$start, 
          "End:", bounds$end, "\n")
    }, error = function(e) {
      cat("Error processing", name, ":", e$message, "\n")
      failed_specimens <- c(failed_specimens, name)
    })
  }
  summary_df <- data.frame(
    transect = names(results),
    start_index = sapply(results, `[[`, "start"),
    end_index = sapply(results, `[[`, "end"),
    original_length = sapply(otolith_list[names(results)], nrow),
    filtered_length = sapply(results, function(x) length(x$filtered_data))
  )
  summary_df$points_removed <- summary_df$original_length - summary_df$filtered_length
  attr(summary_df, "failed_specimens") <- failed_specimens
  return(summary_df)
}
```
#### outlier from analysis
```{r}
# Documentation for anomalous specimen oto10-2
ANOMALY_NOTE <- list(
  specimen_id = "oto10-2",
  issue = "Mid-transect elemental data disruption",
  details = list(
    observation = "All elemental data show aberrant values in middle section",
    impact = "Creates artificial truncation in data processing",
    location = "Around point 288 in transect",
    comparison = "Other oto10 series (0,1,3) show normal patterns"
  ),
  recommendation = "Should be flagged for separate analysis or potential exclusion from bulk processing"
)

```

### Filter DF

```{r}
# Process all specimens
results <- process_otoliths(df)
# Create filtered version of original dataset
filtered_df <- lapply(names(df), function(specimen) {
  # Get boundaries for this specimen
  bounds <- results[results$transect == specimen, ]
  
  # Skip if specimen is the documented anomaly
  if(specimen == "oto10-2") {
    warning(paste("Skipping known anomalous specimen:", specimen))
    return(NULL)
  }
  
  # Extract relevant portion of data
  df[[specimen]][bounds$start_index:bounds$end_index, ]
})
# Name the list elements
names(filtered_df) <- names(df)
# Remove NULL entries (anomalous specimens)
filtered_df <- filtered_df[!sapply(filtered_df, is.null)]
# saveRDS(filtered_df, "RDS_dataframes/ICP-MS_cutoffs_df.RDS")
```

### Validation checks
```{r validation}
# Check filtered dataset
cat("Original specimens:", length(df), "\n")
cat("Filtered specimens:", length(filtered_df), "\n")
cat("Removed specimens:", setdiff(names(df), names(filtered_df)), "\n")

# Check lengths match results
all_match <- all(sapply(names(filtered_df), function(specimen) {
  nrow(filtered_df[[specimen]]) == 
    results$filtered_length[results$transect == specimen]
}))
cat("All lengths match results:", all_match, "\n")
```

### generate_otolith_stats()
summary stats from filtered DF, both by transect and per otolith
```{r}
generate_otolith_stats <- function(filtered_df) {
  elements <- c("Ca43", "Li7_ppm", "B11_ppm", "Mg24_ppm", "Si29_ppm", 
                "Mn55_ppm", "Fe56_ppm", "Co59_ppm", "Ni60_ppm", 
                "Cu63_ppm", "Cu65_ppm", "Zn66_ppm", "Rb85_ppm", 
                "Sr86_ppm", "Sr87_ppm", "Sr88_ppm", "Cd111_ppm", 
                "Ba138_ppm", "Ce140_ppm", "Pb206_ppm", "Pb207_ppm", 
                "Pb208_ppm", "Th232_ppm", "PbTotal_ppm")
  
  format_summary_tables <- function(stats_list, elements) {
    transect_summary <- do.call(rbind, lapply(stats_list$transect_stats, function(x) {
      regions <- c("core", "middle", "edge")
      
      do.call(rbind, lapply(regions, function(r) {
        data.frame(
          specimen = x$specimen,
          region = r,
          element = elements,
          mean = x$stats[[r]]["mean", ],
          sd = x$stats[[r]]["sd", ],
          n = x$stats[[r]]["n", ],
          stringsAsFactors = FALSE
        )
      }))
    }))
    
    otolith_summary <- do.call(rbind, lapply(stats_list$otolith_stats, function(x) {
      regions <- c("core", "middle", "edge")
      
      do.call(rbind, lapply(regions, function(r) {
        data.frame(
          otolith = x$otolith,
          region = r,
          element = elements,
          mean = x$stats[[r]]$means["mean", ],
          rsd = x$stats[[r]]$rsd,
          stringsAsFactors = FALSE
        )
      }))
    }))
    
    return(list(
      transect_summary = transect_summary,
      otolith_summary = otolith_summary
    ))
  }
  
  transect_stats <- lapply(names(filtered_df), function(specimen) {
    df <- filtered_df[[specimen]]
    n_points <- nrow(df)
    
    core_end <- ceiling(n_points * 0.2)
    middle_start <- core_end + 1
    middle_end <- floor(n_points * 0.8)
    edge_start <- middle_end + 1
    
    regions <- list(
      core = df[1:core_end, ],
      middle = df[middle_start:middle_end, ],
      edge = df[edge_start:n_points, ]
    )
    
    region_stats <- lapply(regions, function(region) {
      sapply(elements, function(elem) {
        data <- region[[elem]]
        c(
          mean = mean(data, na.rm = TRUE),
          sd = sd(data, na.rm = TRUE),
          median = median(data, na.rm = TRUE),
          q25 = quantile(data, 0.25, na.rm = TRUE),
          q75 = quantile(data, 0.75, na.rm = TRUE),
          n = sum(!is.na(data))
        )
      })
    })
    
    return(list(
      specimen = specimen,
      stats = region_stats,
      n_points = n_points
    ))
  })
  
  otolith_ids <- unique(sub("-\\d+$", "", names(filtered_df)))
  
  otolith_stats <- lapply(otolith_ids, function(otolith) {
    transects <- grep(paste0("^", otolith, "-"), names(filtered_df), value = TRUE)
    regions <- c("core", "middle", "edge")
    
    combined_stats <- lapply(regions, function(region) {
      region_stats <- lapply(transects, function(t) {
        idx <- match(t, names(filtered_df))
        transect_stats[[idx]]$stats[[region]]
      })
      
      element_means <- Reduce(`+`, region_stats) / length(region_stats)
      
      element_rsd <- sapply(elements, function(elem) {
        values <- sapply(region_stats, function(x) x[1, elem])
        sd(values, na.rm = TRUE) / mean(values, na.rm = TRUE) * 100
      })
      
      list(
        means = element_means,
        rsd = element_rsd
      )
    })
    
    names(combined_stats) <- regions
    
    return(list(
      otolith = otolith,
      stats = combined_stats
    ))
  })
  
  stats_list <- list(
    transect_stats = transect_stats,
    otolith_stats = otolith_stats
  )
  
  summary_tables <- format_summary_tables(stats_list, elements)
  
  return(summary_tables)
}

summary_results <- generate_otolith_stats(filtered_df)
results_transect <- summary_results$transect_summary
results_otolith <- summary_results$otolith_summary
# saveRDS(summary_results,"RDS_dataframes/ICP-MS_summarystats")
```

### Data analysis

```{r}
# 1. Basic Variation Check
# Calculate RSD (Relative Standard Deviation) between transects for each otolith-element-region
transect_variation <- function(summary_results) {
  # Get transect summary data
  ts <- summary_results$transect_summary
  
  # Group by otolith (remove transect number), element, and region
  variation_check <- aggregate(mean ~ sub("-\\d+$", "", specimen) + element + region, 
                               data = ts,
                               FUN = function(x) {
                                 c(mean = mean(x),
                                   rsd = sd(x)/mean(x) * 100,
                                   n_transects = length(x))
                               })
  
  # Filter for high variation (e.g., RSD > 20%)
  high_variation <- subset(variation_check, variation_check$mean[, "rsd"] > 20)
  
  return(list(
    all_variation = variation_check,
    high_variation = high_variation
  ))
}

# 2. Statistical Test
# Perform ANOVA for each otolith-element-region combination
transect_anova <- function(summary_results) {
  ts <- summary_results$transect_summary
  
  # Split data by otolith, element, and region
  unique_combinations <- unique(paste(sub("-\\d+$", "", ts$specimen), 
                                      ts$element, 
                                      ts$region))
  
  # Run ANOVA for each combination
  anova_results <- lapply(unique_combinations, function(combo) {
    split_combo <- strsplit(combo, " ")[[1]]
    otolith <- split_combo[1]
    element <- split_combo[2]
    region <- split_combo[3]
    
    subset_data <- subset(ts, 
                          sub("-\\d+$", "", specimen) == otolith & 
                            element == element & 
                            region == region)
    
    if(length(unique(subset_data$mean)) > 1) {
      anova_result <- summary(aov(mean ~ specimen, data = subset_data))
      p_value <- anova_result[[1]]$"Pr(>F)"[1]
      
      return(data.frame(
        otolith = otolith,
        element = element,
        region = region,
        p_value = p_value,
        significant = p_value < 0.05
      ))
    }
    return(NULL)
  })
  
  # Combine results
  anova_results <- do.call(rbind, anova_results[!sapply(anova_results, is.null)])
  return(anova_results)
}

# Usage:
variation_results <- transect_variation(summary_results)
anova_results <- transect_anova(summary_results)

# Quick summary of problematic measurements
print("Elements with high variation between transects:")
print(variation_results$high_variation)

print("Statistically significant differences between transects:")
print(subset(anova_results, significant == TRUE))

# Comparative Analysis of Elemental Composition Across Otolith Regions

compare_otolith_regions <- function(summary_results) {
  # Prepare otolith summary data
  os <- summary_results$otolith_summary
  
  # ANOVA-like comparison between regions for each element
  region_comparisons <- lapply(unique(os$element), function(elem) {
    elem_data <- subset(os, element == elem)
    
    # Reshape data for statistical testing
    region_means <- reshape2::dcast(elem_data, otolith ~ region, value.var = "mean")
    
    # Perform statistical tests
    pairwise_tests <- list(
      core_vs_middle = t.test(region_means$core, region_means$middle),
      core_vs_edge = t.test(region_means$core, region_means$edge),
      middle_vs_edge = t.test(region_means$middle, region_means$edge)
    )
    
    # Summarize results
    data.frame(
      element = elem,
      core_mean = mean(region_means$core),
      middle_mean = mean(region_means$middle),
      edge_mean = mean(region_means$edge),
      core_vs_middle_p = pairwise_tests$core_vs_middle$p.value,
      core_vs_edge_p = pairwise_tests$core_vs_edge$p.value,
      middle_vs_edge_p = pairwise_tests$middle_vs_edge$p.value
    )
  })
  
  # Combine results
  region_comparison_summary <- do.call(rbind, region_comparisons)
  
  # Adjust p-values for multiple comparisons
  region_comparison_summary$core_vs_middle_padj <- p.adjust(region_comparison_summary$core_vs_middle_p, method = "fdr")
  region_comparison_summary$core_vs_edge_padj <- p.adjust(region_comparison_summary$core_vs_edge_p, method = "fdr")
  region_comparison_summary$middle_vs_edge_padj <- p.adjust(region_comparison_summary$middle_vs_edge_p, method = "fdr")
  
  # Identify significant differences
  region_comparison_summary$significant_differences <- apply(
    region_comparison_summary[, c("core_vs_middle_padj", "core_vs_edge_padj", "middle_vs_edge_padj")], 
    1, 
    function(x) any(x < 0.05)
  )
  
  return(region_comparison_summary)
}
region_comparisons <- compare_otolith_regions(summary_results)
# Print significant differences
print(subset(region_comparisons, significant_differences == TRUE))


```


### compare otoliths

```{r}
compare_otoliths <- function(summary_results) {
  # Prepare otolith summary data
  os <- summary_results$otolith_summary
  
  # Perform statistical comparisons between otoliths
  otolith_comparisons <- lapply(unique(os$element), function(elem) {
    # Extract data for specific element
    elem_data <- subset(os, element == elem)
    
    # Reshape data for statistical analysis
    region_means <- reshape2::dcast(elem_data, otolith + region ~ element, value.var = "mean")
    
    # Perform ANOVA to test overall differences between otoliths
    aov_results <- lapply(c("core", "middle", "edge"), function(reg) {
      region_subset <- subset(region_means, region == reg)
      
      # Check if we have enough unique values to perform ANOVA
      if(length(unique(region_subset$otolith)) > 1) {
        tryCatch({
          aov_model <- aov(get(elem) ~ otolith, data = region_subset)
          aov_summary <- summary(aov_model)
          
          data.frame(
            element = elem,
            region = reg,
            f_value = aov_summary[[1]]$"F value"[1],
            p_value = aov_summary[[1]]$"Pr(>F)"[1]
          )
        }, error = function(e) {
          NULL  # Return NULL if ANOVA fails
        })
      } else {
        NULL  # Not enough unique values for ANOVA
      }
    })
    
    # Remove NULL results
    aov_results <- aov_results[!sapply(aov_results, is.null)]
    
    # Combine results if any exist
    if(length(aov_results) > 0) {
      do.call(rbind, aov_results)
    } else {
      NULL
    }
  })
  
  # Remove NULL results and combine
  otolith_comparisons <- otolith_comparisons[!sapply(otolith_comparisons, is.null)]
  
  if(length(otolith_comparisons) > 0) {
    otolith_comparison_summary <- do.call(rbind, otolith_comparisons)
    
    # Adjust p-values for multiple comparisons
    otolith_comparison_summary$p_adjusted <- p.adjust(
      otolith_comparison_summary$p_value, 
      method = "fdr"
    )
    
    # Identify significant differences
    otolith_comparison_summary$significant <- otolith_comparison_summary$p_adjusted < 0.05
    
    return(otolith_comparison_summary)
  } else {
    warning("No valid comparisons could be made. Check your data.")
    return(NULL)
  }
}

# Visualization function remains the same
visualize_otolith_differences <- function(summary_results) {
  library(ggplot2)
  library(reshape2)
  
  os <- summary_results$otolith_summary
  
  # Prepare data for heatmap
  plot_data <- dcast(os, otolith + region ~ element, value.var = "mean")
  
  # Melt data for ggplot
  plot_data_long <- melt(plot_data, id.vars = c("otolith", "region"))
  
  # Create heatmap
  ggplot(plot_data_long, aes(x = otolith, y = variable, fill = value)) +
    geom_tile() +
    facet_wrap(~ region) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
    theme_minimal() +
    labs(title = "Elemental Composition Across Otoliths",
         x = "Otolith", y = "Element", fill = "Mean Value")
}

# Usage
otolith_comparisons <- compare_otoliths(summary_results)
otolith_plot <- visualize_otolith_differences(summary_results)

# Print significant differences
if(!is.null(otolith_comparisons)) {
  print("Significant differences between otoliths:")
  print(subset(otolith_comparisons, significant == TRUE))
} else {
  print("No significant differences could be detected.")
}
```

# SHINY - mark lines but better


```{r}

library(shiny)
library(plotly)
library(magick)
library(base64enc)

ui <- fluidPage(
  titlePanel("Otolith Ablation Line Quality Assessment Tool"),
  
  sidebarLayout(
    sidebarPanel(
      fileInput("imageFile", "Upload Otolith Image", accept = c('image/png', 'image/jpeg')),
      
      hr(),
      h4("Step 1: Draw Line"),
      textInput("lineName", "Line Name/ID:", placeholder = "Enter specimen ID or line name"),
      actionButton("startLine", "Start New Line", class = "btn-primary"),
      actionButton("clearLine", "Clear Line", class = "btn-danger"),
      
      hr(),
      h4("Step 2: Mark Features"),
      radioButtons("annotationType", "Select Feature to Mark:",
                   choices = c("Start of Otolith" = "start",
                               "Core" = "core",
                               "End of Otolith" = "end",
                               "Start of Useful Data" = "data_start",
                               "End of Useful Data" = "data_end"),
                   selected = "start"),
      actionButton("clearAnnotations", "Clear Annotations", class = "btn-warning"),
      
      hr(),
      h4("Step 3: Transect Quality"),
      selectInput("transectQuality", "Transect Quality:",
                  choices = c("Good" = "good",
                              "Marginal" = "marginal",
                              "Poor (Exclude)" = "poor"),
                  selected = "good"),
      textAreaInput("notes", "Notes:", rows = 3, 
                    placeholder = "Add notes about data quality or reasons for exclusion"),
      
      hr(),
      h4("Line Management"),
      actionButton("saveLine", "Save Current Line", class = "btn-success"),
      selectInput("lineSelector", "Select Line:", choices = NULL),
      actionButton("deleteLine", "Delete Selected Line", class = "btn-danger"),
      
      hr(),
      downloadButton("downloadData", "Download All Data", class = "btn-info"),
      
      hr(),
      verbatimTextOutput("statusText")
    ),
    
    mainPanel(
      plotOutput("imagePlot", height = "600px", click = "plot_click"),
      verbatimTextOutput("lineInfo")
    )
  )
)

server <- function(input, output, session) {
  # Reactive values
  values <- reactiveValues(
    # Image data
    img = NULL,
    img_width = 0,
    img_height = 0,
    
    # Current line being drawn
    drawing = FALSE,
    current_line = list(
      name = "",
      start = NULL,
      end = NULL,
      annotations = list(),
      quality = "good",
      notes = ""
    ),
    
    # Saved lines
    lines = list(),
    selected_line_index = 0
  )
  
  # Display the image with annotations
  output$imagePlot <- renderPlot({
    req(values$img)
    
    # Plot the image
    plot(values$img, axes = FALSE)
    
    # Draw the current line if it exists
    if (!is.null(values$current_line$start)) {
      points(values$current_line$start$x, values$current_line$start$y, 
             col = "red", pch = 16, cex = 2)
      
      # If end point exists, draw the line
      if (!is.null(values$current_line$end)) {
        # Line style based on quality
        line_style <- switch(values$current_line$quality,
                             "good" = 1,    # solid
                             "marginal" = 2, # dashed
                             "poor" = 3)     # dotted
        
        lines(
          c(values$current_line$start$x, values$current_line$end$x),
          c(values$current_line$start$y, values$current_line$end$y),
          col = "red", lwd = 2, lty = line_style
        )
        points(values$current_line$end$x, values$current_line$end$y, 
               col = "red", pch = 16, cex = 2)
        
        # Add line name text near the middle of the line if available
        if (values$current_line$name != "") {
          mid_x <- (values$current_line$start$x + values$current_line$end$x) / 2
          mid_y <- (values$current_line$start$y + values$current_line$end$y) / 2
          text(mid_x, mid_y, values$current_line$name, col = "red", pos = 3, offset = 0.5, cex = 0.8)
        }
        
        # Draw annotations on the current line
        for (type in names(values$current_line$annotations)) {
          point <- values$current_line$annotations[[type]]
          color <- switch(type,
                          "start" = "green",
                          "core" = "purple",
                          "end" = "orange",
                          "data_start" = "blue",
                          "data_end" = "darkred")
          
          symbol <- switch(type,
                           "start" = 15,    # Square
                           "core" = 18,     # Diamond
                           "end" = 17,      # Triangle
                           "data_start" = 23, # Diamond plus
                           "data_end" = 25)   # Square plus
          
          points(point$x, point$y, col = color, pch = symbol, cex = 2)
          
          # Add label for the annotation
          label <- switch(type,
                          "start" = "Start",
                          "core" = "Core",
                          "end" = "End",
                          "data_start" = "Data Start",
                          "data_end" = "Data End")
          text(point$x, point$y, label, col = color, pos = 4, cex = 0.7)
        }
        
        # If data start and end are marked, highlight the useful data section
        if (!is.null(values$current_line$annotations$data_start) && 
            !is.null(values$current_line$annotations$data_end)) {
          
          # Get the positions on the line
          start_point <- list(x = values$current_line$start$x, y = values$current_line$start$y)
          end_point <- list(x = values$current_line$end$x, y = values$current_line$end$y)
          data_start <- values$current_line$annotations$data_start
          data_end <- values$current_line$annotations$data_end
          
          # Calculate position parameters
          t_start <- calculate_position_on_line(data_start, start_point, end_point)
          t_end <- calculate_position_on_line(data_end, start_point, end_point)
          
          # Draw a thicker line segment for the useful data
          lines(
            c(start_point$x + t_start * (end_point$x - start_point$x), 
              start_point$x + t_end * (end_point$x - start_point$x)),
            c(start_point$y + t_start * (end_point$y - start_point$y), 
              start_point$y + t_end * (end_point$y - start_point$y)),
            col = "cyan", lwd = 4
          )
        }
      }
    }
    
    # Draw saved lines
    for (i in seq_along(values$lines)) {
      line <- values$lines[[i]]
      
      # Line color - highlight selected line
      line_color <- ifelse(i == values$selected_line_index, "blue", "gray")
      
      # Line style based on quality
      line_style <- switch(line$quality,
                           "good" = 1,    # solid
                           "marginal" = 2, # dashed
                           "poor" = 3)     # dotted
      
      # Draw line
      lines(
        c(line$start$x, line$end$x),
        c(line$start$y, line$end$y),
        col = line_color, lwd = 2, lty = line_style
      )
      
      # Draw endpoints
      points(line$start$x, line$start$y, col = line_color, pch = 16, cex = 1.5)
      points(line$end$x, line$end$y, col = line_color, pch = 16, cex = 1.5)
      
      # Add line name text near the middle of the line if available
      if (!is.null(line$name) && line$name != "") {
        mid_x <- (line$start$x + line$end$x) / 2
        mid_y <- (line$start$y + line$end$y) / 2
        text(mid_x, mid_y, line$name, col = line_color, pos = 3, offset = 0.5, cex = 0.8)
      }
      
      # Draw annotations
      for (type in names(line$annotations)) {
        point <- line$annotations[[type]]
        color <- switch(type,
                        "start" = "green",
                        "core" = "purple",
                        "end" = "orange",
                        "data_start" = "blue",
                        "data_end" = "darkred")
        
        symbol <- switch(type,
                         "start" = 15,    # Square
                         "core" = 18,     # Diamond
                         "end" = 17,      # Triangle
                         "data_start" = 23, # Diamond plus
                         "data_end" = 25)   # Square plus
        
        points(point$x, point$y, col = color, pch = symbol, cex = 1.5)
      }
      
      # If data start and end are marked, highlight the useful data section
      if (!is.null(line$annotations$data_start) && 
          !is.null(line$annotations$data_end)) {
        
        # Get the positions on the line
        start_point <- list(x = line$start$x, y = line$start$y)
        end_point <- list(x = line$end$x, y = line$end$y)
        data_start <- line$annotations$data_start
        data_end <- line$annotations$data_end
        
        # Calculate position parameters
        t_start <- calculate_position_on_line(data_start, start_point, end_point)
        t_end <- calculate_position_on_line(data_end, start_point, end_point)
        
        # Draw a thicker line segment for the useful data
        lines(
          c(start_point$x + t_start * (end_point$x - start_point$x), 
            start_point$x + t_end * (end_point$x - start_point$x)),
          c(start_point$y + t_start * (end_point$y - start_point$y), 
            start_point$y + t_end * (end_point$y - start_point$y)),
          col = "cyan", lwd = 3
        )
      }
    }
  })
  
  # Function to calculate position along line (0 to 1)
  calculate_position_on_line <- function(point, start, end) {
    # Calculate vectors
    line_vector <- c(end$x - start$x, end$y - start$y)
    point_vector <- c(point$x - start$x, point$y - start$y)
    
    # Calculate projection
    projection <- sum(point_vector * line_vector) / sum(line_vector^2)
    
    # Clamp to [0, 1]
    return(max(0, min(1, projection)))
  }
  
  # Handle image upload
  observeEvent(input$imageFile, {
    # Read the image
    values$img <- image_read(input$imageFile$datapath)
    info <- image_info(values$img)
    values$img_width <- info$width
    values$img_height <- info$height
    
    # Reset current line
    values$current_line <- list(
      name = "",
      start = NULL,
      end = NULL,
      annotations = list(),
      quality = "good",
      notes = ""
    )
    values$drawing <- FALSE
  })
  
  # Start a new line
  observeEvent(input$startLine, {
    values$current_line <- list(
      name = input$lineName,
      start = NULL,
      end = NULL,
      annotations = list(),
      quality = input$transectQuality,
      notes = input$notes
    )
    values$drawing <- TRUE
    output$statusText <- renderText("Click on the image to set the START point of the line")
  })
  
  # Clear current line
  observeEvent(input$clearLine, {
    values$current_line <- list(
      name = input$lineName,
      start = NULL,
      end = NULL,
      annotations = list(),
      quality = input$transectQuality,
      notes = input$notes
    )
    values$drawing <- FALSE
    output$statusText <- renderText("Line cleared")
  })
  
  # Clear annotations on current line
  observeEvent(input$clearAnnotations, {
    values$current_line$annotations <- list()
    output$statusText <- renderText("Annotations cleared")
  })
  
  # Update quality and notes when changed
  observeEvent(input$transectQuality, {
    values$current_line$quality <- input$transectQuality
  })
  
  observeEvent(input$notes, {
    values$current_line$notes <- input$notes
  })
  
  # Save current line
  observeEvent(input$saveLine, {
    if (!is.null(values$current_line$start) && !is.null(values$current_line$end)) {
      # Make sure the line has a name - use default if empty
      if (is.null(values$current_line$name) || values$current_line$name == "") {
        values$current_line$name <- paste("Line", length(values$lines) + 1)
      }
      
      # Add the current line to saved lines
      values$lines[[length(values$lines) + 1]] <- values$current_line
      
      # Update line selector with names
      updateLineSelector()
      
      # Set as selected line
      values$selected_line_index <- length(values$lines)
      
      # Reset current line but keep the name for convenience
      current_name <- values$current_line$name
      values$current_line <- list(
        name = current_name,
        start = NULL,
        end = NULL,
        annotations = list(),
        quality = input$transectQuality,
        notes = input$notes
      )
      values$drawing <- FALSE
      
      output$statusText <- renderText(paste("Line saved as", current_name))
    } else {
      output$statusText <- renderText("Cannot save: Line is incomplete")
    }
  })
  
  # Update the line selector dropdown
  updateLineSelector <- function() {
    line_names <- sapply(values$lines, function(line) {
      if (!is.null(line$name) && line$name != "") {
        return(line$name)
      } else {
        return(paste("Unnamed Line", which(values$lines == line)))
      }
    })
    
    updateSelectInput(session, "lineSelector", 
                      choices = line_names,
                      selected = if (length(line_names) > 0) line_names[length(line_names)] else NULL)
  }
  
  # Handle line selection
  observeEvent(input$lineSelector, {
    if (!is.null(input$lineSelector) && input$lineSelector != "" && length(values$lines) > 0) {
      # Find the line with this name
      for (i in seq_along(values$lines)) {
        if (values$lines[[i]]$name == input$lineSelector) {
          values$selected_line_index <- i
          
          # Load the selected line as current line
          values$current_line <- values$lines[[i]]
          values$drawing <- FALSE
          
          # Update the UI controls
          updateTextInput(session, "lineName", value = values$current_line$name)
          updateSelectInput(session, "transectQuality", selected = values$current_line$quality)
          updateTextAreaInput(session, "notes", value = values$current_line$notes)
          
          output$statusText <- renderText(paste("Loaded", input$lineSelector))
          break
        }
      }
    }
  })
  
  # Delete selected line
  observeEvent(input$deleteLine, {
    if (values$selected_line_index > 0 && values$selected_line_index <= length(values$lines)) {
      # Get the name before deleting
      line_name <- values$lines[[values$selected_line_index]]$name
      
      # Remove the line
      values$lines <- values$lines[-values$selected_line_index]
      
      # Update selected index
      if (length(values$lines) == 0) {
        values$selected_line_index <- 0
      } else if (values$selected_line_index > length(values$lines)) {
        values$selected_line_index <- length(values$lines)
      }
      
      # Update the line selector
      updateLineSelector()
      
      # Clear current line if it was the deleted one
      values$current_line <- list(
        name = input$lineName,
        start = NULL,
        end = NULL,
        annotations = list(),
        quality = input$transectQuality,
        notes = input$notes
      )
      values$drawing <- FALSE
      
      output$statusText <- renderText(paste("Deleted", line_name))
    }
  })
  
  # Handle clicks on the image
  observeEvent(input$plot_click, {
    click <- input$plot_click
    
    # If drawing a new line
    if (values$drawing) {
      if (is.null(values$current_line$start)) {
        # Set start point
        values$current_line$start <- list(x = click$x, y = click$y)
        output$statusText <- renderText("Click on the image to set the END point of the line")
      } else if (is.null(values$current_line$end)) {
        # Set end point
        values$current_line$end <- list(x = click$x, y = click$y)
        values$drawing <- FALSE
        output$statusText <- renderText("Line complete. Now select annotation type and click on line to add markers.")
      }
    } 
    # If line exists and we're adding annotations
    else if (!is.null(values$current_line$start) && !is.null(values$current_line$end)) {
      # Find nearest point on the line
      start_x <- values$current_line$start$x
      start_y <- values$current_line$start$y
      end_x <- values$current_line$end$x
      end_y <- values$current_line$end$y
      
      # Calculate parameter t for closest point on the line
      numerator <- (click$x - start_x) * (end_x - start_x) + 
        (click$y - start_y) * (end_y - start_y)
      denominator <- (end_x - start_x)^2 + (end_y - start_y)^2
      t <- numerator / denominator
      
      # Clamp t to [0, 1] to keep point on the line segment
      t <- max(0, min(1, t))
      
      # Calculate the point on the line
      point_x <- start_x + t * (end_x - start_x)
      point_y <- start_y + t * (end_y - start_y)
      
      # Add the annotation
      type <- input$annotationType
      values$current_line$annotations[[type]] <- list(x = point_x, y = point_y)
      
      output$statusText <- renderText(paste("Added", 
                                            switch(type,
                                                   "start" = "Start of Otolith",
                                                   "core" = "Core",
                                                   "end" = "End of Otolith",
                                                   "data_start" = "Start of Useful Data",
                                                   "data_end" = "End of Useful Data"), 
                                            "marker"))
    }
  })
  
  # Update line name when text input changes
  observeEvent(input$lineName, {
    values$current_line$name <- input$lineName
  })
  
  # Display line information
  output$lineInfo <- renderPrint({
    if (!is.null(values$current_line$start)) {
      cat("Current Line:", if(values$current_line$name != "") values$current_line$name else "Unnamed", "\n")
      cat("Quality:", values$current_line$quality, "\n")
      if (values$current_line$notes != "") {
        cat("Notes:", values$current_line$notes, "\n")
      }
      
      # Start point
      cat("  Start: (", round(values$current_line$start$x, 1), ", ", 
          round(values$current_line$start$y, 1), ")\n", sep="")
      
      # End point
      if (!is.null(values$current_line$end)) {
        cat("  End: (", round(values$current_line$end$x, 1), ", ", 
            round(values$current_line$end$y, 1), ")\n", sep="")
      }
      
      # Annotations
      if (length(values$current_line$annotations) > 0) {
        cat("  Annotations:\n")
        for (type in names(values$current_line$annotations)) {
          point <- values$current_line$annotations[[type]]
          label <- switch(type,
                          "start" = "Start of Otolith",
                          "core" = "Core",
                          "end" = "End of Otolith",
                          "data_start" = "Start of Useful Data",
                          "data_end" = "End of Useful Data")
          cat("    ", label, ": (", round(point$x, 1), ", ", 
              round(point$y, 1), ")\n", sep="")
        }
      }
    } else {
      cat("No current line\n")
    }
    
    # Information about saved lines
    if (length(values$lines) > 0) {
      cat("\nSaved Lines:\n")
      for (i in seq_along(values$lines)) {
        line <- values$lines[[i]]
        cat("  ", if(!is.null(line$name) && line$name != "") line$name else paste("Line", i), 
            " (", line$quality, "):\n", sep="")
        
        if (line$notes != "") {
          cat("    Notes: ", line$notes, "\n", sep="")
        }
        
        cat("    Start: (", round(line$start$x, 1), ", ", 
            round(line$start$y, 1), ")\n", sep="")
        cat("    End: (", round(line$end$x, 1), ", ", 
            round(line$end$y, 1), ")\n", sep="")
        
        if (length(line$annotations) > 0) {
          cat("    Annotations:\n")
          for (type in names(line$annotations)) {
            point <- line$annotations[[type]]
            label <- switch(type,
                            "start" = "Start of Otolith",
                            "core" = "Core",
                            "end" = "End of Otolith",
                            "data_start" = "Start of Useful Data",
                            "data_end" = "End of Useful Data")
            cat("      ", label, ": (", round(point$x, 1), ", ", 
                round(point$y, 1), ")\n", sep="")
          }
        }
      }
    }
  })
  
  # Initialize status text
  output$statusText <- renderText("Upload an image to begin")
  
  # Download data
  output$downloadData <- downloadHandler(
    filename = function() {
      paste("otolith-lines-", Sys.Date(), ".csv", sep="")
    },
    content = function(file) {
      # Create a data frame with all line information
      result <- data.frame()
      
      for (i in seq_along(values$lines)) {
        line <- values$lines[[i]]
        
        # Basic line info
        line_data <- data.frame(
          line_number = i,
          line_name = if(!is.null(line$name)) line$name else paste("Line", i),
          quality = line$quality,
          notes = line$notes,
          start_x = line$start$x,
          start_y = line$start$y,
          end_x = line$end$x,
          end_y = line$end$y
        )
        
        # Add annotation info
        for (type in c("start", "core", "end", "data_start", "data_end")) {
          if (!is.null(line$annotations[[type]])) {
            line_data[[paste0(type, "_annotation_x")]] <- line$annotations[[type]]$x
            line_data[[paste0(type, "_annotation_y")]] <- line$annotations[[type]]$y
          } else {
            line_data[[paste0(type, "_annotation_x")]] <- NA
            line_data[[paste0(type, "_annotation_y")]] <- NA
          }
        }
        
        result <- rbind(result, line_data)
      }
      
      write.csv(result, file, row.names = FALSE)
    }
  )
}

shinyApp(ui, server)

```


# Old Code
# Otolith Data Processing

#### detect_otolith_boundaries()
Function to detect boundaries in Cu65 measurements. This function identifies the start and end points of valid otolith measurements based on Cu65 patterns.  Cu65 levels appear to be quite high in portions of the transect outside of the otolith.  This function specifically tries to identify the start and end of the otolith based on sustained low values of Cu65, taking an average for the
```{r}
detect_otolith_boundaries <- function(cu65_data, 
                                    high_threshold = 100,    
                                    stable_threshold = 100,
                                    min_stable_points = 3) { 
  # Initialize variables
  start_idx <- NULL
  end_idx <- NULL
  # Remove NAs
  clean_data <- cu65_data[!is.na(cu65_data)]
  # Find start (look for sustained low values)
  for(i in min_stable_points:length(clean_data)) {
    window <- clean_data[(i-min_stable_points+1):i]
    if(all(abs(window) < stable_threshold)) {
      start_idx <- i-min_stable_points+1
      break
    }
  }
  # Check if start was found
  if(is.null(start_idx)) {
    stop("Could not detect start boundary")
  }
  # Find end (where values start increasing consistently)
  for(i in start_idx:(length(clean_data)-2)) {  
    if(i + 2 <= length(clean_data)) {  
      window_mean <- mean(abs(clean_data[i:(i+2)]))
      if(window_mean > stable_threshold) {
        end_idx <- i-1
        break
      }
    }
  }
  # Check if end was found
  if(is.null(end_idx)) {
    end_idx <- length(clean_data)
  }
  return(list(
    start = start_idx,
    end = end_idx,
    clean_data = clean_data[start_idx:end_idx],
    mean_value = mean(clean_data[start_idx:end_idx]),
    sd_value = sd(clean_data[start_idx:end_idx])
  ))
}
```

#### process_single_otolith()
Function to process individual otolith dataframes.  Takes a single otolith dataframe and returns filtered data with boundaries.  Returns a list containing filtered data and boundary information
```{r}
process_single_otolith <- function(otolith_df) {
  boundaries <- detect_otolith_boundaries(otolith_df$Cu65_ppm)
  filtered_df <- otolith_df[boundaries$start:boundaries$end, ]
  return(list(
    filtered_data = filtered_df,
    boundaries = boundaries
  ))
}
```

### process_all_otoliths
Function to process the entire list of otoliths.  Includes error handling and progress reporting.  Returns a list of processed results for all otoliths
```{r}
process_all_otoliths <- function(otolith_list) {
  results <- list()
  for(otolith_name in names(otolith_list)) {
    tryCatch({
      results[[otolith_name]] <- process_single_otolith(otolith_list[[otolith_name]])
      cat("Successfully processed:", otolith_name, "\n")
    }, error = function(e) {
      cat("Error processing", otolith_name, ":", e$message, "\n")
    })
  }
  return(results)
}
```


# Execute processing on all otoliths, generate summary

Create summary dataframe with additional metrics including observations removed.  Calculates original observations, final observations, and number filtered out for each otolith
```{r}
filtered_otoliths <- process_all_otoliths(df)
processing_summary <- data.frame(
  otolith = names(filtered_otoliths),
  start_index = sapply(filtered_otoliths, function(x) x$boundaries$start),
  end_index = sapply(filtered_otoliths, function(x) x$boundaries$end),
  total_indices = sapply(df, nrow),  # Get total rows from original df
  filtered_indices = sapply(filtered_otoliths, function(x) (x$boundaries$end - x$boundaries$start + 1))
)
processing_summary$index_diff <- processing_summary$total_indices - processing_summary$filtered_indices
# View enhanced summary
head(processing_summary)
```







### find_optimal_thresholds()
Function to find optimal thresholds using middle portion of data.  Returns suggested stable threshold based on statistical properties
```{r}
find_optimal_thresholds <- function(cu65_data) {
  # Remove NAs
  clean_data <- cu65_data[!is.na(cu65_data)]
  
  # Get the middle portion of data (likely stable region)
  mid_start <- floor(length(clean_data) * 0.2)  # Skip first 20%
  mid_end <- floor(length(clean_data) * 0.8)    # Skip last 20%
  stable_region <- clean_data[mid_start:mid_end]
  
  # Calculate statistics
  stable_mean <- mean(stable_region)
  stable_sd <- sd(stable_region)
  
  # Suggest threshold based on mean + 2*SD
  suggested_threshold <- stable_mean + 2*stable_sd
  
  return(list(
    mean = stable_mean,
    sd = stable_sd,
    suggested_threshold = suggested_threshold,
    stable_region_values = summary(stable_region)
  ))
}
```

### testing threshold finding and boundary detection

```{r}
# Test the threshold finding function on first otolith
sample_thresholds <- find_optimal_thresholds(df[[1]]$Cu65_ppm)
print(sample_thresholds)

# Apply to detection function with new threshold
test_boundaries <- detect_otolith_boundaries(
  df[[1]]$Cu65_ppm,
  stable_threshold = sample_thresholds$suggested_threshold,
  min_stable_points = 3
)

```


### Modified Process Function with Optimal Thresholds

```{r}
# Updated process function incorporating threshold detection
process_single_otolith <- function(otolith_df) {
  # Find optimal threshold
  thresholds <- find_optimal_thresholds(otolith_df$Cu65_ppm)
  
  # Apply boundaries with optimal threshold
  boundaries <- detect_otolith_boundaries(
    otolith_df$Cu65_ppm,
    stable_threshold = thresholds$suggested_threshold,
    min_stable_points = 3
  )
  
  filtered_df <- otolith_df[boundaries$start:boundaries$end, ]
  
  return(list(
    filtered_data = filtered_df,
    boundaries = boundaries,
    thresholds = thresholds
  ))
}

# Execute processing with new function
filtered_otoliths <- process_all_otoliths(df) 
```

### Enhanced Summary with Threshold Information


```{r}
detect_otolith_boundaries <- function(cu65_data, 
                                    high_threshold = 100,    
                                    stable_threshold = 100,
                                    min_stable_points = 40) { 
  # Input validation
  if(length(cu65_data) < min_stable_points) {
    stop(sprintf("Data length (%d) is shorter than min_stable_points (%d)", 
                 length(cu65_data), min_stable_points))
  }
  
  # Initialize variables
  start_idx <- NULL
  end_idx <- NULL
  
  # Remove NAs
  clean_data <- cu65_data[!is.na(cu65_data)]
  
  # Modified start detection - look for first set of stable low values
  for(i in min_stable_points:length(clean_data)) {
    window <- clean_data[(i-min_stable_points+1):i]
    # Check if values are decreasing and below threshold
    if(all(diff(window) <= 0) && max(window) < stable_threshold) {
      start_idx <- i-min_stable_points+1
      break
    }
  }
  
  if(is.null(start_idx)) {
    # Fallback: look for first value below threshold
    for(i in 1:length(clean_data)) {
      if(clean_data[i] < stable_threshold) {
        start_idx <- i
        break
      }
    }
  }
  
  # Check if start was found
  if(is.null(start_idx)) {
    stop(sprintf("Could not detect start boundary. Min value: %f, Max value: %f", 
                 min(clean_data, na.rm=TRUE), max(clean_data, na.rm=TRUE)))
  }
  
  # End detection
  end_idx <- length(clean_data)
  
  # Look for last stable section before NAs or increases
  for(i in (length(clean_data)-5):start_idx) {
    if(!is.na(clean_data[i])) {
      next_points <- clean_data[i:(min(i+2, length(clean_data)))]
      if(mean(abs(next_points)) > stable_threshold || any(is.na(next_points))) {
        end_idx <- i
        break
      }
    }
  }
  
  return(list(
    start = start_idx,
    end = end_idx,
    clean_data = clean_data[start_idx:end_idx],
    mean_value = mean(clean_data[start_idx:end_idx], na.rm=TRUE),
    sd_value = sd(clean_data[start_idx:end_idx], na.rm=TRUE)
  ))
}
```





