# Load packages
Install and load necessary packages
```{r}
packages <- c("tidyverse")
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages) > 0) {
  install.packages(new_packages)
}
invisible(lapply(packages, library, character.only = TRUE))
rm(packages, new_packages)
```

# Data Import
Read and prepare initial dataframe for otolith elemental analysis from saved RDS file

```{r}
# Load ICP-MS data from RDS file
df <- readRDS("RDS_dataframes/ICP-MS_clean.RDS")

# Select only otolith specimens (rows 30-184)
df <- df[c(30:184)]
```

### create single dataframe (remove lists)
```{r}
# no specimen 14
# specimen 24 missing a third transect
for(i in 1:155){
  spec <- names(df[i])
  df[[i]]$specimen <- as.factor(str_sub(spec, start= 0,end = 5))
  df[[i]]$transect <- as.factor(as.numeric(str_sub(spec, start= -1)) + 1)
  rm(spec,i)
}
df <- do.call("rbind", df)
df <- rename(df, "time" = `Elapsed Time`)
# saveRDS(df, "RDS_dataframes/ICP-MS_df.RDS")
rm(ICP_MS, df)
```

### outliers
```{r}
bad <- c(10,14,30,32,34,35)
good <- 1:40
good <- good[!good %in% c(bad)]
```


# Best Approach



### find_otolith_boundaries()
Detect Otolith Boundaries using Ca43 and Cu65 signals
ca43_data: Numeric vector of Ca43 measurements
cu65_data: Numeric vector of Cu65 measurements
window_size:  Size of moving average window (default: 7)
```{r}
find_otolith_boundaries <- function(ca43_data, cu65_data, window_size = 7) {
  # Data preparation
  n_points <- length(ca43_data)
  # Calculate rolling medians with larger window for stability
  ma_ca43 <- stats::runmed(ca43_data, k = window_size)
  ma_cu65 <- stats::runmed(cu65_data, k = window_size)
  # Calculate percentiles using middle portion of data for better baseline
  mid_start <- ceiling(n_points * 0.3)
  mid_end <- floor(n_points * 0.7)
  mid_data_ca43 <- ma_ca43[mid_start:mid_end]
  # Get reference levels
  ca43_median <- median(mid_data_ca43, na.rm = TRUE)
  ca43_start_threshold <- ca43_median * 0.4  # More lenient threshold
  # Start detection with minimum signal requirements
  min_stable_points <- 5
  start_idx <- NULL
  # Look for sustained rise in Ca43
  for(i in window_size:(n_points-min_stable_points)) {
    window <- ma_ca43[i:(i+min_stable_points)]
    if(all(window > ca43_start_threshold) && 
       mean(window) > mean(ma_ca43[(i-window_size):i])) {
      start_idx <- i
      break
    }
  }
  # Fallback if no clear start found
  if(is.null(start_idx)) {
    # Use first major rise in Ca43
    for(i in 2:n_points) {
      if(ma_ca43[i] > ca43_start_threshold && 
         ma_ca43[i] > mean(ma_ca43[max(1, i-5):i]) * 1.5) {
        start_idx <- i
        break
      }
    }
  }
  # If still no start found, use conservative approach
  if(is.null(start_idx)) {
    start_idx <- ceiling(n_points * 0.1)  # Use first 10% as default
  }
  # End detection
  end_idx <- n_points
  tail_length <- min(50, ceiling(n_points * 0.1))
  # Only look for end if we have enough points
  if(n_points - start_idx > window_size * 3) {
    tail_median <- median(ma_ca43[(n_points-tail_length):n_points])
    for(i in (n_points-tail_length):start_idx) {
      if(mean(ma_ca43[i:(i+min(4, n_points-i))]) < ca43_median * 0.3) {
        end_idx <- i
        break
      }
    }
  }
  # Validate and adjust boundaries
  if(end_idx - start_idx < 100) {  # Minimum reasonable length
    end_idx <- n_points
    start_idx <- min(start_idx, ceiling(n_points * 0.15))  # Max 15% from start
  }
  return(list(
    start = start_idx,
    end = end_idx,
    filtered_data = ca43_data[start_idx:end_idx],
    diagnostics = list(
      ca43_median = ca43_median,
      start_threshold = ca43_start_threshold
    )
  ))
}
```
### process_otoliths()
Process all otolith transects
```{r}
#'  otolith_list List of dataframes containing Ca43 and Cu65_ppm measurements
#'  Summary dataframe of processing results
process_otoliths <- function(otolith_list) {
  results <- list()
  failed_specimens <- character()
  
  for(name in names(otolith_list)) {
    tryCatch({
      bounds <- find_otolith_boundaries(
        otolith_list[[name]]$Ca43,
        otolith_list[[name]]$Cu65_ppm
      )
      results[[name]] <- bounds
      cat("Processed", name, 
          "- Start:", bounds$start, 
          "End:", bounds$end, "\n")
    }, error = function(e) {
      cat("Error processing", name, ":", e$message, "\n")
      failed_specimens <- c(failed_specimens, name)
    })
  }
  summary_df <- data.frame(
    transect = names(results),
    start_index = sapply(results, `[[`, "start"),
    end_index = sapply(results, `[[`, "end"),
    original_length = sapply(otolith_list[names(results)], nrow),
    filtered_length = sapply(results, function(x) length(x$filtered_data))
  )
  summary_df$points_removed <- summary_df$original_length - summary_df$filtered_length
  attr(summary_df, "failed_specimens") <- failed_specimens
  return(summary_df)
}
```
#### outlier from analysis
```{r}
# Documentation for anomalous specimen oto10-2
ANOMALY_NOTE <- list(
  specimen_id = "oto10-2",
  issue = "Mid-transect elemental data disruption",
  details = list(
    observation = "All elemental data show aberrant values in middle section",
    impact = "Creates artificial truncation in data processing",
    location = "Around point 288 in transect",
    comparison = "Other oto10 series (0,1,3) show normal patterns"
  ),
  recommendation = "Should be flagged for separate analysis or potential exclusion from bulk processing"
)

```

### Filter DF

```{r}
# Process all specimens
results <- process_otoliths(df)
# Create filtered version of original dataset
filtered_df <- lapply(names(df), function(specimen) {
  # Get boundaries for this specimen
  bounds <- results[results$transect == specimen, ]
  
  # Skip if specimen is the documented anomaly
  if(specimen == "oto10-2") {
    warning(paste("Skipping known anomalous specimen:", specimen))
    return(NULL)
  }
  
  # Extract relevant portion of data
  df[[specimen]][bounds$start_index:bounds$end_index, ]
})
# Name the list elements
names(filtered_df) <- names(df)
# Remove NULL entries (anomalous specimens)
filtered_df <- filtered_df[!sapply(filtered_df, is.null)]
# saveRDS(filtered_df, "RDS_dataframes/ICP-MS_cutoffs_df.RDS")
```

### Validation checks
```{r validation}
# Check filtered dataset
cat("Original specimens:", length(df), "\n")
cat("Filtered specimens:", length(filtered_df), "\n")
cat("Removed specimens:", setdiff(names(df), names(filtered_df)), "\n")

# Check lengths match results
all_match <- all(sapply(names(filtered_df), function(specimen) {
  nrow(filtered_df[[specimen]]) == 
    results$filtered_length[results$transect == specimen]
}))
cat("All lengths match results:", all_match, "\n")
```

### generate_otolith_stats()
summary stats from filtered DF, both by transect and per otolith
```{r}
generate_otolith_stats <- function(filtered_df) {
  elements <- c("Ca43", "Li7_ppm", "B11_ppm", "Mg24_ppm", "Si29_ppm", 
                "Mn55_ppm", "Fe56_ppm", "Co59_ppm", "Ni60_ppm", 
                "Cu63_ppm", "Cu65_ppm", "Zn66_ppm", "Rb85_ppm", 
                "Sr86_ppm", "Sr87_ppm", "Sr88_ppm", "Cd111_ppm", 
                "Ba138_ppm", "Ce140_ppm", "Pb206_ppm", "Pb207_ppm", 
                "Pb208_ppm", "Th232_ppm", "PbTotal_ppm")
  
  format_summary_tables <- function(stats_list, elements) {
    transect_summary <- do.call(rbind, lapply(stats_list$transect_stats, function(x) {
      regions <- c("core", "middle", "edge")
      
      do.call(rbind, lapply(regions, function(r) {
        data.frame(
          specimen = x$specimen,
          region = r,
          element = elements,
          mean = x$stats[[r]]["mean", ],
          sd = x$stats[[r]]["sd", ],
          n = x$stats[[r]]["n", ],
          stringsAsFactors = FALSE
        )
      }))
    }))
    
    otolith_summary <- do.call(rbind, lapply(stats_list$otolith_stats, function(x) {
      regions <- c("core", "middle", "edge")
      
      do.call(rbind, lapply(regions, function(r) {
        data.frame(
          otolith = x$otolith,
          region = r,
          element = elements,
          mean = x$stats[[r]]$means["mean", ],
          rsd = x$stats[[r]]$rsd,
          stringsAsFactors = FALSE
        )
      }))
    }))
    
    return(list(
      transect_summary = transect_summary,
      otolith_summary = otolith_summary
    ))
  }
  
  transect_stats <- lapply(names(filtered_df), function(specimen) {
    df <- filtered_df[[specimen]]
    n_points <- nrow(df)
    
    core_end <- ceiling(n_points * 0.2)
    middle_start <- core_end + 1
    middle_end <- floor(n_points * 0.8)
    edge_start <- middle_end + 1
    
    regions <- list(
      core = df[1:core_end, ],
      middle = df[middle_start:middle_end, ],
      edge = df[edge_start:n_points, ]
    )
    
    region_stats <- lapply(regions, function(region) {
      sapply(elements, function(elem) {
        data <- region[[elem]]
        c(
          mean = mean(data, na.rm = TRUE),
          sd = sd(data, na.rm = TRUE),
          median = median(data, na.rm = TRUE),
          q25 = quantile(data, 0.25, na.rm = TRUE),
          q75 = quantile(data, 0.75, na.rm = TRUE),
          n = sum(!is.na(data))
        )
      })
    })
    
    return(list(
      specimen = specimen,
      stats = region_stats,
      n_points = n_points
    ))
  })
  
  otolith_ids <- unique(sub("-\\d+$", "", names(filtered_df)))
  
  otolith_stats <- lapply(otolith_ids, function(otolith) {
    transects <- grep(paste0("^", otolith, "-"), names(filtered_df), value = TRUE)
    regions <- c("core", "middle", "edge")
    
    combined_stats <- lapply(regions, function(region) {
      region_stats <- lapply(transects, function(t) {
        idx <- match(t, names(filtered_df))
        transect_stats[[idx]]$stats[[region]]
      })
      
      element_means <- Reduce(`+`, region_stats) / length(region_stats)
      
      element_rsd <- sapply(elements, function(elem) {
        values <- sapply(region_stats, function(x) x[1, elem])
        sd(values, na.rm = TRUE) / mean(values, na.rm = TRUE) * 100
      })
      
      list(
        means = element_means,
        rsd = element_rsd
      )
    })
    
    names(combined_stats) <- regions
    
    return(list(
      otolith = otolith,
      stats = combined_stats
    ))
  })
  
  stats_list <- list(
    transect_stats = transect_stats,
    otolith_stats = otolith_stats
  )
  
  summary_tables <- format_summary_tables(stats_list, elements)
  
  return(summary_tables)
}

summary_results <- generate_otolith_stats(filtered_df)
results_transect <- summary_results$transect_summary
results_otolith <- summary_results$otolith_summary
# saveRDS(summary_results,"RDS_dataframes/ICP-MS_summarystats")
```

### Data analysis

```{r}
# 1. Basic Variation Check
# Calculate RSD (Relative Standard Deviation) between transects for each otolith-element-region
transect_variation <- function(summary_results) {
  # Get transect summary data
  ts <- summary_results$transect_summary
  
  # Group by otolith (remove transect number), element, and region
  variation_check <- aggregate(mean ~ sub("-\\d+$", "", specimen) + element + region, 
                               data = ts,
                               FUN = function(x) {
                                 c(mean = mean(x),
                                   rsd = sd(x)/mean(x) * 100,
                                   n_transects = length(x))
                               })
  
  # Filter for high variation (e.g., RSD > 20%)
  high_variation <- subset(variation_check, variation_check$mean[, "rsd"] > 20)
  
  return(list(
    all_variation = variation_check,
    high_variation = high_variation
  ))
}

# 2. Statistical Test
# Perform ANOVA for each otolith-element-region combination
transect_anova <- function(summary_results) {
  ts <- summary_results$transect_summary
  
  # Split data by otolith, element, and region
  unique_combinations <- unique(paste(sub("-\\d+$", "", ts$specimen), 
                                      ts$element, 
                                      ts$region))
  
  # Run ANOVA for each combination
  anova_results <- lapply(unique_combinations, function(combo) {
    split_combo <- strsplit(combo, " ")[[1]]
    otolith <- split_combo[1]
    element <- split_combo[2]
    region <- split_combo[3]
    
    subset_data <- subset(ts, 
                          sub("-\\d+$", "", specimen) == otolith & 
                            element == element & 
                            region == region)
    
    if(length(unique(subset_data$mean)) > 1) {
      anova_result <- summary(aov(mean ~ specimen, data = subset_data))
      p_value <- anova_result[[1]]$"Pr(>F)"[1]
      
      return(data.frame(
        otolith = otolith,
        element = element,
        region = region,
        p_value = p_value,
        significant = p_value < 0.05
      ))
    }
    return(NULL)
  })
  
  # Combine results
  anova_results <- do.call(rbind, anova_results[!sapply(anova_results, is.null)])
  return(anova_results)
}

# Usage:
variation_results <- transect_variation(summary_results)
anova_results <- transect_anova(summary_results)

# Quick summary of problematic measurements
print("Elements with high variation between transects:")
print(variation_results$high_variation)

print("Statistically significant differences between transects:")
print(subset(anova_results, significant == TRUE))

# Comparative Analysis of Elemental Composition Across Otolith Regions

compare_otolith_regions <- function(summary_results) {
  # Prepare otolith summary data
  os <- summary_results$otolith_summary
  
  # ANOVA-like comparison between regions for each element
  region_comparisons <- lapply(unique(os$element), function(elem) {
    elem_data <- subset(os, element == elem)
    
    # Reshape data for statistical testing
    region_means <- reshape2::dcast(elem_data, otolith ~ region, value.var = "mean")
    
    # Perform statistical tests
    pairwise_tests <- list(
      core_vs_middle = t.test(region_means$core, region_means$middle),
      core_vs_edge = t.test(region_means$core, region_means$edge),
      middle_vs_edge = t.test(region_means$middle, region_means$edge)
    )
    
    # Summarize results
    data.frame(
      element = elem,
      core_mean = mean(region_means$core),
      middle_mean = mean(region_means$middle),
      edge_mean = mean(region_means$edge),
      core_vs_middle_p = pairwise_tests$core_vs_middle$p.value,
      core_vs_edge_p = pairwise_tests$core_vs_edge$p.value,
      middle_vs_edge_p = pairwise_tests$middle_vs_edge$p.value
    )
  })
  
  # Combine results
  region_comparison_summary <- do.call(rbind, region_comparisons)
  
  # Adjust p-values for multiple comparisons
  region_comparison_summary$core_vs_middle_padj <- p.adjust(region_comparison_summary$core_vs_middle_p, method = "fdr")
  region_comparison_summary$core_vs_edge_padj <- p.adjust(region_comparison_summary$core_vs_edge_p, method = "fdr")
  region_comparison_summary$middle_vs_edge_padj <- p.adjust(region_comparison_summary$middle_vs_edge_p, method = "fdr")
  
  # Identify significant differences
  region_comparison_summary$significant_differences <- apply(
    region_comparison_summary[, c("core_vs_middle_padj", "core_vs_edge_padj", "middle_vs_edge_padj")], 
    1, 
    function(x) any(x < 0.05)
  )
  
  return(region_comparison_summary)
}
region_comparisons <- compare_otolith_regions(summary_results)
# Print significant differences
print(subset(region_comparisons, significant_differences == TRUE))


```


### compare otoliths

```{r}
compare_otoliths <- function(summary_results) {
  # Prepare otolith summary data
  os <- summary_results$otolith_summary
  
  # Perform statistical comparisons between otoliths
  otolith_comparisons <- lapply(unique(os$element), function(elem) {
    # Extract data for specific element
    elem_data <- subset(os, element == elem)
    
    # Reshape data for statistical analysis
    region_means <- reshape2::dcast(elem_data, otolith + region ~ element, value.var = "mean")
    
    # Perform ANOVA to test overall differences between otoliths
    aov_results <- lapply(c("core", "middle", "edge"), function(reg) {
      region_subset <- subset(region_means, region == reg)
      
      # Check if we have enough unique values to perform ANOVA
      if(length(unique(region_subset$otolith)) > 1) {
        tryCatch({
          aov_model <- aov(get(elem) ~ otolith, data = region_subset)
          aov_summary <- summary(aov_model)
          
          data.frame(
            element = elem,
            region = reg,
            f_value = aov_summary[[1]]$"F value"[1],
            p_value = aov_summary[[1]]$"Pr(>F)"[1]
          )
        }, error = function(e) {
          NULL  # Return NULL if ANOVA fails
        })
      } else {
        NULL  # Not enough unique values for ANOVA
      }
    })
    
    # Remove NULL results
    aov_results <- aov_results[!sapply(aov_results, is.null)]
    
    # Combine results if any exist
    if(length(aov_results) > 0) {
      do.call(rbind, aov_results)
    } else {
      NULL
    }
  })
  
  # Remove NULL results and combine
  otolith_comparisons <- otolith_comparisons[!sapply(otolith_comparisons, is.null)]
  
  if(length(otolith_comparisons) > 0) {
    otolith_comparison_summary <- do.call(rbind, otolith_comparisons)
    
    # Adjust p-values for multiple comparisons
    otolith_comparison_summary$p_adjusted <- p.adjust(
      otolith_comparison_summary$p_value, 
      method = "fdr"
    )
    
    # Identify significant differences
    otolith_comparison_summary$significant <- otolith_comparison_summary$p_adjusted < 0.05
    
    return(otolith_comparison_summary)
  } else {
    warning("No valid comparisons could be made. Check your data.")
    return(NULL)
  }
}

# Visualization function remains the same
visualize_otolith_differences <- function(summary_results) {
  library(ggplot2)
  library(reshape2)
  
  os <- summary_results$otolith_summary
  
  # Prepare data for heatmap
  plot_data <- dcast(os, otolith + region ~ element, value.var = "mean")
  
  # Melt data for ggplot
  plot_data_long <- melt(plot_data, id.vars = c("otolith", "region"))
  
  # Create heatmap
  ggplot(plot_data_long, aes(x = otolith, y = variable, fill = value)) +
    geom_tile() +
    facet_wrap(~ region) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
    theme_minimal() +
    labs(title = "Elemental Composition Across Otoliths",
         x = "Otolith", y = "Element", fill = "Mean Value")
}

# Usage
otolith_comparisons <- compare_otoliths(summary_results)
otolith_plot <- visualize_otolith_differences(summary_results)

# Print significant differences
if(!is.null(otolith_comparisons)) {
  print("Significant differences between otoliths:")
  print(subset(otolith_comparisons, significant == TRUE))
} else {
  print("No significant differences could be detected.")
}
```










# Old Code
# Otolith Data Processing

#### detect_otolith_boundaries()
Function to detect boundaries in Cu65 measurements. This function identifies the start and end points of valid otolith measurements based on Cu65 patterns.  Cu65 levels appear to be quite high in portions of the transect outside of the otolith.  This function specifically tries to identify the start and end of the otolith based on sustained low values of Cu65, taking an average for the
```{r}
detect_otolith_boundaries <- function(cu65_data, 
                                    high_threshold = 100,    
                                    stable_threshold = 100,
                                    min_stable_points = 3) { 
  # Initialize variables
  start_idx <- NULL
  end_idx <- NULL
  # Remove NAs
  clean_data <- cu65_data[!is.na(cu65_data)]
  # Find start (look for sustained low values)
  for(i in min_stable_points:length(clean_data)) {
    window <- clean_data[(i-min_stable_points+1):i]
    if(all(abs(window) < stable_threshold)) {
      start_idx <- i-min_stable_points+1
      break
    }
  }
  # Check if start was found
  if(is.null(start_idx)) {
    stop("Could not detect start boundary")
  }
  # Find end (where values start increasing consistently)
  for(i in start_idx:(length(clean_data)-2)) {  
    if(i + 2 <= length(clean_data)) {  
      window_mean <- mean(abs(clean_data[i:(i+2)]))
      if(window_mean > stable_threshold) {
        end_idx <- i-1
        break
      }
    }
  }
  # Check if end was found
  if(is.null(end_idx)) {
    end_idx <- length(clean_data)
  }
  return(list(
    start = start_idx,
    end = end_idx,
    clean_data = clean_data[start_idx:end_idx],
    mean_value = mean(clean_data[start_idx:end_idx]),
    sd_value = sd(clean_data[start_idx:end_idx])
  ))
}
```

#### process_single_otolith()
Function to process individual otolith dataframes.  Takes a single otolith dataframe and returns filtered data with boundaries.  Returns a list containing filtered data and boundary information
```{r}
process_single_otolith <- function(otolith_df) {
  boundaries <- detect_otolith_boundaries(otolith_df$Cu65_ppm)
  filtered_df <- otolith_df[boundaries$start:boundaries$end, ]
  return(list(
    filtered_data = filtered_df,
    boundaries = boundaries
  ))
}
```

### process_all_otoliths
Function to process the entire list of otoliths.  Includes error handling and progress reporting.  Returns a list of processed results for all otoliths
```{r}
process_all_otoliths <- function(otolith_list) {
  results <- list()
  for(otolith_name in names(otolith_list)) {
    tryCatch({
      results[[otolith_name]] <- process_single_otolith(otolith_list[[otolith_name]])
      cat("Successfully processed:", otolith_name, "\n")
    }, error = function(e) {
      cat("Error processing", otolith_name, ":", e$message, "\n")
    })
  }
  return(results)
}
```


# Execute processing on all otoliths, generate summary

Create summary dataframe with additional metrics including observations removed.  Calculates original observations, final observations, and number filtered out for each otolith
```{r}
filtered_otoliths <- process_all_otoliths(df)
processing_summary <- data.frame(
  otolith = names(filtered_otoliths),
  start_index = sapply(filtered_otoliths, function(x) x$boundaries$start),
  end_index = sapply(filtered_otoliths, function(x) x$boundaries$end),
  total_indices = sapply(df, nrow),  # Get total rows from original df
  filtered_indices = sapply(filtered_otoliths, function(x) (x$boundaries$end - x$boundaries$start + 1))
)
processing_summary$index_diff <- processing_summary$total_indices - processing_summary$filtered_indices
# View enhanced summary
head(processing_summary)
```







### find_optimal_thresholds()
Function to find optimal thresholds using middle portion of data.  Returns suggested stable threshold based on statistical properties
```{r}
find_optimal_thresholds <- function(cu65_data) {
  # Remove NAs
  clean_data <- cu65_data[!is.na(cu65_data)]
  
  # Get the middle portion of data (likely stable region)
  mid_start <- floor(length(clean_data) * 0.2)  # Skip first 20%
  mid_end <- floor(length(clean_data) * 0.8)    # Skip last 20%
  stable_region <- clean_data[mid_start:mid_end]
  
  # Calculate statistics
  stable_mean <- mean(stable_region)
  stable_sd <- sd(stable_region)
  
  # Suggest threshold based on mean + 2*SD
  suggested_threshold <- stable_mean + 2*stable_sd
  
  return(list(
    mean = stable_mean,
    sd = stable_sd,
    suggested_threshold = suggested_threshold,
    stable_region_values = summary(stable_region)
  ))
}
```

### testing threshold finding and boundary detection

```{r}
# Test the threshold finding function on first otolith
sample_thresholds <- find_optimal_thresholds(df[[1]]$Cu65_ppm)
print(sample_thresholds)

# Apply to detection function with new threshold
test_boundaries <- detect_otolith_boundaries(
  df[[1]]$Cu65_ppm,
  stable_threshold = sample_thresholds$suggested_threshold,
  min_stable_points = 3
)

```


### Modified Process Function with Optimal Thresholds

```{r}
# Updated process function incorporating threshold detection
process_single_otolith <- function(otolith_df) {
  # Find optimal threshold
  thresholds <- find_optimal_thresholds(otolith_df$Cu65_ppm)
  
  # Apply boundaries with optimal threshold
  boundaries <- detect_otolith_boundaries(
    otolith_df$Cu65_ppm,
    stable_threshold = thresholds$suggested_threshold,
    min_stable_points = 3
  )
  
  filtered_df <- otolith_df[boundaries$start:boundaries$end, ]
  
  return(list(
    filtered_data = filtered_df,
    boundaries = boundaries,
    thresholds = thresholds
  ))
}

# Execute processing with new function
filtered_otoliths <- process_all_otoliths(df) 
```

### Enhanced Summary with Threshold Information


```{r}
detect_otolith_boundaries <- function(cu65_data, 
                                    high_threshold = 100,    
                                    stable_threshold = 100,
                                    min_stable_points = 40) { 
  # Input validation
  if(length(cu65_data) < min_stable_points) {
    stop(sprintf("Data length (%d) is shorter than min_stable_points (%d)", 
                 length(cu65_data), min_stable_points))
  }
  
  # Initialize variables
  start_idx <- NULL
  end_idx <- NULL
  
  # Remove NAs
  clean_data <- cu65_data[!is.na(cu65_data)]
  
  # Modified start detection - look for first set of stable low values
  for(i in min_stable_points:length(clean_data)) {
    window <- clean_data[(i-min_stable_points+1):i]
    # Check if values are decreasing and below threshold
    if(all(diff(window) <= 0) && max(window) < stable_threshold) {
      start_idx <- i-min_stable_points+1
      break
    }
  }
  
  if(is.null(start_idx)) {
    # Fallback: look for first value below threshold
    for(i in 1:length(clean_data)) {
      if(clean_data[i] < stable_threshold) {
        start_idx <- i
        break
      }
    }
  }
  
  # Check if start was found
  if(is.null(start_idx)) {
    stop(sprintf("Could not detect start boundary. Min value: %f, Max value: %f", 
                 min(clean_data, na.rm=TRUE), max(clean_data, na.rm=TRUE)))
  }
  
  # End detection
  end_idx <- length(clean_data)
  
  # Look for last stable section before NAs or increases
  for(i in (length(clean_data)-5):start_idx) {
    if(!is.na(clean_data[i])) {
      next_points <- clean_data[i:(min(i+2, length(clean_data)))]
      if(mean(abs(next_points)) > stable_threshold || any(is.na(next_points))) {
        end_idx <- i
        break
      }
    }
  }
  
  return(list(
    start = start_idx,
    end = end_idx,
    clean_data = clean_data[start_idx:end_idx],
    mean_value = mean(clean_data[start_idx:end_idx], na.rm=TRUE),
    sd_value = sd(clean_data[start_idx:end_idx], na.rm=TRUE)
  ))
}
```





